### 2023-09-27 manuscript/manuscript.Rmd
starting to write:

---
title: "To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  This paper focuses on analyzing sequential versus batched testing strategies in an emergency department (ED) at Mayo Clinic Arizona, with respect to their associations impacts on patient length of stay, hospital readmission, and healthcare resource utilization. A theoretical model was developed and tested to identify patient and hospital features that may influence the decision to batch or sequentially order tests, such as patient complexity, physician experience, occupancy level and complexity. Finally, we performed retrospective analysis of ED operational data to investigate the impact of batching on our key outcomes. The overall result was that batch ordering tests was associated with greater patient length of stay and resource utilization, even when we control for variation in patient complexity, physician experience, and hospital occupancy.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
---

# Introduction {#sec:1}

Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization. Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events. Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality @bernstein2009.

One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests @naseim2015. A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests such as electrocardiograms (ECGs) and point-of-care (POC) testing. These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.

A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially. This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing. Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results @koch2018. Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier. Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged. Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements @saghafian2015, however, research surrounding test ordering strategies to improve the patient flow processes remains limited.

In this paper, we use data from over 41,000 patient visits to the ED that occur during our study period to quantify the benefits and consequences of batching versus sequentially ordering diagnostic tests on patient length of stay, re-admission, and resource utilization. Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests. When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side. Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests. We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period. The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.

We start by evaluating the reduced-form effects of ED provider batch ordering tendency on downstream patient outcomes and turnaround time. We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences. Being treated by a provider in the top decile of the leniency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].

Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes. This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and hospital effeciency. It is important to note that this paper studies the impact of an batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.

The remainder of this paper is structured as follows. The next section describes the data source and outlines our baseline sample. The empirical strategy and its accompanying identifying assumptions are laid out in Section II. Section III presents the results. Section IV draws implications for batch-ordering policies. The last section concludes.

# Data and Definitions {#sec:2}

Our analytical lens is focused on the Emergency Department at the Mayo Clinic of Arizona, a distinguished tertiary care establishment. During our study's timeframe, the ED recorded an annual visitation of approximately 41,000 patients. The department is singularly staffed by board-eligible or board-certified emergency physicians, abstaining from the services of nurse practitioners or physician assistants. A notable observation was that residents in rotation oversaw a low fraction, roughly 10%, of the patient volume. Comprehensive patient data, encompassing demographics, chief complaints, vital signs, emergency severity, length of stay, and resource utilization metrics, were meticulously logged during the study period.

## Sample Construction

We identified three distinct forms of test batching:

- Lab/Image Batch: The first test ordered for a patient was either an imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an additional imaging or lab test order within 5 minutes.

- Image/Image Batch: The first test ordered for a patient was an imaging test (CT scan, X-ray, Ultrasound), followed by an additional imaging test order within 5 minutes.

- Any Batch: This refers to instances where either a Lab/Image Batch or an Image/Image Batch occurred.

We conducted a sensitivity analysis for this definition, documented in Appendix 2. Table 1 highlights the considerable variation in patient complexity between those who had their tests batch ordered versus those where tests were ordered sequentially.

## Statistics

Table 1 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered either individually (Sequential) or as part of a group (Batch) during an ED encounter from our baseline sample. Around 43% of patient encounters involved diagnostic tests ordered as part of a batch, the majority of which (60%) included two tests.

Differences were noted between the two cohorts. Patients who had batched tests ordered tended to have a longer ED length of stay, a lower Emergency Severity Index (ESI), and were generally older than those who had their tests ordered individually. Patients in the batch order group were also more likely to present with symptoms at triage, including tachycardia, tachypnea, fever, and hypotension. A deeper dive into the ESI data reveals a higher proportion of patients with more urgent conditions (ESI 1 and 2) in the batched group than in the sequential group. This finding might be indicative of the complexity and severity of the conditions of patients who necessitate batch testing.

Age distributions between the two groups demonstrate a greater proportion of patients aged 65 and above in the batch group, suggesting that older patients are more likely to have multiple tests ordered at once, possibly due to the presence of multiple comorbidities. The batched group showed a longer average ED length of stay compared to the sequential group. This extended stay may be reflective of the additional time required to conduct and evaluate the results of multiple tests. Symptoms presentation at triage also differed between the two groups, with the batched group being more likely to present with tachypnea, fever, and hypotension. However, tachycardia was found to be similar across both groups.

\begin{table}[htbp]
\centering
\tiny
\caption{Summary Statistics for Sequential vs. Batched Test Groups}
\begin{tabular}{lcccc}
\hline
Variable & \multicolumn{1}{c}{Overall Visits, N = 41,197\textsuperscript{1}} & \multicolumn{1}{c}{Sequential Arm, N = 23,4331} & \multicolumn{1}{c}{Batching Arm, N = 17,7641} & \multicolumn{1}{c}{p-value\textsuperscript{2}} \\
\hline
\textbf{ESI} & & & & $<$0.001 \\
\hspace{0.5cm} 1 & 474 (1.2\%) & 130 (0.6\%) & 344 (1.9\%) & \\
\hspace{0.5cm} 2 & 13,911 (34\%) & 6,502 (28\%) & 7,409 (42\%) & \\
\hspace{0.5cm} 3 & 23,655 (57\%) & 14,350 (61\%) & 9,305 (52\%) & \\
\hspace{0.5cm} 4 & 3,123 (7.6\%) & 2,422 (10\%) & 701 (3.9\%) & \\
\hspace{0.5cm} 5 & 34 ($<$0.1\%) & 29 (0.1\%) & 5 ($<$0.1\%) & \\
\hline
\textbf{Age} & & & & $<$0.001 \\
\hspace{0.5cm} $<$20 & 790 (1.9\%) & 605 (2.6\%) & 185 (1.0\%) & \\
\hspace{0.5cm} 20-45 & 9,412 (23\%) & 6,242 (27\%) & 3,170 (18\%) & \\
\hspace{0.5cm} 45-65 & 12,826 (31\%) & 7,393 (32\%) & 5,433 (31\%) & \\
\hspace{0.5cm} 65+ & 18,169 (44\%) & 9,193 (39\%) & 8,976 (51\%) & \\
\hline
\textbf{ED LOS} & & & & $<$0.001 \\
\hspace{0.5cm} Mean (SD) & 270 (382) & 258 (177) & 286 (545) & \\
\hline
\textbf{72hr Return} & 1,482 (3.6\%) & 957 (4.1\%) & 525 (3.0\%) & $<$0.001 \\
\hline
\textbf{Number of Tests} & & & & $<$0.001 \\
\hspace{0.5cm} 1 & 12,961 (31\%) & 11,937 (51\%) & 1,024 (5.8\%) & \\
\hspace{0.5cm} 2 & 19,112 (46\%) & 8,457 (36\%) & 10,655 (60\%) & \\
\hspace{0.5cm} 3 & 7,688 (19\%) & 2,581 (11\%) & 5,107 (29\%) & \\
\hspace{0.5cm} 4 & 1,342 (3.3\%) & 432 (1.8\%) & 910 (5.1\%) & \\
\hspace{0.5cm} 5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
\hline
\textbf{Disposition} & & & & $<$0.001 \\
\hspace{0.5cm} Discharge & 25,302 (61\%) & 15,922 (68\%) & 9,380 (53\%) & \\
\hspace{0.5cm} Admit & 9,097 (22\%) & 4,209 (18\%) & 4,888 (28\%) & \\
\hspace{0.5cm} Other & 6,798 (17\%) & 3,302 (14\%) & 3,496 (20\%) & \\
\hline
\textbf{Tachycardic} & 8,115 (20\%) & 4,562 (19\%) & 3,553 (20\%) & 0.18 \\
\hline
\textbf{Tachypneic} & 3,822 (9.3\%) & 1,669 (7.1\%) & 2,153 (12\%) & $<$0.001 \\
\hline
\textbf{Febrile} & 969 (2.4\%) & 353 (1.5\%) & 616 (3.5\%) & $<$0.001 \\
\hline
\textbf{Hypotensive} & 651 (1.6\%) & 305 (1.3\%) & 346 (1.9\%) & $<$0.001 \\
\hline
\textbf{Gender} & & & & $<$0.001 \\
\hspace{0.5cm} Female & 22,105 (54\%) & 12,986 (55\%) & 9,119 (51\%) & \\
\hspace{0.5cm} Male & 19,092 (46\%) & 10,447 (45\%) & 8,645 (49\%) & \\
\hline
\textbf{Race} & & & & $<$0.001 \\
\hspace{0.5cm} White & 36,554 (89\%) & 20,665 (88\%) & 15,889 (89\%) & \\
\hspace{0.5cm} Black & 1,682 (4.1\%) & 974 (4.2\%) & 708 (4.0\%) & \\
\hspace{0.5cm} Asian & 1,217 (3.0\%) & 710 (3.0\%) & 507 (2.9\%) & \\
\hspace{0.5cm} Other & 737 (1.8\%) & 449 (1.9\%) & 288 (1.6\%) & \\
\hspace{0.5cm} Native & 538 (1.3\%) & 340 (1.5\%) & 198 (1.1\%) & \\
\hspace{0.5cm} Unknown & 469 (1.1\%) & 295 (1.3\%) & 174 (1.0\%) & \\
\hline
\hline
\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
\end{tabular}
\end{table}

![Frequency of Sample Cases and Batches](../Manuscript/figures/fig1.png){width=40%, height=30%}

## Institutional Details on Patient-Physician Assignment

Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED. In most EDs, however, physicians have discretion in picking their patients. In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm (Traub et al., 2016), which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Exhibit 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Wald Test Results}
    \begin{tabular}{lcc}
        \toprule
        Chief Complaints & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 1.37 & 0.106 \\
        Back or Flank Pain & 1.00 & 0.451 \\
        Chest Pain & 0.98 & 0.476 \\
        Extremity Complaints & 0.97 & 0.495 \\
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
        Gastrointestinal Issues & 0.98 & 0.480 \\
        Neurological Issue & 0.75 & 0.793 \\
        Shortness of Breath & 1.23 & 0.199 \\
        Skin Complaints & 1.05 & 0.388 \\
        Upper Respiratory Symptoms & 1.21 & 0.218 \\
        \midrule
        Emergency Severity & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 1.09 & 0.346 \\
        ESI 3, 4, or 5 & 1.24 & 0.196 \\
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\tiny
\item Table [blank] reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.
\end{tablenotes}
\end{table}

# Empirical Strategy {#sec:3}

Our study employs the ``judges design", a quasi-random assignment strategy widely used in literature to investigate causal effects. Typically, this design exploits variations in sentencing leniency among judges working in the same court. We adopt a similar approach by examining batching variations across physicians working in the same emergency department. This method enables us to identify the causal impact of being assigned to different types of physicians with differing batch-ordering tendencies.

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \varepsilon_{i,t}
\end{equation}

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on data $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or  seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. As stated earlier, these controls are what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

In step two, the leniency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:

\begin{equation}
Tendency_{i,j}^{phys} =
\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{e}_{i',k}
\end{equation}

where $\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure 2 graphs the histogram of the leniency measure along the x-axis and the left y-axis. A local-linear regression of the fitted probability of batched ordering on tendency after residualizing is overlaid and displayed on the right y-axis. Table 2 presents the same ``first stage" in a regression table: being assigned to a 10 pp more lenient physician is associated with a 17 pp increase in the likelihood of being prescribed an opioid in the ED. The F-statistic is 25 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rarely prescribed diagnoses.

To estimate the reduced-form effects of being treated by a batch-preferring physician, we estimate the following equation:

\begin{equation}
Y_i = \mu_0 + \mu_1 Tendency_{i,j}^{phys} + \gamma X_i + \nu_i
\end{equation}

This reduced form will allow us to check that our instrument is a strong instrument. To study the effects of test batching in the ED on an outcome $Y_i$, we estimate the following 2SLS equations using our baseline sample:

\begin{equation}
Y_i = \beta_0 + \beta_1 Batched_i + \theta X_i + \varepsilon_i
\end{equation}

\begin{equation}
Batched_i = \delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i + \nu_i
\end{equation}

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization.

\begin{table}[htbp]
\centering
\caption{First Stage Regression Results}
\begin{tabular}{lccc}
\toprule
 & Model 1 & Model 2 & Model 3 \\
\midrule
Lab + Image Batching Tendency & 0.99*** & 1.02*** & 1.02*** \\
 & (0.01) & (0.01) & (0.01) \\
\\
F statistic (proj model) & 13206.68 & 6702.01 & 6655.86  \\
F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
Image + Image Batching Tendency & 0.92*** & 0.94*** & 0.94*** \\
 & (0.02) & (0.04) & (0.04) \\
\\
F statistic (proj model) & 1645.35 & 577.97 & 577.09  \\
F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
\midrule
\midrule
Num. obs. & 41197 & 41197 & 41197 \\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & No & Yes & Yes \\
Patient observables? & No & No & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Notes:}  Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$ We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

### Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, if assignment of patients to ED physicians is random, conditional on seasonality, shift, and diagnosis (``conditional independence"). One can test for whether batch-likely physicians are routinely assigned patients with particular characteristics. Figure 4 shows evidence of balance on observables. The left panel is the output of a regression of Prescribed on observables including patient demographics, priority groups, and ED diagnoses. Unsurprisingly, many observables predict length of stay and 72 hour return. The right panel displays the coefficients from a regression of physician tendency instrument on the same set of observables. With so many coefficients, we fail to reject that the instrument violates conditional independence (the joint F-stat drops by two orders of magnitude, from 280 to 2.4). In fact, some of the variables that are significant in the second column go in the opposite direction of the decision to batch

The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch. In Appendix Figure [XXX], residualizing for only seasonality and shift (left panel) and also diagnosis (right panel) in equation (1). This lends support to our assumption that assignment in the ED is nearly random at the seasonality, shift  level.

Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity. We reported a strong first stage (i.e., relevance) at the end of the previous Section. The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching. This is perhaps our strongest assumption and is at its core, untestable. However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings. First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher. Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest. The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis. Nevertheless, we take this assumption seriously and perform a placebo check in Section IIIB as well as various robustness checks in Section IIIE.

## Results

### Reduced-Form Results

In this section, we examine the causal effects of provider batching tendency. We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).

# Conclusion

\clearpage

::: {.APPENDIX latex="{General appendix}"}

\begin{longtable}{|p{5cm}|p{12cm}|}
\caption{Chief Complaints} \\
\hline
\textbf{Complaint Area} & \textbf{Complaints} \\
\hline
Abdominal Complaints & Abdominal Cramping, Abdominal Distention, Dyspepsia, Abdominal Pain, Ascites, Hernia, Abdominal Aortic Aneurysm, Abdominal Injury, Pancreatitis, Umbilical Hernia \\
\hline
Abnormal Test Results & Abnormal Lab, Abnormal Potassium, Abnormal Calcium, ECG Changes, Abnormal ECG, Abnormal Test Result, Blood Infection, Acute Renal Failure, Hypocalcemia, Chronic Renal Failure, Pulmonary Embolism, Abnormal X-ray, Hypoglycemic Unawareness, Elevated Blood Pressure, Abnormal Sodium, Hyperglycemia, Hyponatremia, Platelet Disorders, Anemia, Hypoglycemia, Hypertension, Hypotension, Abnormal Chest Imaging, Abnormal Oximetry, Abnormal Stress Test, Blood Sugar Problem, Hypocalcemia, Hyponatremia \\
\hline
Allergic Reaction & Allergic Reaction, Anaphylaxis \\
\hline
Back or Flank Pain & Back Pain, Back Problem, Flank Pain, Sciatica, Back Injury, Disc Disorder \\
\hline
Breast Complaints & Breast Mass, Breast Pain, Breast Problem, Breast Discharge, Breast Cancer, Breast Discharge, Breast Inflammation \\
\hline
Cardiac Arrhythmias & Atrial Fibrillation, Atrial Flutter, Cardiac Valve Problem, Bradycardia, Irregular Heart Beat, Palpitations, POTS, Ventricular Tachycardia, Rapid Heart Rate, Heart Problem, Cardiac Arrest, Congestive Heart Failure, Circulatory Problem, Transient Ischemic Attack, Ventricular Tachycardia \\
\hline
Chest Pain & Chest Injury, Chest Pain, Chest Wall Pain, Angina, Collarbone Injury, Rib Injury, Heart Pain \\
\hline
Dizziness / Lightheadedness / Syncope & Dizziness, Near Syncope, Syncope, Vertigo, Spells, Hypotension, Paroxysmal Positional Vertigo, Paroxysmal Positional Vertig \\
\hline
Ear Complaints & Cerumen Impaction, Ear Drainage, Ear Fullness, Ear Laceration, Ear Problem, Earache, Hearing Problem, Tinnitus, Ear Injury, Hearing Loss, Nasal Trauma \\
\hline
Epistaxis & Epistaxis, Epistaxis (Nose Bleed), Nose Problem \\
\hline
Exposures, Bites, and Envenomations & Animal Bite, Body Fluid Exposure, Chemical Exposure, Poisoning, Exposure to STD, Insect Bite, Smoke Inhalation, Radiation, Snake Bite, Toxic Inhalation \\
\hline
Extremity Complaints & Ankle Injury, Ankle Pain, Arm Injury, Arm Pain, Cold Extremity, Arm Swelling, Arthritis, Elbow Injury, Elbow Pain, Pseudogout, Extremity Pain, Extremity Weakness, Finger Injury, Hip Injury, Extremity Weakness, Finger Injury, Finger Pain, Dislocation, Foot Infection, Foot Injury, Foot Numbness, Foot Pain, Foot Swelling, Foot Ulcer, Foot Wound Check, Hand Injury, Hand Pain \\
\hline
Eye Complaints & Blurred Vision, Decreased Visual Acuity, Diplopia, Detached Retina, Eye Drainage, Eye Exposure, Eye Pain, Eye Problem, Eye Swelling, Eye Trauma, Foreign Body Eye, Flashes / Light, Loss of Vision, Red Eye, Visual Field Change, Eyelid Problem, Itchy Eye, Eye Exam, Burning Eyes, Eye Twitching, Eyelid/brow Lift Evaluation, Strabismus, Glaucoma, Spots / Floaters \\
\hline
Falls, Motor Vehicle Crashes, Assaults, and Trauma & Assault Victim, Concussion, Facial Injury, Fall, Nasal Trauma, Head Injury, Head Laceration, Motor Vehicle Crash, Puncture Wound, Sexual Assault, Trauma, Domestic Violence, Gun Shot Wound, Work Related Injury, Motorcycle Crash, Injury, Bicycle Accident, Near Drowning, Lip Laceration \\
\hline
Fatigue and Weakness & Difficulty Walking, Fatigue, Gait Problem, Weakness-Generalized, Chronic Fatique, Weakness- Generalized \\
\hline
Fevers, Sweats or Chills & Chills, Diaphoresis, Fever, Night Sweats, Diaphoretic, Diapohresis, Hoarseness, Laryngitis \\
\hline
Foreign Body & Food Bolus, Foreign Body, Foreign Body in Ear, Foreign Body in Skin, Foreign Body in Vagina, Swallowed Foreign Body, Foreign Body in Nose, Foreign Body, FB eye, Foreign Body in Rectum \\
\hline
Gastrointestinal Issues & Anal Fissure, Black or Bloody Stool, Constipation, GERD, Anal Fistula, Diarrhea, Dysphagia, Fecal Impaction, Fistula Follow Up, GIbleeding, GI Problem, Hemorrhoids, Morning Sickness, Nausea, Ostomy Care, Rectal Bleeding, Rectal Pain, Vomiting, Vomiting Blood, Vomiting During Pregnancy, GI Bleeding, Fecal Incontinence, Bloated, Hematochezia, Urine Leakage, Heartburn, Rectal Discharge, Urolithiasis, Ulcerative Colitis, Irritable Bowel Syndrome, Rectal Prolapse, Fistula Evaluation, Rectal Problems, Perianal Abscess, Fisula Evaluation, Stoma Dysfunction \\
\hline
Genital Complaints & Groin Burn, Groin Pain, Groin Swelling, Inguinal Hernia, Menstrual Problem, Pelvic Pain, Penis Pain, Priapism, Testicle Pain, Menorrhagia, Vaginal Bleed, Vaginal Bleeding, Vaginal Itching, Bartholin's Cyst, Genital Warts, Groin Injury, Vaginal Bleeding-Pregnant, Vag Bleed Pregnant, Female Genital Issue, Penis Injury, Vaginal Discharge, Vaginal Pain, Erectile Dysfunction, Vaginal Prolapse, Urethral Stricture, Penile Discharge, Menorrhagia, Gynecologic Exam, Menstrual Problem, Vaginitis/Bacterial Vaginosis, Ovarian Cyst, Vaginitis / Bacterial Vaginosi \\
\hline
Medical Device or Treatment Issue & Cast Problem, Device Check, Dressing Change, Feeding Tube, AICD Problem, Insulin Pump Visit, Gastrostomy Tube Change, Medication Reaction, Shunt, Appliance Removal, Tube Problem, Urinary Catheter Change, Vascular Access Problem, Enteral Nutrition Evaluation, Device Malfunction, Pacemaker Problem, Remova /  Exchange Catheter, Drain Removal, Outpatient Infusion, Treatment, Heart Assist Device, Stoma Dysfunction, Tracheostomy Tube Change, Ureteral Stent Exchange \\
\hline
Medication Request & Immunizations, Infusion / Injection Administration, IV Medication, Infusion/ Injection Administ, Med Refill, Medication Visit, Pain Management, Blood Product Administration, Labs Only, Tetanus (Td \& Tdap), Wound Care \\
\hline
Neurological Issue & Altered Mental Status, Cognitive Concerns, Facial Droop, Pre Syncope, Focal Weakness, Headache, Memory Loss, Migraine, Dementia, Dysphasia, Neuro Problem, Numbness, Paralysis, Seizures, Slurred Speech, Spasms, Stroke Like Symptoms, Tingling, Tremors, Trigeminal Neuralgia, Unable to Speak, Seizure Disorder, Insomnia, Parkinson's Disease, Loss of Consciousness, Neuropathy, Ataxia, Unable to speak, Peripheral Neuropathy, Stroke, Cerebrovascular Accident, Speech Problem, Acute Neurological Problem, Flashes, Light, Unresponsive, Multiple Sclerosis, Parkinson's Disease, Febrile Seizure, Paresthesia, Peripheral Neuropathy, Hydrocephalus, Spasticity, Neuroendocrine Tumor \\
\hline
Other & Dehydration, Fisula Evaluation, Follow-Up, Illness, Letter for School/Work, Aneurysm, Lung Eval, Error, Mass, Oral Swelling, Other, Advice Only, Deformity, Electric Shock, Personal Problem, Shaking, Swelling, Swollen Glands, Adenopathy, Adrenal Problem, Thrombophilia, Weight Gain, Weight Loss, Hiccups, , Chemo Related Symptoms, Hot Flashes, Follow-up, Non Healing Wound, (Other), Mouth Injury, Xerostomia, Prostate Check, Suture / Staple Removal, Wellness, Voice Changes, Vital Sign Check, Coagulation Disorder, Cold Exposure, Consult, Dental Problem, Tetanus (Td \& Tdap), Infusion/ Injection Administ, Tracheostomy Tube Change, Medical Information, Neutropenic Fever, Infection, Leukemia, Heat Exposure, Poor Appetite, Gingivitis, Pre-op Exam, gingivitis, Loss of appetite, Failure To Thrive, Referral, Lymphoma, Hot Flashes, Neutropenia, Radiation, Ingestion, TB Test, Fussy, Lupus, Toxic Inhalation, Lung Screening, Leakage/Loss of Fluid, Liver Eval, Hepatic Cancer, Lung Mass, Venous Thromboembolic Disease, Insulin Pump Visit, Preventive Visit, Avulsion, Peripheral Edema, Hypoglycemic Unawareness, Immobility, Giant Cell Arteritis, Polydipsia, Platelet Disorders, Post-procedure, Lung Follow-up, Poisoning, Injections, POTS, Insulin Reaction, Liver Transplant, Labs Only \\
\hline
Other Pain & Dental Pain, Facial Pain, Generalized Body Aches, Myalgia, Dental Injury, Jaw Pain, Muscle Pain, Neck Pain, Pain, Sickle Cell Pain Crisis, Paresthesia, Torticollis, Chronic Pain, Cancer Pain, Incisional Pain, Bone Pain, Tailbone Pain, Gout, Muscle pain/Weakness, Pseudogout \\
\hline
Post-Op Issue & Post-Op, Post-Procedure, Post-Op Problem, Post-op, Post-Op Issue, Wound Dehiscence, Post-op Problems, Post-op Problem \\
\hline
Psychiatric Complaints & Anxiety, Auditory Hallucinations, Depression, Panic Attack, Homicidal, PTSD (Post-Traumatic Stress, Delusional, Fussy, Paranoia, Suicide Attempt, Hallucinations, Manic Behavior, Eating Disorder, Suicidal, Agitation, Psychiatric Evaluation, Aggressive Behavior, Mental Health Problem, Inappropriate Words \\
\hline
Shortness of Breath & Airway Obstruction, Aspiration, Pain With Breathing, Near Drowning, Respiratory Distress, Shortness of Breath, Wheezing, Increased Work Of Breathing, Difficulty Breathing, Choking, Oxygen Dependence, Hyperventilating, Orthopnea \\
\hline
Skin Complaints & Abrasion, Abscess, Bleeding/Bruising, Blister, Angioedema, Lip Laceration, Burn, Cellulitis, Cyst, Drainage from Incision, Disturb of Skin Sens, Edema, Extremity Laceration, Facial Burn, Cyanosis, Impetigo, Facial Laceration, Facial Swelling, Finger Laceration, Leg Rash, Herpes Zoster, Hives, Itching, Jaundice, Diabetic Ulcer, Diabetic Wound, Laceration, Mouth Lesions, Non-Healing Wound, Rash, Recurrent Skin Infections, Skin Problem, Sore, Scabies, Suture \textbackslash Staple Removal, Wound Check, Wound Infection, Lesion, Skin Check, Minor Skin Infection, Skin Ulcer, Skin Discoloration, Sunburn, Head Lice, Scabies, Fungal Infection, Leg Rash, Impetigo \\
\hline
Substance Abuse Issues & Alcohol Intoxication, Alcohol Problem, Withdrawal, Drug Overdose, Drug / Alcohol Dependency, Addiction Problem, Addiction Assessment, Delirium Tremens (DTS) \\
\hline
Upper Respiratory Symptoms & Congestion, Cough, Coughing Up Blood, Flu Symptoms, Enlarged Tonsils, Peritonsillar Abscess, Nasal Congestion, Sinus Symptoms, Sinusitis, Sore Throat, Hoarseness, Throat Problem, Upper Respiratory Infection, Influenza, Laryngitis, Respiratory Arrest, Pneumonia, Pleural Effusion, Asthma, Croup, URI, Peritonsillar Abscess \\
\hline
Pregnancy Related & Pregnancy Problem, Miscarriage, Contractions, Ectopic Pregnancy, Laboring, Possible Pregnancy, Pregnancy Related \\
\hline
Renal & Av Fistula, Kidney Transplant, Elevated Serum Creatinine, End-Stage Liver Disease, Hemodialysis Access, Nephritis, Ureteral Stent Exchange \\
\hline
Urinary Complaints & Bladder Problem, Blood in Urine, Cystitis, Difficulty Urinating, Dysuria, Gross Hematuria, Painful Urination, Urinary Frequency, Urinary Symptom, Urinary Incontinence, Urinary Problem, Urinary Retention, Slowing Urinary Stream, Urinary Tract Infection, Urinary Urgency, Voiding Dysfunction, Hesitancy Urinary \\
\hline
\end{longtable}

\begin{table}[ht]
\centering
\caption{Regression Results by Chief Complaint} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & -0.72** & 0.03 \\ 
  2 & Upper Respiratory Symptoms & 0.426** & 0.02 \\ 
  3 & Abnormal Test Results & -0.067 & 0.84 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.423* & 0.10 \\ 
  5 & Shortness of Breath & -0.065 & 0.58 \\ 
  6 & Gastrointestinal Issues & 0.071 & 0.67 \\ 
  7 & Chest Pain & 0.107 & 0.68 \\ 
  8 & Dizziness/Lightheadedness/Syncope & -0.163 & 0.32 \\ 
  9 & Neurological Issue & -0.353 & 0.25 \\ 
  10 & Fatigue and Weakness & -0.112 & 0.49 \\ 
  11 & Back or Flank Pain & -0.602** & 0.02 \\ 
  12 & Abdominal Complaints & -0.105 & 0.13 \\ 
  13 & Fevers, Sweats or Chills & 0.01 & 0.95 \\ 
  14 & Skin Complaints & 1.442 & 0.35 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Regression Results by Chief Complaint (2)} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & 0.176 & 0.50 \\ 
  2 & Upper Respiratory Symptoms & 0.332*** & 0.00 \\ 
  3 & Abnormal Test Results & 0.695 & 0.14 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.247 & 0.20 \\ 
  5 & Shortness of Breath & -0.253** & 0.01 \\ 
  6 & Gastrointestinal Issues & 0.421*** & 0.00 \\ 
  7 & Chest Pain & -0.531*** & 0.00 \\ 
  8 & Dizziness/Lightheadedness/Syncope & 0.075 & 0.60 \\ 
  9 & Neurological Issue & -0.305 & 0.15 \\ 
  10 & Fatigue and Weakness & -0.125 & 0.40 \\ 
  11 & Back or Flank Pain & 0.288 & 0.13 \\ 
  12 & Abdominal Complaints & -0.023 & 0.66 \\ 
  13 & Fevers, Sweats or Chills & -0.044 & 0.71 \\ 
  14 & Skin Complaints & 0.328 & 0.35 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{First Stage over Subsamples} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & 0.761*** & 0.00 \\ 
  2 & Upper Respiratory Symptoms & 1.406*** & 0.00 \\ 
  3 & Abnormal Test Results & 1.036*** & 0.00 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.831*** & 0.00 \\ 
  5 & Shortness of Breath & 1.258*** & 0.00 \\ 
  6 & Gastrointestinal Issues & 1.043*** & 0.00 \\ 
  7 & Chest Pain & 0.866*** & 0.00 \\ 
  8 & Dizziness/Lightheadedness/Syncope & 1.668*** & 0.00 \\ 
  9 & Neurological Issue & 0.723*** & 0.00 \\ 
  10 & Fatigue and Weakness & 1.651*** & 0.00 \\ 
  11 & Back or Flank Pain & 0.752*** & 0.01 \\ 
  12 & Abdominal Complaints & 1.109*** & 0.00 \\ 
  13 & Fevers, Sweats or Chills & 1.23*** & 0.00 \\ 
  14 & Skin Complaints & 0.448* & 0.07 \\ 
   \hline
\end{tabular}
\end{table}
:::

### 2023-09-27 manuscript/manuscript.tex
starting to write:

--- 2023-09-27 manuscript/manuscript.Rmd
@@
----
-title: "To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED"
-authors:
-- familyname: Jameson
-  othernames: Jacob
-  affiliation: Interfaculty Initiative in Health Policy, Harvard University
-  email: jacobjameson@g.harvard.edu
-- familyname: Saghafian
-  othernames:  Soroush
-  affiliation: Kennedy School of Government, Harvard University 
-  email: soroush_saghafian@hks.harvard.edu
-- familyname: Huckman
-  othernames:  Robert
-  affiliation: Harvard Business School 
-  email: rhuckman@hbs.edu
-- familyname: Hodgson
-  othernames:  Nicole
-  affiliation: Mayo Clinic
-  email: Hodgson.Nicole@mayo.edu
-keywords:
-  - Emergency Department, Operational Effeciency, Diagnostic Testing
-abstract: |
-  This paper focuses on analyzing sequential versus batched testing strategies in an emergency department (ED) at Mayo Clinic Arizona, with respect to their associations impacts on patient length of stay, hospital readmission, and healthcare resource utilization.
-A theoretical model was developed and tested to identify patient and hospital features that may influence the decision to batch or sequentially order tests, such as patient complexity, physician experience, occupancy level and complexity.
-Finally, we performed retrospective analysis of ED operational data to investigate the impact of batching on our key outcomes.
-The overall result was that batch ordering tests was associated with greater patient length of stay and resource utilization, even when we control for variation in patient complexity, physician experience, and hospital occupancy.
-acknowledgment: |
-  The authors would like to thank ..
-bibliography: references.bib
-blind: false
-output: rticles::informs_article
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{siunitx}
-  - \usepackage{tablefootnote}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \usepackage{natbib}
----
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
+%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
+%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
+%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[,,nonblindrev]{informs}
 
-# Introduction {#sec:1}
+\OneAndAHalfSpacedXI
+%%\OneAndAHalfSpacedXII % Current default line spacing
+%%\DoubleSpacedXII
+%%\DoubleSpacedXI
 
-Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization.
-Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events.
-Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality @bernstein2009.
+%% BEGIN MY ADDITIONS %%
+\usepackage{hyperref}
 
-One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests @naseim2015.
-A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests such as electrocardiograms (ECGs) and point-of-care (POC) testing.
-These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially.
-This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing.
-Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results @koch2018.
-Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier.
-Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged.
-Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements @saghafian2015, however, research surrounding test ordering strategies to improve the patient flow processes remains limited.
+\usepackage{booktabs}
+\usepackage{siunitx}
+\usepackage{tablefootnote}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\usepackage{natbib}
 
-In this paper, we use data from over 41,000 patient visits to the ED that occur during our study period to quantify the benefits and consequences of batching versus sequentially ordering diagnostic tests on patient length of stay, re-admission, and resource utilization.
-Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests.
-When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side.
-Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests.
-We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period.
-The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.
+%% END MY ADDITIONS %%
 
-We start by evaluating the reduced-form effects of ED provider batch ordering tendency on downstream patient outcomes and turnaround time.
-We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences.
-Being treated by a provider in the top decile of the leniency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].
+% Natbib setup for author-year style
+\usepackage{natbib}
+ \bibpunct[, ]{(}{)}{,}{a}{}{,}%
+ \def\bibfont{\small}%
+ \def\bibsep{\smallskipamount}%
+ \def\bibhang{24pt}%
+ \def\newblock{\ }%
+ \def\BIBand{and}%
 
-Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm.
-In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings.
-Furthermore, physicians exhibit wide variation in practice behavior in batch ordering, even within the same hospital, while following the same guidelines.
-Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.
+%% Setup of theorem styles. Outcomment only one.
+%% Preferred default is the first option.
+\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
+%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
+\ECRepeatTheorems
 
-In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes.
-This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and hospital effeciency.
-It is important to note that this paper studies the impact of an batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.
+%% Setup of the equation numbering system. Outcomment only one.
+%% Preferred default is the first option.
+\EquationsNumberedThrough    % Default: (1), (2), ...
+%\EquationsNumberedBySection % (1.1), (1.2), ...
 
-The remainder of this paper is structured as follows.
-The next section describes the data source and outlines our baseline sample.
-The empirical strategy and its accompanying identifying assumptions are laid out in Section II.
-Section III presents the results.
-Section IV draws implications for batch-ordering policies.
-The last section concludes.
+% For new submissions, leave this number blank.
+% For revisions, input the manuscript number assigned by the on-line
+% system along with a suffix ".Rx" where x is the revision number.
+\MANUSCRIPTNO{}
 
-# Data and Definitions {#sec:2}
+%%%%%%%%%%%%%%%%
+\begin{document}
+%%%%%%%%%%%%%%%%
 
-Our analytical lens is focused on the Emergency Department at the Mayo Clinic of Arizona, a distinguished tertiary care establishment.
-During our study's timeframe, the ED recorded an annual visitation of approximately 41,000 patients.
-The department is singularly staffed by board-eligible or board-certified emergency physicians, abstaining from the services of nurse practitioners or physician assistants.
-A notable observation was that residents in rotation oversaw a low fraction, roughly 10%, of the patient volume.
-Comprehensive patient data, encompassing demographics, chief complaints, vital signs, emergency severity, length of stay, and resource utilization metrics, were meticulously logged during the study period.
+% Outcomment only when entries are known. Otherwise leave as is and
+%   default values will be used.
+%\setcounter{page}{1}
+%\VOLUME{00}%
+%\NO{0}%
+%\MONTH{Xxxxx}% (month or a similar seasonal id)
+%\YEAR{0000}% e.g., 2005
+%\FIRSTPAGE{000}%
+%\LASTPAGE{000}%
+%\SHORTYEAR{00}% shortened year (two-digit)
+%\ISSUE{0000} %
+%\LONGFIRSTPAGE{0001} %
+%\DOI{10.1287/xxxx.0000.0000}%
 
-## Sample Construction
+% Author's names for the running heads
+% Sample depending on the number of authors;
+\RUNAUTHOR{%
+Jameson, Saghafian, Huckman
+ and Hodgson
+}
+% \RUNAUTHOR{Jones and Wilson}
+% \RUNAUTHOR{Jones, Miller, and Wilson}
+% \RUNAUTHOR{Jones et al.} % for four or more authors
+% Enter authors following the given pattern:
+%\RUNAUTHOR{}
+
+\RUNTITLE{To Batch or Not to Batch: Sequential vs.~Batched Testing
+Strategies in the ED}
+
+\TITLE{To Batch or Not to Batch: Sequential vs.~Batched Testing
+Strategies in the ED}
+
+\ARTICLEAUTHORS{%
+\AUTHOR{Jacob Jameson}
+\AFF{Interfaculty Initiative in Health Policy, Harvard
+University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}
+
+\AUTHOR{Soroush Saghafian}
+\AFF{Kennedy School of Government, Harvard
+University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}
+
+\AUTHOR{Robert Huckman}
+\AFF{Harvard Business
+School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}
+
+\AUTHOR{Nicole Hodgson}
+\AFF{Mayo
+Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}
+
+%
+}
+
+\ABSTRACT{This paper focuses on analyzing sequential versus batched
+testing strategies in an emergency department (ED) at Mayo Clinic
+Arizona, with respect to their associations impacts on patient length of
+stay, hospital readmission, and healthcare resource utilization. A
+theoretical model was developed and tested to identify patient and
+hospital features that may influence the decision to batch or
+sequentially order tests, such as patient complexity, physician
+experience, occupancy level and complexity. Finally, we performed
+retrospective analysis of ED operational data to investigate the impact
+of batching on our key outcomes. The overall result was that batch
+ordering tests was associated with greater patient length of stay and
+resource utilization, even when we control for variation in patient
+complexity, physician experience, and hospital occupancy.}
+
+\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
+Testing}
+
+\maketitle
+
+\hypertarget{sec:1}{%
+\section{Introduction}\label{sec:1}}
+
+Healthcare delivery, particularly in the emergency department (ED), is a
+delicate balance that involves ensuring optimal patient outcomes while
+optimizing resource utilization. Achieving these twin goals requires
+timely and accurate diagnosis, which in turn enables prompt and
+appropriate treatment, consequently improving patient prognosis and
+reducing the likelihood of adverse events. Furthermore, efficient
+patient discharge from the ED can help alleviate overcrowding, a severe
+issue with potential consequences including higher complication rates
+and increased mortality \citet{bernstein2009}.
+
+One important factor that can impact the speed and effectiveness of
+diagnosis in the ED is the availability and performance of diagnostic
+tests \citet{naseim2015}. A variety of diagnostic tests are used in the
+ED, including laboratory tests, imaging studies, and specialized tests
+such as electrocardiograms (ECGs) and point-of-care (POC) testing. These
+tests can provide valuable information about a patient's condition and
+help to guide treatment decisions.
+
+A critical question in this context pertains to whether physicians in
+the ED should batch order diagnostic tests or order them sequentially.
+This decision essentially represents a tradeoff between reducing patient
+length of stay and risk of over-testing. Over-testing, or performing
+unnecessary tests, can lead to increased costs, unnecessary patient
+anxiety, and potential harm from follow-up of false-positive results
+\citet{koch2018}. Conversely, keeping a patient for an extended time to
+perform all possible tests could lead to ED overcrowding, an issue
+associated with severe consequences, as mentioned earlier. Instead, what
+is needed is a reasonable balance between the number of diagnostic tests
+performed and the total time the patient is kept in the ED before either
+being admitted or discharged. Several studies have demonstrated that
+optimizing the ED patient flow process can result in significant
+improvements \citet{saghafian2015}, however, research surrounding test
+ordering strategies to improve the patient flow processes remains
+limited.
+
+In this paper, we use data from over 41,000 patient visits to the ED
+that occur during our study period to quantify the benefits and
+consequences of batching versus sequentially ordering diagnostic tests
+on patient length of stay, re-admission, and resource utilization. Our
+empirical strategy exploits random assignment of patients to ED
+physicians who differ in their propensity to batch-order diagnostic
+tests. When patients arrive at the ED, they are assigned to a physician
+based on availability, with no discretion on either side. Thus, patients
+who arrive at the ED at similar times are randomly assigned to
+physicians who vary in their willingness to batch order diagnostic
+tests. We measure physician tendency to batch using a leave-out,
+residualized measure based on all other patients the physician has seen
+in the ED in the study period. The tendency measure strongly predicts
+the ED test batch outcome but is uncorrelated with patient and ED visit
+characteristics.
+
+We start by evaluating the reduced-form effects of ED provider batch
+ordering tendency on downstream patient outcomes and turnaround time. We
+find that practice variation as captured by physician batch-ordering
+tendency has large and significant consequences. Being treated by a
+provider in the top decile of the leniency distribution, compared to
+being treated by someone in the bottom decile {[}INSERT RESULTS{]}.
+
+Because of the institutional features of the ED, our research design
+closely approximates an RCT that assigns patients to batch-ordering or
+sequential-ordering arm. In the ED, patients have no discretion over
+choosing providers, and in our specific ED, physicians have discretion
+over choosing patients, alleviating major selection issues present in
+other health care settings. Furthermore, physicians exhibit wide
+variation in practice behavior in batch ordering, even within the same
+hospital, while following the same guidelines. Finally,
+patient-physician interactions in the ED are typically well documented,
+short, and one-off, constraining physician decision-making to a more
+limited, better-observed choice set than present in settings such as
+specialty or primary care.
+
+In sum, exploiting practice variation in ED settings shuts down other
+(but not all) potential channels besides test batching that are present
+in other settings, determine length of stay, and impact patient
+outcomes. This approach allows us to move closer to identifying the
+causal impact of batch-ordering diagnostic tests on patient outcomes and
+hospital effeciency. It is important to note that this paper studies the
+impact of an batch-ordering through a batching decision requiring
+clinical judgment (within practice norms) rather than through specific
+hospital policies, differences in adherence to clinical practice
+guidelines, or substandard care.
+
+The remainder of this paper is structured as follows. The next section
+describes the data source and outlines our baseline sample. The
+empirical strategy and its accompanying identifying assumptions are laid
+out in Section II. Section III presents the results. Section IV draws
+implications for batch-ordering policies. The last section concludes.
+
+\hypertarget{sec:2}{%
+\section{Data and Definitions}\label{sec:2}}
+
+Our analytical lens is focused on the Emergency Department at the Mayo
+Clinic of Arizona, a distinguished tertiary care establishment. During
+our study's timeframe, the ED recorded an annual visitation of
+approximately 41,000 patients. The department is singularly staffed by
+board-eligible or board-certified emergency physicians, abstaining from
+the services of nurse practitioners or physician assistants. A notable
+observation was that residents in rotation oversaw a low fraction,
+roughly 10\%, of the patient volume. Comprehensive patient data,
+encompassing demographics, chief complaints, vital signs, emergency
+severity, length of stay, and resource utilization metrics, were
+meticulously logged during the study period.
+
+\hypertarget{sample-construction}{%
+\subsection{Sample Construction}\label{sample-construction}}
 
@@
 
-- Lab/Image Batch: The first test ordered for a patient was either an imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an additional imaging or lab test order within 5 minutes.
+\begin{itemize}
+\item
+  Lab/Image Batch: The first test ordered for a patient was either an
+  imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an
+  additional imaging or lab test order within 5 minutes.
+\item
+  Image/Image Batch: The first test ordered for a patient was an imaging
+  test (CT scan, X-ray, Ultrasound), followed by an additional imaging
+  test order within 5 minutes.
+\item
+  Any Batch: This refers to instances where either a Lab/Image Batch or
+  an Image/Image Batch occurred.
+\end{itemize}
 
-- Image/Image Batch: The first test ordered for a patient was an imaging test (CT scan, X-ray, Ultrasound), followed by an additional imaging test order within 5 minutes.
+We conducted a sensitivity analysis for this definition, documented in
+Appendix 2. Table 1 highlights the considerable variation in patient
+complexity between those who had their tests batch ordered versus those
+where tests were ordered sequentially.
 
-- Any Batch: This refers to instances where either a Lab/Image Batch or an Image/Image Batch occurred.
+\hypertarget{statistics}{%
+\subsection{Statistics}\label{statistics}}
 
-We conducted a sensitivity analysis for this definition, documented in Appendix 2.
-Table 1 highlights the considerable variation in patient complexity between those who had their tests batch ordered versus those where tests were ordered sequentially.
+Table 1 provides a comprehensive breakdown of the patient encounters
+involving diagnostic tests ordered either individually (Sequential) or
+as part of a group (Batch) during an ED encounter from our baseline
+sample. Around 43\% of patient encounters involved diagnostic tests
+ordered as part of a batch, the majority of which (60\%) included two
+tests.
 
-## Statistics
+Differences were noted between the two cohorts. Patients who had batched
+tests ordered tended to have a longer ED length of stay, a lower
+Emergency Severity Index (ESI), and were generally older than those who
+had their tests ordered individually. Patients in the batch order group
+were also more likely to present with symptoms at triage, including
+tachycardia, tachypnea, fever, and hypotension. A deeper dive into the
+ESI data reveals a higher proportion of patients with more urgent
+conditions (ESI 1 and 2) in the batched group than in the sequential
+group. This finding might be indicative of the complexity and severity
+of the conditions of patients who necessitate batch testing.
 
-Table 1 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered either individually (Sequential) or as part of a group (Batch) during an ED encounter from our baseline sample.
-Around 43% of patient encounters involved diagnostic tests ordered as part of a batch, the majority of which (60%) included two tests.
-
-Differences were noted between the two cohorts.
-Patients who had batched tests ordered tended to have a longer ED length of stay, a lower Emergency Severity Index (ESI), and were generally older than those who had their tests ordered individually.
-Patients in the batch order group were also more likely to present with symptoms at triage, including tachycardia, tachypnea, fever, and hypotension.
-A deeper dive into the ESI data reveals a higher proportion of patients with more urgent conditions (ESI 1 and 2) in the batched group than in the sequential group.
-This finding might be indicative of the complexity and severity of the conditions of patients who necessitate batch testing.
-
-Age distributions between the two groups demonstrate a greater proportion of patients aged 65 and above in the batch group, suggesting that older patients are more likely to have multiple tests ordered at once, possibly due to the presence of multiple comorbidities.
-The batched group showed a longer average ED length of stay compared to the sequential group.
-This extended stay may be reflective of the additional time required to conduct and evaluate the results of multiple tests.
-Symptoms presentation at triage also differed between the two groups, with the batched group being more likely to present with tachypnea, fever, and hypotension.
-However, tachycardia was found to be similar across both groups.
+Age distributions between the two groups demonstrate a greater
+proportion of patients aged 65 and above in the batch group, suggesting
+that older patients are more likely to have multiple tests ordered at
+once, possibly due to the presence of multiple comorbidities. The
+batched group showed a longer average ED length of stay compared to the
+sequential group. This extended stay may be reflective of the additional
+time required to conduct and evaluate the results of multiple tests.
+Symptoms presentation at triage also differed between the two groups,
+with the batched group being more likely to present with tachypnea,
+fever, and hypotension. However, tachycardia was found to be similar
+across both groups.
 
@@
 
-![Frequency of Sample Cases and Batches](../Manuscript/figures/fig1.png){width=40%, height=30%}
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth,height=0.3\textheight]{../Manuscript/figures/fig1.png}
+\caption{Frequency of Sample Cases and Batches}
+\end{figure}
 
-## Institutional Details on Patient-Physician Assignment
+\hypertarget{institutional-details-on-patient-physician-assignment}{%
+\subsection{Institutional Details on Patient-Physician
+Assignment}\label{institutional-details-on-patient-physician-assignment}}
 
-Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED.
-In most EDs, however, physicians have discretion in picking their patients.
-In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm (Traub et al., 2016), which removes potential selection bias concerns for our analyses.
-In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Exhibit 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+Contrary to most healthcare settings where patients exhibit choice, they
+are predominantly passive in their physician assignment in the ED. In
+most EDs, however, physicians have discretion in picking their patients.
+In contrast, patients arriving at the Mayo Clinic ED are randomly
+assigned to physicians via a rotational patient assignment algorithm
+(Traub et al., 2016), which removes potential selection bias concerns
+for our analyses. In essence, barring arrival time and shift-level
+variation, the physician-to-patient matching can be deemed random.
+Exhibit 2 displays that patient encounters (regarding chief complaints
+and emergency severity) are equitably distributed across physicians
+within our study's cohort.
 
@@
 
-# Empirical Strategy {#sec:3}
+\hypertarget{sec:3}{%
+\section{Empirical Strategy}\label{sec:3}}
 
-Our study employs the ``judges design", a quasi-random assignment strategy widely used in literature to investigate causal effects.
-Typically, this design exploits variations in sentencing leniency among judges working in the same court.
-We adopt a similar approach by examining batching variations across physicians working in the same emergency department.
-This method enables us to identify the causal impact of being assigned to different types of physicians with differing batch-ordering tendencies.
+Our study employs the ``judges design'', a quasi-random assignment
+strategy widely used in literature to investigate causal effects.
+Typically, this design exploits variations in sentencing leniency among
+judges working in the same court. We adopt a similar approach by
+examining batching variations across physicians working in the same
+emergency department. This method enables us to identify the causal
+impact of being assigned to different types of physicians with differing
+batch-ordering tendencies.
 
-To measure physician batch tendency, we use the physician's residualized leave-out average batch rate.
-This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways.
-First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.
+To measure physician batch tendency, we use the physician's residualized
+leave-out average batch rate. This measure is derived from two steps
+following the approaches taken by \citet{doyle2015measuring},
+\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
+obtain residuals from a regression model, which includes all ED
+encounters in our sample period.
 
@@
 
-Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on data $t$.
-Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g.
-initiatives to eliminate excess testing) or  seasonality in ED visits.
-We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$.
-As stated earlier, these controls are what is required for our quasi-random assignment assumption.
-Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.
+Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
+\(i\) had their diagnostic tests batched on encounter that took place on
+data \(t\). Fixed effects include year-month fixed effects,
+\(\alpha_{ym}\), to control for time and seasonal variation in batching,
+such as hospital-specific policies (e.g.~initiatives to eliminate excess
+testing) or seasonality in ED visits. We also control for
+``shift-level'' variations that include both physician scheduling and
+patient arrival with day of week-time of day fixed effects,
+\(\alpha_{dt}\). As stated earlier, these controls are what is required
+for our quasi-random assignment assumption. Under the assumption that we
+have captured the observables under which quasi-random assignment occurs
+in the ED, the unexplained variation-- the physician's contribution--
+resides in the error term, \(\varepsilon_{i,t}\).
 
-In step two, the leniency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:
+In step two, the leniency measure for patient \(i\) seen by physician
+\(j\) is computed as the average residual across all other patients seen
+by the physician that year:
 
@@
 
-where $\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$.
-This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument.
-The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.
+where \(\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}\) is the
+residual from equation (1); \(J\) is the set of all ED encounters
+treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
+number of cases that physician has seen that year, excluding patient
+\(i\). This leave-out mean eliminates the mechanical bias that stems
+from patient \(i\)'s own case entering into the instrument. The measure
+is interpreted as the average (leave-out) batch rate of patient \(i\)'s
+physician, relative to other physicians in that hospital-year-month,
+hospital-day of week-time of day.
 
-We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure 2 graphs the histogram of the leniency measure along the x-axis and the left y-axis.
-A local-linear regression of the fitted probability of batched ordering on tendency after residualizing is overlaid and displayed on the right y-axis.
-Table 2 presents the same ``first stage" in a regression table: being assigned to a 10 pp more lenient physician is associated with a 17 pp increase in the likelihood of being prescribed an opioid in the ED.
-The F-statistic is 25 when all controls and fixed effects are included.
-The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rarely prescribed diagnoses.
+We document that the Mayo Clinic ED physicians exhibit wide, systematic
+variation in their propensity to batch order diagnostic tests. Figure 2
+graphs the histogram of the leniency measure along the x-axis and the
+left y-axis. A local-linear regression of the fitted probability of
+batched ordering on tendency after residualizing is overlaid and
+displayed on the right y-axis. Table 2 presents the same ``first stage''
+in a regression table: being assigned to a 10 pp more lenient physician
+is associated with a 17 pp increase in the likelihood of being
+prescribed an opioid in the ED. The F-statistic is 25 when all controls
+and fixed effects are included. The coefficient is greater than one
+because all emergency visits are used to construct the leniency
+instrument, while the first stage is calculated using the baseline
+sample only, which excludes the rarely prescribed diagnoses.
 
-To estimate the reduced-form effects of being treated by a batch-preferring physician, we estimate the following equation:
+To estimate the reduced-form effects of being treated by a
+batch-preferring physician, we estimate the following equation:
 
@@
 
-This reduced form will allow us to check that our instrument is a strong instrument.
-To study the effects of test batching in the ED on an outcome $Y_i$, we estimate the following 2SLS equations using our baseline sample:
+This reduced form will allow us to check that our instrument is a strong
+instrument. To study the effects of test batching in the ED on an
+outcome \(Y_i\), we estimate the following 2SLS equations using our
+baseline sample:
 
@@
 
-Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization.
+Where \(Y_i\) represents our main outcomes of interest: length of stay,
+72 hour readmission, and resource utilization.
 
@@
 
-and $X_i$ is the same as in the reduced-form approach.
-$Batched_i$ variable suffers from potential endogeneity concerns.
-For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay.
-Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$ We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.
+and \(X_i\) is the same as in the reduced-form approach. \(Batched_i\)
+variable suffers from potential endogeneity concerns. For example,
+injury severity may be unobserved and correlated with need to run
+multiple tests, which in turn also affects length of stay. Hence, we
+instrument \(Batched_i\) with the assigned physician \(j\)'s underlying
+tendency to batch, \(Tendency_{i,j}^{phys}\) We cluster robust standard
+errors at the physician level to account for the assignment process of
+patients to physicians.
 
-### Identifying Assumptions
+\hypertarget{identifying-assumptions}{%
+\subsubsection{Identifying Assumptions}\label{identifying-assumptions}}
 
-The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, if assignment of patients to ED physicians is random, conditional on seasonality, shift, and diagnosis (``conditional independence").
-One can test for whether batch-likely physicians are routinely assigned patients with particular characteristics.
-Figure 4 shows evidence of balance on observables.
-The left panel is the output of a regression of Prescribed on observables including patient demographics, priority groups, and ED diagnoses.
-Unsurprisingly, many observables predict length of stay and 72 hour return.
-The right panel displays the coefficients from a regression of physician tendency instrument on the same set of observables.
-With so many coefficients, we fail to reject that the instrument violates conditional independence (the joint F-stat drops by two orders of magnitude, from 280 to 2.4).
-In fact, some of the variables that are significant in the second column go in the opposite direction of the decision to batch
+The reduced-form approach delivers an unbiased estimate of the causal
+effect of being treated by a higher tendency to batch physician, if
+assignment of patients to ED physicians is random, conditional on
+seasonality, shift, and diagnosis (``conditional independence''). One
+can test for whether batch-likely physicians are routinely assigned
+patients with particular characteristics. Figure 4 shows evidence of
+balance on observables. The left panel is the output of a regression of
+Prescribed on observables including patient demographics, priority
+groups, and ED diagnoses. Unsurprisingly, many observables predict
+length of stay and 72 hour return. The right panel displays the
+coefficients from a regression of physician tendency instrument on the
+same set of observables. With so many coefficients, we fail to reject
+that the instrument violates conditional independence (the joint F-stat
+drops by two orders of magnitude, from 280 to 2.4). In fact, some of the
+variables that are significant in the second column go in the opposite
+direction of the decision to batch
 
-The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.
-In Appendix Figure [XXX], residualizing for only seasonality and shift (left panel) and also diagnosis (right panel) in equation (1).
-This lends support to our assumption that assignment in the ED is nearly random at the seasonality, shift  level.
+The residualization in equation (1) controls for more controls than
+required to achieve quasi-random assignment; they are included for
+statistical precision in measuring physician tendency to batch. In
+Appendix Figure {[}XXX{]}, residualizing for only seasonality and shift
+(left panel) and also diagnosis (right panel) in equation (1). This
+lends support to our assumption that assignment in the ED is nearly
+random at the seasonality, shift level.
 
-Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity.
-We reported a strong first stage (i.e., relevance) at the end of the previous Section.
-The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching.
-This is perhaps our strongest assumption and is at its core, untestable.
-However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings.
-First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher.
-Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest.
-The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis.
-Nevertheless, we take this assumption seriously and perform a placebo check in Section IIIB as well as various robustness checks in Section IIIE.
+Our instrumental variable approach, which aims to recover the causal
+effect of having diagnostic tests batch ordered, relies on three
+additional assumptions: relevance, exclusion, and monotonicity. We
+reported a strong first stage (i.e., relevance) at the end of the
+previous Section. The exclusion restriction requires that the instrument
+must influence the outcome of interest only through its effect on test
+batching. This is perhaps our strongest assumption and is at its core,
+untestable. However, several features of the ED setting suggest that
+such violation may likely only have a small impact and may be less
+concerning than in other health care settings. First, unlike in primary
+care settings, where the patient and primary care provider have many
+repeat encounters, the scope of what the emergency physician can do to
+impact medium-term outcomes is limited and well-observed by the
+researcher. Second, any violation of the exclusion restriction needs to
+directly affect the specific outcome of interest. The channel by which
+ED physicians can influence length of stay relative outcomes is likely
+through testing and diagnosis. Nevertheless, we take this assumption
+seriously and perform a placebo check in Section IIIB as well as various
+robustness checks in Section IIIE.
 
-## Results
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
 
-### Reduced-Form Results
+\hypertarget{reduced-form-results}{%
+\subsubsection{Reduced-Form Results}\label{reduced-form-results}}
 
-In this section, we examine the causal effects of provider batching tendency.
-We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).
+In this section, we examine the causal effects of provider batching
+tendency. We first present the effects on our main outcomes of length of
+stay, 72hr readmission, and healthcare resource utilization (number of
+tests).
 
-# Conclusion
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 
-::: {.APPENDIX latex="{General appendix}"}
+\begin{APPENDIX}{General appendix}
 
@@
 \end{table}
-:::
+
+\end{APPENDIX}
+
+% Appendix here
+% Options are (1) APPENDIX (with or without general title) or
+%             (2) APPENDICES (if it has more than one unrelated sections)
+% Outcomment the appropriate case if necessary
+%
+% \begin{APPENDIX}{<Title of the Appendix>}
+% \end{APPENDIX}
+%
+%   or
+%
+% \begin{APPENDICES}
+% \section{<Title of Section A>}
+% \section{<Title of Section B>}
+% etc
+% \end{APPENDICES}
+
+% Acknowledgments here
+\ACKNOWLEDGMENT{The authors would like to thank ..}
+
+\bibliographystyle{informs2014}
+\bibliography{references.bib}
+
+\end{document}

### 2023-09-27 src/IV Validation.R

@@ -11,25 +11,6 @@ library(xtable)
 
 ##########################################################################
 #=========================================================================
-# IV Construction --------------------------------------------------------
-#=========================================================================
-##########################################################################
-
-# Step 1: leave-out residualize at the ED encounter level
-## conditional on shift-level variation, random assignment
-## residual from regression represents physician tendency to batch
-final$residual_batch <- resid(
-  felm(any.batch ~ 0 | dayofweekt + month, data=final))
-
-# Step 2: get batch tendency for each provider
-final <- final %>%
-  group_by(ED_PROVIDER) %>%
-  mutate(Sum_Resid=sum(residual_batch, na.rm=T),
-         batch.tendency = (Sum_Resid - residual_batch) / (n() - 1)) %>% 
-  ungroup()
-
-##########################################################################
-#=========================================================================
 # Establishing IV Validity -----------------------------------------------
 #=========================================================================
 ## Relevance

### 2023-09-27 src/clean.R

@@ -593,5 +593,24 @@ final <- final %>%
 
 rm(list = setdiff(ls(), "final"))
 #=========================================================================
+##########################################################################
+#=========================================================================
+# IV Construction --------------------------------------------------------
+#=========================================================================
+##########################################################################
+
+# Step 1: leave-out residualize at the ED encounter level
+## conditional on shift-level variation, random assignment
+## residual from regression represents physician tendency to batch
+final$residual_batch <- resid(
+  felm(any.batch ~ 0 | dayofweekt + month, data=final))
+
+# Step 2: get batch tendency for each provider
+final <- final %>%
+  group_by(ED_PROVIDER) %>%
+  mutate(Sum_Resid=sum(residual_batch, na.rm=T),
+         batch.tendency = (Sum_Resid - residual_batch) / (n() - 1)) %>% 
+  ungroup()
+
 write.csv(final, 'final.csv')
 #=========================================================================

### 2023-09-27 src/figures.R

@@ -1,0 +1,11 @@
+#=========================================================================
+# Purpose: Main R file for Generating Figures
+# Author: Jacob Jameson 
+#=========================================================================
+rm(list = ls()) 
+
+library(tidyverse)
+library(stringr)
+
+
+

### 2023-09-27 src/IV Validation.R
deciding on the sample

@@ -1,5 +1,5 @@
 #=========================================================================
-# Purpose: Construct main IV and Validate
+# Purpose: Validate IV
 # Author: Jacob Jameson 
 #=========================================================================
 

### 2023-09-27 src/clean.R

@@ -6,6 +6,7 @@ rm(list = ls())
 
 library(tidyverse)
 library(stringr)
+library(lfe)
 
 #=========================================================================
 # Determine test times
@@ -553,8 +554,7 @@ final <- df %>%
          rel_minutes_triage, lab_image_batch, image_image_batch, imaging,
          overlap_per_min, overlap_count, complaint_esi, dayofweekt, month) %>%
   mutate(RTN_72_HR = ifelse(RTN_72_HR == 'Y', 1, 0),
-         RTN_72_HR_ADMIT = ifelse(RTN_72_HR_ADMIT == 'Y', 1, 0)) %>%
-  filter(nEDTests > 0, ED_LOS != 0, ED_LOS < 2880)
+         RTN_72_HR_ADMIT = ifelse(RTN_72_HR_ADMIT == 'Y', 1, 0)) 
 
 final$age_groups <- cut(final$ARRIVAL_AGE_DI, 
                         c(-Inf, 20, 45, 65, Inf),
@@ -592,6 +592,7 @@ final <- final %>%
 
 
 rm(list = setdiff(ls(), "final"))
+
 #=========================================================================
 ##########################################################################
 #=========================================================================

### 2023-09-27 src/figures.R

@@ -1,11 +1,146 @@
 #=========================================================================
-# Purpose: Main R file for Generating Figures
+# Purpose: Main R file for Generating Figures and Tables
 # Author: Jacob Jameson 
 #=========================================================================
 rm(list = ls()) 
+source('src/clean.R')
 
-library(tidyverse)
-library(stringr)
+library(caret)
+library(lmtest)
+library(sandwich)
+library(xtable) # Output to LaTeX table format
+
+##########################################################################
+#=========================================================================
+# Table []
+#   - Show that across physician, there is random assignment and patient
+#     characteristics are balanced
+#=========================================================================
+
+fig.1 <- final
+
+chief_complaint_freq <- table(fig.1$CHIEF_COMPLAINT)
+
+top_10_chief_complaints <- names(chief_complaint_freq)[order(chief_complaint_freq, 
+                                                             decreasing = TRUE)][1:10]
+
+fig.1$CHIEF_COMPLAINT <- ifelse(fig.1$CHIEF_COMPLAINT %in% top_10_chief_complaints, 
+                                fig.1$CHIEF_COMPLAINT, "DROP")
+
+dummy <- dummyVars(" ~ CHIEF_COMPLAINT", data = fig.1)
+one.hot <- data.frame(predict(dummy, newdata = fig.1))
+vars <- setdiff(names(one.hot), 'CHIEF_COMPLAINTDROP')
+
+fig.1 <- cbind(fig.1, one.hot)
+
+fig.1$ESI1.2 <- ifelse(fig.1$ESI == 1 | fig.1$ESI == 2, 1, 0)
+fig.1$ESI3.4.5 <- ifelse(fig.1$ESI %in% c(3, 4, 5), 1, 0)
+
+vars <- c(vars, 'ESI1.2', 'ESI3.4.5')
+
+# Model and collect balance statistics
+balance <- data.frame(Df = numeric(), 
+                      F = numeric(),
+                      Pr..F. = numeric(), 
+                      dummy = character())
+
+for (dummy in vars) {
+  model_pos_1 <- lm(as.formula(paste(dummy, '~ 1')), fig.1)
+  model_pos_2 <- lm(as.formula(paste(dummy, '~ ED_PROVIDER')), fig.1)
+  
+  wald_pos <- waldtest(model_pos_1, model_pos_2, 
+                       vcov = vcovHC(model_pos_2, type = "HC1"))
+  
[…]

### 2023-09-28 src/IV Validation.R
figure 2 updated

@@ -9,6 +9,9 @@ library(stargazer)
 library(texreg)
 library(xtable)
 
+final <- final %>%
+  filter(nEDTests > 0, ED_LOS != 0, ED_LOS < 2880)
+
 ##########################################################################
 #=========================================================================
 # Establishing IV Validity -----------------------------------------------

### 2023-09-28 src/figures.R

@@ -8,6 +8,7 @@ source('src/clean.R')
 library(caret)
 library(lmtest)
 library(sandwich)
+library(ggsci)
 library(xtable) # Output to LaTeX table format
 
 ##########################################################################
@@ -58,25 +59,17 @@ for (dummy in vars) {
 }
 
 print(xtable(balance, 
-             caption = "Exhibit 2: Wald Test Results"), type = "latex")
+             caption = "Wald Test Results"), type = "latex")
 
 ##########################################################################
 
 ##########################################################################
 #=========================================================================
-# Table []
-#   - Show that across physician, there is random assignment and patient
-#     characteristics are balanced
+# Figure []
+#   - Show that frequency of batching that occurs across complaints
 #=========================================================================
 
-#=========================================================================
-# Figure 2
-#   - Show that across physician, there is variation in batching and 
-#     propensity to test
-#=========================================================================
-
-fig.2 <- final %>%
-  filter(nEDTests > 0, ED_LOS != 0, ED_LOS < 2880)
+fig.2 <- final 
 
 collapsed_df <- fig.2 %>%
   group_by(CHIEF_COMPLAINT) %>%
@@ -114,10 +107,7 @@ collapsed_df$type <- factor(collapsed_df$type,
                                        "Lab + Image Batching Rate",
                                        "Image + Image Batching Rate"))
 
-
-
 collapsed_df <- unique(collapsed_df)
-
 
 collapsed_df %>%
   filter(type != "Fraction of Complaint") %>%
@@ -126,9 +116,8 @@ collapsed_df %>%
   labs(x = "Chief Complaint\n",
        y = '\nFrequency',
        title = 'Frequency of Batched Cases by Complaint') +
-  scale_fill_manual(values = c("Fraction of Complaint" = "blue",
-                               "Image + Image Batching Rate" = "grey40",
-                               "Lab + Image Batching Rate" = "grey60")) +
+  scale_fill_manual(values = c("Image + Image Batching Rate" = "#0E65A3",
+                               "Lab + Image Batching Rate" = "#F79500")) +
   theme_minimal() +  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
   scale_y_continuous(labels = scales::percent) +
   theme(axis.text.x = element_text(color = "black", size = 11, angle=0),
@@ -136,9 +125,22 @@ collapsed_df %>%
         axis.title =element_blank(),
[…]

### 2023-10-06 manuscript/manuscript.Rmd
aking updated to manuscript

@@
   - \usepackage{booktabs}
+  - \usepackage{tabularx}
   - \usepackage{siunitx}
@@
 
-# Introduction {#sec:1}
+# Introduction {#sec:I}
 
@@
 
-We start by evaluating the reduced-form effects of ED provider batch ordering tendency on downstream patient outcomes and turnaround time.
+We start by evaluating the reduced-form effects of ED provider batch-ordering tendency on downstream patient outcomes and turnaround time.
 We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences.
-Being treated by a provider in the top decile of the leniency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].
+Being treated by a provider in the top decile of the tendency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].
 
@@
 In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings.
-Furthermore, physicians exhibit wide variation in practice behavior in batch ordering, even within the same hospital, while following the same guidelines.
+Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines.
 Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.
@@
 In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes.
-This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and hospital effeciency.
-It is important to note that this paper studies the impact of an batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.
+This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization.
+It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.
 
@@
 The next section describes the data source and outlines our baseline sample.
-The empirical strategy and its accompanying identifying assumptions are laid out in Section II.
-Section III presents the results.
-Section IV draws implications for batch-ordering policies.
+The empirical strategy and its accompanying identifying assumptions are laid out in Section III.
+Section IV presents the results.
+Section V draws implications for batch-ordering policies.
 The last section concludes.
 
-# Data and Definitions {#sec:2}
+# Data and Definitions {#sec:II}
 
@@
 
-We identified three distinct forms of test batching:
+In the context of diagnostic test ordering, we defined batching as placing multiple test orders within 5 minutes of one another.
+Sensitivity analyses on this cutoff point showed that our results are robust to this definition.
+Our study differentiated between two distinct batching strategies:
 
-- Lab/Image Batch: The first test ordered for a patient was either an imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an additional imaging or lab test order within 5 minutes.
+\begin{itemize}
+\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.
 
-- Image/Image Batch: The first test ordered for a patient was an imaging test (CT scan, X-ray, Ultrasound), followed by an additional imaging test order within 5 minutes.
+\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
+\end{itemize}
 
-- Any Batch: This refers to instances where either a Lab/Image Batch or an Image/Image Batch occurred.
+The distinction is necessary because laboratory tests may be processed in parallel with imaging tests, which have a much more limited effect on patient waiting time.
+However, a batch of multiple imaging tests requires that the patient is present for both tests.
+Thus, waiting times are much more likely to be impacted.
+Therefore, the specific tests that make up a batch may provide insight into how the physician trades off between waiting times and uncertainty, given the various complexity levels of patient encounters.
+Figure 1 shows the rate at which each batching type occurs for the top ten most frequently occurring chief complaint fields.
 
-We conducted a sensitivity analysis for this definition, documented in Appendix 2.
-Table 1 highlights the considerable variation in patient complexity between those who had their tests batch ordered versus those where tests were ordered sequentially.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
+  \tiny
+  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
+\end{figure}
 
@@
 
-Table 1 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered either individually (Sequential) or as part of a group (Batch) during an ED encounter from our baseline sample.
-Around 43% of patient encounters involved diagnostic tests ordered as part of a batch, the majority of which (60%) included two tests.
+Table 2 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered individually in a batched or sequential way during an ED encounter from our baseline sample.
+About 39% of patient encounters involved diagnostic tests ordered as part of a batch, most of which (64%) included two tests.
 
-Differences were noted between the two cohorts.
-Patients who had batched tests ordered tended to have a longer ED length of stay, a lower Emergency Severity Index (ESI), and were generally older than those who had their tests ordered individually.
-Patients in the batch order group were also more likely to present with symptoms at triage, including tachycardia, tachypnea, fever, and hypotension.
-A deeper dive into the ESI data reveals a higher proportion of patients with more urgent conditions (ESI 1 and 2) in the batched group than in the sequential group.
-This finding might be indicative of the complexity and severity of the conditions of patients who necessitate batch testing.
+Notable distinctions emerge when comparing the two cohorts.
+Patients subjected to batched test ordering exhibited a notably prolonged stay in the emergency department (ED) (p<0.001) and were more frequently assigned a lower Emergency Severity Index (ESI).
+Furthermore, this group skewed towards older demographics, with half of them falling into the 65+ age category.
+Patients in the batch-order group also displayed a higher likelihood of presenting symptoms at triage, including tachycardia (p<0.001), tachypnea (p<0.001), fever (p<0.001), and hypotension (p<0.001).
+Delving deeper into the ESI data unveils a higher concentration of patients with more critical conditions (ESI 1 and 2) within the batched group, in contrast to the sequential group (45.1% with ESI of 1 or 2 in the batched group compared to 27.5% with ESI of 1 or 2 in the sequential group).
+This finding suggests that factors related to patient complexity and severity (Saghafian et al., 2014) significantly influence physicians' decisions to opt for batch-ordering of diagnostic tests.
+However, it is important to note that our investigation revealed substantial variation among physicians in their choices between batching and sequencing for a given patient.
 
-Age distributions between the two groups demonstrate a greater proportion of patients aged 65 and above in the batch group, suggesting that older patients are more likely to have multiple tests ordered at once, possibly due to the presence of multiple comorbidities.
-The batched group showed a longer average ED length of stay compared to the sequential group.
-This extended stay may be reflective of the additional time required to conduct and evaluate the results of multiple tests.
-Symptoms presentation at triage also differed between the two groups, with the batched group being more likely to present with tachypnea, fever, and hypotension.
-However, tachycardia was found to be similar across both groups.
-
-\begin{table}[htbp]
+\begin{table}
 \centering
-\tiny
-\caption{Summary Statistics for Sequential vs. Batched Test Groups}
-\begin{tabular}{lcccc}
-\hline
-Variable & \multicolumn{1}{c}{Overall Visits, N = 41,197\textsuperscript{1}} & \multicolumn{1}{c}{Sequential Arm, N = 23,4331} & \multicolumn{1}{c}{Batching Arm, N = 17,7641} & \multicolumn{1}{c}{p-value\textsuperscript{2}} \\
-\hline
-\textbf{ESI} & & & & $<$0.001 \\
-\hspace{0.5cm} 1 & 474 (1.2\%) & 130 (0.6\%) & 344 (1.9\%) & \\
-\hspace{0.5cm} 2 & 13,911 (34\%) & 6,502 (28\%) & 7,409 (42\%) & \\
-\hspace{0.5cm} 3 & 23,655 (57\%) & 14,350 (61\%) & 9,305 (52\%) & \\
-\hspace{0.5cm} 4 & 3,123 (7.6\%) & 2,422 (10\%) & 701 (3.9\%) & \\
-\hspace{0.5cm} 5 & 34 ($<$0.1\%) & 29 (0.1\%) & 5 ($<$0.1\%) & \\
-\hline
-\textbf{Age} & & & & $<$0.001 \\
-\hspace{0.5cm} $<$20 & 790 (1.9\%) & 605 (2.6\%) & 185 (1.0\%) & \\
-\hspace{0.5cm} 20-45 & 9,412 (23\%) & 6,242 (27\%) & 3,170 (18\%) & \\
-\hspace{0.5cm} 45-65 & 12,826 (31\%) & 7,393 (32\%) & 5,433 (31\%) & \\
-\hspace{0.5cm} 65+ & 18,169 (44\%) & 9,193 (39\%) & 8,976 (51\%) & \\
-\hline
-\textbf{ED LOS} & & & & $<$0.001 \\
-\hspace{0.5cm} Mean (SD) & 270 (382) & 258 (177) & 286 (545) & \\
-\hline
-\textbf{72hr Return} & 1,482 (3.6\%) & 957 (4.1\%) & 525 (3.0\%) & $<$0.001 \\
-\hline
-\textbf{Number of Tests} & & & & $<$0.001 \\
-\hspace{0.5cm} 1 & 12,961 (31\%) & 11,937 (51\%) & 1,024 (5.8\%) & \\
-\hspace{0.5cm} 2 & 19,112 (46\%) & 8,457 (36\%) & 10,655 (60\%) & \\
-\hspace{0.5cm} 3 & 7,688 (19\%) & 2,581 (11\%) & 5,107 (29\%) & \\
-\hspace{0.5cm} 4 & 1,342 (3.3\%) & 432 (1.8\%) & 910 (5.1\%) & \\
-\hspace{0.5cm} 5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
-\hline
-\textbf{Disposition} & & & & $<$0.001 \\
-\hspace{0.5cm} Discharge & 25,302 (61\%) & 15,922 (68\%) & 9,380 (53\%) & \\
-\hspace{0.5cm} Admit & 9,097 (22\%) & 4,209 (18\%) & 4,888 (28\%) & \\
-\hspace{0.5cm} Other & 6,798 (17\%) & 3,302 (14\%) & 3,496 (20\%) & \\
-\hline
-\textbf{Tachycardic} & 8,115 (20\%) & 4,562 (19\%) & 3,553 (20\%) & 0.18 \\
-\hline
-\textbf{Tachypneic} & 3,822 (9.3\%) & 1,669 (7.1\%) & 2,153 (12\%) & $<$0.001 \\
-\hline
-\textbf{Febrile} & 969 (2.4\%) & 353 (1.5\%) & 616 (3.5\%) & $<$0.001 \\
-\hline
-\textbf{Hypotensive} & 651 (1.6\%) & 305 (1.3\%) & 346 (1.9\%) & $<$0.001 \\
-\hline
-\textbf{Gender} & & & & $<$0.001 \\
-\hspace{0.5cm} Female & 22,105 (54\%) & 12,986 (55\%) & 9,119 (51\%) & \\
-\hspace{0.5cm} Male & 19,092 (46\%) & 10,447 (45\%) & 8,645 (49\%) & \\
-\hline
-\textbf{Race} & & & & $<$0.001 \\
-\hspace{0.5cm} White & 36,554 (89\%) & 20,665 (88\%) & 15,889 (89\%) & \\
-\hspace{0.5cm} Black & 1,682 (4.1\%) & 974 (4.2\%) & 708 (4.0\%) & \\
-\hspace{0.5cm} Asian & 1,217 (3.0\%) & 710 (3.0\%) & 507 (2.9\%) & \\
-\hspace{0.5cm} Other & 737 (1.8\%) & 449 (1.9\%) & 288 (1.6\%) & \\
-\hspace{0.5cm} Native & 538 (1.3\%) & 340 (1.5\%) & 198 (1.1\%) & \\
-\hspace{0.5cm} Unknown & 469 (1.1\%) & 295 (1.3\%) & 174 (1.0\%) & \\
-\hline
-\hline
+\caption{Summary Statistics}
+\begin{tabular}{p{3cm} *{4}{p{2.5cm}}}
+\toprule
+Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
+&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
+\midrule
+ESI & & & & $<0.001$ \\
+1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
+2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
+3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
+4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
+5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
+(Missing) & 27 & 16 & 11 & \\
+\midrule
+Age & & & & $<0.001$ \\
+<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
+20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
+45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
+65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
+\midrule
+ED LOS (min) & & & & $<0.001$ \\
+Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
+\midrule
+Disposition & & & & $<0.001$ \\
+Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
+Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
+Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
+\midrule
+Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
+Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
+Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
+Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
+\midrule
+Number of Tests & & & & $<0.001$ \\
+1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
+2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
+3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
+4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
+5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
+\midrule
+Gender & & & & $<0.001$ \\
+Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
+Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
+\midrule
+Race & & & & $0.002$ \\
+White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
+Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
+Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
+Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
+Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
+Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
+\bottomrule
 \multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
@@
 \end{table}
-
-![Frequency of Sample Cases and Batches](../Manuscript/figures/fig1.png){width=40%, height=30%}
 
@@
 In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Exhibit 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+Table 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
 
@@
 \tiny
-\item Table [blank] reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
+\item Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
 A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization.
@@
 We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure 2 graphs the histogram of the leniency measure along the x-axis and the left y-axis.
-A local-linear regression of the fitted probability of batched ordering on tendency after residualizing is overlaid and displayed on the right y-axis.
-Table 2 presents the same ``first stage" in a regression table: being assigned to a 10 pp more lenient physician is associated with a 17 pp increase in the likelihood of being prescribed an opioid in the ED.
-The F-statistic is 25 when all controls and fixed effects are included.
-The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rarely prescribed diagnoses.
+Figure 2 graphs the histogram of batch-ordering frequency by physician for populat chief complaints, highlighting that the variation in batching differs systematically.
+Table 3 presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 10 pp increase in the likelihood of having tests batch-ordered in the ED.
+The F-statistic is 51 when all controls and fixed effects are included.
+
+\begin{table}[htbp]
+\centering
+\caption{First Stage Regression Results}
+\begin{threeparttable}
+\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
+\toprule
+ & Model 1 & Model 2 & Model 3 \\
+\midrule
+Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
+ & (0.01) & (0.01) & (0.01) \\
+\midrule
+F statistic (full model)&7.32&52.03&50.93\\  
+F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
+\midrule
+Num. obs. & 43300 & 43300 & 43300 \\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & No & Yes & Yes \\
+Patient observables? & No & No & Yes \\
+\bottomrule
+\end{tabularx}
+\begin{tablenotes}
+\tiny
+\item Estimates of the first stage for the baseline sample described in the text.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Column 3 corresponds to the baseline controls.
+Robust standard errors are clustered at the physician level.
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}\tiny
+\end{threeparttable}
+\label{table:regression}
+\end{table}
 
@@
 Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization. 
-
-\begin{table}[htbp]
-\centering
-\caption{First Stage Regression Results}
-\begin{tabular}{lccc}
-\toprule
- & Model 1 & Model 2 & Model 3 \\
-\midrule
-Lab + Image Batching Tendency & 0.99*** & 1.02*** & 1.02*** \\
- & (0.01) & (0.01) & (0.01) \\
-\\
-F statistic (proj model) & 13206.68 & 6702.01 & 6655.86  \\
-F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
-\midrule
-Image + Image Batching Tendency & 0.92*** & 0.94*** & 0.94*** \\
- & (0.02) & (0.04) & (0.04) \\
-\\
-F statistic (proj model) & 1645.35 & 577.97 & 577.09  \\
-F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
-\midrule
-\midrule
-Num. obs. & 41197 & 41197 & 41197 \\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & No & Yes & Yes \\
-Patient observables? & No & No & Yes \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\small
-\item \textit{Notes:}  Estimates of the first stage for the baseline sample described in the text.
-Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
-Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
-Column 3 corresponds to the baseline controls.
-Robust standard errors are clustered at the physician level.
-\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
-\end{tablenotes}
-\end{table}
-
 and $X_i$ is the same as in the reduced-form approach.
@@
 
-### Identifying Assumptions
+## Identifying Assumptions
 
-The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, if assignment of patients to ED physicians is random, conditional on seasonality, shift, and diagnosis (``conditional independence").
-One can test for whether batch-likely physicians are routinely assigned patients with particular characteristics.
-Figure 4 shows evidence of balance on observables.
-The left panel is the output of a regression of Prescribed on observables including patient demographics, priority groups, and ED diagnoses.
-Unsurprisingly, many observables predict length of stay and 72 hour return.
-The right panel displays the coefficients from a regression of physician tendency instrument on the same set of observables.
-With so many coefficients, we fail to reject that the instrument violates conditional independence (the joint F-stat drops by two orders of magnitude, from 280 to 2.4).
-In fact, some of the variables that are significant in the second column go in the opposite direction of the decision to batch
-
+The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence").
 The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.
-In Appendix Figure [XXX], residualizing for only seasonality and shift (left panel) and also diagnosis (right panel) in equation (1).
-This lends support to our assumption that assignment in the ED is nearly random at the seasonality, shift  level.
 
@@
 The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis.
-Nevertheless, we take this assumption seriously and perform a placebo check in Section IIIB as well as various robustness checks in Section IIIE.
+Nevertheless, we take this assumption seriously and perform a placebo check in Section 4.2 as well as various robustness checks in Section 4.4.
 
-## Results
+Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects.
+It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician.
+The literature leveraging the judges design typically performs two informal tests for its implications.
+The first one provides that the first stage should be weakly positive for all subsamples (Dobbie, Goldin, and Yang 2018).
+The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (Bhuller et al.
+2020).
+Appendix Table 2 presents both of these tests in the two columns for various subsamples of interest.
+Finally, we check whether our main results hold using differential, mutually exclusive leniency measures (i.e., by complaint category) in Section 4.4.
 
-### Reduced-Form Results
+# Results {#sec:4}
+
+## Reduced-Form Results
 
@@
 We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).
+
+## Placebo Check
+
+## Instrumental Variables Results
+
+## Robustness
 
@@
 
-\begin{table}[ht]
-\centering
-\caption{Regression Results by Chief Complaint} 
-\begin{tabular}{llrr}
-  \hline
- & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
-  \hline
-1 & Extremity Complaints & -0.72** & 0.03 \\ 
-  2 & Upper Respiratory Symptoms & 0.426** & 0.02 \\ 
-  3 & Abnormal Test Results & -0.067 & 0.84 \\ 
-  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.423* & 0.10 \\ 
-  5 & Shortness of Breath & -0.065 & 0.58 \\ 
-  6 & Gastrointestinal Issues & 0.071 & 0.67 \\ 
-  7 & Chest Pain & 0.107 & 0.68 \\ 
-  8 & Dizziness/Lightheadedness/Syncope & -0.163 & 0.32 \\ 
-  9 & Neurological Issue & -0.353 & 0.25 \\ 
-  10 & Fatigue and Weakness & -0.112 & 0.49 \\ 
-  11 & Back or Flank Pain & -0.602** & 0.02 \\ 
-  12 & Abdominal Complaints & -0.105 & 0.13 \\ 
-  13 & Fevers, Sweats or Chills & 0.01 & 0.95 \\ 
-  14 & Skin Complaints & 1.442 & 0.35 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\caption{Regression Results by Chief Complaint (2)} 
-\begin{tabular}{llrr}
-  \hline
- & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
-  \hline
-1 & Extremity Complaints & 0.176 & 0.50 \\ 
-  2 & Upper Respiratory Symptoms & 0.332*** & 0.00 \\ 
-  3 & Abnormal Test Results & 0.695 & 0.14 \\ 
-  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.247 & 0.20 \\ 
-  5 & Shortness of Breath & -0.253** & 0.01 \\ 
-  6 & Gastrointestinal Issues & 0.421*** & 0.00 \\ 
-  7 & Chest Pain & -0.531*** & 0.00 \\ 
-  8 & Dizziness/Lightheadedness/Syncope & 0.075 & 0.60 \\ 
-  9 & Neurological Issue & -0.305 & 0.15 \\ 
-  10 & Fatigue and Weakness & -0.125 & 0.40 \\ 
-  11 & Back or Flank Pain & 0.288 & 0.13 \\ 
-  12 & Abdominal Complaints & -0.023 & 0.66 \\ 
-  13 & Fevers, Sweats or Chills & -0.044 & 0.71 \\ 
-  14 & Skin Complaints & 0.328 & 0.35 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\caption{First Stage over Subsamples} 
-\begin{tabular}{llrr}
-  \hline
- & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
-  \hline
-1 & Extremity Complaints & 0.761*** & 0.00 \\ 
-  2 & Upper Respiratory Symptoms & 1.406*** & 0.00 \\ 
-  3 & Abnormal Test Results & 1.036*** & 0.00 \\ 
-  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.831*** & 0.00 \\ 
-  5 & Shortness of Breath & 1.258*** & 0.00 \\ 
-  6 & Gastrointestinal Issues & 1.043*** & 0.00 \\ 
-  7 & Chest Pain & 0.866*** & 0.00 \\ 
-  8 & Dizziness/Lightheadedness/Syncope & 1.668*** & 0.00 \\ 
-  9 & Neurological Issue & 0.723*** & 0.00 \\ 
-  10 & Fatigue and Weakness & 1.651*** & 0.00 \\ 
-  11 & Back or Flank Pain & 0.752*** & 0.01 \\ 
-  12 & Abdominal Complaints & 1.109*** & 0.00 \\ 
-  13 & Fevers, Sweats or Chills & 1.23*** & 0.00 \\ 
-  14 & Skin Complaints & 0.448* & 0.07 \\ 
-   \hline
-\end{tabular}
-\end{table}
 :::

### 2023-10-06 manuscript/manuscript.tex
aking updated to manuscript

@@
 \usepackage{booktabs}
+\usepackage{tabularx}
 \usepackage{siunitx}
@@
 
-\hypertarget{sec:1}{%
-\section{Introduction}\label{sec:1}}
+\hypertarget{sec:I}{%
+\section{Introduction}\label{sec:I}}
 
@@
 
-We start by evaluating the reduced-form effects of ED provider batch
-ordering tendency on downstream patient outcomes and turnaround time. We
-find that practice variation as captured by physician batch-ordering
-tendency has large and significant consequences. Being treated by a
-provider in the top decile of the leniency distribution, compared to
-being treated by someone in the bottom decile {[}INSERT RESULTS{]}.
+We start by evaluating the reduced-form effects of ED provider
+batch-ordering tendency on downstream patient outcomes and turnaround
+time. We find that practice variation as captured by physician
+batch-ordering tendency has large and significant consequences. Being
+treated by a provider in the top decile of the tendency distribution,
+compared to being treated by someone in the bottom decile {[}INSERT
+RESULTS{]}.
 
@@
 other health care settings. Furthermore, physicians exhibit wide
-variation in practice behavior in batch ordering, even within the same
+variation in practice behavior in batch-ordering, even within the same
 hospital, while following the same guidelines. Finally,
@@
 causal impact of batch-ordering diagnostic tests on patient outcomes and
-hospital effeciency. It is important to note that this paper studies the
-impact of an batch-ordering through a batching decision requiring
+resource utilization. It is important to note that this paper studies
+the impact of batch-ordering through a batching decision requiring
 clinical judgment (within practice norms) rather than through specific
@@
 empirical strategy and its accompanying identifying assumptions are laid
-out in Section II. Section III presents the results. Section IV draws
+out in Section III. Section IV presents the results. Section V draws
 implications for batch-ordering policies. The last section concludes.
 
-\hypertarget{sec:2}{%
-\section{Data and Definitions}\label{sec:2}}
+\hypertarget{sec:II}{%
+\section{Data and Definitions}\label{sec:II}}
 
@@
 
-We identified three distinct forms of test batching:
+In the context of diagnostic test ordering, we defined batching as
+placing multiple test orders within 5 minutes of one another.
+Sensitivity analyses on this cutoff point showed that our results are
+robust to this definition. Our study differentiated between two distinct
+batching strategies:
 
 \begin{itemize}
-\item
-  Lab/Image Batch: The first test ordered for a patient was either an
-  imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an
-  additional imaging or lab test order within 5 minutes.
-\item
-  Image/Image Batch: The first test ordered for a patient was an imaging
-  test (CT scan, X-ray, Ultrasound), followed by an additional imaging
-  test order within 5 minutes.
-\item
-  Any Batch: This refers to instances where either a Lab/Image Batch or
-  an Image/Image Batch occurred.
+\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.
+
+\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
 \end{itemize}
 
-We conducted a sensitivity analysis for this definition, documented in
-Appendix 2. Table 1 highlights the considerable variation in patient
-complexity between those who had their tests batch ordered versus those
-where tests were ordered sequentially.
+The distinction is necessary because laboratory tests may be processed
+in parallel with imaging tests, which have a much more limited effect on
+patient waiting time. However, a batch of multiple imaging tests
+requires that the patient is present for both tests. Thus, waiting times
+are much more likely to be impacted. Therefore, the specific tests that
+make up a batch may provide insight into how the physician trades off
+between waiting times and uncertainty, given the various complexity
+levels of patient encounters. Figure 1 shows the rate at which each
+batching type occurs for the top ten most frequently occurring chief
+complaint fields.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
+  \tiny
+  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
+\end{figure}
 
@@
 
-Table 1 provides a comprehensive breakdown of the patient encounters
-involving diagnostic tests ordered either individually (Sequential) or
-as part of a group (Batch) during an ED encounter from our baseline
-sample. Around 43\% of patient encounters involved diagnostic tests
-ordered as part of a batch, the majority of which (60\%) included two
-tests.
+Table 2 provides a comprehensive breakdown of the patient encounters
+involving diagnostic tests ordered individually in a batched or
+sequential way during an ED encounter from our baseline sample. About
+39\% of patient encounters involved diagnostic tests ordered as part of
+a batch, most of which (64\%) included two tests.
 
-Differences were noted between the two cohorts. Patients who had batched
-tests ordered tended to have a longer ED length of stay, a lower
-Emergency Severity Index (ESI), and were generally older than those who
-had their tests ordered individually. Patients in the batch order group
-were also more likely to present with symptoms at triage, including
-tachycardia, tachypnea, fever, and hypotension. A deeper dive into the
-ESI data reveals a higher proportion of patients with more urgent
-conditions (ESI 1 and 2) in the batched group than in the sequential
-group. This finding might be indicative of the complexity and severity
-of the conditions of patients who necessitate batch testing.
+Notable distinctions emerge when comparing the two cohorts. Patients
+subjected to batched test ordering exhibited a notably prolonged stay in
+the emergency department (ED) (p\textless0.001) and were more frequently
+assigned a lower Emergency Severity Index (ESI). Furthermore, this group
+skewed towards older demographics, with half of them falling into the
+65+ age category. Patients in the batch-order group also displayed a
+higher likelihood of presenting symptoms at triage, including
+tachycardia (p\textless0.001), tachypnea (p\textless0.001), fever
+(p\textless0.001), and hypotension (p\textless0.001). Delving deeper
+into the ESI data unveils a higher concentration of patients with more
+critical conditions (ESI 1 and 2) within the batched group, in contrast
+to the sequential group (45.1\% with ESI of 1 or 2 in the batched group
+compared to 27.5\% with ESI of 1 or 2 in the sequential group). This
+finding suggests that factors related to patient complexity and severity
+(Saghafian et al., 2014) significantly influence physicians' decisions
+to opt for batch-ordering of diagnostic tests. However, it is important
+to note that our investigation revealed substantial variation among
+physicians in their choices between batching and sequencing for a given
+patient.
 
-Age distributions between the two groups demonstrate a greater
-proportion of patients aged 65 and above in the batch group, suggesting
-that older patients are more likely to have multiple tests ordered at
-once, possibly due to the presence of multiple comorbidities. The
-batched group showed a longer average ED length of stay compared to the
-sequential group. This extended stay may be reflective of the additional
-time required to conduct and evaluate the results of multiple tests.
-Symptoms presentation at triage also differed between the two groups,
-with the batched group being more likely to present with tachypnea,
-fever, and hypotension. However, tachycardia was found to be similar
-across both groups.
-
-\begin{table}[htbp]
+\begin{table}
 \centering
-\tiny
-\caption{Summary Statistics for Sequential vs. Batched Test Groups}
-\begin{tabular}{lcccc}
-\hline
-Variable & \multicolumn{1}{c}{Overall Visits, N = 41,197\textsuperscript{1}} & \multicolumn{1}{c}{Sequential Arm, N = 23,4331} & \multicolumn{1}{c}{Batching Arm, N = 17,7641} & \multicolumn{1}{c}{p-value\textsuperscript{2}} \\
-\hline
-\textbf{ESI} & & & & $<$0.001 \\
-\hspace{0.5cm} 1 & 474 (1.2\%) & 130 (0.6\%) & 344 (1.9\%) & \\
-\hspace{0.5cm} 2 & 13,911 (34\%) & 6,502 (28\%) & 7,409 (42\%) & \\
-\hspace{0.5cm} 3 & 23,655 (57\%) & 14,350 (61\%) & 9,305 (52\%) & \\
-\hspace{0.5cm} 4 & 3,123 (7.6\%) & 2,422 (10\%) & 701 (3.9\%) & \\
-\hspace{0.5cm} 5 & 34 ($<$0.1\%) & 29 (0.1\%) & 5 ($<$0.1\%) & \\
-\hline
-\textbf{Age} & & & & $<$0.001 \\
-\hspace{0.5cm} $<$20 & 790 (1.9\%) & 605 (2.6\%) & 185 (1.0\%) & \\
-\hspace{0.5cm} 20-45 & 9,412 (23\%) & 6,242 (27\%) & 3,170 (18\%) & \\
-\hspace{0.5cm} 45-65 & 12,826 (31\%) & 7,393 (32\%) & 5,433 (31\%) & \\
-\hspace{0.5cm} 65+ & 18,169 (44\%) & 9,193 (39\%) & 8,976 (51\%) & \\
-\hline
-\textbf{ED LOS} & & & & $<$0.001 \\
-\hspace{0.5cm} Mean (SD) & 270 (382) & 258 (177) & 286 (545) & \\
-\hline
-\textbf{72hr Return} & 1,482 (3.6\%) & 957 (4.1\%) & 525 (3.0\%) & $<$0.001 \\
-\hline
-\textbf{Number of Tests} & & & & $<$0.001 \\
-\hspace{0.5cm} 1 & 12,961 (31\%) & 11,937 (51\%) & 1,024 (5.8\%) & \\
-\hspace{0.5cm} 2 & 19,112 (46\%) & 8,457 (36\%) & 10,655 (60\%) & \\
-\hspace{0.5cm} 3 & 7,688 (19\%) & 2,581 (11\%) & 5,107 (29\%) & \\
-\hspace{0.5cm} 4 & 1,342 (3.3\%) & 432 (1.8\%) & 910 (5.1\%) & \\
-\hspace{0.5cm} 5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
-\hline
-\textbf{Disposition} & & & & $<$0.001 \\
-\hspace{0.5cm} Discharge & 25,302 (61\%) & 15,922 (68\%) & 9,380 (53\%) & \\
-\hspace{0.5cm} Admit & 9,097 (22\%) & 4,209 (18\%) & 4,888 (28\%) & \\
-\hspace{0.5cm} Other & 6,798 (17\%) & 3,302 (14\%) & 3,496 (20\%) & \\
-\hline
-\textbf{Tachycardic} & 8,115 (20\%) & 4,562 (19\%) & 3,553 (20\%) & 0.18 \\
-\hline
-\textbf{Tachypneic} & 3,822 (9.3\%) & 1,669 (7.1\%) & 2,153 (12\%) & $<$0.001 \\
-\hline
-\textbf{Febrile} & 969 (2.4\%) & 353 (1.5\%) & 616 (3.5\%) & $<$0.001 \\
-\hline
-\textbf{Hypotensive} & 651 (1.6\%) & 305 (1.3\%) & 346 (1.9\%) & $<$0.001 \\
-\hline
-\textbf{Gender} & & & & $<$0.001 \\
-\hspace{0.5cm} Female & 22,105 (54\%) & 12,986 (55\%) & 9,119 (51\%) & \\
-\hspace{0.5cm} Male & 19,092 (46\%) & 10,447 (45\%) & 8,645 (49\%) & \\
-\hline
-\textbf{Race} & & & & $<$0.001 \\
-\hspace{0.5cm} White & 36,554 (89\%) & 20,665 (88\%) & 15,889 (89\%) & \\
-\hspace{0.5cm} Black & 1,682 (4.1\%) & 974 (4.2\%) & 708 (4.0\%) & \\
-\hspace{0.5cm} Asian & 1,217 (3.0\%) & 710 (3.0\%) & 507 (2.9\%) & \\
-\hspace{0.5cm} Other & 737 (1.8\%) & 449 (1.9\%) & 288 (1.6\%) & \\
-\hspace{0.5cm} Native & 538 (1.3\%) & 340 (1.5\%) & 198 (1.1\%) & \\
-\hspace{0.5cm} Unknown & 469 (1.1\%) & 295 (1.3\%) & 174 (1.0\%) & \\
-\hline
-\hline
+\caption{Summary Statistics}
+\begin{tabular}{p{3cm} *{4}{p{2.5cm}}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}}
+\toprule
+Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
+&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
+\midrule
+ESI & & & & $<0.001$ \\
+1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
+2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
+3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
+4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
+5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
+(Missing) & 27 & 16 & 11 & \\
+\midrule
+Age & & & & $<0.001$ \\
+<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
+20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
+45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
+65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
+\midrule
+ED LOS (min) & & & & $<0.001$ \\
+Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
+\midrule
+Disposition & & & & $<0.001$ \\
+Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
+Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
+Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
+\midrule
+Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
+Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
+Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
+Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
+\midrule
+Number of Tests & & & & $<0.001$ \\
+1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
+2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
+3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
+4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
+5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
+\midrule
+Gender & & & & $<0.001$ \\
+Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
+Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
+\midrule
+Race & & & & $0.002$ \\
+White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
+Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
+Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
+Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
+Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
+Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
+\bottomrule
 \multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
@@
 \end{table}
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth,height=0.3\textheight]{../Manuscript/figures/fig1.png}
-\caption{Frequency of Sample Cases and Batches}
-\end{figure}
 
@@
 for our analyses. In essence, barring arrival time and shift-level
-variation, the physician-to-patient matching can be deemed random.
-Exhibit 2 displays that patient encounters (regarding chief complaints
-and emergency severity) are equitably distributed across physicians
-within our study's cohort.
+variation, the physician-to-patient matching can be deemed random. Table
+2 displays that patient encounters (regarding chief complaints and
+emergency severity) are equitably distributed across physicians within
+our study's cohort.
 
@@
 variation in their propensity to batch order diagnostic tests. Figure 2
-graphs the histogram of the leniency measure along the x-axis and the
-left y-axis. A local-linear regression of the fitted probability of
-batched ordering on tendency after residualizing is overlaid and
-displayed on the right y-axis. Table 2 presents the same ``first stage''
-in a regression table: being assigned to a 10 pp more lenient physician
-is associated with a 17 pp increase in the likelihood of being
-prescribed an opioid in the ED. The F-statistic is 25 when all controls
-and fixed effects are included. The coefficient is greater than one
-because all emergency visits are used to construct the leniency
-instrument, while the first stage is calculated using the baseline
-sample only, which excludes the rarely prescribed diagnoses.
+graphs the histogram of batch-ordering frequency by physician for
+populat chief complaints, highlighting that the variation in batching
+differs systematically. Table 3 presents the ``first stage'' in a
+regression table: being assigned to a 10 pp higher tendency physician is
+associated with a 10 pp increase in the likelihood of having tests
+batch-ordered in the ED. The F-statistic is 51 when all controls and
+fixed effects are included.
+
+\begin{table}[htbp]
+\centering
+\caption{First Stage Regression Results}
+\begin{threeparttable}
+\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
+\toprule
+ & Model 1 & Model 2 & Model 3 \\
+\midrule
+Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
+ & (0.01) & (0.01) & (0.01) \\
+\midrule
+F statistic (full model)&7.32&52.03&50.93\\  
+F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
+\midrule
+Num. obs. & 43300 & 43300 & 43300 \\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & No & Yes & Yes \\
+Patient observables? & No & No & Yes \\
+\bottomrule
+\end{tabularx}
+\begin{tablenotes}
+\tiny
+\item Estimates of the first stage for the baseline sample described in the text.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Column 3 corresponds to the baseline controls.
+Robust standard errors are clustered at the physician level.
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}\tiny
+\end{threeparttable}
+\label{table:regression}
+\end{table}
 
@@
 Where \(Y_i\) represents our main outcomes of interest: length of stay,
-72 hour readmission, and resource utilization.
-
-\begin{table}[htbp]
-\centering
-\caption{First Stage Regression Results}
-\begin{tabular}{lccc}
-\toprule
- & Model 1 & Model 2 & Model 3 \\
-\midrule
-Lab + Image Batching Tendency & 0.99*** & 1.02*** & 1.02*** \\
- & (0.01) & (0.01) & (0.01) \\
-\\
-F statistic (proj model) & 13206.68 & 6702.01 & 6655.86  \\
-F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
-\midrule
-Image + Image Batching Tendency & 0.92*** & 0.94*** & 0.94*** \\
- & (0.02) & (0.04) & (0.04) \\
-\\
-F statistic (proj model) & 1645.35 & 577.97 & 577.09  \\
-F (proj model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
-\midrule
-\midrule
-Num. obs. & 41197 & 41197 & 41197 \\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & No & Yes & Yes \\
-Patient observables? & No & No & Yes \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\small
-\item \textit{Notes:}  Estimates of the first stage for the baseline sample described in the text.
-Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
-Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
-Column 3 corresponds to the baseline controls.
-Robust standard errors are clustered at the physician level.
-\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
-\end{tablenotes}
-\end{table}
-
-and \(X_i\) is the same as in the reduced-form approach. \(Batched_i\)
-variable suffers from potential endogeneity concerns. For example,
-injury severity may be unobserved and correlated with need to run
-multiple tests, which in turn also affects length of stay. Hence, we
-instrument \(Batched_i\) with the assigned physician \(j\)'s underlying
-tendency to batch, \(Tendency_{i,j}^{phys}\) We cluster robust standard
-errors at the physician level to account for the assignment process of
-patients to physicians.
+72 hour readmission, and resource utilization. and \(X_i\) is the same
+as in the reduced-form approach. \(Batched_i\) variable suffers from
+potential endogeneity concerns. For example, injury severity may be
+unobserved and correlated with need to run multiple tests, which in turn
+also affects length of stay. Hence, we instrument \(Batched_i\) with the
+assigned physician \(j\)'s underlying tendency to batch,
+\(Tendency_{i,j}^{phys}\) We cluster robust standard errors at the
+physician level to account for the assignment process of patients to
+physicians.
 
 \hypertarget{identifying-assumptions}{%
-\subsubsection{Identifying Assumptions}\label{identifying-assumptions}}
+\subsection{Identifying Assumptions}\label{identifying-assumptions}}
 
 The reduced-form approach delivers an unbiased estimate of the causal
-effect of being treated by a higher tendency to batch physician, if
+effect of being treated by a higher tendency to batch physician, since
 assignment of patients to ED physicians is random, conditional on
-seasonality, shift, and diagnosis (``conditional independence''). One
-can test for whether batch-likely physicians are routinely assigned
-patients with particular characteristics. Figure 4 shows evidence of
-balance on observables. The left panel is the output of a regression of
-Prescribed on observables including patient demographics, priority
-groups, and ED diagnoses. Unsurprisingly, many observables predict
-length of stay and 72 hour return. The right panel displays the
-coefficients from a regression of physician tendency instrument on the
-same set of observables. With so many coefficients, we fail to reject
-that the instrument violates conditional independence (the joint F-stat
-drops by two orders of magnitude, from 280 to 2.4). In fact, some of the
-variables that are significant in the second column go in the opposite
-direction of the decision to batch
-
-The residualization in equation (1) controls for more controls than
-required to achieve quasi-random assignment; they are included for
-statistical precision in measuring physician tendency to batch. In
-Appendix Figure {[}XXX{]}, residualizing for only seasonality and shift
-(left panel) and also diagnosis (right panel) in equation (1). This
-lends support to our assumption that assignment in the ED is nearly
-random at the seasonality, shift level.
+seasonality and shift (``conditional independence''). The
+residualization in equation (1) controls for more controls than required
+to achieve quasi-random assignment; they are included for statistical
+precision in measuring physician tendency to batch.
 
@@
 through testing and diagnosis. Nevertheless, we take this assumption
-seriously and perform a placebo check in Section IIIB as well as various
-robustness checks in Section IIIE.
+seriously and perform a placebo check in Section 4.2 as well as various
+robustness checks in Section 4.4.
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+Finally, the monotonicity assumption is necessary for interpreting the
+coefficient estimates obtained from the IV approach as Local Average
+Treatment Effects (LATEs) if there are heterogeneous treatment effects.
+It requires that any patient who is (not) batched by a sequencer
+(batcher) would also (not) be batched by a batcher (sequencer)
+physician. The literature leveraging the judges design typically
+performs two informal tests for its implications. The first one provides
+that the first stage should be weakly positive for all subsamples
+(Dobbie, Goldin, and Yang 2018). The second implication asserts that the
+instrument constructed by leaving out a particular subsample has
+predictive power over that same left-out subsample (Bhuller et
+al.~2020). Appendix Table 2 presents both of these tests in the two
+columns for various subsamples of interest. Finally, we check whether
+our main results hold using differential, mutually exclusive leniency
+measures (i.e., by complaint category) in Section 4.4.
+
+\hypertarget{sec:4}{%
+\section{Results}\label{sec:4}}
 
 \hypertarget{reduced-form-results}{%
-\subsubsection{Reduced-Form Results}\label{reduced-form-results}}
+\subsection{Reduced-Form Results}\label{reduced-form-results}}
 
@@
 tests).
+
+\hypertarget{placebo-check}{%
+\subsection{Placebo Check}\label{placebo-check}}
+
+\hypertarget{instrumental-variables-results}{%
+\subsection{Instrumental Variables
+Results}\label{instrumental-variables-results}}
+
+\hypertarget{robustness}{%
+\subsection{Robustness}\label{robustness}}
 
@@
 
-\begin{table}[ht]
-\centering
-\caption{Regression Results by Chief Complaint} 
-\begin{tabular}{llrr}
-  \hline
- & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
-  \hline
-1 & Extremity Complaints & -0.72** & 0.03 \\ 
-  2 & Upper Respiratory Symptoms & 0.426** & 0.02 \\ 
-  3 & Abnormal Test Results & -0.067 & 0.84 \\ 
-  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.423* & 0.10 \\ 
-  5 & Shortness of Breath & -0.065 & 0.58 \\ 
-  6 & Gastrointestinal Issues & 0.071 & 0.67 \\ 
-  7 & Chest Pain & 0.107 & 0.68 \\ 
-  8 & Dizziness/Lightheadedness/Syncope & -0.163 & 0.32 \\ 
-  9 & Neurological Issue & -0.353 & 0.25 \\ 
-  10 & Fatigue and Weakness & -0.112 & 0.49 \\ 
-  11 & Back or Flank Pain & -0.602** & 0.02 \\ 
-  12 & Abdominal Complaints & -0.105 & 0.13 \\ 
-  13 & Fevers, Sweats or Chills & 0.01 & 0.95 \\ 
-  14 & Skin Complaints & 1.442 & 0.35 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\caption{Regression Results by Chief Complaint (2)} 
-\begin{tabular}{llrr}
-  \hline
- & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
-  \hline
-1 & Extremity Complaints & 0.176 & 0.50 \\ 
-  2 & Upper Respiratory Symptoms & 0.332*** & 0.00 \\ 
-  3 & Abnormal Test Results & 0.695 & 0.14 \\ 
-  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & -0.247 & 0.20 \\ 
-  5 & Shortness of Breath & -0.253** & 0.01 \\ 
-  6 & Gastrointestinal Issues & 0.421*** & 0.00 \\ 
-  7 & Chest Pain & -0.531*** & 0.00 \\ 
-  8 & Dizziness/Lightheadedness/Syncope & 0.075 & 0.60 \\ 
-  9 & Neurological Issue & -0.305 & 0.15 \\ 
-  10 & Fatigue and Weakness & -0.125 & 0.40 \\ 
-  11 & Back or Flank Pain & 0.288 & 0.13 \\ 
-  12 & Abdominal Complaints & -0.023 & 0.66 \\ 
-  13 & Fevers, Sweats or Chills & -0.044 & 0.71 \\ 
-  14 & Skin Complaints & 0.328 & 0.35 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-\begin{table}[ht]
-\centering
-\caption{First Stage over Subsamples} 
-\begin{tabular}{llrr}
-  \hline
- & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
-  \hline
-1 & Extremity Complaints & 0.761*** & 0.00 \\ 
-  2 & Upper Respiratory Symptoms & 1.406*** & 0.00 \\ 
-  3 & Abnormal Test Results & 1.036*** & 0.00 \\ 
-  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.831*** & 0.00 \\ 
-  5 & Shortness of Breath & 1.258*** & 0.00 \\ 
-  6 & Gastrointestinal Issues & 1.043*** & 0.00 \\ 
-  7 & Chest Pain & 0.866*** & 0.00 \\ 
-  8 & Dizziness/Lightheadedness/Syncope & 1.668*** & 0.00 \\ 
-  9 & Neurological Issue & 0.723*** & 0.00 \\ 
-  10 & Fatigue and Weakness & 1.651*** & 0.00 \\ 
-  11 & Back or Flank Pain & 0.752*** & 0.01 \\ 
-  12 & Abdominal Complaints & 1.109*** & 0.00 \\ 
-  13 & Fevers, Sweats or Chills & 1.23*** & 0.00 \\ 
-  14 & Skin Complaints & 0.448* & 0.07 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
 \end{APPENDIX}

### 2023-10-06 src/clean.R

@@ -592,7 +592,6 @@ final <- final %>%
 
 
 rm(list = setdiff(ls(), "final"))
-
 #=========================================================================
 ##########################################################################
 #=========================================================================

### 2023-10-06 src/figures.R

@@ -11,55 +11,8 @@ library(sandwich)
 library(ggsci)
 library(xtable) # Output to LaTeX table format
 
-##########################################################################
-#=========================================================================
-# Table []
-#   - Show that across physician, there is random assignment and patient
-#     characteristics are balanced
-#=========================================================================
+final <- read_csv('final.csv')
 
-fig.1 <- final
-
-chief_complaint_freq <- table(fig.1$CHIEF_COMPLAINT)
-
-top_10_chief_complaints <- names(chief_complaint_freq)[order(chief_complaint_freq, 
-                                                             decreasing = TRUE)][1:10]
-
-fig.1$CHIEF_COMPLAINT <- ifelse(fig.1$CHIEF_COMPLAINT %in% top_10_chief_complaints, 
-                                fig.1$CHIEF_COMPLAINT, "DROP")
-
-dummy <- dummyVars(" ~ CHIEF_COMPLAINT", data = fig.1)
-one.hot <- data.frame(predict(dummy, newdata = fig.1))
-vars <- setdiff(names(one.hot), 'CHIEF_COMPLAINTDROP')
-
-fig.1 <- cbind(fig.1, one.hot)
-
-fig.1$ESI1.2 <- ifelse(fig.1$ESI == 1 | fig.1$ESI == 2, 1, 0)
-fig.1$ESI3.4.5 <- ifelse(fig.1$ESI %in% c(3, 4, 5), 1, 0)
-
-vars <- c(vars, 'ESI1.2', 'ESI3.4.5')
-
-# Model and collect balance statistics
-balance <- data.frame(Df = numeric(), 
-                      F = numeric(),
-                      Pr..F. = numeric(), 
-                      dummy = character())
-
-for (dummy in vars) {
-  model_pos_1 <- lm(as.formula(paste(dummy, '~ 1')), fig.1)
-  model_pos_2 <- lm(as.formula(paste(dummy, '~ ED_PROVIDER')), fig.1)
-  
-  wald_pos <- waldtest(model_pos_1, model_pos_2, 
-                       vcov = vcovHC(model_pos_2, type = "HC1"))
-  
-  temp <- data.frame(wald_pos)[2, c(2,3,4)]
-  temp$dummy <- dummy
-  
-  balance <- rbind(balance, temp)
-}
-
-print(xtable(balance, 
-             caption = "Wald Test Results"), type = "latex")
 
 ##########################################################################
 
@@ -69,9 +22,9 @@ print(xtable(balance,
[…]

### 2023-11-07 manuscript/manuscript.Rmd
mediation analysis

@@
 Our analytical lens is focused on the Emergency Department at the Mayo Clinic of Arizona, a distinguished tertiary care establishment.
-During our study's timeframe, the ED recorded an annual visitation of approximately 41,000 patients.
+During our study's timeframe, the ED recorded an annual visitation of approximately 45,000 patients.
 The department is singularly staffed by board-eligible or board-certified emergency physicians, abstaining from the services of nurse practitioners or physician assistants.
@@
 
-In the context of diagnostic test ordering, we defined batching as placing multiple test orders within 5 minutes of one another.
-Sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-Our study differentiated between two distinct batching strategies:
+Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
+We observe approximately 45,000 such visits until during the stufy period.
+To improve power, we drop encounters with rare chief complaints (< 1000 total encounters of this kind) and complaints where a batch order occurs less than 10 percent of the time.
+Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem were we to include them.
+Example complaints dropped include urinary complaints and _____.
+Excluding these conditions does not introduce selection bias only if physician test batching tendency is orthogonal to physician diagnosing behavior.
+While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (“Fracture of ankle”), but are unlikely to disagree at the broader chief complaint cateogry.}.
 
-\begin{itemize}
-\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.
+## Variable Definitions
 
-\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
-\end{itemize}
+### Outcomes
 
-The distinction is necessary because laboratory tests may be processed in parallel with imaging tests, which have a much more limited effect on patient waiting time.
-However, a batch of multiple imaging tests requires that the patient is present for both tests.
-Thus, waiting times are much more likely to be impacted.
-Therefore, the specific tests that make up a batch may provide insight into how the physician trades off between waiting times and uncertainty, given the various complexity levels of patient encounters.
-Figure 1 shows the rate at which each batching type occurs for the top ten most frequently occurring chief complaint fields.
+In our instrumental variables (IV) analysis, the key explanatory variable, $Tendency_i$, is an indicator for whether patient $i$ has their tests batch ordered at their ED encounter.
+This variable is crucial for understanding the causal effects of batch testing on various health outcomes and resource utilization.
+Below, we detail our main outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.
+
+*ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
+It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
+Although literature provides a range for long ED-LOS, for the purpose of this study, we operationalize long ED-LOS as stays exceeding the 75th percentile in our dataset.
+This approach allows us to contextualize the definition of long stays according to the specific demographics and case mix of the EDs under study.
+
+*Resource Utilization*-- Resource utilization in the ED context typically refers to the extent of medical services and interventions a patient receives.
+In this study, we quantify resource utilization by the total number of diagnostic tests ordered per patient during their ED stay.
+This encompasses both initial and any subsequent tests.
+The hypothesis is that batch testing may lead to variations in the number of tests ordered, potentially influencing the overall healthcare expenditure and efficiency.
+
+*72-hour Return*-- The rate of patients returning to the ED within 72 hours of their initial visit serves as a proxy for the quality of care received.
+A high 72-hour return rate might indicate inadequate treatment or diagnostic oversight during the initial visit.
+We measure this outcome as the proportion of patients who have a subsequent ED visit within 72 hours post-initial discharge.
+
+### Batching
+
+In the context of our study, batching in diagnostic test ordering is defined as placing multiple test orders for a single patient within a 5-minute interval.
+To ensure the robustness of our analysis, sensitivity tests were conducted on this cutoff point, affirming that our results are consistent and reliable irrespective of minor variations in the batching definition.
+
+Our study identifies and differentiates between two distinct strategies of batching:
+
+* Lab + Image Batch: This category of batch order includes instances where a patient’s batched diagnostic tests comprise a combination of at least one imaging test (such as a contrasted CT scan, non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
+This mix is particularly interesting as it suggests a comprehensive diagnostic approach, possibly catering to complex medical cases.
+
+* Image + Image Batch: This category encompasses batch orders where a patient’s diagnostic tests consist of two or more distinct imaging tests.
+The reliance solely on imaging in these batches may indicate specific diagnostic pathways or a focus on conditions that primarily require visual diagnostic methods.
+
+It is important to note the differing implications of these batching strategies.
+While lab tests often can be processed concurrently with imaging tests, causing minimal additional waiting time, a batch of multiple imaging tests requires the patient’s presence for each test, potentially leading to significant increases in waiting time.
+Thus, the composition of a batch order can provide valuable insights into the physician’s decision-making process, particularly in balancing between patient waiting times and diagnostic uncertainty.
+This balance can be especially crucial in the context of varying complexity levels of patient encounters.
+
+Figure 1 illustrates the occurrence rates of each batching type across the top ten most frequently reported chief complaints, providing a visual representation of how batching practices vary with different patient presentations.
 
@@
   \centering
-  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
-  \tiny
-  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
+  \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
+\begin{tablenotes}
+\small
+\item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
+\end{tablenotes}  
 \end{figure}
 
-## Statistics
+## Summary Statistics
 
-Table 2 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered individually in a batched or sequential way during an ED encounter from our baseline sample.
-About 39% of patient encounters involved diagnostic tests ordered as part of a batch, most of which (64%) included two tests.
+\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview of the emergency department (ED) characteristics, patient demographics, and medical tests for our baseline sample, derived from the data collected during the study period.
+On average, the emergency department manages a volume of approximately 24 patients (Mean = 24.18), indicating a significant but manageable patient load.
+Key physiological markers such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are prevalent among patients, albeit at varying degrees, highlighting critical aspects of emergency care.
+Notably, a small proportion of patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01) conditions, emphasizing the diversity of cases encountered in the ED setting.
+The Emergency Severity Index (ESI) averages at 2.78, suggesting that most cases fall within the moderate acuity level.
 
-Notable distinctions emerge when comparing the two cohorts.
-Patients subjected to batched test ordering exhibited a notably prolonged stay in the emergency department (ED) (p<0.001) and were more frequently assigned a lower Emergency Severity Index (ESI).
-Furthermore, this group skewed towards older demographics, with half of them falling into the 65+ age category.
-Patients in the batch-order group also displayed a higher likelihood of presenting symptoms at triage, including tachycardia (p<0.001), tachypnea (p<0.001), fever (p<0.001), and hypotension (p<0.001).
-Delving deeper into the ESI data unveils a higher concentration of patients with more critical conditions (ESI 1 and 2) within the batched group, in contrast to the sequential group (45.1% with ESI of 1 or 2 in the batched group compared to 27.5% with ESI of 1 or 2 in the sequential group).
-This finding suggests that factors related to patient complexity and severity (Saghafian et al., 2014) significantly influence physicians' decisions to opt for batch-ordering of diagnostic tests.
-However, it is important to note that our investigation revealed substantial variation among physicians in their choices between batching and sequencing for a given patient.
+Common chief complaints in the emergency department paint a vivid picture of patient needs and healthcare demands.
+The data reveals abdominal complaints (Mean = 0.14), extremity complaints (Mean = 0.12), and chest pain (Mean = 0.08) as the most frequently occurring, followed by neurological (Mean = 0.08) and gastrointestinal issues (Mean = 0.08).
+This information is crucial in understanding the primary reasons for ED visits and the required resources for patient care.
 
-\begin{table}
+The patient characteristics in our sample depict a diverse demographic landscape.
+The average arrival age in the ED is approximately 58 years (Mean = 58.33), indicating a predominantly middle-aged population.
+The racial composition is predominantly White (Mean = 0.89), with smaller proportions of Black (Mean = 0.04) and Asian (Mean = 0.03) patients.
+Notably, a slight majority of the patients are female (Mean = 0.54), offering insights into gender dynamics in healthcare utilization.
+
+In terms of diagnostic tests, the data reveals a high reliance on lab tests (Mean = 0.77), indicating their crucial role in patient diagnosis and management.
+X-Ray (Mean = 0.47) and CT scans, both non-contrast (Mean = 0.21) and contrast (Mean = 0.19), are also frequently employed, underscoring the importance of imaging in modern emergency medicine.
+Ultrasound usage (Mean = 0.11) and the practice of batch ordering tests (Mean = 0.38) further reflect the operational aspects of the ED and the strategies employed to manage patient care effectively.
+
+\begin{table}[ht]
 \centering
 \caption{Summary Statistics}
-\begin{tabular}{p{3cm} *{4}{p{2.5cm}}}
+\label{tab:summary_statistics}
+\begin{tabular}{p{9cm}cccc}
 \toprule
-Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
-&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
+\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
 \midrule
-ESI & & & & $<0.001$ \\
-1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
-2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
-3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
-4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
-5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
-(Missing) & 27 & 16 & 11 & \\
+\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
+ED Volume & 24.18 & 15 & 25 & 32 \\
+Tachycardic & 0.19 & & & \\
+Tachypneic & 0.09 & & & \\
+Febrile & 0.02 & & & \\
+Hypotensive & 0.01 & & & \\ 
+ESI & 2.78 & 2 & 3 & 3 \\
+Complaint: Abdominal & 0.14 & & & \\
+Complaint: Extremity & 0.12 & & & \\
+Complaint: Chest Pain & 0.08 & & & \\
+Complaint: Neurological & 0.08 & & & \\
+Complaint: Gastrointestinal & 0.08 & & & \\
 \midrule
-Age & & & & $<0.001$ \\
-<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
-20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
-45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
-65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
+\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
+Arrival Age & 58.33 & 44 & 62 & 75 \\
+Race: White & 0.89 & & & \\
+Race: Black & 0.04 & & & \\
+Race: Asian & 0.03 & & & \\
+Gender: Female & 0.54 & & & \\
 \midrule
-ED LOS (min) & & & & $<0.001$ \\
-Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
-\midrule
-Disposition & & & & $<0.001$ \\
-Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
-Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
-Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
-\midrule
-Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
-Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
-Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
-Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
-\midrule
-Number of Tests & & & & $<0.001$ \\
-1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
-2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
-3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
-4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
-5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
-\midrule
-Gender & & & & $<0.001$ \\
-Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
-Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
-\midrule
-Race & & & & $0.002$ \\
-White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
-Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
-Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
-Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
-Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
-Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
+\multicolumn{5}{l}{\textbf{Tests}} \\
+X-Ray & 0.47 & & & \\
+Ultrasound & 0.11 & & & \\
+Non-Contrast CT & 0.21 & & & \\
+Contrast CT & 0.19 & & & \\
+Lab & 0.77 & & & \\
+Tests were Batch Ordered & 0.38 & & & \\
 \bottomrule
-\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
-\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
 \end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Notes: This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text.}
+\end{tablenotes}
 \end{table}
+
+# Empirical Strategy {#sec:3}
+
+Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court.
+Similarly, we explore batching variation across physicians who work in the same emergency department.
+In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians.
+Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision.
+We employ both approaches and lay out their details in the next subsections.
 
@@
 In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Table 2 displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+\hyperref[tab:summary_statistics]{Table 2} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
 
@@
     \caption{Wald Test Results}
-    \begin{tabular}{lcc}
+    \label{tab:wald_test}
+    \begin{tabular}{p{10cm}cc}
         \toprule
@@
 \begin{tablenotes}
-\tiny
-\item Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
+\small
+\item \textit{Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
 A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization.
 The Wald F-statistic and p-value are reported.
-Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.
+Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
 \end{tablenotes}
@@
 
-# Empirical Strategy {#sec:3}
-
-Our study employs the ``judges design", a quasi-random assignment strategy widely used in literature to investigate causal effects.
-Typically, this design exploits variations in sentencing leniency among judges working in the same court.
-We adopt a similar approach by examining batching variations across physicians working in the same emergency department.
-This method enables us to identify the causal impact of being assigned to different types of physicians with differing batch-ordering tendencies.
+## Batch Tendency Construction
 
@@
 \begin{equation}
-Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \varepsilon_{i,t}
+Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint} + \varepsilon_{i,t}
 \end{equation}
@@
 We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$.
+Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision.
 As stated earlier, these controls are what is required for our quasi-random assignment assumption.
@@
 Tendency_{i,j}^{phys} =
-\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{e}_{i',k}
+\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{\varepsilon}_{i'}
 \end{equation}
 
-where $\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$.
+where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$.
 This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument.
@@
 We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure 2 graphs the histogram of batch-ordering frequency by physician for populat chief complaints, highlighting that the variation in batching differs systematically.
-Table 3 presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 10 pp increase in the likelihood of having tests batch-ordered in the ED.
-The F-statistic is 51 when all controls and fixed effects are included.
+Figure 2 graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically.
+\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 5 pp increase in the likelihood of having tests batch-ordered in the ED.
+The F-statistic is 50 when all controls and fixed effects are included.
+The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
 
-\begin{table}[htbp]
+\begin{table}[ht]
 \centering
 \caption{First Stage Regression Results}
-\begin{threeparttable}
-\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
+\label{tab:regression}
+\begin{tabular}{p{9cm}ccc}
 \toprule
- & Model 1 & Model 2 & Model 3 \\
+\textbf{} & \textbf{Model 1} & \textbf{Model 2} & \textbf{Model 3} \\
 \midrule
-Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
+Batching Tendency & 1.05 *** & 1.04 *** & 1.04 *** \\
  & (0.01) & (0.01) & (0.01) \\
 \midrule
-F statistic (full model)&7.32&52.03&50.93\\  
-F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
+F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
+F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
 \midrule
-Num. obs. & 43300 & 43300 & 43300 \\
+Num. obs. & 43327 & 43327 & 43327 \\
 Seasonality and shift fixed effects? & Yes & Yes & Yes \\
@@
 \bottomrule
-\end{tabularx}
+\end{tabular}
 \begin{tablenotes}
-\tiny
-\item Estimates of the first stage for the baseline sample described in the text.
+\small
+\item \textit{Estimates of the first stage for the baseline sample described in the text.
 Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
@@
 Column 3 corresponds to the baseline controls.
-Robust standard errors are clustered at the physician level.
+Robust standard errors are clustered at the physician level.}
 \item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
-\end{tablenotes}\tiny
-\end{threeparttable}
-\label{table:regression}
+\end{tablenotes}
 \end{table}
@@
 In this section, we examine the causal effects of provider batching tendency.
-We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).
+We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).We find statistically and economically significant positive effects of assignment to a high batching physician on every outcome (Table 3).
+Estimated coefficients are scaled by the difference in leniency going from the nineti- eth to tenth percentile in physician leniency—equal to 11.6 pp—for interpretability.
+Assignment to a physician in the top prescribing decile (relative to one in the bottom decile) is associated with a 0.24 pp increase in the probability of long-term prescrip- tion opioid use—a 4.1 percent increase.
+The effect on proxies for opioid-seeking behavior is 0.50 pp—a 3.4 percent increase.
+Effects on development of opioid use disorder and on opioid overdose mortality are 2.1 percent and 8.8 percent, respec- tively.
+These findings highlight the substantial role ED physicians can play in put- ting patients on a path of long-term opioid use and dependence.
 
 ## Placebo Check
+
+In this section we investigate whether the reduced-form effects observed in Section 4.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency We start by studying reduced-form effects among patients with complaints that are never prescribed opioids, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
+For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
+Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.
+
+To that end, we restrict attention to ED visits for complaints where batching occurs no more than 10 percent of the time (recall that our baseline sample only includes complaints with a >10 percent batching rate).
+We estimate a reduced-form regression of each main outcome on physician prescribing leniency for the subsample, following equation (3).
+The results of this exercise are displayed in Table 4.
+They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
+
+To further probe robustness, we perform the placebo exercise for a set of non-opioid outcomes—in the form of homelessness, suicide, all-cause mortality, and preventable hospitalizations—in online Appendix Table G.7.
+Note that these outcomes are likely to be more sensitive to other dimensions of physician care than our opioid-related outcomes because the causal link to prescription opioid exposure is less tight, while other margins of care may matter relatively more.
+Indeed, we do find a sizable and statistically significant effect of being assigned to a high-prescribing physician on preventable hospitalizations and all-cause mor- tality for our placebo conditions.
+For patients with ED diagnoses that are rarely prescribed an opioid (column 3), assignment to a physician in the top decile of the opioid prescribing range increases the likelihood of a preventable hospitalization within 3 years by 0.27 pp, or 1.8 percent.19
+
+The failure of our placebo check for some non-opioid outcomes suggests that there may be dimensions of care correlated with opioid prescribing tendency that vary across ED physicians and may impact non-opioid patient outcomes.
+While the set of potentially relevant dimensions of care is very large and may include dimen- sions unobservable to the researcher (such as, for example, instructions on medica- tion adherence and use, physician-nurse scheduling that lead to complementarities), we can make progress by accounting for the chief observable dimensions likely to correlate highly with patient health outcomes in ED settings: quality of care as mea- sured by immediate patient mortality, intensity of procedures, and tendency to admit patients for inpatient hospitalizations.
+In moving toward a well-identified IV design that estimates the causal impact of an opioid prescription on patient outcomes, we address these potential violations to the exclusion restriction by estimating physician “propensities” along these “non-focal” (relative to opioid prescribing) dimensions and including them as controls (details are provided in Section IIIE).
+This approach follows recent advances from the judge stringency literature (Mueller-Smith 2015; Bhuller et al.
+2020), which deals with similar concerns.
+
+We recognize that given the multidimensionality of physician behavior, potential for exclusion restriction violations remains even after controlling for chief dimensions of behavior.
+For this reason, we also present reduced-form results, and we keep a tight focus on opioid-related outcomes, for which such concerns are less pronounced relative to general health outcomes.
 

### 2023-11-07 manuscript/manuscript.tex
mediation analysis

@@
 our study's timeframe, the ED recorded an annual visitation of
-approximately 41,000 patients. The department is singularly staffed by
+approximately 45,000 patients. The department is singularly staffed by
 board-eligible or board-certified emergency physicians, abstaining from
@@
 
-In the context of diagnostic test ordering, we defined batching as
-placing multiple test orders within 5 minutes of one another.
-Sensitivity analyses on this cutoff point showed that our results are
-robust to this definition. Our study differentiated between two distinct
-batching strategies:
+Our research design focuses on adults who visit the Mayo Clinic of
+Arizona ED. We observe approximately 45,000 such visits until during the
+stufy period. To improve power, we drop encounters with rare chief
+complaints (\textless{} 1000 total encounters of this kind) and
+complaints where a batch order occurs less than 10 percent of the time.
+Since batch orders are rare for these cases, our physician batch
+tendency instrument could suffer from a weak instrument problem were we
+to include them. Example complaints dropped include urinary complaints
+and \_\_\_\_\_. Excluding these conditions does not introduce selection
+bias only if physician test batching tendency is orthogonal to physician
+diagnosing behavior. While this assumption may be violated if we were to
+use a very detailed level of chief complaint information upon which to
+base our exclusion criterion, it is plausibly satisfied when using broad
+complaint
+categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (“Fracture of ankle”), but are unlikely to disagree at the broader chief complaint cateogry.}.
+
+\hypertarget{variable-definitions}{%
+\subsection{Variable Definitions}\label{variable-definitions}}
+
+\hypertarget{outcomes}{%
+\subsubsection{Outcomes}\label{outcomes}}
+
+In our instrumental variables (IV) analysis, the key explanatory
+variable, \(Tendency_i\), is an indicator for whether patient \(i\) has
+their tests batch ordered at their ED encounter. This variable is
+crucial for understanding the causal effects of batch testing on various
+health outcomes and resource utilization. Below, we detail our main
+outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c)
+72-hour return.
+
+\emph{ED length of stay (LOS).}-- ED-LOS is a critical measure of
+efficiency and patient throughput in emergency care settings. It is
+defined as the duration from a patient's arrival to the ED until their
+departure, whether by discharge or admission to the hospital. Although
+literature provides a range for long ED-LOS, for the purpose of this
+study, we operationalize long ED-LOS as stays exceeding the 75th
+percentile in our dataset. This approach allows us to contextualize the
+definition of long stays according to the specific demographics and case
+mix of the EDs under study.
+
+\emph{Resource Utilization}-- Resource utilization in the ED context
+typically refers to the extent of medical services and interventions a
+patient receives. In this study, we quantify resource utilization by the
+total number of diagnostic tests ordered per patient during their ED
+stay. This encompasses both initial and any subsequent tests. The
+hypothesis is that batch testing may lead to variations in the number of
+tests ordered, potentially influencing the overall healthcare
+expenditure and efficiency.
+
+\emph{72-hour Return}-- The rate of patients returning to the ED within
+72 hours of their initial visit serves as a proxy for the quality of
+care received. A high 72-hour return rate might indicate inadequate
+treatment or diagnostic oversight during the initial visit. We measure
+this outcome as the proportion of patients who have a subsequent ED
+visit within 72 hours post-initial discharge.
+
+\hypertarget{batching}{%
+\subsubsection{Batching}\label{batching}}
+
+In the context of our study, batching in diagnostic test ordering is
+defined as placing multiple test orders for a single patient within a
+5-minute interval. To ensure the robustness of our analysis, sensitivity
+tests were conducted on this cutoff point, affirming that our results
+are consistent and reliable irrespective of minor variations in the
+batching definition.
+
+Our study identifies and differentiates between two distinct strategies
+of batching:
 
 \begin{itemize}
-\item Lab + Image Batch: The batch order for a patient consisted of two or more distinct diagnostic tests where at least one imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) and one lab test were placed.
-
-\item Image + Image Batch: The batch order for a patient consisted of two or more distinct imaging (contrasted CT scan, non-contrasted CT scan, X-ray, ultrasound) diagnostic tests.
+\item
+  Lab + Image Batch: This category of batch order includes instances
+  where a patient's batched diagnostic tests comprise a combination of
+  at least one imaging test (such as a contrasted CT scan,
+  non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
+  This mix is particularly interesting as it suggests a comprehensive
+  diagnostic approach, possibly catering to complex medical cases.
+\item
+  Image + Image Batch: This category encompasses batch orders where a
+  patient's diagnostic tests consist of two or more distinct imaging
+  tests. The reliance solely on imaging in these batches may indicate
+  specific diagnostic pathways or a focus on conditions that primarily
+  require visual diagnostic methods.
 \end{itemize}
 
-The distinction is necessary because laboratory tests may be processed
-in parallel with imaging tests, which have a much more limited effect on
-patient waiting time. However, a batch of multiple imaging tests
-requires that the patient is present for both tests. Thus, waiting times
-are much more likely to be impacted. Therefore, the specific tests that
-make up a batch may provide insight into how the physician trades off
-between waiting times and uncertainty, given the various complexity
-levels of patient encounters. Figure 1 shows the rate at which each
-batching type occurs for the top ten most frequently occurring chief
-complaint fields.
+It is important to note the differing implications of these batching
+strategies. While lab tests often can be processed concurrently with
+imaging tests, causing minimal additional waiting time, a batch of
+multiple imaging tests requires the patient's presence for each test,
+potentially leading to significant increases in waiting time. Thus, the
+composition of a batch order can provide valuable insights into the
+physician's decision-making process, particularly in balancing between
+patient waiting times and diagnostic uncertainty. This balance can be
+especially crucial in the context of varying complexity levels of
+patient encounters.
+
+Figure 1 illustrates the occurrence rates of each batching type across
+the top ten most frequently reported chief complaints, providing a
+visual representation of how batching practices vary with different
+patient presentations.
 
@@
   \centering
-  \includegraphics[width=1\textwidth]{../manuscript/figures/frequency_batches.png}
-  \tiny
-  \caption{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
+  \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
+\begin{tablenotes}
+\small
+\item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
+\end{tablenotes}  
 \end{figure}
 
-\hypertarget{statistics}{%
-\subsection{Statistics}\label{statistics}}
+\hypertarget{summary-statistics}{%
+\subsection{Summary Statistics}\label{summary-statistics}}
 
-Table 2 provides a comprehensive breakdown of the patient encounters
-involving diagnostic tests ordered individually in a batched or
-sequential way during an ED encounter from our baseline sample. About
-39\% of patient encounters involved diagnostic tests ordered as part of
-a batch, most of which (64\%) included two tests.
+\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview
+of the emergency department (ED) characteristics, patient demographics,
+and medical tests for our baseline sample, derived from the data
+collected during the study period. On average, the emergency department
+manages a volume of approximately 24 patients (Mean = 24.18), indicating
+a significant but manageable patient load. Key physiological markers
+such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are
+prevalent among patients, albeit at varying degrees, highlighting
+critical aspects of emergency care. Notably, a small proportion of
+patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01)
+conditions, emphasizing the diversity of cases encountered in the ED
+setting. The Emergency Severity Index (ESI) averages at 2.78, suggesting
+that most cases fall within the moderate acuity level.
 
-Notable distinctions emerge when comparing the two cohorts. Patients
-subjected to batched test ordering exhibited a notably prolonged stay in
-the emergency department (ED) (p\textless0.001) and were more frequently
-assigned a lower Emergency Severity Index (ESI). Furthermore, this group
-skewed towards older demographics, with half of them falling into the
-65+ age category. Patients in the batch-order group also displayed a
-higher likelihood of presenting symptoms at triage, including
-tachycardia (p\textless0.001), tachypnea (p\textless0.001), fever
-(p\textless0.001), and hypotension (p\textless0.001). Delving deeper
-into the ESI data unveils a higher concentration of patients with more
-critical conditions (ESI 1 and 2) within the batched group, in contrast
-to the sequential group (45.1\% with ESI of 1 or 2 in the batched group
-compared to 27.5\% with ESI of 1 or 2 in the sequential group). This
-finding suggests that factors related to patient complexity and severity
-(Saghafian et al., 2014) significantly influence physicians' decisions
-to opt for batch-ordering of diagnostic tests. However, it is important
-to note that our investigation revealed substantial variation among
-physicians in their choices between batching and sequencing for a given
-patient.
+Common chief complaints in the emergency department paint a vivid
+picture of patient needs and healthcare demands. The data reveals
+abdominal complaints (Mean = 0.14), extremity complaints (Mean = 0.12),
+and chest pain (Mean = 0.08) as the most frequently occurring, followed
+by neurological (Mean = 0.08) and gastrointestinal issues (Mean = 0.08).
+This information is crucial in understanding the primary reasons for ED
+visits and the required resources for patient care.
 
-\begin{table}
+The patient characteristics in our sample depict a diverse demographic
+landscape. The average arrival age in the ED is approximately 58 years
+(Mean = 58.33), indicating a predominantly middle-aged population. The
+racial composition is predominantly White (Mean = 0.89), with smaller
+proportions of Black (Mean = 0.04) and Asian (Mean = 0.03) patients.
+Notably, a slight majority of the patients are female (Mean = 0.54),
+offering insights into gender dynamics in healthcare utilization.
+
+In terms of diagnostic tests, the data reveals a high reliance on lab
+tests (Mean = 0.77), indicating their crucial role in patient diagnosis
+and management. X-Ray (Mean = 0.47) and CT scans, both non-contrast
+(Mean = 0.21) and contrast (Mean = 0.19), are also frequently employed,
+underscoring the importance of imaging in modern emergency medicine.
+Ultrasound usage (Mean = 0.11) and the practice of batch ordering tests
+(Mean = 0.38) further reflect the operational aspects of the ED and the
+strategies employed to manage patient care effectively.
+
+\begin{table}[ht]
 \centering
 \caption{Summary Statistics}
-\begin{tabular}{p{3cm} *{4}{p{2.5cm}}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}p{2.5cm}}
+\label{tab:summary_statistics}
+\begin{tabular}{p{9cm}cccc}
 \toprule
-Variable & Overall Visits & Sequentially Arm & Batch Arm & $p$-value \\
-&  ($N = 41,846$) & ($N = 25,354$)  & ($N = 16,492$) & \\
+\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
 \midrule
-ESI & & & & $<0.001$ \\
-1 & 476 (1.1\%) & 136 (0.5\%) & 340 (2.1\%) & \\
-2 & 14,020 (34\%) & 6,865 (27\%) & 7,155 (43\%) & \\
-3 & 24,006 (57\%) & 15,517 (61\%) & 8,489 (52\%) & \\
-4 & 3,283 (7.9\%) & 2,787 (11\%) & 496 (3.0\%) & \\
-5 & 34 (<0.1\%) & 33 (0.1\%) & 1 (<0.1\%) & \\
-(Missing) & 27 & 16 & 11 & \\
+\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
+ED Volume & 24.18 & 15 & 25 & 32 \\
+Tachycardic & 0.19 & & & \\
+Tachypneic & 0.09 & & & \\
+Febrile & 0.02 & & & \\
+Hypotensive & 0.01 & & & \\ 
+ESI & 2.78 & 2 & 3 & 3 \\
+Complaint: Abdominal & 0.14 & & & \\
+Complaint: Extremity & 0.12 & & & \\
+Complaint: Chest Pain & 0.08 & & & \\
+Complaint: Neurological & 0.08 & & & \\
+Complaint: Gastrointestinal & 0.08 & & & \\
 \midrule
-Age & & & & $<0.001$ \\
-<20 & 1,413 (3.4\%) & 1,154 (4.6\%) & 259 (1.6\%) & \\
-20-45 & 9,419 (23\%) & 6,538 (26\%) & 2,881 (17\%) & \\
-45-65 & 12,829 (31\%) & 7,764 (31\%) & 5,065 (31\%) & \\
-65+ & 18,185 (43\%) & 9,898 (39\%) & 8,287 (50\%) & \\
+\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
+Arrival Age & 58.33 & 44 & 62 & 75 \\
+Race: White & 0.89 & & & \\
+Race: Black & 0.04 & & & \\
+Race: Asian & 0.03 & & & \\
+Gender: Female & 0.54 & & & \\
 \midrule
-ED LOS (min) & & & & $<0.001$ \\
-Mean (SD) & 269 (379) & 255 (174) & 290 (564) & \\
-\midrule
-Disposition & & & & $<0.001$ \\
-Admit & 9,124 (22\%) & 4,286 (17\%) & 4,838 (29\%) & \\
-Discharge & 25,861 (62\%) & 17,596 (69\%) & 8,265 (50\%) & \\
-Other & 6,861 (16\%) & 3,472 (14\%) & 3,389 (21\%) & \\
-\midrule
-Tachycardic & 8,328 (20\%) & 4,853 (19\%) & 3,475 (21\%) & $<0.001$ \\
-Tachypneic & 3,953 (9.4\%) & 1,818 (7.2\%) & 2,135 (13\%) & $<0.001$ \\
-Febrile & 1,013 (2.4\%) & 386 (1.5\%) & 627 (3.8\%) & $<0.001$ \\
-Hypotensive & 69 (1.6\%) & 327 (1.3\%) & 342 (2.1\%) & $<0.001$ \\
-\midrule
-Number of Tests & & & & $<0.001$ \\
-1 & 13,406 (32\%) & 13,406 (53\%) & 0 (0\%) & \\
-2 & 19,291 (46\%) & 8,782 (35\%) & 10,509 (64\%) & \\
-3 & 7,707 (18\%) & 2,689 (11\%) & 5,018 (30\%) & \\
-4 & 1,348 (3.2\%) & 451 (1.8\%) & 897 (5.4\%) & \\
-5 & 94 (0.2\%) & 26 (0.1\%) & 68 (0.4\%) & \\
-\midrule
-Gender & & & & $<0.001$ \\
-Female & 22,441 (54\%) & 14,053 (55\%) & 8,388 (51\%) & \\
-Male & 19,405 (46\%) & 11,301 (45\%) & 8,104 (49\%) & \\
-\midrule
-Race & & & & $0.002$ \\
-White & 37,070 (89\%) & 22,355 (88\%) & 14,715 (89\%) & \\
-Black & 1,727 (4.1\%) & 1,057 (4.2\%) & 670 (4.1\%) & \\
-Asian & 1,260 (3.0\%) & 775 (3.1\%) & 485 (2.9\%) & \\
-Native & 546 (1.3\%) & 354 (1.4\%) & 192 (1.2\%) & \\
-Other & 763 (1.8\%) & 493 (1.9\%) & 270 (1.6\%) & \\
-Unknown & 480 (1.1\%) & 320 (1.3\%) & 160 (1.0\%) & \\
+\multicolumn{5}{l}{\textbf{Tests}} \\
+X-Ray & 0.47 & & & \\
+Ultrasound & 0.11 & & & \\
+Non-Contrast CT & 0.21 & & & \\
+Contrast CT & 0.19 & & & \\
+Lab & 0.77 & & & \\
+Tests were Batch Ordered & 0.38 & & & \\
 \bottomrule
-\multicolumn{5}{l}{\textsuperscript{1} n (\%)} \\
-\multicolumn{5}{l}{\textsuperscript{2} Pearson's Chi-squared test; Welch Two Sample t-test} \\
 \end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Notes: This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text.}
+\end{tablenotes}
 \end{table}
+
+\hypertarget{sec:3}{%
+\section{Empirical Strategy}\label{sec:3}}
+
+Our empirical strategy closely follows the literature that relies on
+quasi-random assignment of agents to cases, often referred to as the
+``judges design.'' Papers in this literature typically exploit variation
+in the sentencing leniency of judges who work in the same court.
+Similarly, we explore batching variation across physicians who work in
+the same emergency department. In its reduced form, under the assumption
+of quasi-random assignment, this approach allows researchers to identify
+the causal effect of being assigned to different types of physicians.
+Under additional assumptions, an instrumental variable approach
+identifies the causal effect of a given medical decision. We employ both
+approaches and lay out their details in the next subsections.
 
@@
 for our analyses. In essence, barring arrival time and shift-level
-variation, the physician-to-patient matching can be deemed random. Table
-2 displays that patient encounters (regarding chief complaints and
-emergency severity) are equitably distributed across physicians within
-our study's cohort.
+variation, the physician-to-patient matching can be deemed random.
+\hyperref[tab:summary_statistics]{Table 2} displays that patient
+encounters (regarding chief complaints and emergency severity) are
+equitably distributed across physicians within our study's cohort.
 
@@
 
-\hypertarget{sec:3}{%
-\section{Empirical Strategy}\label{sec:3}}
-
-Our study employs the ``judges design'', a quasi-random assignment
-strategy widely used in literature to investigate causal effects.
-Typically, this design exploits variations in sentencing leniency among
-judges working in the same court. We adopt a similar approach by
-examining batching variations across physicians working in the same
-emergency department. This method enables us to identify the causal
-impact of being assigned to different types of physicians with differing
-batch-ordering tendencies.
+\hypertarget{batch-tendency-construction}{%
+\subsection{Batch Tendency
+Construction}\label{batch-tendency-construction}}
 
@@
 patient arrival with day of week-time of day fixed effects,
-\(\alpha_{dt}\). As stated earlier, these controls are what is required
-for our quasi-random assignment assumption. Under the assumption that we
-have captured the observables under which quasi-random assignment occurs
-in the ED, the unexplained variation-- the physician's contribution--
-resides in the error term, \(\varepsilon_{i,t}\).
+\(\alpha_{dt}\). Chief complaint by severity fixed effects,
+\(\alpha_{complaint}\), were also included to increase precision. As
+stated earlier, these controls are what is required for our quasi-random
+assignment assumption. Under the assumption that we have captured the
+observables under which quasi-random assignment occurs in the ED, the
+unexplained variation-- the physician's contribution-- resides in the
+error term, \(\varepsilon_{i,t}\).
 
@@
 Tendency_{i,j}^{phys} =
-\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{e}_{i',k}
+\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{\varepsilon}_{i'}
 \end{equation}
 
-where \(\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}\) is the
+where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
 residual from equation (1); \(J\) is the set of all ED encounters
@@
 graphs the histogram of batch-ordering frequency by physician for
-populat chief complaints, highlighting that the variation in batching
-differs systematically. Table 3 presents the ``first stage'' in a
-regression table: being assigned to a 10 pp higher tendency physician is
-associated with a 10 pp increase in the likelihood of having tests
-batch-ordered in the ED. The F-statistic is 51 when all controls and
-fixed effects are included.
+popular chief complaints, highlighting that the variation in batching
+differs systematically. \hyperref[tab:regression]{Table 3} presents the
+``first stage'' in a regression table: being assigned to a 10 pp higher
+tendency physician is associated with a 5 pp increase in the likelihood
+of having tests batch-ordered in the ED. The F-statistic is 50 when all
+controls and fixed effects are included. The coefficient is greater than
+one because all emergency visits are used to construct the leniency
+instrument, while the first stage is calculated using the baseline
+sample only, which excludes the rare complaints.
 
-\begin{table}[htbp]
+\begin{table}[ht]
 \centering
 \caption{First Stage Regression Results}
-\begin{threeparttable}
-\begin{tabularx}{0.85\textwidth}{@{}Xccc@{}}
+\label{tab:regression}
+\begin{tabular}{p{9cm}ccc}
 \toprule
- & Model 1 & Model 2 & Model 3 \\
+\textbf{} & \textbf{Model 1} & \textbf{Model 2} & \textbf{Model 3} \\
 \midrule
-Batching Tendency & 0.99*** & 0.99*** & 0.99*** \\
+Batching Tendency & 1.05 *** & 1.04 *** & 1.04 *** \\
  & (0.01) & (0.01) & (0.01) \\
 \midrule
-F statistic (full model)&7.32&52.03&50.93\\  
-F (full model): p-value& $<0.001$ & $<0.001$ & $<0.001$ \\
+F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
+F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
 \midrule
-Num. obs. & 43300 & 43300 & 43300 \\
+Num. obs. & 43327 & 43327 & 43327 \\
 Seasonality and shift fixed effects? & Yes & Yes & Yes \\
@@
 \subsection{Placebo Check}\label{placebo-check}}
+
+In this section we investigate whether the reduced-form effects observed
+in Section 4.1 are due to differences in batch rates across providers or
+due to other provider differences correlated with batch tendency We
+start by studying reduced-form effects among patients with complaints
+that are never prescribed opioids, as a ``placebo/falsification check.''
+By way of example, consider a patient who arrives at the ED with a
+urinary tact infection---a condition for which patients rarely undergo
+imaging testing. For such patients, we should expect to see no impact of
+batch tendency only if high batching and low batching physicians do not
+systematically differ in other dimensions of care relevant to patient
+outcomes. Conversely, if we do find a reduced-form effect for these
+patients, then high batch tendency physicians must systematically differ
+from low batch tendency physicians in other dimensions of care, beyond
+batching.
+
+To that end, we restrict attention to ED visits for complaints where
+batching occurs no more than 10 percent of the time (recall that our
+baseline sample only includes complaints with a \textgreater10 percent
+batching rate). We estimate a reduced-form regression of each main
+outcome on physician prescribing leniency for the subsample, following
+equation (3). The results of this exercise are displayed in Table 4.
+They show that in contrast to results for our main sample, the
+association between physician tendency to batch and a given outcome is
+statistically indistinguishable from zero and much smaller in magnitude
+for the samples of patients who visit the ED with health conditions that
+are rarely batched.
+
+To further probe robustness, we perform the placebo exercise for a set
+of non-opioid outcomes---in the form of homelessness, suicide, all-cause
+mortality, and preventable hospitalizations---in online Appendix Table
+G.7. Note that these outcomes are likely to be more sensitive to other
+dimensions of physician care than our opioid-related outcomes because
+the causal link to prescription opioid exposure is less tight, while
+other margins of care may matter relatively more. Indeed, we do find a
+sizable and statistically significant effect of being assigned to a
+high-prescribing physician on preventable hospitalizations and all-cause
+mor- tality for our placebo conditions. For patients with ED diagnoses
+that are rarely prescribed an opioid (column 3), assignment to a
+physician in the top decile of the opioid prescribing range increases
+the likelihood of a preventable hospitalization within 3 years by 0.27
+pp, or 1.8 percent.19
+
+The failure of our placebo check for some non-opioid outcomes suggests
+that there may be dimensions of care correlated with opioid prescribing
+tendency that vary across ED physicians and may impact non-opioid
+patient outcomes. While the set of potentially relevant dimensions of
+care is very large and may include dimen- sions unobservable to the
+researcher (such as, for example, instructions on medica- tion adherence
+and use, physician-nurse scheduling that lead to complementarities), we
+can make progress by accounting for the chief observable dimensions
+likely to correlate highly with patient health outcomes in ED settings:
+quality of care as mea- sured by immediate patient mortality, intensity
+of procedures, and tendency to admit patients for inpatient
+hospitalizations. In moving toward a well-identified IV design that
+estimates the causal impact of an opioid prescription on patient
+outcomes, we address these potential violations to the exclusion
+restriction by estimating physician ``propensities'' along these
+``non-focal'' (relative to opioid prescribing) dimensions and including
+them as controls (details are provided in Section IIIE). This approach
+follows recent advances from the judge stringency literature
+(Mueller-Smith 2015; Bhuller et al.~2020), which deals with similar
+concerns.
+
+We recognize that given the multidimensionality of physician behavior,
+potential for exclusion restriction violations remains even after
+controlling for chief dimensions of behavior. For this reason, we also
+present reduced-form results, and we keep a tight focus on
+opioid-related outcomes, for which such concerns are less pronounced
+relative to general health outcomes.
 

### 2023-11-07 src/IV Validation.R

@@ -3,14 +3,13 @@
 # Author: Jacob Jameson 
 #=========================================================================
 
-source('src/clean.R')
 library(lfe)
 library(stargazer)
 library(texreg)
 library(xtable)
+library(tidyverse)
 
-final <- final %>%
-  filter(nEDTests > 0, ED_LOS != 0, ED_LOS < 2880)
+data <- read_csv('final.csv')
 
 ##########################################################################
 #=========================================================================
@@ -32,38 +31,110 @@ final <- final %>%
 
 # Shift-level FE
 first_stage_final <- felm(any.batch ~ batch.tendency | 
-                          dayofweekt + month |
+                          dayofweekt + month_of_year |
                           0| ED_PROVIDER, 
-                          data = final)
+                          data = data)
 
 # Shift-level + complaint FE
 first_stage_final_d <- felm(any.batch ~ batch.tendency | 
-                            dayofweekt + month + age_groups + 
+                            dayofweekt + month_of_year + age_groups + 
                             complaint_esi |
                             0|ED_PROVIDER,
-                            data = final)
+                            data = data)
 
 # Shift-level + complaint + individual FE
 first_stage_final_d_i <- felm(any.batch ~ batch.tendency | 
-                              dayofweekt + month + age_groups + 
+                              dayofweekt + month_of_year + age_groups + 
                               complaint_esi + GENDER + race |
                               0| ED_PROVIDER , 
-                              data = final)
-# Latex First-Stage
-stargazer(first_stage_final, 
-          first_stage_final_d,
-          first_stage_final_d_i, title='')
+                              data = data)
 
-# Print First-Stage
+# Save the results to a .txt file
+sink("outputs/tables/First Stage.txt")
+
 screenreg(list(first_stage_final,
                first_stage_final_d,
                first_stage_final_d_i), 
           include.fstatistic = T)
 
+stargazer(list(first_stage_final,first_stage_final_d,
+               first_stage_final_d_i), type = "text", 
+          header = FALSE, title = "First Stage", style = 'QJE')
+
+sink()
+
+
+##########################################################################
+#=========================================================================
[…]

### 2023-11-07 src/Results.R

@@ -1,0 +1,440 @@
+#=========================================================================
+# Purpose: Validate IV
+# Author: Jacob Jameson 
+#=========================================================================
+
+source('src/clean.R')
+library(lfe)
+library(stargazer)
+library(texreg)
+library(xtable)
+library(tidyverse)
+
+final <- read_csv('final.csv')
+
+final <- final %>%
+  filter(ED_LOS != 0, ED_LOS < 2880)
+
+##########################################################################
+#=========================================================================
+# Main Results -----------------------------------------------------------
+#=========================================================================
+
+#=========================================================================
+##########################################################################
+
+
+summary(felm(ln_ED_LOS ~ avg_nEDTests |
+                 dayofweekt + month + complaint_esi | 
+               (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final))
+
+summary(felm(RTN_72_HR ~ avg_nEDTests |
+              dayofweekt + month + complaint_esi | 
+              (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final))
+
+summary(felm(nEDTests ~ avg_nEDTests |
+               dayofweekt + month + complaint_esi | 
+               (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final))
+
+ t<- felm(nEDTests ~ avg_nEDTests |
+           dayofweekt + month + complaint_esi | 
+           (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final)
+# Sub Group Analysis 1
+## Use tendency constructed from all complaints
+ subgroup_analysis <- function(dependent_var, caption_text){
+   
+   unique_complaints <- unique(final$CHIEF_COMPLAINT)
+   
+   results_df <- data.frame(
+     CHIEF_COMPLAINT = character(),
+     Coefficient = numeric(),
+     P_value = numeric(),
+     Cases = integer(),  # New column to store the number of cases
+     stringsAsFactors = FALSE
+   )
+   
+   for(complaint in unique_complaints){
+     
+     sub_data <- final[final$CHIEF_COMPLAINT == complaint, ]
+     result <- summary(
+       felm(as.formula(
+         paste(dependent_var,"~ avg_nEDTests | 
+                            dayofweekt + month  | 
[…]

### 2023-11-07 src/clean.R

@@ -7,6 +7,7 @@ rm(list = ls())
 library(tidyverse)
 library(stringr)
 library(lfe)
+library(lubridate)
 
 #=========================================================================
 # Determine test times
@@ -19,19 +20,15 @@ df <- read.csv(paste0(path, 'deidentified_FINAL.csv'))
 test_columns = c("US_PERF", "NON_CON_CT_PERF", "CON_CT_PERF", 
                  "LAB_PERF", "XR_PERF")
 
-test_time_columns = c("US_ORDER_DTTM", "NON_CON_CT_ORDER_DTTM",
-                      "CON_CT_ORDER_DTTM", "LAB_ORDER_DTTM", 
-                      "XR_ORDER_DTTM")
 
 colnames(df)[colnames(df) == "PLAIN_XRAY"] = "XR_PERF"
-colnames(df)[colnames(df) == "US_ORDER_DTTM_REL"] ="US_ORDER_DTTM"
-colnames(df)[colnames(df) == "CT_WITHOUT_CONTR_ORDER_DTTM_REL"] = "NON_CON_CT_ORDER_DTTM"
-colnames(df)[colnames(df) == "CT_WITH_CONTR_ORDER_DTTM_REL"] ="CON_CT_ORDER_DTTM"
-colnames(df)[colnames(df) == "LAB_ORDER_DTTM_REL"] ="LAB_ORDER_DTTM"
-colnames(df)[colnames(df) == "PLAIN_XRAY_ORDER_DTTM_REL"] ="XR_ORDER_DTTM"
+colnames(df)[colnames(df) == "US_ORDER_DTTM_REL"] ="US_ORDER_DTTM_REL"
+colnames(df)[colnames(df) == "CT_WITHOUT_CONTR_ORDER_DTTM_REL"] = "NON_CON_CT_ORDER_REL"
+colnames(df)[colnames(df) == "CT_WITH_CONTR_ORDER_DTTM_REL"] ="CON_CT_ORDER_REL"
+colnames(df)[colnames(df) == "LAB_ORDER_DTTM_REL"] ="LAB_ORDER_REL"
+colnames(df)[colnames(df) == "PLAIN_XRAY_ORDER_DTTM_REL"] ="XR_ORDER_REL"
 
-df$CT_PERF = ifelse(df$NON_CON_CT_PERF=='Y' | 
-                      df$CON_CT_PERF=='Y', 1, 0)
[…]

### 2023-11-07 src/figures.R

@@ -3,7 +3,6 @@
 # Author: Jacob Jameson 
 #=========================================================================
 rm(list = ls()) 
-source('src/clean.R')
 
 library(caret)
 library(lmtest)
@@ -11,10 +10,44 @@ library(sandwich)
 library(ggsci)
 library(xtable) # Output to LaTeX table format
 
-final <- read_csv('final.csv')
+data <- read_csv('outputs/data/final.csv')
 
 
 ##########################################################################
+
+
+library(ggplot2)
+library(lfe)
+
+
+# Run your fixed effects model
+FS <- lm(any.batch ~ batch.tendency + as.factor(dayofweekt) + 
+           as.factor(month_of_year), data)
+
+data$predlm <- predict(FS)
+predslm <- predict(FS, interval = "confidence")
+data <- cbind(data, predslm)
+
+predslm
+# Extract the fitted values
+data$fitted_values <- first_stage_final$fitted.values
+data$predlm = predict(first_stage_final)
+
+
+# Create the histogram with the regression curve
+ggplot(data, aes(x = batch.tendency)) +
+  geom_histogram(aes(y = ..density..), binwidth = 0.05, fill = "blue", alpha = 0.7) +
+  geom_ribbon( aes(ymin = lwr, ymax = upr, color = NULL), alpha = .15) +
+  geom_line( aes(y = fit), size = 1) +
+  labs(x = "Physician Leniency", y = "Density (Histogram) / Fitted Probability (Curve)") +
+  theme_minimal() +
+  theme(axis.title.y.right = element_text(angle = 0)) +
+  scale_y_continuous(sec.axis = sec_axis(~., name = "Fitted Probability of Prescribed Opioids"))
+
+# Note: Adjust binwidth in geom_histogram and span in geom_smooth as needed
+
+
+
 
 ##########################################################################
 #=========================================================================

### 2023-11-07 src/summary tables.R

@@ -1,0 +1,100 @@
+#=========================================================================
+# Purpose: Produce Summary Tables for Manuscript
+# Author: Jacob Jameson 
+#=========================================================================
+
+library(lfe)
+library(stargazer)
+library(texreg)
+library(xtable)
+library(tidyverse)
+
+data <- read_csv('outputs/data/final.csv')
+
+##########################################################################
+##########################################################################
+#=========================================================================
+# Summary Table: Table 1
+#=========================================================================
+##########################################################################
+
+# Identify the five most frequent chief complaints
+top_chief_complaints <- data %>%
+  count(CHIEF_COMPLAINT, sort = TRUE) %>%
+  slice_max(n = 5, order_by = n) %>%
+  pull(CHIEF_COMPLAINT)
+
+# Calculate the mean for each of the top chief complaints
+means_chief_complaints <- sapply(
+  top_chief_complaints, 
+  function(cc) mean(data$CHIEF_COMPLAINT == cc, na.rm = TRUE))
+
+# Panel 1: Emergency department characteristics ===================================
+panel1 <- data %>%
+  reframe(
+    Variable = c('Patients in ED', "Tachycardic", "Tachypneic", 
+                 "Febrile", "Hypotensive", "ESI", 
+                 paste("Complaint: ", top_chief_complaints)),
+    Mean = c(mean(patients_in_hospital, na.rm = TRUE), 
+             mean(tachycardic, na.rm = TRUE), 
+             mean(tachypneic, na.rm = TRUE),
+             mean(febrile, na.rm = TRUE), mean(hypotensive, na.rm = TRUE),
+             mean(ESI, na.rm = TRUE), means_chief_complaints),
+    Q1 = c(quantile(patients_in_hospital, 0.25, na.rm = TRUE),NA, NA, NA, NA, 
+           quantile(ESI, 0.25, na.rm = TRUE), rep(NA, 5)),
+    Median = c(median(patients_in_hospital, na.rm = TRUE), NA, NA, NA, NA, 
+               median(ESI, na.rm = TRUE), rep(NA, 5)),
+    Q3 = c(quantile(patients_in_hospital, 0.75, na.rm = TRUE), NA, NA, NA, NA, 
+           quantile(ESI, 0.75, na.rm = TRUE), rep(NA, 5))
+  ) %>%
[…]

### 2023-11-08 manuscript/manuscript.Rmd
working on manuscript

@@
 
-In this section, we examine the causal effects of provider batching tendency.
-We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).We find statistically and economically significant positive effects of assignment to a high batching physician on every outcome (Table 3).
-Estimated coefficients are scaled by the difference in leniency going from the nineti- eth to tenth percentile in physician leniency—equal to 11.6 pp—for interpretability.
-Assignment to a physician in the top prescribing decile (relative to one in the bottom decile) is associated with a 0.24 pp increase in the probability of long-term prescrip- tion opioid use—a 4.1 percent increase.
+In this section, we examine the causal effects of provider batching tendency. One thing that we note in the
+
+We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).
+We find statistically and economically significant positive effects of assignment to a high batching physician on every outcome (Table 3).
+Estimated coefficients are scaled by the difference in leniency going from the ninetieth to tenth percentile in physician leniency—equal to 11.6 pp—for interpretability.
+Assignment to a physician in the top prescribing decile (relative to one in the bottom decile) is associated with a 0.24 pp increase in the probability of long-term prescription opioid use— a 4.1 percent increase.
 The effect on proxies for opioid-seeking behavior is 0.50 pp—a 3.4 percent increase.
 Effects on development of opioid use disorder and on opioid overdose mortality are 2.1 percent and 8.8 percent, respec- tively.
-These findings highlight the substantial role ED physicians can play in put- ting patients on a path of long-term opioid use and dependence.
+These findings highlight the substantial role ED physicians can play in putting patients on a path of long-term opioid use and dependence.
+
+\begin{table}[htbp]
+\centering
+\caption{Reduced Form Regression Results}
+\label{tab:reducedform}
+\begin{tabular}{p{6cm}ccc}
+\toprule
+& \multicolumn{1}{c}{\textbf{Log ED LOS}}
+& \multicolumn{1}{c}{\textbf{Number of Tests}}
+& \multicolumn{1}{c}{\textbf{72hr Return}} \\
+& \multicolumn{1}{c}{(1)}
+& \multicolumn{1}{c}{(2)}
+& \multicolumn{1}{c}{(3)} \\
+\midrule
+\textit{Model 1} & & & \\
+Batch Tendency      & $0.905^{***}$ & $1.715^{***}$ & $-0.025$  \\
+                    & $(0.308)$  & $(0.329)$     & $(0.019)$ \\
+\\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+\midrule
+\midrule
+\textit{Model 2} & & & \\
+Batch Tendency      & $-0.103$      & $0.120^{***}$ & $0.020$   \\
+                    & $(0.462)$     & $(0.046)$     & $(0.021)$ \\
+                    &     &     &  \\
+Testing Inclination    & $0.608^{**}$  & $0.963^{***}$ & $-0.027^{**}$ \\
+                    & $(0.242)$  & $(0.021)$     & $(0.011)$ \\
+\\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the reduced form are for the baseline sample described in the text.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
 

### 2023-11-08 manuscript/manuscript.tex
working on manuscript

@@
 stay, 72hr readmission, and healthcare resource utilization (number of
-tests).
+tests). We find statistically and economically significant positive
+effects of assignment to a high batching physician on every outcome
+(Table 3). Estimated coefficients are scaled by the difference in
+leniency going from the ninetieth to tenth percentile in physician
+leniency---equal to 11.6 pp---for interpretability. Assignment to a
+physician in the top prescribing decile (relative to one in the bottom
+decile) is associated with a 0.24 pp increase in the probability of
+long-term prescription opioid use--- a 4.1 percent increase. The effect
+on proxies for opioid-seeking behavior is 0.50 pp---a 3.4 percent
+increase. Effects on development of opioid use disorder and on opioid
+overdose mortality are 2.1 percent and 8.8 percent, respec- tively.
+These findings highlight the substantial role ED physicians can play in
+putting patients on a path of long-term opioid use and dependence.
+
+\begin{table}[htbp]
+\centering
+\caption{Reduced Form Regression Results}
+\label{tab:reducedform}
+\begin{tabular}{p{7cm}ccc}
+\toprule
+& \multicolumn{1}{c}{\textbf{Log ED LOS}}
+& \multicolumn{1}{c}{\textbf{Number of Tests}}
+& \multicolumn{1}{c}{\textbf{72hr Return}} \\
+& \multicolumn{1}{c}{(1)}
+& \multicolumn{1}{c}{(2)}
+& \multicolumn{1}{c}{(3)} \\
+\midrule
+\textit{Model 1} & & & \\
+Batch Tendency      & $0.905^{***}$ & $1.715^{***}$ & $-0.025$  \\
+                    & $(0.308)$  & $(0.329)$     & $(0.019)$ \\
+\\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+\midrule
+\midrule
+\textit{Model 2} & & & \\
+Batch Tendency      & $-0.103$      & $0.120^{***}$ & $0.020$   \\
+                    & $(0.462)$     & $(0.046)$     & $(0.021)$ \\
+                    &     &     &  \\
+Testing Inclination    & $0.608^{**}$  & $0.963^{***}$ & $-0.027^{**}$ \\
+                    & $(0.242)$  & $(0.021)$     & $(0.011)$ \\
+\\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the first stage for the baseline sample described in the text.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Column 3 corresponds to the baseline controls.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
 

### 2023-11-08 src/IV Validation.R

@@ -9,7 +9,7 @@ library(texreg)
 library(xtable)
 library(tidyverse)
 
-data <- read_csv('final.csv')
+data <- read_csv('outputs/data/final.csv')
 
 ##########################################################################
 #=========================================================================
@@ -30,20 +30,20 @@ data <- read_csv('final.csv')
 ##########################################################################
 
 # Shift-level FE
-first_stage_final <- felm(any.batch ~ batch.tendency | 
+first_stage_final <- felm(any.batch ~ batch.tendency + test.inclination | 
                           dayofweekt + month_of_year |
                           0| ED_PROVIDER, 
                           data = data)
 
 # Shift-level + complaint FE
-first_stage_final_d <- felm(any.batch ~ batch.tendency | 
+first_stage_final_d <- felm(any.batch ~ batch.tendency + test.inclination | 
                             dayofweekt + month_of_year + age_groups + 
                             complaint_esi |
                             0|ED_PROVIDER,
                             data = data)
 
 # Shift-level + complaint + individual FE
-first_stage_final_d_i <- felm(any.batch ~ batch.tendency | 
+first_stage_final_d_i <- felm(any.batch ~ batch.tendency + test.inclination | 
                               dayofweekt + month_of_year + age_groups + 
                               complaint_esi + GENDER + race |
                               0| ED_PROVIDER , 
@@ -97,44 +97,108 @@ summary(mediation)
 #=========================================================================
 ##########################################################################
 
+reduced.form.LOS.0 <- 
+  felm(ln_ED_LOS ~ 
+         batch.tendency | 
+         dayofweekt + month_of_year + complaint_esi |
+         0|ED_PROVIDER, 
+       data = data
+  )
+
+reduced.form.ntest.0 <- 
+  felm(nEDTests ~ 
+         batch.tendency | 
+         dayofweekt + month_of_year + complaint_esi |
+         0|ED_PROVIDER, 
+       data = data
+  )
+
+reduced.form.72.0 <- 
+  felm(RTN_72_HR ~ 
+         batch.tendency  | 
+         dayofweekt + month_of_year + complaint_esi |
+         0|ED_PROVIDER, 
+       data = data
+  )
+
 reduced.form.LOS <- 
   felm(ln_ED_LOS ~ 
-         batch.tendency + avg_nEDTests | 
[…]

### 2023-11-08 src/clean.R

@@ -552,6 +552,9 @@ final$residual_batch_li <- resid(
 final$residual_batch_ii <- resid(
   felm(image_image_batch ~ 0 | dayofweekt + month_of_year + complaint_esi, data=final))
 
+final$residual_ntests <- resid(
+  felm(nEDTests ~ 0 | dayofweekt + month_of_year + complaint_esi, data=final))
+
 # Step 2: get batch tendency for each provider
 final <- final %>%
   group_by(ED_PROVIDER) %>%
@@ -564,7 +567,19 @@ final <- final %>%
          Sum_Resid_ii=sum(residual_batch_ii, na.rm=T),
          batch.tendency_ii = (Sum_Resid_ii - residual_batch_ii) / (n() - 1),
          
+         Sum_Resid_ntests=sum(residual_ntests, na.rm=T),
+         test.inclination = (Sum_Resid_ntests - residual_ntests) / (n() - 1),
+         
          avg.batch.tendency = mean(batch.tendency)) %>%
+  ungroup()
+
+final$residual_batch <- resid(
+  felm(any.batch ~ 0 | dayofweekt + month_of_year + complaint_esi + test.inclination, data=final))
+
+final <- final %>%
+  group_by(ED_PROVIDER) %>%
+  mutate(Sum_Resid=sum(residual_batch, na.rm=T),
+         batch.tendency.2 = (Sum_Resid - residual_batch) / (n() - 1)) %>%
   ungroup()
 
 write.csv(final, 'outputs/data/all_clean.csv')

### 2023-11-09 manuscript/manuscript.Rmd
breaktthough on the manuscript interpretation

@@
 
-In this section, we examine the causal effects of provider batching tendency. One thing that we note in the
-
-We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests).
-We find statistically and economically significant positive effects of assignment to a high batching physician on every outcome (Table 3).
-Estimated coefficients are scaled by the difference in leniency going from the ninetieth to tenth percentile in physician leniency—equal to 11.6 pp—for interpretability.
-Assignment to a physician in the top prescribing decile (relative to one in the bottom decile) is associated with a 0.24 pp increase in the probability of long-term prescription opioid use— a 4.1 percent increase.
-The effect on proxies for opioid-seeking behavior is 0.50 pp—a 3.4 percent increase.
-Effects on development of opioid use disorder and on opioid overdose mortality are 2.1 percent and 8.8 percent, respec- tively.
-These findings highlight the substantial role ED physicians can play in putting patients on a path of long-term opioid use and dependence.
+In this section, we examine the causal effects of provider batching tendency.
+Our reduced form regression results, as presented in Table \ref{tab:reducedform}, indicate that the physician's batch tendency is associated with longer emergency department length of stay (ED LOS), a greater number of tests ordered, and does not significantly influence the likelihood of a 72-hour return.
+However, the relationship between batch tendency and our outcomes of interest is nuanced by the presence of testing inclination among physicians.
 
@@
 \centering
-\caption{Reduced Form Regression Results}
+\caption{Reduced Form Regression Results with Scaled Coefficients}
 \label{tab:reducedform}
@@
 \textit{Model 1} & & & \\
-Batch Tendency      & $0.905^{***}$ & $1.715^{***}$ & $-0.025$  \\
-                    & $(0.308)$  & $(0.329)$     & $(0.019)$ \\
+Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
+                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
 \\
@@
 Patient observables? & Yes & Yes & Yes \\
+Hospital volume? & Yes & Yes & Yes \\
 \midrule
@@
 \textit{Model 2} & & & \\
-Batch Tendency      & $-0.103$      & $0.120^{***}$ & $0.020$   \\
-                    & $(0.462)$     & $(0.046)$     & $(0.021)$ \\
+Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
+                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
                     &     &     &  \\
-Testing Inclination    & $0.608^{**}$  & $0.963^{***}$ & $-0.027^{**}$ \\
-                    & $(0.242)$  & $(0.021)$     & $(0.011)$ \\
+Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
+                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
 \\
@@
 Patient observables? & Yes & Yes & Yes \\
+Hospital volume? & Yes & Yes & Yes \\
+\midrule
+Mean number of encounters per physician & 1805.3 &  &  \\
+\midrule
 \bottomrule
@@
 \item \textit{Estimates of the reduced form are for the baseline sample described in the text.
+Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability.
 Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
@@
 \end{table}
+
+Similar to our method used to measure physician batch tendency, we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test (which we call testing inclination).
+A key consideration in our analysis is the strong positive correlation (r = 0.79) between batch tendency and testing inclination.
+Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often because as a result of ordering more tests.
+Failure to control for this inclination could result in biased estimates of the effect of batch tendency.
+We note this assumption made on the causal pathway and further explore this in the section on mediation analysis.
+
+In Model 2, we control for testing inclination and observe notable changes in the coefficients for batch tendency.
+The previously significant positive association with Log ED LOS diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
+Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency—equal to 15.8 pp—for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination—equal to 34.9 pp.
+
+Assignment to a physician in the top batching decile (relative to one in the bottom decile) is associated with a 0.2 unit increase in the number of diagnostic tests being ordered per patient encounter, without any additional change in the liklihood of a 72 hour return.
+With an average of 1,805 patient encounters per physician in our study period, this translates to an additional 8,667 diagnostic tests with no perceived benefit in terms of reducing returns or length of stay.
+
+The effects of testing inclination are also striking.
+We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 0.96 unit increase in the number of diagnostic tests being ordered per patient encounter, and -0.03 pp change in the probability of a patient returning within 72 hours.
+Given the average number of patient encounters per physician in our study period this translates to an additional 1088 patient hours in LOS, 41,594 diagnostic tests, and 47 avoided 72 hour returns.
 

### 2023-11-09 manuscript/manuscript.tex
breaktthough on the manuscript interpretation

--- 2023-11-07 manuscript/manuscript.tex
@@
 In this section, we examine the causal effects of provider batching
-tendency. We first present the effects on our main outcomes of length of
-stay, 72hr readmission, and healthcare resource utilization (number of
-tests).
+tendency. Our reduced form regression results, as presented in Table
+\ref{tab:reducedform}, indicate that the physician's batch tendency is
+associated with longer emergency department length of stay (ED LOS), a
+greater number of tests ordered, and does not significantly influence
+the likelihood of a 72-hour return. However, the relationship between
+batch tendency and our outcomes of interest is nuanced by the presence
+of testing inclination among physicians.
+
+\begin{table}[htbp]
+\centering
+\caption{Reduced Form Regression Results with Scaled Coefficients}
+\label{tab:reducedform}
+\begin{tabular}{p{6cm}ccc}
+\toprule
+& \multicolumn{1}{c}{\textbf{Log ED LOS}}
+& \multicolumn{1}{c}{\textbf{Number of Tests}}
+& \multicolumn{1}{c}{\textbf{72hr Return}} \\
+& \multicolumn{1}{c}{(1)}
+& \multicolumn{1}{c}{(2)}
+& \multicolumn{1}{c}{(3)} \\
+\midrule
+\textit{Model 1} & & & \\
+Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
+                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
+\\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+Hospital volume? & Yes & Yes & Yes \\
+\midrule
+\midrule
+\textit{Model 2} & & & \\
+Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
+                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
+                    &     &     &  \\
+Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
+                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
+\\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+Hospital volume? & Yes & Yes & Yes \\
+\midrule
+Mean number of encounters per physician & 1805.3 &  &  \\
+\midrule
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the reduced form are for the baseline sample described in the text.
+Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
+
+Similar to our method used to measure physician batch tendency, we use
+the physician's residualized leave-out average test rate to get a
+measure of a physicians propensity to test (which we call testing
+inclination). A key consideration in our analysis is the strong positive
+correlation (r = 0.79) between batch tendency and testing inclination.
+Such a high degree of correlation raises the concern that physicians
+with a higher testing inclination may batch more often because as a
+result of ordering more tests. Failure to control for this inclination
+could result in biased estimates of the effect of batch tendency. We
+note this assumption made on the causal pathway and further explore this
+in the section on mediation analysis.
+
+In Model 2, we control for testing inclination and observe notable
+changes in the coefficients for batch tendency. The previously
+significant positive association with Log ED LOS diminishes, suggesting
+that the initial estimate may have been capturing the combined effects
+of batching and testing inclination. Estimated coefficients for batch
+tendency are scaled by the difference in batch tendency going from the
+ninetieth to tenth percentile in physician batch tendency---equal to
+15.8 pp---for interpretability, and coefficients for testing inclination
+are scaled by the difference in testing inclination going from the
+ninetieth to tenth percentile in physician testing inclination---equal
+to 34.9 pp.
+
+Assignment to a physician in the top batching decile (relative to one in
+the bottom decile) is associated with a 0.2 unit increase in the number
+of diagnostic tests being ordered per patient encounter, without any
+additional change in the liklihood of a 72 hour return. With an average
+of 1,805 patient encounters per physician in our study period, this
+translates to an additional 8,667 diagnostic tests with no perceived
+benefit in terms of reducing returns or length of stay.
+
+The effects of testing inclination are also striking. We find that a
+movement from a physician in the top testing decile (relative to one in
+the bottom decile) is associated with 0.6\% change in the length of stay
+in the emergency department, a 0.96 unit increase in the number of
+diagnostic tests being ordered per patient encounter, and -0.03 pp
+change in the probability of a patient returning within 72 hours. Given
+the average number of patient encounters per physician in our study
+period this translates to an additional 1088 patient hours in LOS,
+41,594 diagnostic tests, and 47 avoided 72 hour returns.
 

### 2023-11-09 src/IV Validation.R

@@ -99,15 +99,15 @@ summary(mediation)
 
 reduced.form.LOS.0 <- 
   felm(ln_ED_LOS ~ 
-         batch.tendency | 
-         dayofweekt + month_of_year + complaint_esi |
+         batch.tendency + patients_in_hospital | 
+         dayofweekt + month_of_year + complaint_esi  |
          0|ED_PROVIDER, 
        data = data
   )
 
 reduced.form.ntest.0 <- 
   felm(nEDTests ~ 
-         batch.tendency | 
+         batch.tendency + patients_in_hospital | 
          dayofweekt + month_of_year + complaint_esi |
          0|ED_PROVIDER, 
        data = data
@@ -115,15 +115,15 @@ reduced.form.ntest.0 <-
 
 reduced.form.72.0 <- 
   felm(RTN_72_HR ~ 
-         batch.tendency  | 
-         dayofweekt + month_of_year + complaint_esi |
+         batch.tendency  + patients_in_hospital| 
+         dayofweekt + month_of_year + complaint_esi  |
          0|ED_PROVIDER, 
        data = data
   )
 
 reduced.form.LOS <- 
   felm(ln_ED_LOS ~ 
-         batch.tendency + test.inclination | 
+         batch.tendency + test.inclination + patients_in_hospital| 
          dayofweekt + month_of_year + complaint_esi |
          0|ED_PROVIDER, 
        data = data
@@ -131,7 +131,7 @@ reduced.form.LOS <-
 
 reduced.form.ntest <- 
   felm(nEDTests ~ 
-         batch.tendency + test.inclination | 
+         batch.tendency + test.inclination + patients_in_hospital | 
          dayofweekt + month_of_year + complaint_esi |
          0|ED_PROVIDER, 
        data = data
@@ -139,7 +139,7 @@ reduced.form.ntest <-
 
 reduced.form.72 <- 
   felm(RTN_72_HR ~ 
-         batch.tendency + test.inclination | 
+         batch.tendency + test.inclination  + patients_in_hospital| 
          dayofweekt + month_of_year + complaint_esi |
          0|ED_PROVIDER, 
        data = data
@@ -176,6 +176,20 @@ ggplot()  +
   labs(x='\nBatch-Ordering Tendency',
        y='Test Ordering Inclination\n')
 
+
+# All coefficients are scaled by the 
+# difference in batch tendency between the ninetieth and tenth lenient physicians  
+# for interpretability. 
[…]

### 2023-11-10 manuscript/manuscript.Rmd
cleaning up repo and reduced form results

@@
 
-In this section we investigate whether the reduced-form effects observed in Section 4.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency We start by studying reduced-form effects among patients with complaints that are never prescribed opioids, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
+In this section we investigate whether the reduced-form effects observed in Section 4.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
 For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
@@
 We estimate a reduced-form regression of each main outcome on physician prescribing leniency for the subsample, following equation (3).
-The results of this exercise are displayed in Table 4.
+The results of this exercise are displayed in Table __.
 They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
 
-To further probe robustness, we perform the placebo exercise for a set of non-opioid outcomes—in the form of homelessness, suicide, all-cause mortality, and preventable hospitalizations—in online Appendix Table G.7.
-Note that these outcomes are likely to be more sensitive to other dimensions of physician care than our opioid-related outcomes because the causal link to prescription opioid exposure is less tight, while other margins of care may matter relatively more.
-Indeed, we do find a sizable and statistically significant effect of being assigned to a high-prescribing physician on preventable hospitalizations and all-cause mor- tality for our placebo conditions.
-For patients with ED diagnoses that are rarely prescribed an opioid (column 3), assignment to a physician in the top decile of the opioid prescribing range increases the likelihood of a preventable hospitalization within 3 years by 0.27 pp, or 1.8 percent.19
+## Instrumental Variables Results
 
-The failure of our placebo check for some non-opioid outcomes suggests that there may be dimensions of care correlated with opioid prescribing tendency that vary across ED physicians and may impact non-opioid patient outcomes.
-While the set of potentially relevant dimensions of care is very large and may include dimen- sions unobservable to the researcher (such as, for example, instructions on medica- tion adherence and use, physician-nurse scheduling that lead to complementarities), we can make progress by accounting for the chief observable dimensions likely to correlate highly with patient health outcomes in ED settings: quality of care as mea- sured by immediate patient mortality, intensity of procedures, and tendency to admit patients for inpatient hospitalizations.
-In moving toward a well-identified IV design that estimates the causal impact of an opioid prescription on patient outcomes, we address these potential violations to the exclusion restriction by estimating physician “propensities” along these “non-focal” (relative to opioid prescribing) dimensions and including them as controls (details are provided in Section IIIE).
-This approach follows recent advances from the judge stringency literature (Mueller-Smith 2015; Bhuller et al.
-2020), which deals with similar concerns.
-
-We recognize that given the multidimensionality of physician behavior, potential for exclusion restriction violations remains even after controlling for chief dimensions of behavior.
-For this reason, we also present reduced-form results, and we keep a tight focus on opioid-related outcomes, for which such concerns are less pronounced relative to general health outcomes.
-
-## Instrumental Variables Results
+\begin{table}[!htbp] \centering 
+  \caption{IV Results} 
+  \label{} 
+\begin{tabular}{p{6cm}ccc}
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & ln\_ED\_LOS & RTN\_72\_HR & nEDTests \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Batch & $-$0.097 & 0.019 & 0.113$^{***}$ \\ 
+  & (0.435) & (0.020) & (0.043) \\ 
+  & & & \\ 
+ Testing Inclination & 0.607$^{**}$ & $-$0.027$^{**}$ & 0.964$^{***}$ \\ 
+  & (0.238) & (0.011) & (0.020) \\ 
+  & & & \\ 
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+Hospital volume? & Yes & Yes & Yes \\
+\midrule
+\textit{N} & 43,328 & 43,328 & 43,328 \\ 
+R$^{2}$ & 0.195 & 0.009 & 0.306 \\ 
+Adjusted R$^{2}$ & 0.190 & 0.002 & 0.302 \\ 
+Residual Std. Error (df = 43061) & 0.498 & 0.187 & 0.818 \\ 
+\\
+\midrule
+\midrule
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the IV model are for the baseline sample described in the text.
+Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
 

### 2023-11-10 manuscript/manuscript.tex
cleaning up repo and reduced form results

@@
 start by studying reduced-form effects among patients with complaints
-that are never prescribed opioids, as a ``placebo/falsification check.''
-By way of example, consider a patient who arrives at the ED with a
-urinary tact infection---a condition for which patients rarely undergo
-imaging testing. For such patients, we should expect to see no impact of
-batch tendency only if high batching and low batching physicians do not
+that are never batched, as a ``placebo/falsification check.'' By way of
+example, consider a patient who arrives at the ED with a urinary tact
+infection---a condition for which patients rarely undergo imaging
+testing. For such patients, we should expect to see no impact of batch
+tendency only if high batching and low batching physicians do not
 systematically differ in other dimensions of care relevant to patient
@@
 outcome on physician prescribing leniency for the subsample, following
-equation (3). The results of this exercise are displayed in Table 4.
+equation (3). The results of this exercise are displayed in Table \_\_.
 They show that in contrast to results for our main sample, the
@@
 
-To further probe robustness, we perform the placebo exercise for a set
-of non-opioid outcomes---in the form of homelessness, suicide, all-cause
-mortality, and preventable hospitalizations---in online Appendix Table
-G.7. Note that these outcomes are likely to be more sensitive to other
-dimensions of physician care than our opioid-related outcomes because
-the causal link to prescription opioid exposure is less tight, while
-other margins of care may matter relatively more. Indeed, we do find a
-sizable and statistically significant effect of being assigned to a
-high-prescribing physician on preventable hospitalizations and all-cause
-mor- tality for our placebo conditions. For patients with ED diagnoses
-that are rarely prescribed an opioid (column 3), assignment to a
-physician in the top decile of the opioid prescribing range increases
-the likelihood of a preventable hospitalization within 3 years by 0.27
-pp, or 1.8 percent.19
-
-The failure of our placebo check for some non-opioid outcomes suggests
-that there may be dimensions of care correlated with opioid prescribing
-tendency that vary across ED physicians and may impact non-opioid
-patient outcomes. While the set of potentially relevant dimensions of
-care is very large and may include dimen- sions unobservable to the
-researcher (such as, for example, instructions on medica- tion adherence
-and use, physician-nurse scheduling that lead to complementarities), we
-can make progress by accounting for the chief observable dimensions
-likely to correlate highly with patient health outcomes in ED settings:
-quality of care as mea- sured by immediate patient mortality, intensity
-of procedures, and tendency to admit patients for inpatient
-hospitalizations. In moving toward a well-identified IV design that
-estimates the causal impact of an opioid prescription on patient
-outcomes, we address these potential violations to the exclusion
-restriction by estimating physician ``propensities'' along these
-``non-focal'' (relative to opioid prescribing) dimensions and including
-them as controls (details are provided in Section IIIE). This approach
-follows recent advances from the judge stringency literature
-(Mueller-Smith 2015; Bhuller et al.~2020), which deals with similar
-concerns.
-
-We recognize that given the multidimensionality of physician behavior,
-potential for exclusion restriction violations remains even after
-controlling for chief dimensions of behavior. For this reason, we also
-present reduced-form results, and we keep a tight focus on
-opioid-related outcomes, for which such concerns are less pronounced
-relative to general health outcomes.
-
 \hypertarget{instrumental-variables-results}{%
@@
 Results}\label{instrumental-variables-results}}
+
+\begin{table}[!htbp] \centering 
+  \caption{IV Results} 
+  \label{} 
+\begin{tabular}{p{6cm}ccc}
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & ln\_ED\_LOS & RTN\_72\_HR & nEDTests \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Batch & $-$0.097 & 0.019 & 0.113$^{***}$ \\ 
+  & (0.435) & (0.020) & (0.043) \\ 
+  & & & \\ 
+ Testing Inclination & 0.607$^{**}$ & $-$0.027$^{**}$ & 0.964$^{***}$ \\ 
+  & (0.238) & (0.011) & (0.020) \\ 
+  & & & \\ 
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+Hospital volume? & Yes & Yes & Yes \\
+\midrule
+\textit{N} & 43,328 & 43,328 & 43,328 \\ 
+R$^{2}$ & 0.195 & 0.009 & 0.306 \\ 
+Adjusted R$^{2}$ & 0.190 & 0.002 & 0.302 \\ 
+Residual Std. Error (df = 43061) & 0.498 & 0.187 & 0.818 \\ 
+\\
+\midrule
+\midrule
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the IV model are for the baseline sample described in the text.
+Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
 

### 2023-11-10 src/IV Validation.R

@@ -1,5 +1,5 @@
 #=========================================================================
-# Purpose: Validate IV
+# Purpose: Validate IV + Main Results
 # Author: Jacob Jameson 
 #=========================================================================
 
@@ -21,7 +21,6 @@ data <- read_csv('outputs/data/final.csv')
 #=========================================================================
 ##########################################################################
 
-
 ##########################################################################
 #=========================================================================
 # Relevance --------------------------------------------------------------
@@ -30,24 +29,24 @@ data <- read_csv('outputs/data/final.csv')
 ##########################################################################
 
 # Shift-level FE
-first_stage_final <- felm(any.batch ~ batch.tendency + test.inclination | 
-                          dayofweekt + month_of_year |
-                          0| ED_PROVIDER, 
-                          data = data)
+first_stage_final <- felm(
+  any.batch ~ batch.tendency + test.inclination | 
+              dayofweekt + month_of_year |0| ED_PROVIDER, 
+  data = data)
 
 # Shift-level + complaint FE
-first_stage_final_d <- felm(any.batch ~ batch.tendency + test.inclination | 
-                            dayofweekt + month_of_year + age_groups + 
-                            complaint_esi |
-                            0|ED_PROVIDER,
-                            data = data)
+first_stage_final_d <- felm(
+  any.batch ~ batch.tendency + test.inclination | 
+              dayofweekt + month_of_year + age_groups + 
+              complaint_esi | 0 | ED_PROVIDER,
+  data = data)
 
 # Shift-level + complaint + individual FE
-first_stage_final_d_i <- felm(any.batch ~ batch.tendency + test.inclination | 
-                              dayofweekt + month_of_year + age_groups + 
-                              complaint_esi + GENDER + race |
-                              0| ED_PROVIDER , 
-                              data = data)
+first_stage_final_d_i <- felm(
+  any.batch ~ batch.tendency + test.inclination | 
+              dayofweekt + month_of_year + age_groups + 
[…]

### 2023-11-10 src/Results.R

@@ -24,21 +24,21 @@ final <- final %>%
 ##########################################################################
 
 
-summary(felm(ln_ED_LOS ~ avg_nEDTests |
-                 dayofweekt + month + complaint_esi | 
-               (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final))
+m1 <- (felm(ln_ED_LOS ~ test.inclination |
+                 dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)|ED_PROVIDER, data = data))
 
-summary(felm(RTN_72_HR ~ avg_nEDTests |
-              dayofweekt + month + complaint_esi | 
-              (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final))
+m2 <-(felm(RTN_72_HR ~ test.inclination |
+              dayofweekt + month_of_year + complaint_esi | 
+              (any.batch ~ batch.tendency + test.inclination)|ED_PROVIDER, data = data))
 
-summary(felm(nEDTests ~ avg_nEDTests |
-               dayofweekt + month + complaint_esi | 
-               (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final))
+m3 <- (felm(nEDTests ~ test.inclination |
+               dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)|ED_PROVIDER, data = data))
 
- t<- felm(nEDTests ~ avg_nEDTests |
-           dayofweekt + month + complaint_esi | 
-           (any.batch ~ batch.tendency + avg_nEDTests)|0, data = final)
+stargazer(list(m1, m2, m3), type = "latex", 
+          header = FALSE, title = "IV Results", style = 'QJE')
+
 # Sub Group Analysis 1
 ## Use tendency constructed from all complaints
  subgroup_analysis <- function(dependent_var, caption_text){

### 2023-11-10 src/IV Validation.R
monotonicity check in progress

@@ -131,11 +131,6 @@ stargazer(list(model2.LOS,
           header = FALSE, title = "Reduced Form", style = 'QJE')
 
 
-# Scale your models
-
-scale_factors.1 <- c(coeffecient.scale.b, coeffecient.scale.t)
-scale_factors.2 <- c(coeffecient.scale.b)
-
 
 # Function for scaling the coefficients in regressions for 
 # interpretability
@@ -160,6 +155,9 @@ scale_coefficients <- function(model, scale_factors) {
   
   return(scaled_model)
 }
+
+scale_factors.1 <- c(coeffecient.scale.b, coeffecient.scale.t)
+scale_factors.2 <- c(coeffecient.scale.b)
 
 # Scaled regression results
 summary(model1.LOS)$call
@@ -206,90 +204,94 @@ data <- read_csv('outputs/data/all_clean.csv')
 placebo.complaints <- data %>%
   group_by(CHIEF_COMPLAINT) %>%
   summarise(mean.batch = mean(any.batch), n=n()) %>%
-  filter(mean.batch <= 0.1) 
+  filter(mean.batch <= 0.1)
 
 placebo.complaints <- placebo.complaints$CHIEF_COMPLAINT
 
 placebo <- data %>%
   filter(CHIEF_COMPLAINT %in% placebo.complaints)
 
-# The idea here is that we want to show that being a high-batching
-# physician is not associated with other physician traits 
-# associated with quality. 
-# We are going to show that for Urinary Complaints,
-# a complaint area where batching is rare, that there are not preferable
-# outcomes associated with being a batcher
-
-#=========================================================================
-# LOS
-#=========================================================================
-
 # Save the results to a .txt file
 sink("outputs/tables/Placebo Check.txt")
 
-#
+#=========================================================================
+# Time controls only
+
 placebo1.LOS <- felm(
   ln_ED_LOS ~ batch.tendency + test.inclination | 
               dayofweekt + month_of_year |0| ED_PROVIDER, 
-  data = data)
+  data = placebo)
 
 placebo1.ntest <- felm(
   nEDTests ~ batch.tendency + test.inclination | 
               dayofweekt + month_of_year |0| ED_PROVIDER, 
-  data = data)
+  data = placebo)
 
 placebo1.72 <- felm(
   RTN_72_HR ~ batch.tendency + test.inclination | 
               dayofweekt + month_of_year |0| ED_PROVIDER, 
-  data = data)
[…]

### 2023-11-10 src/clean.R

@@ -61,7 +61,8 @@ df <- df %>%
 # Determine which was the first test ordered -----------------------------
 # Columns of interest
 
-cols_of_interest <- c('US_ORDER_DTTM_REL', 'NON_CON_CT_ORDER_REL', 'CON_CT_ORDER_REL',
+cols_of_interest <- c('US_ORDER_DTTM_REL', 'NON_CON_CT_ORDER_REL', 
+                      'CON_CT_ORDER_REL',
                       'LAB_ORDER_REL', 'XR_ORDER_REL')
 
 # Function to determine batch
@@ -533,6 +534,7 @@ final <- final[!(final$ED_PROVIDER %in% providers_less_than_500), ]
 final$complaint_esi <- paste(final$CHIEF_COMPLAINT, final$ESI)
 
 rm(list = setdiff(ls(), "final"))
+
 #=========================================================================
 ##########################################################################
 #=========================================================================
@@ -571,6 +573,7 @@ final <- final %>%
          test.inclination = (Sum_Resid_ntests - residual_ntests) / (n() - 1),
          
          avg.batch.tendency = mean(batch.tendency)) %>%
+  
   ungroup()
 
 final$residual_batch <- resid(
@@ -580,14 +583,17 @@ final <- final %>%
   group_by(ED_PROVIDER) %>%
   mutate(Sum_Resid=sum(residual_batch, na.rm=T),
          batch.tendency.2 = (Sum_Resid - residual_batch) / (n() - 1)) %>%
-  ungroup()
+  ungroup() %>%
+  filter(ED_LOS > 0)
 
 write.csv(final, 'outputs/data/all_clean.csv')
 
 # Limit to prevalent complaints only
 final <- final %>%
   group_by(CHIEF_COMPLAINT) %>%
-  filter(n() > 1000)
+  filter(n() > 1000) %>%
+  ungroup() %>%
+  filter(CHIEF_COMPLAINT != 'Urinary Complaints')
 
 write.csv(final, 'outputs/data/final.csv')
 #=========================================================================

### 2023-11-11 manuscript/manuscript.Rmd
adding additional results related to IV

@@
 
-In this paper, we use data from over 41,000 patient visits to the ED that occur during our study period to quantify the benefits and consequences of batching versus sequentially ordering diagnostic tests on patient length of stay, re-admission, and resource utilization.
+In this paper, we use data from over 45,000 patient visits to the ED that occur during our study period to quantify the benefits and consequences of batching versus sequentially ordering diagnostic tests on patient length of stay, re-admission, and resource utilization.
 Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests.
@@
 Excluding these conditions does not introduce selection bias only if physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (“Fracture of ankle”), but are unlikely to disagree at the broader chief complaint cateogry.}.
+While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint cateogry.}.
 
@@
 
-In our instrumental variables (IV) analysis, the key explanatory variable, $Tendency_i$, is an indicator for whether patient $i$ has their tests batch ordered at their ED encounter.
-This variable is crucial for understanding the causal effects of batch testing on various health outcomes and resource utilization.
+In our instrumental variables (IV) analysis, the key explanatory variable, $Batched_i$, is an indicator for whether patient $i$ has their tests batch ordered at their ED encounter.
+This variable is crucial for understanding the causal effects of batch testing on various patient outcomes and resource utilization.
 Below, we detail our main outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.
@@
 \centering
-\caption{First Stage Regression Results}
+\caption{First Stage: Effect of Batch Tendency on Test Batching}
 \label{tab:regression}
@@
 \centering
-\caption{Reduced Form Regression Results with Scaled Coefficients}
+\caption{Reduced-Form Results: Length of Stay, Number of Tests, and 72-Hour Return}
 \label{tab:reducedform}
@@
 \toprule
-& \multicolumn{1}{c}{\textbf{Log ED LOS}}
+& \multicolumn{1}{c}{\textbf{Log LOS}}
 & \multicolumn{1}{c}{\textbf{Number of Tests}}
-& \multicolumn{1}{c}{\textbf{72hr Return}} \\
+& \multicolumn{1}{c}{\textbf{72-Hour Return}} \\
 & \multicolumn{1}{c}{(1)}
@@
 
+In this section we examine the causal effects of batch ordering in the ED.
+Mirroring our presentation of the reduced-form results, we turn to presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED LOS, number of tests, and 72-hour return.
+
 \begin{table}[!htbp] \centering 
-  \caption{IV Results} 
-  \label{} 
+  \caption{2SLS Results: Length of Stay, Number of Tests, and 72-Hour Return} 
+  \label{tab:2SLS} 
 \begin{tabular}{p{6cm}ccc}
@@
 \hline \\[-1.8ex] 
-\\[-1.8ex] & ln\_ED\_LOS & RTN\_72\_HR & nEDTests \\ 
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
 \\[-1.8ex] & (1) & (2) & (3)\\ 
 \hline \\[-1.8ex] 
- Batch & $-$0.097 & 0.019 & 0.113$^{***}$ \\ 
-  & (0.435) & (0.020) & (0.043) \\ 
+ Batch & $-$0.086 & 0.135$^{***}$ & 0.018 \\ 
+  & (0.419) & (0.046) & (0.018) \\ 
   & & & \\ 
- Testing Inclination & 0.607$^{**}$ & $-$0.027$^{**}$ & 0.964$^{***}$ \\ 
-  & (0.238) & (0.011) & (0.020) \\ 
+ Testing Inclination & 0.210$^{**}$ & 0.335$^{***}$ & $-$0.009$^{***}$ \\ 
+  & (0.082) & (0.008) & (0.003) \\ 
   & & & \\ 
@@
 \midrule
-\textit{N} & 43,328 & 43,328 & 43,328 \\ 
-R$^{2}$ & 0.195 & 0.009 & 0.306 \\ 
-Adjusted R$^{2}$ & 0.190 & 0.002 & 0.302 \\ 
-Residual Std. Error (df = 43061) & 0.498 & 0.187 & 0.818 \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.199 & 0.312 & 0.008 \\ 
+Adjusted R$^{2}$ & 0.194 & 0.307 & 0.002 \\ 
+Residual Std. Error (df = 41667) & 0.496 & 0.817 & 0.184 \\ 
 \\
-\midrule
-\midrule
 \bottomrule
@@
 
+Length of Stay.—We find that batch ordering has no effect on reducing length of stay.
+This estimate is robust to inclusion and exclusion of controls for baseline characteristics.
+This implies that the anticipated efficiency gains from ordering tests in a batch, as opposed to waiting for the results of one test to come back before ordering a subsequent test, are likely washed out by the inefficiencies associated with ordering additional unnecessary tests.
+
+Number of Tests.—Our 2SLS findings on the resource utilization are presented in column 2 of Table \ref{tab:2SLS}. We find large effects. The effect of tests batching was associated
+
+\begin{table}[!htbp] \centering 
+  \caption{2SLS Results: Specific Tests} 
+  \label{tab:2SLStests} 
+\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & X-Ray & Ultrasound & Non-Con CT & Con CT & Lab \\ 
+\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
+\hline \\[-1.8ex] 
+ Batch & 0.288$^{**}$ & $-$0.003 & 0.095 & $-$0.242$^{***}$ & $-$0.003 \\ 
+  & (0.126) & (0.051) & (0.087) & (0.056)  & (0.080) \\ 
+  & & & & &\\ 
+ Testing Inclination & 0.240$^{***}$ & 0.086$^{***}$ & 0.096$^{**}$ & 0.253$^{***}$ & 0.286$^{***}$ \\ 
+  & (0.056) & (0.024) & (0.039) & (0.027) & (0.038) \\ 
+  & & & & & \\ 
+Test Ordering Frequency & 0.47 & 0.11 & 0.21 & 0.19 & 0.77 \\ 
+  & & & & &\\ 
+\textit{N} & 41,929 & 41,929 & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.350 & 0.077 & 0.236 & 0.017 \\ 
+\hline 
+\hline \\[-1.8ex] 
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the IV model are for the baseline sample described in the text.
+Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
+
 ## Robustness
+
+The 2SLS results presented in the previous sections are robust to several alter- native specifications probing the key identifying assumptions of independence, exclusion, and monotonicity.
+We describe each robustness check in detail in online Appendix D and provide a brief overview here.
+To probe the independence assump- tion, we perform two checks.
+First, we take into account potential endogeneity in diagnosing behavior by no longer excluding any diagnosis from our analysis sample and by replacing the original diagnosis control with the most recent diag- nosis code observed prior to the ED visit (results are presented in online Appendix Table G.8, column 2).
+Second, we replace the physician leniency instrument with a shift-level instrument that captures the average leniency across all physicians on shift (column 3).
+To provide robustness checks to potential exclusion restriction violations, we adjust for chief observable margins of care that may be correlated with leniency, namely (i) decision to admit patients to the hospital, (ii) intensity of procedures performed, (iii) quality of care provided, and (iv) amount of opi- oids prescribed—i.e., intensive margin (columns 4–6).
+To address monotonicity, we construct physician-diagnosis-specific leniency instruments (column 7), and we check whether our instrument has a positive first stage across key demographic- and health-related subsamples (online Appendix Table G.12).
+
+Across all robustness checks and main outcomes, we find magnitudes of our estimates broadly unchanged.
+Precision of estimates remains unchanged for the outcomes of long-term use and opioid-seeking behavior, while estimates for the outcomes of opioid use disorder and opioid overdose mortality become noisier in some specifications.
 
@@
 
+\newpage
+
+\begin{table}[htbp]
+\centering
+\caption{First Stage over all Subsamples}
+\label{tab:monotonicitycheck}
+\begin{tabular}{p{7cm}cccc}
+\toprule
+Complaint & \multicolumn{2}{c}{Batch Tendency} & N & R$^2$ \\
+\midrule
+Extremity Complaints & 0.662$^{***}$ & (0.079) & 5,259 & 0.126 \\
+Upper Respiratory Symptoms & 1.205$^{***}$ & (0.445) & 1,915 & 0.262 \\
+Abnormal Test Results & 0.670$^{**}$ & (0.325) & 1,816 & 0.141 \\
+Falls, Motor Vehicle Crashes, and Trauma & 0.640$^{**}$ & (0.264) & 2,381 & 0.220 \\
+Shortness of Breath & 1.574$^{***}$ & (0.468) &  2,963 & 0.147 \\
+Gastrointestinal Issues & 1.080$^{***}$ & (0.241) & 3,323 & 0.094 \\
+Chest Pain & 0.790$^{***}$ & (0.253) & 3,521 & 0.140 \\
+Dizziness/Lightheadedness/Syncope & 1.343$^{***}$ & (0.220) & 1,894 & 0.208 \\
+Back or Flank Pain & 0.859$^{**}$ & (0.385) &  2,550 & 0.123 \\
+Neurological Issue & 0.416$^{*}$ & (0.232) &  3,492 & 0.232 \\
+Fatigue and Weakness & 1.728$^{***}$ & (0.414) & 1,494 & 0.193 \\
+Cardiac Arrhythmias & 2.432$^{***}$ & (0.548) & 1,044 & 0.327 \\
+Abdominal Complaints & 1.775$^{***}$ & (0.350) & 6,231 & 0.086 \\
+Skin Complaints & 0.205 & (0.212) & 2,176 & 0.133 \\
+Fevers, Sweats or Chills & 1.650$^{***}$ & (0.576) &  1,842 & 0.211 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
 :::

### 2023-11-11 manuscript/manuscript.tex
adding additional results related to IV

@@
 
-In this paper, we use data from over 41,000 patient visits to the ED
+In this paper, we use data from over 45,000 patient visits to the ED
 that occur during our study period to quantify the benefits and
@@
 complaint
-categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (“Fracture of ankle”), but are unlikely to disagree at the broader chief complaint cateogry.}.
+categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint cateogry.}.
 
@@
 In our instrumental variables (IV) analysis, the key explanatory
-variable, \(Tendency_i\), is an indicator for whether patient \(i\) has
+variable, \(Batched_i\), is an indicator for whether patient \(i\) has
 their tests batch ordered at their ED encounter. This variable is
 crucial for understanding the causal effects of batch testing on various
-health outcomes and resource utilization. Below, we detail our main
+patient outcomes and resource utilization. Below, we detail our main
 outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c)
@@
 
+In this section we examine the causal effects of batch ordering in the
+ED. Mirroring our presentation of the reduced-form results, we turn to
+presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED
+LOS, number of tests, and 72-hour return.
+
 \begin{table}[!htbp] \centering 
-  \caption{IV Results} 
-  \label{} 
+  \caption{2SLS Results: Length of Stay, Number of Tests, and 72-Hour Return} 
+  \label{tab:2SLS} 
 \begin{tabular}{p{6cm}ccc}
@@
 
+Length of Stay.---We find that batch ordering has no effect on reducing
+length of stay. This estimate is robust to inclusion and exclusion of
+controls for baseline characteristics. This implies that the anticipated
+efficiency gains from ordering tests in a batch, as opposed to waiting
+for the results of one test to come back before ordering a subsequent
+test, are likely washed out by the inefficiencies associated with
+ordering additional unnecessary tests.
+
+Number of Tests.---Our 2SLS findings on the resource utilization are
+presented in column 2 of Table \ref{tab:2SLS}. We find large effects.
+The effect of tests batching
+
+\begin{table}[!htbp] \centering 
+  \caption{2SLS Results: Specific Tests} 
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & X-Ray & Ultrasound & Non-Con CT & Con CT & Lab \\ 
+\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.240$^{***}$ & 0.086$^{***}$ & 0.096$^{**}$ & 0.253$^{***}$ & 0.286$^{***}$ \\ 
+  & (0.056) & (0.024) & (0.039) & (0.027) & (0.038) \\ 
+  & & & & & \\ 
+ Batch & 0.288$^{**}$ & $-$0.003 & 0.095 & $-$0.242$^{***}$ & $-$0.003 \\ 
+  & (0.126) & (0.051) & (0.087) & (0.056)  & (0.080) \\ 
+  & & & & &\\ 
+Test Ordering Frequency & 0.47 & 0.11 & 0.21 & 0.19 & 0.77 \\ 
+  & & & & &\\ 
+\textit{N} & 41,929 & 41,929 & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.350 & 0.077 & 0.236 & 0.017 \\ 
+\hline 
+\hline \\[-1.8ex] 
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{Estimates of the IV model are for the baseline sample described in the text.
+Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
+Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
+Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
+Patient observables include sex dummy, race/ethnicity, and age bins.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
+
 \hypertarget{robustness}{%
@@
 
+The 2SLS results presented in the previous sections are robust to
+several alter- native specifications probing the key identifying
+assumptions of independence, exclusion, and monotonicity. We describe
+each robustness check in detail in online Appendix D and provide a brief
+overview here. To probe the independence assump- tion, we perform two
+checks. First, we take into account potential endogeneity in diagnosing
+behavior by no longer excluding any diagnosis from our analysis sample
+and by replacing the original diagnosis control with the most recent
+diag- nosis code observed prior to the ED visit (results are presented
+in online Appendix Table G.8, column 2). Second, we replace the
+physician leniency instrument with a shift-level instrument that
+captures the average leniency across all physicians on shift (column 3).
+To provide robustness checks to potential exclusion restriction
+violations, we adjust for chief observable margins of care that may be
+correlated with leniency, namely (i) decision to admit patients to the
+hospital, (ii) intensity of procedures performed, (iii) quality of care
+provided, and (iv) amount of opi- oids prescribed---i.e., intensive
+margin (columns 4--6). To address monotonicity, we construct
+physician-diagnosis-specific leniency instruments (column 7), and we
+check whether our instrument has a positive first stage across key
+demographic- and health-related subsamples (online Appendix Table G.12).
+
+Across all robustness checks and main outcomes, we find magnitudes of
+our estimates broadly unchanged. Precision of estimates remains
+unchanged for the outcomes of long-term use and opioid-seeking behavior,
+while estimates for the outcomes of opioid use disorder and opioid
+overdose mortality become noisier in some specifications.
+
 \hypertarget{conclusion}{%
@@
 \end{longtable}
+
+\newpage
+
+\begin{table}[htbp]
+\centering
+\caption{First Stage over all Subsamples}
+\label{tab:monotonicitycheck}
+\begin{tabular}{p{7cm}cccc}
+\toprule
+Complaint & \multicolumn{2}{c}{Batch Tendency} & N & R$^2$ \\
+\midrule
+Extremity Complaints & 0.662$^{***}$ & (0.079) & 5,259 & 0.126 \\
+Upper Respiratory Symptoms & 1.205$^{***}$ & (0.445) & 1,915 & 0.262 \\
+Abnormal Test Results & 0.670$^{**}$ & (0.325) & 1,816 & 0.141 \\
+Falls, Motor Vehicle Crashes, and Trauma & 0.640$^{**}$ & (0.264) & 2,381 & 0.220 \\
+Shortness of Breath & 1.574$^{***}$ & (0.468) &  2,963 & 0.147 \\
+Gastrointestinal Issues & 1.080$^{***}$ & (0.241) & 3,323 & 0.094 \\
+Chest Pain & 0.790$^{***}$ & (0.253) & 3,521 & 0.140 \\
+Dizziness/Lightheadedness/Syncope & 1.343$^{***}$ & (0.220) & 1,894 & 0.208 \\
+Back or Flank Pain & 0.859$^{**}$ & (0.385) &  2,550 & 0.123 \\
+Neurological Issue & 0.416$^{*}$ & (0.232) &  3,492 & 0.232 \\
+Fatigue and Weakness & 1.728$^{***}$ & (0.414) & 1,494 & 0.193 \\
+Cardiac Arrhythmias & 2.432$^{***}$ & (0.548) & 1,044 & 0.327 \\
+Abdominal Complaints & 1.775$^{***}$ & (0.350) & 6,231 & 0.086 \\
+Skin Complaints & 0.205 & (0.212) & 2,176 & 0.133 \\
+Fevers, Sweats or Chills & 1.650$^{***}$ & (0.576) &  1,842 & 0.211 \\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2023-11-11 src/IV Validation.R

@@ -216,6 +216,7 @@ sink("outputs/tables/Placebo Check.txt")
 
 #=========================================================================
 # Time controls only
+#=========================================================================
 
 placebo1.LOS <- felm(
   ln_ED_LOS ~ batch.tendency + test.inclination | 
@@ -241,6 +242,7 @@ stargazer(list(placebo1.LOS,
 
 #=========================================================================
 # Time controls + complaint severity
+#=========================================================================
 
 placebo2.LOS <- felm(
   ln_ED_LOS ~ batch.tendency + test.inclination | 
@@ -265,6 +267,7 @@ stargazer(list(placebo2.LOS,
 
 #=========================================================================
 # Time controls + complaint + person level + volums
+#=========================================================================
 
 placebo3.LOS <- felm(
   ln_ED_LOS ~ batch.tendency + test.inclination + 
@@ -347,28 +350,170 @@ stargazer(model_results[10:15], type = "text",
 sink()
 
 
+
 ##########################################################################
 #=========================================================================
 # IV Results ------------------------------------------------------------
 #=========================================================================
 ##########################################################################
 
-m1 <- felm(ln_ED_LOS ~ test.inclination |
-              dayofweekt + month_of_year + complaint_esi | 
-              (any.batch ~ batch.tendency + test.inclination)|ED_PROVIDER, data = data)
+# Main Results
+model.IV.LOS <- felm(
+  ln_ED_LOS ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+              (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+  data = data)
 
-m2 <- felm(nEDTests ~ test.inclination |
-              dayofweekt + month_of_year + complaint_esi | 
-              (any.batch ~ batch.tendency + test.inclination)|ED_PROVIDER, data = data)
+model.IV.ntest <- felm(
+  nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+            (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+  data = data)
        
-m3 <- felm(RTN_72_HR ~ test.inclination |
[…]

### 2023-12-04 manuscript/manuscript.Rmd
last commit on this computer

@@
 Figure 2 graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically.
-\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 5 pp increase in the likelihood of having tests batch-ordered in the ED.
+\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 8 pp increase in the likelihood of having tests batch-ordered in the ED.
 The F-statistic is 50 when all controls and fixed effects are included.
@@
 \midrule
-Batching Tendency & 1.05 *** & 1.04 *** & 1.04 *** \\
+Batching Tendency & 1.06 *** & 1.08 *** & 1.08 *** \\
  & (0.01) & (0.01) & (0.01) \\
@@
 \toprule
-& \multicolumn{1}{c}{\textbf{Log LOS}}
+& \multicolumn{1}{c}{\textbf{ln(LOS)}}
 & \multicolumn{1}{c}{\textbf{Number of Tests}}
@@
 Similar to our method used to measure physician batch tendency, we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test (which we call testing inclination).
-A key consideration in our analysis is the strong positive correlation (r = 0.79) between batch tendency and testing inclination.
+A key consideration in our analysis is the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
 Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often because as a result of ordering more tests.
@@
 In Model 2, we control for testing inclination and observe notable changes in the coefficients for batch tendency.
-The previously significant positive association with Log ED LOS diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
+The previously significant positive association with ln(LOS) diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
 Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency—equal to 15.8 pp—for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination—equal to 34.9 pp.
 
-Assignment to a physician in the top batching decile (relative to one in the bottom decile) is associated with a 0.2 unit increase in the number of diagnostic tests being ordered per patient encounter, without any additional change in the liklihood of a 72 hour return.
-With an average of 1,805 patient encounters per physician in our study period, this translates to an additional 8,667 diagnostic tests with no perceived benefit in terms of reducing returns or length of stay.
+Assignment to a physician in the top batching decile (relative to one in the bottom decile) is associated with a 0.2 unit increase in the number of diagnostic tests being ordered per patient encounter, without any additional change in the likelihood of a 72-hour return.
+With an average of 1,805 patient encounters per physician in our study period, this translates to an additional 361 diagnostic tests with no perceived benefit in terms of reducing returns or length of stay.
 
@@
 We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 0.96 unit increase in the number of diagnostic tests being ordered per patient encounter, and -0.03 pp change in the probability of a patient returning within 72 hours.
-Given the average number of patient encounters per physician in our study period this translates to an additional 1088 patient hours in LOS, 41,594 diagnostic tests, and 47 avoided 72 hour returns.
+Given the average number of patient encounters per physician in our study period this translates to an additional [...] patient hours in LOS, [...] diagnostic tests, and [...] avoided 72-hour returns.
 
@@
 We estimate a reduced-form regression of each main outcome on physician prescribing leniency for the subsample, following equation (3).
-The results of this exercise are displayed in Table __.
+The results of this exercise are displayed in Table \ref{tab:placebo_check}.
 They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
+
+\begin{table}[htbp]
+\centering
+\caption{Placebo Exercise: Reduced-Form Results for Rarely Batched Samples}
+\label{tab:placebo_check}
+\begin{tabular}{p{8cm}ccc}
+\toprule
+& ln(LOS) & Number of Tests & 72 Hour Return \\ 
+& (1) & (2) & (3) \\
+\midrule
+\multicolumn{4}{l}{\textbf{Controls: Time}} \\
+Batch Tendency & $-0.529$ & $-0.760$ & $-0.006$ \\
+& $(0.710)$ & $(0.490)$ & $(0.094)$ \\
+                    &     &     &  \\
+Testing Inclination & $0.812^{**}$ & $0.892^{***}$ & $-0.001$ \\
+& $(0.343)$ & $(0.194)$ & $(0.045)$ \\
+\\
+\addlinespace
+\multicolumn{4}{l}{\textbf{Controls: Time + Complaint}} \\
+Batch Tendency & $-0.517$ & $-0.714$ & $-0.015$ \\
+& $(0.660)$ & $(0.446)$ & $(0.092)$ \\
+                    &     &     &  \\
+Testing Inclination & $0.848^{**}$ & $1.031^{***}$ & $-0.010$ \\
+& $(0.334)$ & $(0.189)$ & $(0.045)$ \\
+\\
+\addlinespace
+\multicolumn{4}{l}{\textbf{Controls: All Controls}} \\
+Batch Tendency & $-0.465$ & $-0.699$ & $-0.011$ \\
+& $(0.645)$ & $(0.427)$ & $(0.095)$ \\
+                    &     &     &  \\
+Testing Inclination & $0.818^{**}$ & $1.011^{***}$ & $-0.009$ \\
+& $(0.322)$ & $(0.183)$ & $(0.046)$ \\
+                    &     &     &  \\
+Observations & & 2,361 & \\
+\addlinespace
+\midrule
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{This table reports the estimated coefficients of a reduced-form regression of our main outcomes on physician batching tendency for samples based on the patient’s chief complaint field.
+Estimates come from regression on conditions that are batched less than 10 percent of the time.
+See text for residualization fixed effects and baseline controls.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
 
@@
 \hline \\[-1.8ex] 
-\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & ln(LOS) & Number of Tests & 72 Hour Return \\ 
 \\[-1.8ex] & (1) & (2) & (3)\\ 
@@
 
-Length of Stay.—We find that batch ordering has no effect on reducing length of stay.
-This estimate is robust to inclusion and exclusion of controls for baseline characteristics.
-This implies that the anticipated efficiency gains from ordering tests in a batch, as opposed to waiting for the results of one test to come back before ordering a subsequent test, are likely washed out by the inefficiencies associated with ordering additional unnecessary tests.
+*ED length of stay (LOS).*-- Our analysis reveals that batch ordering does not significantly reduce the length of stay in the ED.
+This finding holds even after accounting for various control variables.
+Contrary to common assumptions that batching tests could expedite patient throughput, our results suggest that the efficiency gains are offset.
+A possible explanation lies in the additional time needed to process the increased volume of tests ordered through batching.
+These findings challenge prevailing perceptions and call for a reevaluation of batching as a time-saving strategy in ED operations.
 
-Number of Tests.—Our 2SLS findings on the resource utilization are presented in column 2 of Table \ref{tab:2SLS}. We find large effects. The effect of tests batching was associated
+*Resource Utilization.*-- The impact of batching on resource utilization is pronounced.
+We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 2,206 additional tests in our dataset.
+Particularly noteworthy is the substantial rise in X-Ray orders (by 28.8 percentage points), as shown in Table \ref{tab:2SLStests}.
+This trend towards increased X-Ray orders raises questions about potential over-testing and its implications for healthcare resources and patient care, especially considering the risks associated with unnecessary radiation exposure.
 
@@
 
+*72-hour Return*-- Our findings indicate that batch ordering does not significantly impact the likelihood of a 72-hour return.
+This suggests that despite the increased number of tests, there is no corresponding decrease in short-term readmission rates.
+This highlights the need for more targeted testing strategies that balance comprehensive assessment with resource efficiency.
+
+The instrumental variable analysis presented here underscores the complexities underlying batch ordering practices in the ED.
+While batching is often seen as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization without corresponding gains in patient throughput or reductions in short-term return rates.
+These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
+
+## Investigating Mechanisms: Excess X-Rays
+
 ## Robustness
 
-The 2SLS results presented in the previous sections are robust to several alter- native specifications probing the key identifying assumptions of independence, exclusion, and monotonicity.
-We describe each robustness check in detail in online Appendix D and provide a brief overview here.
-To probe the independence assump- tion, we perform two checks.
-First, we take into account potential endogeneity in diagnosing behavior by no longer excluding any diagnosis from our analysis sample and by replacing the original diagnosis control with the most recent diag- nosis code observed prior to the ED visit (results are presented in online Appendix Table G.8, column 2).
-Second, we replace the physician leniency instrument with a shift-level instrument that captures the average leniency across all physicians on shift (column 3).
-To provide robustness checks to potential exclusion restriction violations, we adjust for chief observable margins of care that may be correlated with leniency, namely (i) decision to admit patients to the hospital, (ii) intensity of procedures performed, (iii) quality of care provided, and (iv) amount of opi- oids prescribed—i.e., intensive margin (columns 4–6).
-To address monotonicity, we construct physician-diagnosis-specific leniency instruments (column 7), and we check whether our instrument has a positive first stage across key demographic- and health-related subsamples (online Appendix Table G.12).
+The 2SLS results presented in the previous sections are robust to several alternative specifications probing the key identifying assumptions of independence, exclusion, and monotonicity.
+We describe each robustness check in detail in Appendix ___ and provide a brief overview here ....
 
 Across all robustness checks and main outcomes, we find magnitudes of our estimates broadly unchanged.
-Precision of estimates remains unchanged for the outcomes of long-term use and opioid-seeking behavior, while estimates for the outcomes of opioid use disorder and opioid overdose mortality become noisier in some specifications.
 
@@
 
+\begin{table}[!htbp] \centering 
+  \caption{Regression Results by Complaint} 
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
+ & Dizziness/Syncope & Extremity Complaints & Upper Respiratory Symptoms \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 1.156$^{***}$ & 0.901$^{***}$ & 0.516$^{***}$ \\ 
+  & (0.291) & (0.111) & (0.194) \\ 
+  & & & \\ 
+ Batch & 0.655$^{*}$ & 0.479$^{**}$ & 0.501$^{*}$ \\ 
+  & (0.368) & (0.242) & (0.283) \\ 
+  & & & \\ 
+\textit{N} & 1,895 & 5,265 & 1,917 \\ 
+R$^{2}$ & 0.400 & 0.348 & 0.461 \\ 
+Adjusted R$^{2}$ & 0.338 & 0.324 & 0.404 \\ 
+Residual Std. Error & 0.801 (df = 1715) & 0.765 (df = 5079) & 0.741 (df = 1733) \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
+\end{tabular} 
+\end{table}
+
+\begin{table}[!htbp] \centering 
+  \caption{Regression Results by Complaint} 
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
+ & Abdominal Complaints & Cardiac Arrhythmias & Fatigue and Weakness \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.913$^{***}$ & 0.695$^{***}$ & 0.870$^{***}$ \\ 
+  & (0.068) & (0.180) & (0.195) \\ 
+  & & & \\ 
+ Batch & $-$0.135$^{**}$ & 0.708$^{***}$ & 0.333$^{*}$ \\ 
+  & (0.066) & (0.129) & (0.189) \\ 
+  & & & \\ 
+\textit{N} & 6,232 & 1,045 & 1,494 \\ 
+R$^{2}$ & 0.038 & 0.525 & 0.294 \\ 
+Adjusted R$^{2}$ & 0.009 & 0.428 & 0.203 \\ 
+Residual Std. Error & 0.752 (df = 6046) & 0.595 (df = 866) & 0.828 (df = 1322) \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
+\end{tabular} 
+\end{table} 
 :::

### 2023-12-04 manuscript/manuscript.tex
last commit on this computer

@@
 ``first stage'' in a regression table: being assigned to a 10 pp higher
-tendency physician is associated with a 5 pp increase in the likelihood
+tendency physician is associated with a 8 pp increase in the likelihood
 of having tests batch-ordered in the ED. The F-statistic is 50 when all
@@
 inclination). A key consideration in our analysis is the strong positive
-correlation (r = 0.79) between batch tendency and testing inclination.
-Such a high degree of correlation raises the concern that physicians
-with a higher testing inclination may batch more often because as a
-result of ordering more tests. Failure to control for this inclination
-could result in biased estimates of the effect of batch tendency. We
-note this assumption made on the causal pathway and further explore this
-in the section on mediation analysis.
+correlation (\(r = 0.79\)) between batch tendency and testing
+inclination. Such a high degree of correlation raises the concern that
+physicians with a higher testing inclination may batch more often
+because as a result of ordering more tests. Failure to control for this
+inclination could result in biased estimates of the effect of batch
+tendency. We note this assumption made on the causal pathway and further
+explore this in the section on mediation analysis.
 
@@
 changes in the coefficients for batch tendency. The previously
-significant positive association with Log ED LOS diminishes, suggesting
+significant positive association with ln(LOS) diminishes, suggesting
 that the initial estimate may have been capturing the combined effects
@@
 of diagnostic tests being ordered per patient encounter, without any
-additional change in the liklihood of a 72 hour return. With an average
+additional change in the likelihood of a 72-hour return. With an average
 of 1,805 patient encounters per physician in our study period, this
-translates to an additional 8,667 diagnostic tests with no perceived
+translates to an additional 361 diagnostic tests with no perceived
 benefit in terms of reducing returns or length of stay.
@@
 the average number of patient encounters per physician in our study
-period this translates to an additional 1088 patient hours in LOS,
-41,594 diagnostic tests, and 47 avoided 72 hour returns.
+period this translates to an additional {[}\ldots{]} patient hours in
+LOS, {[}\ldots{]} diagnostic tests, and {[}\ldots{]} avoided 72-hour
+returns.
 
@@
 outcome on physician prescribing leniency for the subsample, following
-equation (3). The results of this exercise are displayed in Table \_\_.
-They show that in contrast to results for our main sample, the
-association between physician tendency to batch and a given outcome is
-statistically indistinguishable from zero and much smaller in magnitude
-for the samples of patients who visit the ED with health conditions that
-are rarely batched.
+equation (3). The results of this exercise are displayed in Table
+\ref{tab:placebo_check}. They show that in contrast to results for our
+main sample, the association between physician tendency to batch and a
+given outcome is statistically indistinguishable from zero and much
+smaller in magnitude for the samples of patients who visit the ED with
+health conditions that are rarely batched.
+
+\begin{table}[htbp]
+\centering
+\caption{Placebo Exercise: Reduced-Form Results for Rarely Batched Samples}
+\label{tab:placebo_check}
+\begin{tabular}{p{8cm}ccc}
+\toprule
+& ln(LOS) & Number of Tests & 72 Hour Return \\ 
+& (1) & (2) & (3) \\
+\midrule
+\multicolumn{4}{l}{\textbf{Controls: Time}} \\
+Batch Tendency & $-0.529$ & $-0.760$ & $-0.006$ \\
+& $(0.710)$ & $(0.490)$ & $(0.094)$ \\
+                    &     &     &  \\
+Testing Inclination & $0.812^{**}$ & $0.892^{***}$ & $-0.001$ \\
+& $(0.343)$ & $(0.194)$ & $(0.045)$ \\
+\\
+\addlinespace
+\multicolumn{4}{l}{\textbf{Controls: Time + Complaint}} \\
+Batch Tendency & $-0.517$ & $-0.714$ & $-0.015$ \\
+& $(0.660)$ & $(0.446)$ & $(0.092)$ \\
+                    &     &     &  \\
+Testing Inclination & $0.848^{**}$ & $1.031^{***}$ & $-0.010$ \\
+& $(0.334)$ & $(0.189)$ & $(0.045)$ \\
+\\
+\addlinespace
+\multicolumn{4}{l}{\textbf{Controls: All Controls}} \\
+Batch Tendency & $-0.465$ & $-0.699$ & $-0.011$ \\
+& $(0.645)$ & $(0.427)$ & $(0.095)$ \\
+                    &     &     &  \\
+Testing Inclination & $0.818^{**}$ & $1.011^{***}$ & $-0.009$ \\
+& $(0.322)$ & $(0.183)$ & $(0.046)$ \\
+                    &     &     &  \\
+Observations & & 2,361 & \\
+\addlinespace
+\midrule
+\bottomrule
+\end{tabular}
+\begin{tablenotes}
+\small
+\item \textit{This table reports the estimated coefficients of a reduced-form regression of our main outcomes on physician batching tendency for samples based on the patient’s chief complaint field.
+Estimates come from regression on conditions that are batched less than 10 percent of the time.
+See text for residualization fixed effects and baseline controls.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
 
@@
 
-Length of Stay.---We find that batch ordering has no effect on reducing
-length of stay. This estimate is robust to inclusion and exclusion of
-controls for baseline characteristics. This implies that the anticipated
-efficiency gains from ordering tests in a batch, as opposed to waiting
-for the results of one test to come back before ordering a subsequent
-test, are likely washed out by the inefficiencies associated with
-ordering additional unnecessary tests.
-
-Number of Tests.---Our 2SLS findings on the resource utilization are
-presented in column 2 of Table \ref{tab:2SLS}. We find large effects.
-The effect of tests batching
+\emph{ED length of stay (LOS).}-- Our analysis reveals that batch
+ordering does not significantly reduce the length of stay in the ED.
+This finding holds even after accounting for various control variables.
+Contrary to common assumptions that batching tests could expedite
+patient throughput, our results suggest that the efficiency gains are
+offset. A possible explanation lies in the additional time needed to
+process the increased volume of tests ordered through batching. These
+findings challenge prevailing perceptions and call for a reevaluation of
+batching as a time-saving strategy in ED operations.
+
+\emph{Resource Utilization.}-- The impact of batching on resource
+utilization is pronounced. We observe a 0.135 increase in the number of
+tests ordered due to batching, corresponding to approximately 2,206
+additional tests in our dataset. Particularly noteworthy is the
+substantial rise in X-Ray orders (by 28.8 percentage points), as shown
+in Table \ref{tab:2SLStests}. This trend towards increased X-Ray orders
+raises questions about potential over-testing and its implications for
+healthcare resources and patient care, especially considering the risks
+associated with unnecessary radiation exposure.
 
@@
   \caption{2SLS Results: Specific Tests} 
-  \label{} 
+  \label{tab:2SLStests} 
 \begin{tabular}{@{\extracolsep{5pt}}lccccc} 
@@
 \hline \\[-1.8ex] 
+ Batch & 0.288$^{**}$ & $-$0.003 & 0.095 & $-$0.242$^{***}$ & $-$0.003 \\ 
+  & (0.126) & (0.051) & (0.087) & (0.056)  & (0.080) \\ 
+  & & & & &\\ 
  Testing Inclination & 0.240$^{***}$ & 0.086$^{***}$ & 0.096$^{**}$ & 0.253$^{***}$ & 0.286$^{***}$ \\ 
@@
   & & & & & \\ 
- Batch & 0.288$^{**}$ & $-$0.003 & 0.095 & $-$0.242$^{***}$ & $-$0.003 \\ 
-  & (0.126) & (0.051) & (0.087) & (0.056)  & (0.080) \\ 
-  & & & & &\\ 
 Test Ordering Frequency & 0.47 & 0.11 & 0.21 & 0.19 & 0.77 \\ 
@@
 
+\emph{72-hour Return}-- Our findings indicate that batch ordering does
+not significantly impact the likelihood of a 72-hour return. This
+suggests that despite the increased number of tests, there is no
+corresponding decrease in short-term readmission rates. This highlights
+the need for more targeted testing strategies that balance comprehensive
+assessment with resource efficiency.
+
+The instrumental variable analysis presented here underscores the
+complexities underlying batch ordering practices in the ED. While
+batching is often seen as a method to enhance efficiency, our findings
+suggest that it may lead to increased resource utilization without
+corresponding gains in patient throughput or reductions in short-term
+return rates. These insights point to the need for a more nuanced
+approach to test ordering in emergency care, one that carefully weighs
+the benefits of comprehensive assessment against the risks and costs of
+potential over-testing.
+
+\hypertarget{investigating-mechanisms-excess-x-rays}{%
+\subsection{Investigating Mechanisms: Excess
+X-Rays}\label{investigating-mechanisms-excess-x-rays}}
+
 \hypertarget{robustness}{%
@@
 The 2SLS results presented in the previous sections are robust to
-several alter- native specifications probing the key identifying
+several alternative specifications probing the key identifying
 assumptions of independence, exclusion, and monotonicity. We describe
-each robustness check in detail in online Appendix D and provide a brief
-overview here. To probe the independence assump- tion, we perform two
-checks. First, we take into account potential endogeneity in diagnosing
-behavior by no longer excluding any diagnosis from our analysis sample
-and by replacing the original diagnosis control with the most recent
-diag- nosis code observed prior to the ED visit (results are presented
-in online Appendix Table G.8, column 2). Second, we replace the
-physician leniency instrument with a shift-level instrument that
-captures the average leniency across all physicians on shift (column 3).
-To provide robustness checks to potential exclusion restriction
-violations, we adjust for chief observable margins of care that may be
-correlated with leniency, namely (i) decision to admit patients to the
-hospital, (ii) intensity of procedures performed, (iii) quality of care
-provided, and (iv) amount of opi- oids prescribed---i.e., intensive
-margin (columns 4--6). To address monotonicity, we construct
-physician-diagnosis-specific leniency instruments (column 7), and we
-check whether our instrument has a positive first stage across key
-demographic- and health-related subsamples (online Appendix Table G.12).
+each robustness check in detail in Appendix \_\_\_ and provide a brief
+overview here \ldots.
 
 Across all robustness checks and main outcomes, we find magnitudes of
-our estimates broadly unchanged. Precision of estimates remains
-unchanged for the outcomes of long-term use and opioid-seeking behavior,
-while estimates for the outcomes of opioid use disorder and opioid
-overdose mortality become noisier in some specifications.
+our estimates broadly unchanged.
 
@@
 
+\begin{table}[!htbp] \centering 
+  \caption{Regression Results by Complaint} 
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
+ & Dizziness/Syncope & Extremity Complaints & Upper Respiratory Symptoms \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 1.156$^{***}$ & 0.901$^{***}$ & 0.516$^{***}$ \\ 
+  & (0.291) & (0.111) & (0.194) \\ 
+  & & & \\ 
+ Batch & 0.655$^{*}$ & 0.479$^{**}$ & 0.501$^{*}$ \\ 
+  & (0.368) & (0.242) & (0.283) \\ 
+  & & & \\ 
+\textit{N} & 1,895 & 5,265 & 1,917 \\ 
+R$^{2}$ & 0.400 & 0.348 & 0.461 \\ 
+Adjusted R$^{2}$ & 0.338 & 0.324 & 0.404 \\ 
+Residual Std. Error & 0.801 (df = 1715) & 0.765 (df = 5079) & 0.741 (df = 1733) \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
+\end{tabular} 
+\end{table}
+
+\begin{table}[!htbp] \centering 
+  \caption{Regression Results by Complaint} 
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
+ & Abdominal Complaints & Cardiac Arrhythmias & Fatigue and Weakness \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.913$^{***}$ & 0.695$^{***}$ & 0.870$^{***}$ \\ 
+  & (0.068) & (0.180) & (0.195) \\ 
+  & & & \\ 
+ Batch & $-$0.135$^{**}$ & 0.708$^{***}$ & 0.333$^{*}$ \\ 
+  & (0.066) & (0.129) & (0.189) \\ 
+  & & & \\ 
+\textit{N} & 6,232 & 1,045 & 1,494 \\ 
+R$^{2}$ & 0.038 & 0.525 & 0.294 \\ 
+Adjusted R$^{2}$ & 0.009 & 0.428 & 0.203 \\ 
+Residual Std. Error & 0.752 (df = 6046) & 0.595 (df = 866) & 0.828 (df = 1322) \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
+\end{tabular} 
+\end{table}
+
 \end{APPENDIX}

### 2023-12-04 src/IV Validation.R

@@ -432,13 +432,21 @@ sink("outputs/tables/2SLS Results Tests Only.txt")
 stargazer(list(model.IV.xray,
                model.IV.us,
                model.IV.nct,
-               model.IV.lab), type = "text", 
+               model.IV.ct), type = "text", 
           covariate.labels = c('Testing Inclination', 'Batch'),
           dep.var.labels = c('X-Ray', 'Ultrasound', 'Non Contrast CT',
-                             'Contrast CT', 'Lab'),
+                             'Contrast CT'),
           header = F, 
           title = "2SLS Results: Specific Tests", 
           style = 'QJE')
+
+
+stargazer(list(model.IV.lab), type = "text", 
+          covariate.labels = c('Testing Inclination', 'Batch'),
+          header = F, 
+          title = "2SLS Results: Specific Tests", 
+          style = 'QJE')
+
 
 stargazer(list(model.IV.xray,
                model.IV.us,
@@ -452,9 +460,18 @@ stargazer(list(model.IV.xray,
           style = 'QJE')
 
 sink()
+
+data %>%
+  filter('XR_ORDER_REL' %in% batch) %>%
+  group_by(batch, CHIEF_COMPLAINT) %>%
+  summarize(n = n()) %>%
+  arrange(desc(n))
 #=========================================================================
 # Specific Types of Test Batching: Lab+Image Batch
 #=========================================================================
+
+# Save the results to a .txt file
+sink("outputs/tables/2SLS Results Specific Batch Types.txt")
 
 model.IV.LOS <- felm(
   ln_ED_LOS ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
@@ -470,6 +487,16 @@ model.IV.72 <- felm(
   RTN_72_HR ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
     (lab_image_batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
   data = data)
+
+
+stargazer(list(model.IV.LOS,
+               model.IV.ntest,
+               model.IV.72), type = "text", 
+          covariate.labels = c('Testing Inclination', 'Batch'),
+          dep.var.labels = c('Log LOS', 'Number of Tests', '72 Hour Return'),
+          header = F, 
+          title = "2SLS Results for Lab+Image Batch: Length of Stay, Number of Tests, and 72-Hour", 
+          style = 'QJE')
 
 
[…]

### 2023-12-04 src/clean.R

@@ -20,7 +20,6 @@ df <- read.csv(paste0(path, 'deidentified_FINAL.csv'))
 test_columns = c("US_PERF", "NON_CON_CT_PERF", "CON_CT_PERF", 
                  "LAB_PERF", "XR_PERF")
 
-
 colnames(df)[colnames(df) == "PLAIN_XRAY"] = "XR_PERF"
 colnames(df)[colnames(df) == "US_ORDER_DTTM_REL"] ="US_ORDER_DTTM_REL"
 colnames(df)[colnames(df) == "CT_WITHOUT_CONTR_ORDER_DTTM_REL"] = "NON_CON_CT_ORDER_REL"

### 2023-12-19 manuscript/manuscript.Rmd
manuscript updates

@@
 
-In this section, we examine the causal effects of provider batching tendency.
-Our reduced form regression results, as presented in Table \ref{tab:reducedform}, indicate that the physician's batch tendency is associated with longer emergency department length of stay (ED LOS), a greater number of tests ordered, and does not significantly influence the likelihood of a 72-hour return.
-However, the relationship between batch tendency and our outcomes of interest is nuanced by the presence of testing inclination among physicians.
+In this section, we explore the causal influence of physician batch tendency on patient outcomes in the emergency department.
+Our reduced form regression results, as presented in Model 1 of Table \ref{tab:reducedform}, reveal that batch tendency correlates with prolonged ED length of stay (LOS), increased diagnostic test ordering, and does not materially alter the likelihood of a 72-hour return.
+Notably, these associations are nuanced by an underlying factor: the physician's testing inclination.
+
+We construct a physician's testing inclination using the same method used to measure physician batch tendency, but instead we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test.
+A key consideration in our analysis is the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
+Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often because as a result of ordering more tests.
+Failure to control for this inclination could result in biased estimates of the effect of batch tendency.
+
+The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG}) elucidates the interconnected pathways between batch tendency, batch ordering, and primary outcomes.
+We posit that while batch tendency directly influences the practice of batch ordering tests, batch tendency and batch ordering are concurrently influenced by testing inclination.
+This is critical as testing inclination has a direct effect on primary outcomes, necessitating its inclusion as a control variable to disentangle its confounding effects.
 
@@
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{Estimates of the reduced form are for the baseline sample described in the text.
@@
 
-Similar to our method used to measure physician batch tendency, we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test (which we call testing inclination).
-A key consideration in our analysis is the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
-Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often because as a result of ordering more tests.
-Failure to control for this inclination could result in biased estimates of the effect of batch tendency.
-We note this assumption made on the causal pathway and further explore this in the section on mediation analysis.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=4in]{../outputs/figures/DAG.png}
+\begin{tablenotes}
+\small
+\item \textit{}
+\end{tablenotes}  
+  \caption{Directed Acyclic Graph}
+  \label{fig:DAG}
+\end{figure}
 
-In Model 2, we control for testing inclination and observe notable changes in the coefficients for batch tendency.
+In Model 2 of Table \ref{tab:reducedform}, we control for testing inclination and observe notable changes in the coefficients for batch tendency.
 The previously significant positive association with ln(LOS) diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
@@
 
-In this section we investigate whether the reduced-form effects observed in Section 4.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
+In this section we investigate whether the reduced-form effects observed in Section 4.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency.
+We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
 For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
@@
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{This table reports the estimated coefficients of a reduced-form regression of our main outcomes on physician batching tendency for samples based on the patient’s chief complaint field.
@@
   \label{tab:2SLS} 
-\begin{tabular}{p{6cm}ccc}
+\begin{tabular}{p{7cm}ccc}
 \\[-1.8ex]\hline 
@@
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{Estimates of the IV model are for the baseline sample described in the text.
@@
 *ED length of stay (LOS).*-- Our analysis reveals that batch ordering does not significantly reduce the length of stay in the ED.
-This finding holds even after accounting for various control variables.
-Contrary to common assumptions that batching tests could expedite patient throughput, our results suggest that the efficiency gains are offset.
-A possible explanation lies in the additional time needed to process the increased volume of tests ordered through batching.
-These findings challenge prevailing perceptions and call for a reevaluation of batching as a time-saving strategy in ED operations.
+This finding holds even after accounting for various control variables, including ED volume and patient acuity.
+Contrary to common assumptions that batching tests could expedite patient throughput, our results suggest that the efficiency gains could be offset.
+A possible explanation lies in excess tests being ordered due to batching, which could lead to increased wait times for patients.
+This is consistent with our finding that batching increases the number of tests ordered.
 
 *Resource Utilization.*-- The impact of batching on resource utilization is pronounced.
-We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 2,206 additional tests in our dataset.
-Particularly noteworthy is the substantial rise in X-Ray orders (by 28.8 percentage points), as shown in Table \ref{tab:2SLStests}.
-This trend towards increased X-Ray orders raises questions about potential over-testing and its implications for healthcare resources and patient care, especially considering the risks associated with unnecessary radiation exposure.
-
-\begin{table}[!htbp] \centering 
-  \caption{2SLS Results: Specific Tests} 
-  \label{tab:2SLStests} 
-\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & X-Ray & Ultrasound & Non-Con CT & Con CT & Lab \\ 
-\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
-\hline \\[-1.8ex] 
- Batch & 0.288$^{**}$ & $-$0.003 & 0.095 & $-$0.242$^{***}$ & $-$0.003 \\ 
-  & (0.126) & (0.051) & (0.087) & (0.056)  & (0.080) \\ 
-  & & & & &\\ 
- Testing Inclination & 0.240$^{***}$ & 0.086$^{***}$ & 0.096$^{**}$ & 0.253$^{***}$ & 0.286$^{***}$ \\ 
-  & (0.056) & (0.024) & (0.039) & (0.027) & (0.038) \\ 
-  & & & & & \\ 
-Test Ordering Frequency & 0.47 & 0.11 & 0.21 & 0.19 & 0.77 \\ 
-  & & & & &\\ 
-\textit{N} & 41,929 & 41,929 & 41,929 & 41,929 & 41,929 \\ 
-R$^{2}$ & 0.350 & 0.077 & 0.236 & 0.017 \\ 
-\hline 
-\hline \\[-1.8ex] 
-\end{tabular}
-\begin{tablenotes}
-\small
-\item \textit{Estimates of the IV model are for the baseline sample described in the text.
-Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
-Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
-Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
-Robust standard errors are clustered at the physician level.}
-\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
-\end{tablenotes}
-\end{table}
+We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 2,206 additional tests during our study period.
+This is likely due to the information gain that occurs from sequential test ordering, where a physician may order additional tests after receiving the results of the first test.
 
@@
 The instrumental variable analysis presented here underscores the complexities underlying batch ordering practices in the ED.
-While batching is often seen as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization without corresponding gains in patient throughput or reductions in short-term return rates.
+While batching is often seen by physicians as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization without corresponding gains in patient throughput or reductions in short-term return rates.
 These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
-
-## Investigating Mechanisms: Excess X-Rays
 

### 2023-12-19 manuscript/manuscript.tex
manuscript updates

@@
 
-In this section, we examine the causal effects of provider batching
-tendency. Our reduced form regression results, as presented in Table
-\ref{tab:reducedform}, indicate that the physician's batch tendency is
-associated with longer emergency department length of stay (ED LOS), a
-greater number of tests ordered, and does not significantly influence
-the likelihood of a 72-hour return. However, the relationship between
-batch tendency and our outcomes of interest is nuanced by the presence
-of testing inclination among physicians.
+In this section, we explore the causal influence of physician batch
+tendency on patient outcomes in the emergency department. Our reduced
+form regression results, as presented in Model 1 of Table
+\ref{tab:reducedform}, reveal that batch tendency correlates with
+prolonged ED length of stay (LOS), increased diagnostic test ordering,
+and does not materially alter the likelihood of a 72-hour return.
+Notably, these associations are nuanced by an underlying factor: the
+physician's testing inclination.
+
+We construct a physician's testing inclination using the same method
+used to measure physician batch tendency, but instead we use the
+physician's residualized leave-out average test rate to get a measure of
+a physicians propensity to test. A key consideration in our analysis is
+the strong positive correlation (\(r = 0.79\)) between batch tendency
+and testing inclination. Such a high degree of correlation raises the
+concern that physicians with a higher testing inclination may batch more
+often because as a result of ordering more tests. Failure to control for
+this inclination could result in biased estimates of the effect of batch
+tendency.
+
+The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG})
+elucidates the interconnected pathways between batch tendency, batch
+ordering, and primary outcomes. We posit that while batch tendency
+directly influences the practice of batch ordering tests, batch tendency
+and batch ordering are concurrently influenced by testing inclination.
+This is critical as testing inclination has a direct effect on primary
+outcomes, necessitating its inclusion as a control variable to
+disentangle its confounding effects.
 
@@
 
-Similar to our method used to measure physician batch tendency, we use
-the physician's residualized leave-out average test rate to get a
-measure of a physicians propensity to test (which we call testing
-inclination). A key consideration in our analysis is the strong positive
-correlation (\(r = 0.79\)) between batch tendency and testing
-inclination. Such a high degree of correlation raises the concern that
-physicians with a higher testing inclination may batch more often
-because as a result of ordering more tests. Failure to control for this
-inclination could result in biased estimates of the effect of batch
-tendency. We note this assumption made on the causal pathway and further
-explore this in the section on mediation analysis.
-
-In Model 2, we control for testing inclination and observe notable
-changes in the coefficients for batch tendency. The previously
-significant positive association with ln(LOS) diminishes, suggesting
-that the initial estimate may have been capturing the combined effects
-of batching and testing inclination. Estimated coefficients for batch
-tendency are scaled by the difference in batch tendency going from the
-ninetieth to tenth percentile in physician batch tendency---equal to
-15.8 pp---for interpretability, and coefficients for testing inclination
-are scaled by the difference in testing inclination going from the
-ninetieth to tenth percentile in physician testing inclination---equal
-to 34.9 pp.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=4in]{../outputs/figures/DAG.png}
+\begin{tablenotes}
+\small
+\item \textit{}
+\end{tablenotes}  
+  \caption{Directed Acyclic Graph}
+  \label{fig:DAG}
+\end{figure}
+
+In Model 2 of Table \ref{tab:reducedform}, we control for testing
+inclination and observe notable changes in the coefficients for batch
+tendency. The previously significant positive association with ln(LOS)
+diminishes, suggesting that the initial estimate may have been capturing
+the combined effects of batching and testing inclination. Estimated
+coefficients for batch tendency are scaled by the difference in batch
+tendency going from the ninetieth to tenth percentile in physician batch
+tendency---equal to 15.8 pp---for interpretability, and coefficients for
+testing inclination are scaled by the difference in testing inclination
+going from the ninetieth to tenth percentile in physician testing
+inclination---equal to 34.9 pp.
 
@@
 in Section 4.1 are due to differences in batch rates across providers or
-due to other provider differences correlated with batch tendency We
+due to other provider differences correlated with batch tendency. We
 start by studying reduced-form effects among patients with complaints
@@
 ordering does not significantly reduce the length of stay in the ED.
-This finding holds even after accounting for various control variables.
-Contrary to common assumptions that batching tests could expedite
-patient throughput, our results suggest that the efficiency gains are
-offset. A possible explanation lies in the additional time needed to
-process the increased volume of tests ordered through batching. These
-findings challenge prevailing perceptions and call for a reevaluation of
-batching as a time-saving strategy in ED operations.
+This finding holds even after accounting for various control variables,
+including ED volume and patient acuity. Contrary to common assumptions
+that batching tests could expedite patient throughput, our results
+suggest that the efficiency gains could be offset. A possible
+explanation lies in excess tests being ordered due to batching, which
+could lead to increased wait times for patients. This is consistent with
+our finding that batching increases the number of tests ordered.
 
@@
 tests ordered due to batching, corresponding to approximately 2,206
-additional tests in our dataset. Particularly noteworthy is the
-substantial rise in X-Ray orders (by 28.8 percentage points), as shown
-in Table \ref{tab:2SLStests}. This trend towards increased X-Ray orders
-raises questions about potential over-testing and its implications for
-healthcare resources and patient care, especially considering the risks
-associated with unnecessary radiation exposure.
-
-\begin{table}[!htbp] \centering 
-  \caption{2SLS Results: Specific Tests} 
-  \label{tab:2SLStests} 
-\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & X-Ray & Ultrasound & Non-Con CT & Con CT & Lab \\ 
-\\[-1.8ex] & (1) & (2) & (3) & (4) & (5)\\ 
-\hline \\[-1.8ex] 
- Batch & 0.288$^{**}$ & $-$0.003 & 0.095 & $-$0.242$^{***}$ & $-$0.003 \\ 
-  & (0.126) & (0.051) & (0.087) & (0.056)  & (0.080) \\ 
-  & & & & &\\ 
- Testing Inclination & 0.240$^{***}$ & 0.086$^{***}$ & 0.096$^{**}$ & 0.253$^{***}$ & 0.286$^{***}$ \\ 
-  & (0.056) & (0.024) & (0.039) & (0.027) & (0.038) \\ 
-  & & & & & \\ 
-Test Ordering Frequency & 0.47 & 0.11 & 0.21 & 0.19 & 0.77 \\ 
-  & & & & &\\ 
-\textit{N} & 41,929 & 41,929 & 41,929 & 41,929 & 41,929 \\ 
-R$^{2}$ & 0.350 & 0.077 & 0.236 & 0.017 \\ 
-\hline 
-\hline \\[-1.8ex] 
-\end{tabular}
-\begin{tablenotes}
-\small
-\item \textit{Estimates of the IV model are for the baseline sample described in the text.
-Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
-Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
-Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
-Robust standard errors are clustered at the physician level.}
-\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
-\end{tablenotes}
-\end{table}
+additional tests during our study period. This is likely due to the
+information gain that occurs from sequential test ordering, where a
+physician may order additional tests after receiving the results of the
+first test.
 
@@
 complexities underlying batch ordering practices in the ED. While
-batching is often seen as a method to enhance efficiency, our findings
-suggest that it may lead to increased resource utilization without
-corresponding gains in patient throughput or reductions in short-term
-return rates. These insights point to the need for a more nuanced
-approach to test ordering in emergency care, one that carefully weighs
-the benefits of comprehensive assessment against the risks and costs of
-potential over-testing.
-
-\hypertarget{investigating-mechanisms-excess-x-rays}{%
-\subsection{Investigating Mechanisms: Excess
-X-Rays}\label{investigating-mechanisms-excess-x-rays}}
+batching is often seen by physicians as a method to enhance efficiency,
+our findings suggest that it may lead to increased resource utilization
+without corresponding gains in patient throughput or reductions in
+short-term return rates. These insights point to the need for a more
+nuanced approach to test ordering in emergency care, one that carefully
+weighs the benefits of comprehensive assessment against the risks and
+costs of potential over-testing.
 

### 2023-12-19 src/figures.R

@@ -9,12 +9,14 @@ library(lmtest)
 library(sandwich)
 library(ggsci)
 library(xtable) # Output to LaTeX table format
+library(ggdag)
+library(ggplot2)
+theme_set(theme_dag())
 
 data <- read_csv('outputs/data/final.csv')
 
 
 ##########################################################################
-
 
 library(ggplot2)
 library(lfe)

### 2023-12-30 manuscript/manuscript.Rmd
made changes to manuscript

@@
 ---
-title: "To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED"
+title: "I Got 99 Problems, But a Batch Ain't One"
+subtitle: "Sequential vs. Batched Testing Strategies in the ED"
 authors:
@@
 abstract: |
-  This paper focuses on analyzing sequential versus batched testing strategies in an emergency department (ED) at Mayo Clinic Arizona, with respect to their associations impacts on patient length of stay, hospital readmission, and healthcare resource utilization.
-A theoretical model was developed and tested to identify patient and hospital features that may influence the decision to batch or sequentially order tests, such as patient complexity, physician experience, occupancy level and complexity.
-Finally, we performed retrospective analysis of ED operational data to investigate the impact of batching on our key outcomes.
-The overall result was that batch ordering tests was associated with greater patient length of stay and resource utilization, even when we control for variation in patient complexity, physician experience, and hospital occupancy.
+  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED.
+Using records of over 45,000 Mayo Clinic emergency department visits, we find that quasi-random assignment to a top (versus bottom) decile batching physician significantly increases excess diagnostic testing.
+Instrumental variable results show that batching leads to an additional 13.5 tests per 100 batches, resulting in excess tests that would not have been ordered under a sequential-ordering strategy.
+Contrary to expectations, batch ordering does not reduce ED length of stay or affect 72-hour return rates, challenging the perceived efficiency of batching and highlighting the necessity for more targeted testing strategies in emergency care.
 acknowledgment: |
@@
 Our analytical lens is focused on the Emergency Department at the Mayo Clinic of Arizona, a distinguished tertiary care establishment.
-During our study's timeframe, the ED recorded an annual visitation of approximately 45,000 patients.
+During the timeframe of our study, the ED recorded an annual visitation of approximately 45,000 patients.
 The department is singularly staffed by board-eligible or board-certified emergency physicians, abstaining from the services of nurse practitioners or physician assistants.
@@
 Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
-We observe approximately 45,000 such visits until during the stufy period.
-To improve power, we drop encounters with rare chief complaints (< 1000 total encounters of this kind) and complaints where a batch order occurs less than 10 percent of the time.
-Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem were we to include them.
-Example complaints dropped include urinary complaints and _____.
-Excluding these conditions does not introduce selection bias only if physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint cateogry.}.
+We observe approximately $48,000$ such visits until during the study period.
+To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $10$ percent of the time.
+Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them.
+Examples of complaints dropped include urinary complaints and allergic reactions.
+Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
+While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
+In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving physicians who treat over 500 ED cases per year.
 
@@
 
-### Outcomes
+Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter.
+While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice, where we observe that approximately 97(?) percent of all tests ordered are conducted.
+Below, we detail our primary outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.
 
-In our instrumental variables (IV) analysis, the key explanatory variable, $Batched_i$, is an indicator for whether patient $i$ has their tests batch ordered at their ED encounter.
-This variable is crucial for understanding the causal effects of batch testing on various patient outcomes and resource utilization.
-Below, we detail our main outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.
+  *(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
+It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
+Although the literature provides a range for long ED-LOS, we operationalize long ED-LOS as stays exceeding the 75th percentile in our dataset for this study.
+This approach allows us to contextualize the definition of extended stays according to the specific demographics and case mix of the ED under study.
 
-*ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
-It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
-Although literature provides a range for long ED-LOS, for the purpose of this study, we operationalize long ED-LOS as stays exceeding the 75th percentile in our dataset.
-This approach allows us to contextualize the definition of long stays according to the specific demographics and case mix of the EDs under study.
-
-*Resource Utilization*-- Resource utilization in the ED context typically refers to the extent of medical services and interventions a patient receives.
+  *(b) Resource Utilization*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives.
 In this study, we quantify resource utilization by the total number of diagnostic tests ordered per patient during their ED stay.
-This encompasses both initial and any subsequent tests.
+This encompasses both initial and subsequent tests.
 The hypothesis is that batch testing may lead to variations in the number of tests ordered, potentially influencing the overall healthcare expenditure and efficiency.
 
-*72-hour Return*-- The rate of patients returning to the ED within 72 hours of their initial visit serves as a proxy for the quality of care received.
+  *(c) 72-hour return*-- The rate of patients returning to the ED within 72 hours of their initial visit is a proxy for the quality of care received.
 A high 72-hour return rate might indicate inadequate treatment or diagnostic oversight during the initial visit.
-We measure this outcome as the proportion of patients who have a subsequent ED visit within 72 hours post-initial discharge.
+We measure this outcome as the proportion of patients with a subsequent ED visit within 72 hours post-initial discharge.
 
@@
 In the context of our study, batching in diagnostic test ordering is defined as placing multiple test orders for a single patient within a 5-minute interval.
-To ensure the robustness of our analysis, sensitivity tests were conducted on this cutoff point, affirming that our results are consistent and reliable irrespective of minor variations in the batching definition.
+To ensure the robustness of our analysis, sensitivity tests were conducted on this cutoff point, affirming that our results are consistent and reliable, irrespective of minor variations in the batching definition.
 
@@
 * Lab + Image Batch: This category of batch order includes instances where a patient’s batched diagnostic tests comprise a combination of at least one imaging test (such as a contrasted CT scan, non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
-This mix is particularly interesting as it suggests a comprehensive diagnostic approach, possibly catering to complex medical cases.
+This mix is fascinating as it suggests a comprehensive diagnostic approach, possibly catering to complex medical cases.
 
@@
   \centering
+  \caption{Batch Frequency by Chief Complaint}
   \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
@@
 \label{tab:summary_statistics}
-\begin{tabular}{p{9cm}cccc}
+\begin{tabular}{p{10.5cm}cccc}
 \toprule
@@
 \begin{tablenotes}
-\small
-\item \textit{Notes: This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text.}
+\tiny
+\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text.
+Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
 \end{tablenotes}
@@
 In most EDs, however, physicians have discretion in picking their patients.
-In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm (Traub et al., 2016), which removes potential selection bias concerns for our analyses.
+In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses.
 In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-\hyperref[tab:summary_statistics]{Table 2} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
 
@@
     \centering
-    \caption{Wald Test Results}
+    \caption{Balancing Test: Wald Test for Equality of Means}
     \label{tab:wald_test}
-    \begin{tabular}{p{10cm}cc}
+    \begin{tabular}{p{11cm}cc}
         \toprule
@@
 \begin{tablenotes}
-\small
-\item \textit{Table 2 reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
+\tiny
+\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
 A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization.
@@
 
-Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on data $t$.
+Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on date $t$.
 Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g.
@@
 We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure 2 graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically.
-\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a 10 pp higher tendency physician is associated with a 8 pp increase in the likelihood of having tests batch-ordered in the ED.
+Figure \ref{fig:physician_batching} graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically.
+\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $8$pp increase in the likelihood of having tests batch-ordered in the ED.
 The F-statistic is 50 when all controls and fixed effects are included.
 The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
+
+\begin{figure}[h]
+  \centering
+  \label{fig:physician_batching}
+  \caption{Batch Frequency by Physician for High Volume Complaints}
+  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
+\begin{tablenotes}
+\tiny
+\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.
+Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue).
+The quartile of batchers is calculated based on the batching rate across all complaint areas per physician.
+Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching.
+Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
+\end{tablenotes}  
+\end{figure}
 
@@
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{Estimates of the first stage for the baseline sample described in the text.
@@
 For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay.
-Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$ We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.
+Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$.
+We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.
 
@@
 The literature leveraging the judges design typically performs two informal tests for its implications.
-The first one provides that the first stage should be weakly positive for all subsamples (Dobbie, Goldin, and Yang 2018).
-The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (Bhuller et al.
-2020).
-Appendix Table 2 presents both of these tests in the two columns for various subsamples of interest.
+The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects).
+The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration).
+Appendix Table [...] presents both of these tests in the two columns for various subsamples of interest.
 Finally, we check whether our main results hold using differential, mutually exclusive leniency measures (i.e., by complaint category) in Section 4.4.
@@
 
-In this section, we explore the causal influence of physician batch tendency on patient outcomes in the emergency department.
-Our reduced form regression results, as presented in Model 1 of Table \ref{tab:reducedform}, reveal that batch tendency correlates with prolonged ED length of stay (LOS), increased diagnostic test ordering, and does not materially alter the likelihood of a 72-hour return.
+In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department.
+Our reduced form regression results, as presented in Model 1 of Table \ref{tab:reducedform}, reveal that batch tendency correlates with prolonged ED length of stay (LOS), increased test ordering, and does not materially alter the likelihood of a 72-hour return.
 Notably, these associations are nuanced by an underlying factor: the physician's testing inclination.
-
-We construct a physician's testing inclination using the same method used to measure physician batch tendency, but instead we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test.
-A key consideration in our analysis is the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
-Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often because as a result of ordering more tests.
-Failure to control for this inclination could result in biased estimates of the effect of batch tendency.
-
-The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG}) elucidates the interconnected pathways between batch tendency, batch ordering, and primary outcomes.
-We posit that while batch tendency directly influences the practice of batch ordering tests, batch tendency and batch ordering are concurrently influenced by testing inclination.
-This is critical as testing inclination has a direct effect on primary outcomes, necessitating its inclusion as a control variable to disentangle its confounding effects.
 
@@
 \label{tab:reducedform}
-\begin{tabular}{p{6cm}ccc}
+\begin{tabular}{p{6.5cm}ccc}
 \toprule
@@
 Hospital volume? & Yes & Yes & Yes \\
-\midrule
-Mean number of encounters per physician & 1805.3 &  &  \\
-\midrule
 \bottomrule
@@
 
+We construct a physician's testing inclination using the same method used to measure physician batch tendency, but instead we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test.
+A key consideration in our analysis is the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
+Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often as a result of ordering more tests.
+Failure to control for this inclination could result in biased estimates of the effect of batch tendency.
+
+The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG}) elucidates the interconnected pathways between batch tendency, batch ordering, and primary outcomes.
+We posit that while batch tendency directly influences the practice of batch ordering tests, batch tendency and batch ordering are concurrently influenced by testing inclination.
+This is critical as testing inclination has a direct effect on primary outcomes, necessitating its inclusion as a control variable to disentangle its confounding effects.
+
 \begin{figure}[h]
   \centering
+  \caption{Directed Acyclic Graph}
+  \label{fig:DAG}
   \includegraphics[width=4in]{../outputs/figures/DAG.png}
@@
 \end{tablenotes}  
-  \caption{Directed Acyclic Graph}
-  \label{fig:DAG}
 \end{figure}
@@
 The previously significant positive association with ln(LOS) diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
-Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency—equal to 15.8 pp—for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination—equal to 34.9 pp.
+Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination—equal to 34.9 pp.
 
@@
 These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
+
+## Investigating Mechanisms: Excess Testing
+
+In the context of our study's findings, particularly the pronounced impact of batching on resource utilization, it is essential to condense and ground our discussion in the observed data.
+We reported a significant increase of 0.135 in the number of tests ordered due to batching, translating to roughly 13.5 excess tests per 100 batch orders.
+Here, an 'excess test' is defined as one that likely would not have been ordered if the physician had chosen sequential test ordering.
+This suggests that the information gain from sequential testing, where the need for additional tests is informed by the results of earlier tests, could lead to more efficient and targeted use of diagnostic resources.
+
+To further understand this phenomenon, we probe into high-volume complaint areas in the ED.
+By focusing on these areas, we can discern which specific complaints are more likely to lead to excess testing under a batch ordering system.
+This analysis is crucial as it sheds light on the causal effect of batching in different clinical scenarios.
+
+For instance, let's consider complaints like chest pain, abdominal pain, and shortness of breath – common high-volume areas in the ED.
+In cases of chest pain, batch ordering might lead to simultaneous orders for ECGs, blood tests, and imaging, potentially resulting in unnecessary tests if the initial ECG provides clear diagnostic information.
+In contrast, a sequential approach could have deferred or eliminated the need for some of these tests based on initial findings.
+
+Similarly, in patients presenting with abdominal pain, batched orders for a combination of imaging and laboratory tests might be common.
+However, if initial lab results indicate a clear diagnosis, further imaging may not be necessary, highlighting the potential for excess testing through batching.
+
+Analyzing these high-volume complaint areas allows us to identify patterns in testing practices and their implications.
+It becomes evident that batch ordering, while perhaps intended to expedite diagnosis, may inadvertently lead to the over-utilization of tests.
+This not only has implications for resource utilization but also for patient experience and overall ED efficiency.
+
+In conclusion, our investigation into the mechanisms behind excess testing due to batching reveals a nuanced picture.
+While batching might seem efficient at a glance, the lack of information gain compared to sequential testing leads to a significant increase in the number of tests ordered, particularly in high-volume complaint areas.
+This insight is vital for formulating more effective testing strategies in emergency care, highlighting the need for careful consideration of when to employ batching versus sequential testing approaches.
 

### 2023-12-30 manuscript/manuscript.tex
made changes to manuscript

@@
 
-\RUNTITLE{To Batch or Not to Batch: Sequential vs.~Batched Testing
-Strategies in the ED}
-
-\TITLE{To Batch or Not to Batch: Sequential vs.~Batched Testing
-Strategies in the ED}
+\RUNTITLE{I Got 99 Problems, But a Batch Ain't One}
+
+\TITLE{I Got 99 Problems, But a Batch Ain't One}
 
@@
 
-\ABSTRACT{This paper focuses on analyzing sequential versus batched
-testing strategies in an emergency department (ED) at Mayo Clinic
-Arizona, with respect to their associations impacts on patient length of
-stay, hospital readmission, and healthcare resource utilization. A
-theoretical model was developed and tested to identify patient and
-hospital features that may influence the decision to batch or
-sequentially order tests, such as patient complexity, physician
-experience, occupancy level and complexity. Finally, we performed
-retrospective analysis of ED operational data to investigate the impact
-of batching on our key outcomes. The overall result was that batch
-ordering tests was associated with greater patient length of stay and
-resource utilization, even when we control for variation in patient
-complexity, physician experience, and hospital occupancy.}
+\ABSTRACT{We use practice variation across physicians to uncover the
+role of test-ordering on care delivery in the ED. Using records of over
+45,000 Mayo Clinic emergency department visits, we find that
+quasi-random assignment to a top (versus bottom) decile batching
+physician significantly increases excess diagnostic testing.
+Instrumental variable results show that batching leads to an additional
+13.5 tests per 100 batches, resulting in excess tests that would not
+have been ordered under a sequential-ordering strategy. Contrary to
+expectations, batch ordering does not reduce ED length of stay or affect
+72-hour return rates, challenging the perceived efficiency of batching
+and highlighting the necessity for more targeted testing strategies in
+emergency care.}
 
@@
 Clinic of Arizona, a distinguished tertiary care establishment. During
-our study's timeframe, the ED recorded an annual visitation of
+the timeframe of our study, the ED recorded an annual visitation of
 approximately 45,000 patients. The department is singularly staffed by
@@
 Our research design focuses on adults who visit the Mayo Clinic of
-Arizona ED. We observe approximately 45,000 such visits until during the
-stufy period. To improve power, we drop encounters with rare chief
-complaints (\textless{} 1000 total encounters of this kind) and
-complaints where a batch order occurs less than 10 percent of the time.
-Since batch orders are rare for these cases, our physician batch
-tendency instrument could suffer from a weak instrument problem were we
-to include them. Example complaints dropped include urinary complaints
-and \_\_\_\_\_. Excluding these conditions does not introduce selection
-bias only if physician test batching tendency is orthogonal to physician
-diagnosing behavior. While this assumption may be violated if we were to
-use a very detailed level of chief complaint information upon which to
-base our exclusion criterion, it is plausibly satisfied when using broad
-complaint
-categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint cateogry.}.
+Arizona ED. We observe approximately \(48,000\) such visits until during
+the study period. To improve power, we drop encounters with rare chief
+complaints (\(<1000\) total encounters of this kind) and complaints
+where a batch order occurs less than \(10\) percent of the time. Since
+batch orders are rare for these cases, our physician batch tendency
+instrument could suffer from a weak instrument problem if we included
+them. Examples of complaints dropped include urinary complaints and
+allergic reactions. Excluding these conditions does not introduce
+selection bias unless physician test batching tendency is orthogonal to
+physician diagnosing behavior. While this assumption may be violated if
+we were to use a very detailed level of chief complaint information upon
+which to base our exclusion criterion, it is plausibly satisfied when
+using broad complaint
+categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
+In order to estimate a precise measure of physician-level batch
+tendency, we further restrict our sample to the \(42,000\) encounters
+involving physicians who treat over 500 ED cases per year.
 
@@
 
-\hypertarget{outcomes}{%
-\subsubsection{Outcomes}\label{outcomes}}
-
-In our instrumental variables (IV) analysis, the key explanatory
-variable, \(Batched_i\), is an indicator for whether patient \(i\) has
-their tests batch ordered at their ED encounter. This variable is
-crucial for understanding the causal effects of batch testing on various
-patient outcomes and resource utilization. Below, we detail our main
-outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c)
-72-hour return.
-
-\emph{ED length of stay (LOS).}-- ED-LOS is a critical measure of
+Our explanatory variable in the IV analysis, \(Batched_i\), is an
+indicator for whether patient \(i\) has their tests batch-ordered at
+their ED encounter. While the patient could decide not to undergo the
+tests ordered by the physician, this is rare in practice, where we
+observe that approximately 97(?) percent of all tests ordered are
+conducted. Below, we detail our primary outcomes: (a) ED length of stay
+(LOS), (b) resource utilization, and (c) 72-hour return.
+
+\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
 efficiency and patient throughput in emergency care settings. It is
@@
 departure, whether by discharge or admission to the hospital. Although
-literature provides a range for long ED-LOS, for the purpose of this
-study, we operationalize long ED-LOS as stays exceeding the 75th
-percentile in our dataset. This approach allows us to contextualize the
-definition of long stays according to the specific demographics and case
-mix of the EDs under study.
-
-\emph{Resource Utilization}-- Resource utilization in the ED context
+the literature provides a range for long ED-LOS, we operationalize long
+ED-LOS as stays exceeding the 75th percentile in our dataset for this
+study. This approach allows us to contextualize the definition of
+extended stays according to the specific demographics and case mix of
+the ED under study.
+
+\emph{(b) Resource Utilization}-- Resource utilization in the ED
 typically refers to the extent of medical services and interventions a
@@
 total number of diagnostic tests ordered per patient during their ED
-stay. This encompasses both initial and any subsequent tests. The
-hypothesis is that batch testing may lead to variations in the number of
-tests ordered, potentially influencing the overall healthcare
-expenditure and efficiency.
-
-\emph{72-hour Return}-- The rate of patients returning to the ED within
-72 hours of their initial visit serves as a proxy for the quality of
+stay. This encompasses both initial and subsequent tests. The hypothesis
+is that batch testing may lead to variations in the number of tests
+ordered, potentially influencing the overall healthcare expenditure and
+efficiency.
+
+\emph{(c) 72-hour return}-- The rate of patients returning to the ED
+within 72 hours of their initial visit is a proxy for the quality of
 care received. A high 72-hour return rate might indicate inadequate
 treatment or diagnostic oversight during the initial visit. We measure
-this outcome as the proportion of patients who have a subsequent ED
-visit within 72 hours post-initial discharge.
+this outcome as the proportion of patients with a subsequent ED visit
+within 72 hours post-initial discharge.
 
@@
 tests were conducted on this cutoff point, affirming that our results
-are consistent and reliable irrespective of minor variations in the
+are consistent and reliable, irrespective of minor variations in the
 batching definition.
@@
   non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
-  This mix is particularly interesting as it suggests a comprehensive
-  diagnostic approach, possibly catering to complex medical cases.
+  This mix is fascinating as it suggests a comprehensive diagnostic
+  approach, possibly catering to complex medical cases.
 \item
@@
   \centering
+  \caption{Batch Frequency by Chief Complaint}
   \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
@@
 assigned to physicians via a rotational patient assignment algorithm
-(Traub et al., 2016), which removes potential selection bias concerns
-for our analyses. In essence, barring arrival time and shift-level
-variation, the physician-to-patient matching can be deemed random.
-\hyperref[tab:summary_statistics]{Table 2} displays that patient
-encounters (regarding chief complaints and emergency severity) are
-equitably distributed across physicians within our study's cohort.
+\citet{traub2016emergency}, which removes potential selection bias
+concerns for our analyses. In essence, barring arrival time and
+shift-level variation, the physician-to-patient matching can be deemed
+random. Table \ref{tab:wald_test} displays that patient encounters
+(regarding chief complaints and emergency severity) are equitably
+distributed across physicians within our study's cohort.
 
@@
 \(i\) had their diagnostic tests batched on encounter that took place on
-data \(t\). Fixed effects include year-month fixed effects,
+date \(t\). Fixed effects include year-month fixed effects,
 \(\alpha_{ym}\), to control for time and seasonal variation in batching,
@@
 We document that the Mayo Clinic ED physicians exhibit wide, systematic
-variation in their propensity to batch order diagnostic tests. Figure 2
-graphs the histogram of batch-ordering frequency by physician for
-popular chief complaints, highlighting that the variation in batching
-differs systematically. \hyperref[tab:regression]{Table 3} presents the
-``first stage'' in a regression table: being assigned to a 10 pp higher
-tendency physician is associated with a 8 pp increase in the likelihood
-of having tests batch-ordered in the ED. The F-statistic is 50 when all
+variation in their propensity to batch order diagnostic tests. Figure
+\ref{fig:physician_batching} graphs the histogram of batch-ordering
+frequency by physician for popular chief complaints, highlighting that
+the variation in batching differs systematically.
+\hyperref[tab:regression]{Table 3} presents the ``first stage'' in a
+regression table: being assigned to a \(10\)pp higher batch-tendency
+physician is associated with a \(8\)pp increase in the likelihood of
+having tests batch-ordered in the ED. The F-statistic is 50 when all
 controls and fixed effects are included. The coefficient is greater than
@@
 sample only, which excludes the rare complaints.
+
+\begin{figure}[h]
+  \centering
+  \label{fig:physician_batching}
+  \caption{Batch Frequency by Physician for High Volume Complaints}
+  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
+\begin{tablenotes}
+\tiny
+\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.
+Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue).
+The quartile of batchers is calculated based on the batching rate across all complaint areas per physician.
+Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching.
+Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
+\end{tablenotes}  
+\end{figure}
 
@@
 assigned physician \(j\)'s underlying tendency to batch,
-\(Tendency_{i,j}^{phys}\) We cluster robust standard errors at the
+\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
 physician level to account for the assignment process of patients to
@@
 that the first stage should be weakly positive for all subsamples
-(Dobbie, Goldin, and Yang 2018). The second implication asserts that the
+(\citet{dobbie2018effects}). The second implication asserts that the
 instrument constructed by leaving out a particular subsample has
-predictive power over that same left-out subsample (Bhuller et
-al.~2020). Appendix Table 2 presents both of these tests in the two
-columns for various subsamples of interest. Finally, we check whether
-our main results hold using differential, mutually exclusive leniency
-measures (i.e., by complaint category) in Section 4.4.
+predictive power over that same left-out subsample
+(\citet{bhuller2020incarceration}). Appendix Table {[}\ldots{]} presents
+both of these tests in the two columns for various subsamples of
+interest. Finally, we check whether our main results hold using
+differential, mutually exclusive leniency measures (i.e., by complaint
+category) in Section 4.4.
 
@@
 In this section, we explore the causal influence of physician batch
-tendency on patient outcomes in the emergency department. Our reduced
-form regression results, as presented in Model 1 of Table
-\ref{tab:reducedform}, reveal that batch tendency correlates with
-prolonged ED length of stay (LOS), increased diagnostic test ordering,
-and does not materially alter the likelihood of a 72-hour return.
-Notably, these associations are nuanced by an underlying factor: the
-physician's testing inclination.
-
-We construct a physician's testing inclination using the same method
-used to measure physician batch tendency, but instead we use the
-physician's residualized leave-out average test rate to get a measure of
-a physicians propensity to test. A key consideration in our analysis is
-the strong positive correlation (\(r = 0.79\)) between batch tendency
-and testing inclination. Such a high degree of correlation raises the
-concern that physicians with a higher testing inclination may batch more
-often because as a result of ordering more tests. Failure to control for
-this inclination could result in biased estimates of the effect of batch
-tendency.
-
-The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG})
-elucidates the interconnected pathways between batch tendency, batch
-ordering, and primary outcomes. We posit that while batch tendency
-directly influences the practice of batch ordering tests, batch tendency
-and batch ordering are concurrently influenced by testing inclination.
-This is critical as testing inclination has a direct effect on primary
-outcomes, necessitating its inclusion as a control variable to
-disentangle its confounding effects.
+tendency on patient outcomes and resource utilization in the emergency
+department. Our reduced form regression results, as presented in Model 1
+of Table \ref{tab:reducedform}, reveal that batch tendency correlates
+with prolonged ED length of stay (LOS), increased test ordering, and
+does not materially alter the likelihood of a 72-hour return. Notably,
+these associations are nuanced by an underlying factor: the physician's
+testing inclination.
 
@@
 
+We construct a physician's testing inclination using the same method
+used to measure physician batch tendency, but instead we use the
+physician's residualized leave-out average test rate to get a measure of
+a physicians propensity to test. A key consideration in our analysis is
+the strong positive correlation (\(r = 0.79\)) between batch tendency
+and testing inclination. Such a high degree of correlation raises the
+concern that physicians with a higher testing inclination may batch more
+often as a result of ordering more tests. Failure to control for this
+inclination could result in biased estimates of the effect of batch
+tendency.
+
+The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG})
+elucidates the interconnected pathways between batch tendency, batch
+ordering, and primary outcomes. We posit that while batch tendency
+directly influences the practice of batch ordering tests, batch tendency
+and batch ordering are concurrently influenced by testing inclination.
+This is critical as testing inclination has a direct effect on primary
+outcomes, necessitating its inclusion as a control variable to
+disentangle its confounding effects.
+
 \begin{figure}[h]
   \centering
+  \caption{Directed Acyclic Graph}
+  \label{fig:DAG}
   \includegraphics[width=4in]{../outputs/figures/DAG.png}
@@
 tendency going from the ninetieth to tenth percentile in physician batch
-tendency---equal to 15.8 pp---for interpretability, and coefficients for
-testing inclination are scaled by the difference in testing inclination
-going from the ninetieth to tenth percentile in physician testing
-inclination---equal to 34.9 pp.
+tendency for interpretability, and coefficients for testing inclination
+are scaled by the difference in testing inclination going from the
+ninetieth to tenth percentile in physician testing inclination---equal
+to 34.9 pp.
 
@@
 
+\hypertarget{investigating-mechanisms-excess-testing}{%
+\subsection{Investigating Mechanisms: Excess
+Testing}\label{investigating-mechanisms-excess-testing}}
+
+In the context of our study's findings, particularly the pronounced
+impact of batching on resource utilization, it is essential to condense
+and ground our discussion in the observed data. We reported a
+significant increase of 0.135 in the number of tests ordered due to
+batching, translating to roughly 13.5 excess tests per 100 batch orders.
+Here, an `excess test' is defined as one that likely would not have been
+ordered if the physician had chosen sequential test ordering. This
+suggests that the information gain from sequential testing, where the
+need for additional tests is informed by the results of earlier tests,
+could lead to more efficient and targeted use of diagnostic resources.
+
+To further understand this phenomenon, we probe into high-volume
+complaint areas in the ED. By focusing on these areas, we can discern
+which specific complaints are more likely to lead to excess testing
+under a batch ordering system. This analysis is crucial as it sheds
+light on the causal effect of batching in different clinical scenarios.
+
+For instance, let's consider complaints like chest pain, abdominal pain,
+and shortness of breath -- common high-volume areas in the ED. In cases
+of chest pain, batch ordering might lead to simultaneous orders for
+ECGs, blood tests, and imaging, potentially resulting in unnecessary
+tests if the initial ECG provides clear diagnostic information. In
+contrast, a sequential approach could have deferred or eliminated the
+need for some of these tests based on initial findings.
+
+Similarly, in patients presenting with abdominal pain, batched orders
+for a combination of imaging and laboratory tests might be common.
+However, if initial lab results indicate a clear diagnosis, further
+imaging may not be necessary, highlighting the potential for excess
+testing through batching.
+
+Analyzing these high-volume complaint areas allows us to identify
+patterns in testing practices and their implications. It becomes evident
+that batch ordering, while perhaps intended to expedite diagnosis, may
+inadvertently lead to the over-utilization of tests. This not only has
+implications for resource utilization but also for patient experience
+and overall ED efficiency.
+
+In conclusion, our investigation into the mechanisms behind excess
+testing due to batching reveals a nuanced picture. While batching might
+seem efficient at a glance, the lack of information gain compared to
+sequential testing leads to a significant increase in the number of
+tests ordered, particularly in high-volume complaint areas. This insight
+is vital for formulating more effective testing strategies in emergency
+care, highlighting the need for careful consideration of when to employ
+batching versus sequential testing approaches.
+
 \hypertarget{robustness}{%

### 2023-12-30 src/IV Validation.R

@@ -317,9 +317,12 @@ complaints <- unique(data$CHIEF_COMPLAINT)
 
 model_results <- list()
 # Iterate through each complaint and run the regression
+
 for(complaint in complaints) {
   
-  data_subset <- data %>% filter(CHIEF_COMPLAINT == complaint)
+  data_subset <- data %>% 
+    filter(CHIEF_COMPLAINT == complaint)
+  
   model <- felm(
     any.batch ~ batch.tendency + test.inclination | 
                 dayofweekt + month_of_year + age_groups + ESI | 0 | ED_PROVIDER,
@@ -348,6 +351,23 @@ stargazer(model_results[10:15], type = "text",
 
 
 sink()
+
+#=========================================================================
+
+data.temp <- data
+
+data.temp$residual_batch <- resid(
+  felm(any.batch ~ 0 | dayofweekt + month_of_year + complaint_esi, data=data.temp))
+
+# Step 2: get batch tendency for each provider
+table(data.temp$CHIEF_COMPLAINT, data.temp$any.batch)
+
+data.temp <- data.temp %>%
+  group_by(ED_PROVIDER) %>%
+  mutate(Sum_Resid=sum(residual_batch, na.rm=T),
+         batch.tendency = (Sum_Resid - residual_batch) / (n() - 1)) %>%
+  
+  ungroup()
 
 
 

### 2024-01-02 manuscript/manuscript.Rmd
draft paper

@@
 ---
-title: "I Got 99 Problems, But a Batch Ain't One"
-subtitle: "Sequential vs. Batched Testing Strategies in the ED"
+title: "To Batch or Not to Batch: Test-Ordering Variability in the Emergency Department and the Impact on Care Delivery"
 authors:
@@
   \centering
+  \caption{Batch Frequency by Physician for High Volume Complaints}
   \label{fig:physician_batching}
-  \caption{Batch Frequency by Physician for High Volume Complaints}
   \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
@@
 \toprule
-\textbf{} & \textbf{Model 1} & \textbf{Model 2} & \textbf{Model 3} \\
+\textbf{} & Model 1 & Model 2 & Model 3 \\
 \midrule
@@
 The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration).
-Appendix Table [...] presents both of these tests in the two columns for various subsamples of interest.
-Finally, we check whether our main results hold using differential, mutually exclusive leniency measures (i.e., by complaint category) in Section 4.4.
+Appendix Table  \ref{tab:monotonicity} presents both of these tests in the two columns for various subsamples of interest.
 
@@
 \toprule
-& \multicolumn{1}{c}{\textbf{ln(LOS)}}
-& \multicolumn{1}{c}{\textbf{Number of Tests}}
-& \multicolumn{1}{c}{\textbf{72-Hour Return}} \\
+& \multicolumn{1}{c}{ln(LOS)}
+& \multicolumn{1}{c}{Number of Tests}
+& \multicolumn{1}{c}{72-Hour Return} \\
 & \multicolumn{1}{c}{(1)}
@@
 The previously significant positive association with ln(LOS) diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
-Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination—equal to 34.9 pp.
+Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination.
 
@@
 We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 0.96 unit increase in the number of diagnostic tests being ordered per patient encounter, and -0.03 pp change in the probability of a patient returning within 72 hours.
-Given the average number of patient encounters per physician in our study period this translates to an additional [...] patient hours in LOS, [...] diagnostic tests, and [...] avoided 72-hour returns.
 
@@
 \midrule
-\multicolumn{4}{l}{\textbf{Controls: Time}} \\
-Batch Tendency & $-0.529$ & $-0.760$ & $-0.006$ \\
-& $(0.710)$ & $(0.490)$ & $(0.094)$ \\
-                    &     &     &  \\
-Testing Inclination & $0.812^{**}$ & $0.892^{***}$ & $-0.001$ \\
-& $(0.343)$ & $(0.194)$ & $(0.045)$ \\
-\\
-\addlinespace
-\multicolumn{4}{l}{\textbf{Controls: Time + Complaint}} \\
-Batch Tendency & $-0.517$ & $-0.714$ & $-0.015$ \\
-& $(0.660)$ & $(0.446)$ & $(0.092)$ \\
-                    &     &     &  \\
-Testing Inclination & $0.848^{**}$ & $1.031^{***}$ & $-0.010$ \\
-& $(0.334)$ & $(0.189)$ & $(0.045)$ \\
-\\
-\addlinespace
-\multicolumn{4}{l}{\textbf{Controls: All Controls}} \\
 Batch Tendency & $-0.465$ & $-0.699$ & $-0.011$ \\
@@
                     &     &     &  \\
-Testing Inclination & $0.818^{**}$ & $1.011^{***}$ & $-0.009$ \\
-& $(0.322)$ & $(0.183)$ & $(0.046)$ \\
+Seasonality and shift fixed effects? & Yes & Yes & Yes \\
+Chief Complaint? & Yes & Yes & Yes \\
+Patient observables? & Yes & Yes & Yes \\
+Testing inclination? & Yes & Yes & Yes \\
                     &     &     &  \\
-Observations & & 2,361 & \\
+N & 2,361 & 2,361 & 2,361 \\
 \addlinespace
@@
 
-In the context of our study's findings, particularly the pronounced impact of batching on resource utilization, it is essential to condense and ground our discussion in the observed data.
-We reported a significant increase of 0.135 in the number of tests ordered due to batching, translating to roughly 13.5 excess tests per 100 batch orders.
-Here, an 'excess test' is defined as one that likely would not have been ordered if the physician had chosen sequential test ordering.
+In the previous section we find a particularly pronounced impact of batching on resource utilization.
+We reported a significant coefficient of 0.135 in the number of tests ordered due to batching, translating to 13.5 excess tests per 100 batch orders.
+Here, an ``excess test" is defined as one that would not have been ordered if the physician had chosen to order tests sequentially.
 This suggests that the information gain from sequential testing, where the need for additional tests is informed by the results of earlier tests, could lead to more efficient and targeted use of diagnostic resources.
@@
 
-For instance, let's consider complaints like chest pain, abdominal pain, and shortness of breath – common high-volume areas in the ED.
-In cases of chest pain, batch ordering might lead to simultaneous orders for ECGs, blood tests, and imaging, potentially resulting in unnecessary tests if the initial ECG provides clear diagnostic information.
-In contrast, a sequential approach could have deferred or eliminated the need for some of these tests based on initial findings.
+We consider the top 5 most common broad complaints in the ED, which account for 82% of all visits.
+Appendix table [...] .
+These include Cardiac/Chest-Related Issues, Gastrointestinal/Abdominal Issues, Respiratory-Related Issues, Neurological/Syncope Issues, and Musculoskeletal/Extremity Issues.
+We then examine the impact of batching on the number of tests ordered for each of these complaints.
 
-Similarly, in patients presenting with abdominal pain, batched orders for a combination of imaging and laboratory tests might be common.
-However, if initial lab results indicate a clear diagnosis, further imaging may not be necessary, highlighting the potential for excess testing through batching.
-
-Analyzing these high-volume complaint areas allows us to identify patterns in testing practices and their implications.
-It becomes evident that batch ordering, while perhaps intended to expedite diagnosis, may inadvertently lead to the over-utilization of tests.
-This not only has implications for resource utilization but also for patient experience and overall ED efficiency.
-
-In conclusion, our investigation into the mechanisms behind excess testing due to batching reveals a nuanced picture.
-While batching might seem efficient at a glance, the lack of information gain compared to sequential testing leads to a significant increase in the number of tests ordered, particularly in high-volume complaint areas.
-This insight is vital for formulating more effective testing strategies in emergency care, highlighting the need for careful consideration of when to employ batching versus sequential testing approaches.
-
-## Robustness
-
-The 2SLS results presented in the previous sections are robust to several alternative specifications probing the key identifying assumptions of independence, exclusion, and monotonicity.
-We describe each robustness check in detail in Appendix ___ and provide a brief overview here ....
-
-Across all robustness checks and main outcomes, we find magnitudes of our estimates broadly unchanged.
+[...]
 
 # Conclusion
+
+Our results highlight that changes in resource utilization and costs can arise as a consequence of small variations in physician test ordering strategies in a single medical encounter, at the emergency department.
+Our instrumental variable approach suggests that the causal effects of batching are substantial: 13.5 excess tests arise per 100 batched orders.
+This is particularly pronounced in high-volume complaint areas, where batching leads to a significant increase in the number of tests ordered.
+However, we find no significant impact of batching on ED length of stay or 72-hour return rates.
+This suggests that despite the increased number of tests, there is no corresponding decrease in short-term readmission rates.
+This highlights the need for more targeted testing strategies that balance comprehensive assessment with resource efficiency.
 
@@
 
-\begin{table}[htbp]
+\begin{table}[!htbp]
 \centering 
-\caption{First Stage over all Subsamples}
-\label{tab:monotonicitycheck}
-\begin{tabular}{p{7cm}cccc}
-\toprule
-Complaint & \multicolumn{2}{c}{Batch Tendency} & N & R$^2$ \\
-\midrule
-Extremity Complaints & 0.662$^{***}$ & (0.079) & 5,259 & 0.126 \\
-Upper Respiratory Symptoms & 1.205$^{***}$ & (0.445) & 1,915 & 0.262 \\
-Abnormal Test Results & 0.670$^{**}$ & (0.325) & 1,816 & 0.141 \\
-Falls, Motor Vehicle Crashes, and Trauma & 0.640$^{**}$ & (0.264) & 2,381 & 0.220 \\
-Shortness of Breath & 1.574$^{***}$ & (0.468) &  2,963 & 0.147 \\
-Gastrointestinal Issues & 1.080$^{***}$ & (0.241) & 3,323 & 0.094 \\
-Chest Pain & 0.790$^{***}$ & (0.253) & 3,521 & 0.140 \\
-Dizziness/Lightheadedness/Syncope & 1.343$^{***}$ & (0.220) & 1,894 & 0.208 \\
-Back or Flank Pain & 0.859$^{**}$ & (0.385) &  2,550 & 0.123 \\
-Neurological Issue & 0.416$^{*}$ & (0.232) &  3,492 & 0.232 \\
-Fatigue and Weakness & 1.728$^{***}$ & (0.414) & 1,494 & 0.193 \\
-Cardiac Arrhythmias & 2.432$^{***}$ & (0.548) & 1,044 & 0.327 \\
-Abdominal Complaints & 1.775$^{***}$ & (0.350) & 6,231 & 0.086 \\
-Skin Complaints & 0.205 & (0.212) & 2,176 & 0.133 \\
-Fevers, Sweats or Chills & 1.650$^{***}$ & (0.576) &  1,842 & 0.211 \\
-\bottomrule
+  \caption{Testing the Monotonicity Assumption} 
+\label{tab:monotonicity}
+\begin{tabular}{@{\extracolsep{5.5pt}}lcc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+Sub-sample margin & Baseline Tendency  & Reverse-Sample Tendency \\ 
+\hline \\[-1.8ex] 
+Extremity Complaints & 0.864$^{***}$ (0.128) & 0.893$^{***}$ (0.068) \\ 
+Upper Respiratory Symptoms & 1.235$^{***}$ (0.247) & 0.953$^{***}$ (0.038) \\ 
+Abnormal Test Results & 0.980$^{***}$ (0.219) & 0.766$^{***}$ (0.111) \\
+Falls, Motor Vehicle Crashes, Assaults, and Trauma & 1.061$^{***}$ (0.167) & 0.784$^{***}$ (0.103) \\
+Shortness of Breath & 1.204$^{***}$ (0.232) & 0.944$^{***}$ (0.047) \\
+Gastrointestinal Issues & 1.049$^{***}$ (0.140) & 0.929$^{***}$ (0.052) \\
+Chest Pain & 1.073$^{***}$ (0.223) & 0.943$^{***}$ (0.048) \\
+Dizziness/Lightheadedness/Syncope & 1.899$^{***}$ (0.171) & 0.930$^{***}$ (0.042) \\
+Back or Flank Pain & 0.657$^{***}$ (0.180) & 0.718$^{***}$ (0.106) \\
+Neurological Issue & 0.970$^{***}$ (0.191) & 0.835$^{***}$ (0.067) \\
+Fatigue and Weakness & 1.698$^{***}$ (0.230) & 0.938$^{***}$ (0.058) \\
+Cardiac Arrhythmias & 2.571$^{***}$ (0.243) & 1.107$^{***}$ (0.059) \\
+Abdominal Complaints & 0.903$^{***}$ (0.275) & 0.971$^{***}$ (0.019) \\
+Skin Complaints & 0.448$^{***}$ (0.098) & 0.584$^{***}$ (0.108) \\
+Fevers, Sweats or Chills & 1.033$^{***}$ (0.305) & 0.888$^{***}$ (0.084) \\
+\hline \\[-1.8ex] 
 \end{tabular} 
+\begin{tablenotes}
+\small
+\item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample.
+Column 2 constructs a new physician batch tendency instrument using all emergency visits, excluding the corresponding sub-sample (``reverse-sample"), and displays the coefficient of the first stage regression back on that sub-sample.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
 \end{table}
 
-\begin{table}[!htbp] \centering 
-  \caption{Regression Results by Complaint} 
-  \label{} 
-\begin{tabular}{@{\extracolsep{5pt}}lccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
- & Dizziness/Syncope & Extremity Complaints & Upper Respiratory Symptoms \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
-\hline \\[-1.8ex] 
- Testing Inclination & 1.156$^{***}$ & 0.901$^{***}$ & 0.516$^{***}$ \\ 
-  & (0.291) & (0.111) & (0.194) \\ 
-  & & & \\ 
- Batch & 0.655$^{*}$ & 0.479$^{**}$ & 0.501$^{*}$ \\ 
-  & (0.368) & (0.242) & (0.283) \\ 
-  & & & \\ 
-\textit{N} & 1,895 & 5,265 & 1,917 \\ 
-R$^{2}$ & 0.400 & 0.348 & 0.461 \\ 
-Adjusted R$^{2}$ & 0.338 & 0.324 & 0.404 \\ 
-Residual Std. Error & 0.801 (df = 1715) & 0.765 (df = 5079) & 0.741 (df = 1733) \\ 
+\begin{longtable}{|p{5cm}|p{12cm}|}
+\caption{Condensed Complaints} \\
 \hline
-\hline \\[-1.8ex] 
-\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
- & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
- & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
-\end{tabular} 
-\end{table}
+\textbf{Condensed Complaint Area} & \textbf{Complaints} \\
+\hline
+Gastrointestinal/Abdominal Issues & Abdominal Cramping, Abdominal Distention, Dyspepsia, Abdominal Pain, Ascites, Hernia, Abdominal Aortic Aneurysm, Abdominal Injury, Pancreatitis, Umbilical Hernia, Anal Fissure, Black or Bloody Stool, Constipation, GERD, Anal Fistula, Diarrhea, Dysphagia, Fecal Impaction, Fistula Follow Up, GIbleeding, GI Problem, Hemorrhoids, Morning Sickness, Nausea, Ostomy Care, Rectal Bleeding, Rectal Pain, Vomiting, Vomiting Blood, Vomiting During Pregnancy, GI Bleeding, Fecal Incontinence, Bloated, Hematochezia, Urine Leakage, Heartburn, Rectal Discharge, Urolithiasis, Ulcerative Colitis, Irritable Bowel Syndrome, Rectal Prolapse, Fistula Evaluation, Rectal Problems, Perianal Abscess, Fisula Evaluation, Stoma Dysfunction \\
+\hline
+Cardiac/Chest-Related Issues & Atrial Fibrillation, Atrial Flutter, Cardiac Valve Problem, Bradycardia, Irregular Heart Beat, Palpitations, POTS, Ventricular Tachycardia, Rapid Heart Rate, Heart Problem, Cardiac Arrest, Congestive Heart Failure, Circulatory Problem, Transient Ischemic Attack, Ventricular Tachycardia, Chest Injury, Chest Pain, Chest Wall Pain, Angina, Collarbone Injury, Rib Injury, Heart Pain \\
+\hline
+Musculoskeletal/Extremity Issues & Ankle Injury, Ankle Pain, Arm Injury, Arm Pain, Cold Extremity, Arm Swelling, Arthritis, Elbow Injury, Elbow Pain, Pseudogout, Extremity Pain, Extremity Weakness, Finger Injury, Hip Injury, Extremity Weakness, Finger Injury, Finger Pain, Dislocation, Foot Infection, Foot Injury, Foot Numbness, Foot Pain, Foot Swelling, Foot Ulcer, Foot Wound Check, Hand Injury, Hand Pain, Back Pain, Back Problem, Flank Pain, Sciatica, Back Injury, Disc Disorder, Assault Victim, Concussion, Facial Injury, Fall, Nasal Trauma, Head Injury, Head Laceration, Motor Vehicle Crash, Puncture Wound, Sexual Assault, Trauma, Domestic Violence, Gun Shot Wound, Work Related Injury, Motorcycle Crash, Injury, Bicycle Accident, Near Drowning, Lip Laceration \\
+\hline
+Neurological/Syncope Issues & Altered Mental Status, Cognitive Concerns, Facial Droop, Pre Syncope, Focal Weakness, Headache, Memory Loss, Migraine, Dementia, Dysphasia, Neuro Problem, Numbness, Paralysis, Seizures, Slurred Speech, Spasms, Stroke Like Symptoms, Tingling, Tremors, Trigeminal Neuralgia, Unable to Speak, Seizure Disorder, Insomnia, Parkinson's Disease, Loss of Consciousness, Neuropathy, Ataxia, Unable to speak, Peripheral Neuropathy, Stroke, Cerebrovascular Accident, Speech Problem, Acute Neurological Problem, Flashes, Light, Unresponsive, Multiple Sclerosis, Parkinson's Disease, Febrile Seizure, Paresthesia, Peripheral Neuropathy, Hydrocephalus, Spasticity, Neuroendocrine Tumor, Dizziness, Near Syncope, Syncope, Vertigo, Spells, Hypotension, Paroxysmal Positional Vertigo, Paroxysmal Positional Vertigo \\
+\hline
+Respiratory-Related Issues & Airway Obstruction, Aspiration, Pain With Breathing, Near Drowning, Respiratory Distress, Shortness of Breath, Wheezing, Increased Work Of Breathing, Difficulty Breathing, Choking, Oxygen Dependence, Hyperventilating, Orthopnea, Congestion, Cough, Coughing Up Blood, Flu Symptoms, Enlarged Tonsils, Peritonsillar Abscess, Nasal Congestion, Sinus Symptoms, Sinusitis, Sore Throat, Hoarseness, Throat Problem, Upper Respiratory Infection, Influenza, Laryngitis, Respiratory Arrest, Pneumonia, Pleural Effusion, Asthma, Croup, URI, Peritonsillar Abscess\\
+\hline
+\end{longtable}
 
-\begin{table}[!htbp] \centering 
-  \caption{Regression Results by Complaint} 
-  \label{} 
-\begin{tabular}{@{\extracolsep{5pt}}lccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
- & Abdominal Complaints & Cardiac Arrhythmias & Fatigue and Weakness \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
-\hline \\[-1.8ex] 
- Testing Inclination & 0.913$^{***}$ & 0.695$^{***}$ & 0.870$^{***}$ \\ 
-  & (0.068) & (0.180) & (0.195) \\ 
-  & & & \\ 
- Batch & $-$0.135$^{**}$ & 0.708$^{***}$ & 0.333$^{*}$ \\ 
-  & (0.066) & (0.129) & (0.189) \\ 
-  & & & \\ 
-\textit{N} & 6,232 & 1,045 & 1,494 \\ 
-R$^{2}$ & 0.038 & 0.525 & 0.294 \\ 
-Adjusted R$^{2}$ & 0.009 & 0.428 & 0.203 \\ 
-Residual Std. Error & 0.752 (df = 6046) & 0.595 (df = 866) & 0.828 (df = 1322) \\ 
-\hline 
-\hline \\[-1.8ex] 
-\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
- & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
- & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
-\end{tabular} 
-\end{table} 
 :::

### 2024-01-02 manuscript/manuscript.tex
draft paper

@@
 
-\RUNTITLE{I Got 99 Problems, But a Batch Ain't One}
-
-\TITLE{I Got 99 Problems, But a Batch Ain't One}
+\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the
+Emergency Department and the Impact on Care Delivery}
+
+\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the
+Emergency Department and the Impact on Care Delivery}
 
@@
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
@@
 predictive power over that same left-out subsample
-(\citet{bhuller2020incarceration}). Appendix Table {[}\ldots{]} presents
-both of these tests in the two columns for various subsamples of
-interest. Finally, we check whether our main results hold using
-differential, mutually exclusive leniency measures (i.e., by complaint
-category) in Section 4.4.
+(\citet{bhuller2020incarceration}). Appendix Table
+\ref{tab:monotonicity} presents both of these tests in the two columns
+for various subsamples of interest.
 
@@
 are scaled by the difference in testing inclination going from the
-ninetieth to tenth percentile in physician testing inclination---equal
-to 34.9 pp.
+ninetieth to tenth percentile in physician testing inclination.
 
@@
 diagnostic tests being ordered per patient encounter, and -0.03 pp
-change in the probability of a patient returning within 72 hours. Given
-the average number of patient encounters per physician in our study
-period this translates to an additional {[}\ldots{]} patient hours in
-LOS, {[}\ldots{]} diagnostic tests, and {[}\ldots{]} avoided 72-hour
-returns.
+change in the probability of a patient returning within 72 hours.
 
@@
 
-In the context of our study's findings, particularly the pronounced
-impact of batching on resource utilization, it is essential to condense
-and ground our discussion in the observed data. We reported a
-significant increase of 0.135 in the number of tests ordered due to
-batching, translating to roughly 13.5 excess tests per 100 batch orders.
-Here, an `excess test' is defined as one that likely would not have been
-ordered if the physician had chosen sequential test ordering. This
-suggests that the information gain from sequential testing, where the
-need for additional tests is informed by the results of earlier tests,
-could lead to more efficient and targeted use of diagnostic resources.
+In the previous section we find a particularly pronounced impact of
+batching on resource utilization. We reported a significant coefficient
+of 0.135 in the number of tests ordered due to batching, translating to
+13.5 excess tests per 100 batch orders. Here, an ``excess test'' is
+defined as one that would not have been ordered if the physician had
+chosen to order tests sequentially. This suggests that the information
+gain from sequential testing, where the need for additional tests is
+informed by the results of earlier tests, could lead to more efficient
+and targeted use of diagnostic resources.
 
@@
 
-For instance, let's consider complaints like chest pain, abdominal pain,
-and shortness of breath -- common high-volume areas in the ED. In cases
-of chest pain, batch ordering might lead to simultaneous orders for
-ECGs, blood tests, and imaging, potentially resulting in unnecessary
-tests if the initial ECG provides clear diagnostic information. In
-contrast, a sequential approach could have deferred or eliminated the
-need for some of these tests based on initial findings.
-
-Similarly, in patients presenting with abdominal pain, batched orders
-for a combination of imaging and laboratory tests might be common.
-However, if initial lab results indicate a clear diagnosis, further
-imaging may not be necessary, highlighting the potential for excess
-testing through batching.
-
-Analyzing these high-volume complaint areas allows us to identify
-patterns in testing practices and their implications. It becomes evident
-that batch ordering, while perhaps intended to expedite diagnosis, may
-inadvertently lead to the over-utilization of tests. This not only has
-implications for resource utilization but also for patient experience
-and overall ED efficiency.
-
-In conclusion, our investigation into the mechanisms behind excess
-testing due to batching reveals a nuanced picture. While batching might
-seem efficient at a glance, the lack of information gain compared to
-sequential testing leads to a significant increase in the number of
-tests ordered, particularly in high-volume complaint areas. This insight
-is vital for formulating more effective testing strategies in emergency
-care, highlighting the need for careful consideration of when to employ
-batching versus sequential testing approaches.
-
-\hypertarget{robustness}{%
-\subsection{Robustness}\label{robustness}}
-
-The 2SLS results presented in the previous sections are robust to
-several alternative specifications probing the key identifying
-assumptions of independence, exclusion, and monotonicity. We describe
-each robustness check in detail in Appendix \_\_\_ and provide a brief
-overview here \ldots.
-
-Across all robustness checks and main outcomes, we find magnitudes of
-our estimates broadly unchanged.
+We consider the top 5 most common broad complaints in the ED, which
+account for 82\% of all visits. Appendix table {[}\ldots{]} . These
+include Cardiac/Chest-Related Issues, Gastrointestinal/Abdominal Issues,
+Respiratory-Related Issues, Neurological/Syncope Issues, and
+Musculoskeletal/Extremity Issues. We then examine the impact of batching
+on the number of tests ordered for each of these complaints.
+
+{[}\ldots{]}
 
@@
 
+Our results highlight that changes in resource utilization and costs can
+arise as a consequence of small variations in physician test ordering
+strategies in a single medical encounter, at the emergency department.
+Our instrumental variable approach suggests that the causal effects of
+batching are substantial: 13.5 excess tests arise per 100 batched
+orders. This is particularly pronounced in high-volume complaint areas,
+where batching leads to a significant increase in the number of tests
+ordered. However, we find no significant impact of batching on ED length
+of stay or 72-hour return rates. This suggests that despite the
+increased number of tests, there is no corresponding decrease in
+short-term readmission rates. This highlights the need for more targeted
+testing strategies that balance comprehensive assessment with resource
+efficiency.
+
 \clearpage
@@
 
-\begin{table}[htbp]
+\begin{table}[!htbp]
 \centering 
-\caption{First Stage over all Subsamples}
-\label{tab:monotonicitycheck}
-\begin{tabular}{p{7cm}cccc}
-\toprule
-Complaint & \multicolumn{2}{c}{Batch Tendency} & N & R$^2$ \\
-\midrule
-Extremity Complaints & 0.662$^{***}$ & (0.079) & 5,259 & 0.126 \\
-Upper Respiratory Symptoms & 1.205$^{***}$ & (0.445) & 1,915 & 0.262 \\
-Abnormal Test Results & 0.670$^{**}$ & (0.325) & 1,816 & 0.141 \\
-Falls, Motor Vehicle Crashes, and Trauma & 0.640$^{**}$ & (0.264) & 2,381 & 0.220 \\
-Shortness of Breath & 1.574$^{***}$ & (0.468) &  2,963 & 0.147 \\
-Gastrointestinal Issues & 1.080$^{***}$ & (0.241) & 3,323 & 0.094 \\
-Chest Pain & 0.790$^{***}$ & (0.253) & 3,521 & 0.140 \\
-Dizziness/Lightheadedness/Syncope & 1.343$^{***}$ & (0.220) & 1,894 & 0.208 \\
-Back or Flank Pain & 0.859$^{**}$ & (0.385) &  2,550 & 0.123 \\
-Neurological Issue & 0.416$^{*}$ & (0.232) &  3,492 & 0.232 \\
-Fatigue and Weakness & 1.728$^{***}$ & (0.414) & 1,494 & 0.193 \\
-Cardiac Arrhythmias & 2.432$^{***}$ & (0.548) & 1,044 & 0.327 \\
-Abdominal Complaints & 1.775$^{***}$ & (0.350) & 6,231 & 0.086 \\
-Skin Complaints & 0.205 & (0.212) & 2,176 & 0.133 \\
-Fevers, Sweats or Chills & 1.650$^{***}$ & (0.576) &  1,842 & 0.211 \\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}[!htbp] \centering 
-  \caption{Regression Results by Complaint} 
-  \label{} 
-\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+  \caption{Testing the Monotonicity Assumption} 
+\label{tab:monotonicity}
+\begin{tabular}{@{\extracolsep{5.5pt}}lcc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
- & Dizziness/Syncope & Extremity Complaints & Upper Respiratory Symptoms \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
+Sub-sample margin & Baseline Tendency  & Reverse-Sample Tendency \\ 
 \hline \\[-1.8ex] 
- Testing Inclination & 1.156$^{***}$ & 0.901$^{***}$ & 0.516$^{***}$ \\ 
-  & (0.291) & (0.111) & (0.194) \\ 
-  & & & \\ 
- Batch & 0.655$^{*}$ & 0.479$^{**}$ & 0.501$^{*}$ \\ 
-  & (0.368) & (0.242) & (0.283) \\ 
-  & & & \\ 
-\textit{N} & 1,895 & 5,265 & 1,917 \\ 
-R$^{2}$ & 0.400 & 0.348 & 0.461 \\ 
-Adjusted R$^{2}$ & 0.338 & 0.324 & 0.404 \\ 
-Residual Std. Error & 0.801 (df = 1715) & 0.765 (df = 5079) & 0.741 (df = 1733) \\ 
-\hline 
+Extremity Complaints & 0.864$^{***}$ (0.128) & 0.893$^{***}$ (0.068) \\ 
+Upper Respiratory Symptoms & 1.235$^{***}$ (0.247) & 0.953$^{***}$ (0.038) \\ 
+Abnormal Test Results & 0.980$^{***}$ (0.219) & 0.766$^{***}$ (0.111) \\
+Falls, Motor Vehicle Crashes, Assaults, and Trauma & 1.061$^{***}$ (0.167) & 0.784$^{***}$ (0.103) \\
+Shortness of Breath & 1.204$^{***}$ (0.232) & 0.944$^{***}$ (0.047) \\
+Gastrointestinal Issues & 1.049$^{***}$ (0.140) & 0.929$^{***}$ (0.052) \\
+Chest Pain & 1.073$^{***}$ (0.223) & 0.943$^{***}$ (0.048) \\
+Dizziness/Lightheadedness/Syncope & 1.899$^{***}$ (0.171) & 0.930$^{***}$ (0.042) \\
+Back or Flank Pain & 0.657$^{***}$ (0.180) & 0.718$^{***}$ (0.106) \\
+Neurological Issue & 0.970$^{***}$ (0.191) & 0.835$^{***}$ (0.067) \\
+Fatigue and Weakness & 1.698$^{***}$ (0.230) & 0.938$^{***}$ (0.058) \\
+Cardiac Arrhythmias & 2.571$^{***}$ (0.243) & 1.107$^{***}$ (0.059) \\
+Abdominal Complaints & 0.903$^{***}$ (0.275) & 0.971$^{***}$ (0.019) \\
+Skin Complaints & 0.448$^{***}$ (0.098) & 0.584$^{***}$ (0.108) \\
+Fevers, Sweats or Chills & 1.033$^{***}$ (0.305) & 0.888$^{***}$ (0.084) \\
 \hline \\[-1.8ex] 
-\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
- & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
- & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
 \end{tabular} 
+\begin{tablenotes}
+\small
+\item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample.
+Column 2 constructs a new physician batch tendency instrument using all emergency visits, excluding the corresponding sub-sample (``reverse-sample"), and displays the coefficient of the first stage regression back on that sub-sample.
+Robust standard errors are clustered at the physician level.}
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
 \end{table}
 
-\begin{table}[!htbp] \centering 
-  \caption{Regression Results by Complaint} 
-  \label{} 
-\begin{tabular}{@{\extracolsep{5pt}}lccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & \multicolumn{3}{c}{nEDTests} \\ 
- & Abdominal Complaints & Cardiac Arrhythmias & Fatigue and Weakness \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
-\hline \\[-1.8ex] 
- Testing Inclination & 0.913$^{***}$ & 0.695$^{***}$ & 0.870$^{***}$ \\ 
-  & (0.068) & (0.180) & (0.195) \\ 
-  & & & \\ 
- Batch & $-$0.135$^{**}$ & 0.708$^{***}$ & 0.333$^{*}$ \\ 
-  & (0.066) & (0.129) & (0.189) \\ 
-  & & & \\ 
-\textit{N} & 6,232 & 1,045 & 1,494 \\ 
-R$^{2}$ & 0.038 & 0.525 & 0.294 \\ 
-Adjusted R$^{2}$ & 0.009 & 0.428 & 0.203 \\ 
-Residual Std. Error & 0.752 (df = 6046) & 0.595 (df = 866) & 0.828 (df = 1322) \\ 
-\hline 
-\hline \\[-1.8ex] 
-\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
- & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
- & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
-\end{tabular} 
-\end{table}
+\begin{longtable}{|p{5cm}|p{12cm}|}
+\caption{Condensed Complaints} \\
+\hline
+\textbf{Condensed Complaint Area} & \textbf{Complaints} \\
+\hline
+Gastrointestinal/Abdominal Issues & Abdominal Cramping, Abdominal Distention, Dyspepsia, Abdominal Pain, Ascites, Hernia, Abdominal Aortic Aneurysm, Abdominal Injury, Pancreatitis, Umbilical Hernia, Anal Fissure, Black or Bloody Stool, Constipation, GERD, Anal Fistula, Diarrhea, Dysphagia, Fecal Impaction, Fistula Follow Up, GIbleeding, GI Problem, Hemorrhoids, Morning Sickness, Nausea, Ostomy Care, Rectal Bleeding, Rectal Pain, Vomiting, Vomiting Blood, Vomiting During Pregnancy, GI Bleeding, Fecal Incontinence, Bloated, Hematochezia, Urine Leakage, Heartburn, Rectal Discharge, Urolithiasis, Ulcerative Colitis, Irritable Bowel Syndrome, Rectal Prolapse, Fistula Evaluation, Rectal Problems, Perianal Abscess, Fisula Evaluation, Stoma Dysfunction \\
+\hline
+Cardiac/Chest-Related Issues & Atrial Fibrillation, Atrial Flutter, Cardiac Valve Problem, Bradycardia, Irregular Heart Beat, Palpitations, POTS, Ventricular Tachycardia, Rapid Heart Rate, Heart Problem, Cardiac Arrest, Congestive Heart Failure, Circulatory Problem, Transient Ischemic Attack, Ventricular Tachycardia, Chest Injury, Chest Pain, Chest Wall Pain, Angina, Collarbone Injury, Rib Injury, Heart Pain \\
+\hline
+Musculoskeletal/Extremity Issues & Ankle Injury, Ankle Pain, Arm Injury, Arm Pain, Cold Extremity, Arm Swelling, Arthritis, Elbow Injury, Elbow Pain, Pseudogout, Extremity Pain, Extremity Weakness, Finger Injury, Hip Injury, Extremity Weakness, Finger Injury, Finger Pain, Dislocation, Foot Infection, Foot Injury, Foot Numbness, Foot Pain, Foot Swelling, Foot Ulcer, Foot Wound Check, Hand Injury, Hand Pain, Back Pain, Back Problem, Flank Pain, Sciatica, Back Injury, Disc Disorder, Assault Victim, Concussion, Facial Injury, Fall, Nasal Trauma, Head Injury, Head Laceration, Motor Vehicle Crash, Puncture Wound, Sexual Assault, Trauma, Domestic Violence, Gun Shot Wound, Work Related Injury, Motorcycle Crash, Injury, Bicycle Accident, Near Drowning, Lip Laceration \\
+\hline
+Neurological/Syncope Issues & Altered Mental Status, Cognitive Concerns, Facial Droop, Pre Syncope, Focal Weakness, Headache, Memory Loss, Migraine, Dementia, Dysphasia, Neuro Problem, Numbness, Paralysis, Seizures, Slurred Speech, Spasms, Stroke Like Symptoms, Tingling, Tremors, Trigeminal Neuralgia, Unable to Speak, Seizure Disorder, Insomnia, Parkinson's Disease, Loss of Consciousness, Neuropathy, Ataxia, Unable to speak, Peripheral Neuropathy, Stroke, Cerebrovascular Accident, Speech Problem, Acute Neurological Problem, Flashes, Light, Unresponsive, Multiple Sclerosis, Parkinson's Disease, Febrile Seizure, Paresthesia, Peripheral Neuropathy, Hydrocephalus, Spasticity, Neuroendocrine Tumor, Dizziness, Near Syncope, Syncope, Vertigo, Spells, Hypotension, Paroxysmal Positional Vertigo, Paroxysmal Positional Vertigo \\
+\hline
+Respiratory-Related Issues & Airway Obstruction, Aspiration, Pain With Breathing, Near Drowning, Respiratory Distress, Shortness of Breath, Wheezing, Increased Work Of Breathing, Difficulty Breathing, Choking, Oxygen Dependence, Hyperventilating, Orthopnea, Congestion, Cough, Coughing Up Blood, Flu Symptoms, Enlarged Tonsils, Peritonsillar Abscess, Nasal Congestion, Sinus Symptoms, Sinusitis, Sore Throat, Hoarseness, Throat Problem, Upper Respiratory Infection, Influenza, Laryngitis, Respiratory Arrest, Pneumonia, Pleural Effusion, Asthma, Croup, URI, Peritonsillar Abscess\\
+\hline
+\end{longtable}
 

### 2024-01-02 src/IV Validation.R

@@ -324,8 +324,8 @@ for(complaint in complaints) {
     filter(CHIEF_COMPLAINT == complaint)
   
   model <- felm(
-    any.batch ~ batch.tendency + test.inclination | 
-                dayofweekt + month_of_year + age_groups + ESI | 0 | ED_PROVIDER,
+    any.batch ~ batch.tendency | 
+                dayofweekt + month_of_year + ESI | 0 | ED_PROVIDER,
     data = data_subset
   )
   model_results[[complaint]] <- model
@@ -333,6 +333,9 @@ for(complaint in complaints) {
 
 # Save the results to a .txt file
 sink("outputs/tables/Monotonicity.txt")
+
+print("first stage should be weakly positive for 
+      all subsamples (Dobbie, Goldin, and Yang 2018)")
 
 stargazer(model_results[1:5], type = "text", 
           title = "Regression Results by Complaint",
@@ -350,26 +353,50 @@ stargazer(model_results[10:15], type = "text",
           style = 'QJE')
 
 
-sink()
+model_results <- list()
 
-#=========================================================================
-
-data.temp <- data
-
-data.temp$residual_batch <- resid(
-  felm(any.batch ~ 0 | dayofweekt + month_of_year + complaint_esi, data=data.temp))
-
-# Step 2: get batch tendency for each provider
-table(data.temp$CHIEF_COMPLAINT, data.temp$any.batch)
-
-data.temp <- data.temp %>%
-  group_by(ED_PROVIDER) %>%
-  mutate(Sum_Resid=sum(residual_batch, na.rm=T),
-         batch.tendency = (Sum_Resid - residual_batch) / (n() - 1)) %>%
+for(complaint in complaints) {
   
-  ungroup()
+  data_subset <- data %>% 
+    filter(CHIEF_COMPLAINT == complaint)
+  
+  data_subset$residual_batch <- resid(
+    felm(any.batch ~ 0 | dayofweekt + month_of_year, data=data_subset))
+  
+  data_subset <- data_subset %>%
+    group_by(ED_PROVIDER) %>%
+    mutate(Sum_Resid=sum(residual_batch, na.rm=T),
+           batch.tendency = (Sum_Resid - residual_batch) / (n() - 1))
+  
+  model <- felm(
+    any.batch ~ batch.tendency | 
+      dayofweekt + month_of_year + ESI | 0 | ED_PROVIDER,
+    data = data_subset
+  )
+  model_results[[complaint]] <- model
+}
+
+print("instrument constructed by leaving out a particular subsample
[…]

### 2024-01-02 src/clean.R

@@ -12,8 +12,7 @@ library(lubridate)
 #=========================================================================
 # Determine test times
 #=========================================================================
-
-path <- '~/Dropbox/Batch vs sequential testing/Data/'
+path <- '~/Sue Goldie Dropbox/Jacob Jameson/Batch vs sequential testing/Data/'
 
 df <- read.csv(paste0(path, 'deidentified_FINAL.csv'))
 
@@ -436,6 +435,7 @@ complaints <- list(
       'Voiding Dysfunction',"Hesitancy Urinary")
 )
 
+df$complaint <- df$CHIEF_COMPLAINT
 
 for (i in seq(1,length(complaints))){
   name <- names(complaints[i])

### 2024-01-25 manuscript/manuscript.Rmd
manuscript updates

@@
 ---
-title: "To Batch or Not to Batch: Test-Ordering Variability in the Emergency Department and the Impact on Care Delivery"
+title: "To Batch or Not to Batch: Test-Ordering Variabilityin the ED"
 authors:
@@
 One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests @naseim2015.
-A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests such as electrocardiograms (ECGs) and point-of-care (POC) testing.
+A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests.
 These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.
@@
 We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences.
-Being treated by a provider in the top decile of the tendency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].
+Being treated by a provider in the top decile of the tendency distribution, compared to being treated by someone in the bottom decile results in an additional 2 diagnostic test per 100 patient encounters after controlling for the physician's underlying propensity to test.
+
+Physician care takes on multiple dimensions besides the decision to batch order diagnostic tests.
+Employing a placebo exercise, we find evidence that that excess testing occurs after having seen a high-batching physician is due to batching from this physician, as opposed to other differences across physician in care that correlate with batching tendency: we find no effects of physician batch tendency on key outcomes for a ``placebo sample" of patients who visit the ED for conditions that rarely result in batched orders.
+
+With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly.
+We find that being treated by a top decile batching physician, compared to a bottom decile batching physician, results in an additional 13.5 tests per 100 patient encounters.
+We find no evidence that batch-ordering reduces ED length of stay or 72-hour return rates.
+These results suggest that the benefits of batch-ordering are limited, and that the practice may be associated with significant costs.
+Our findings are robust to analyses probing the key assumptions of our batching definition.
 
@@
 
-Our analytical lens is focused on the Emergency Department at the Mayo Clinic of Arizona, a distinguished tertiary care establishment.
-During the timeframe of our study, the ED recorded an annual visitation of approximately 45,000 patients.
-The department is singularly staffed by board-eligible or board-certified emergency physicians, abstaining from the services of nurse practitioners or physician assistants.
-A notable observation was that residents in rotation oversaw a low fraction, roughly 10%, of the patient volume.
-Comprehensive patient data, encompassing demographics, chief complaints, vital signs, emergency severity, length of stay, and resource utilization metrics, were meticulously logged during the study period.
+Our study was conducted in the Emergency Department (ED) of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation.
+During the study period, the ED recorded approximately 43,000 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces.
+The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume.
+Physicians operate in a unique workflow that includes staggered 8.5-hour shifts and a rotational patient assignment system, minimizing potential selection bias in patient encounters.
+
+We conducted a retrospective review of comprehensive ED operational data from 10/6/2018 through 12/31/2019, coinciding with the initiation of a new electronic medical record.
+The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics.
+This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.
 
@@
 Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
-We observe approximately $48,000$ such visits until during the study period.
+We observe approximately $48,000$ such visits during the study period.
 To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $10$ percent of the time.
@@
 Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter.
-While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice, where we observe that approximately 97(?) percent of all tests ordered are conducted.
+While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice.
 Below, we detail our primary outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.
@@
 It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
-Although the literature provides a range for long ED-LOS, we operationalize long ED-LOS as stays exceeding the 75th percentile in our dataset for this study.
-This approach allows us to contextualize the definition of extended stays according to the specific demographics and case mix of the ED under study.
+The hypothesis is that batch testing may lead to shorter ED-LOS, potentially improving patient throughput and reducing crowding.
 
   *(b) Resource Utilization*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives.
-In this study, we quantify resource utilization by the total number of diagnostic tests ordered per patient during their ED stay.
+In this study, we quantify resource utilization by the total number of distinct diagnostic tests ordered per patient during their ED stay.
 This encompasses both initial and subsequent tests.
@@
 We measure this outcome as the proportion of patients with a subsequent ED visit within 72 hours post-initial discharge.
+The hypothesis is that batch testing may lead to variations in the 72-hour return rate, potentially influencing the quality of care received.
 
@@
 
-In the context of our study, batching in diagnostic test ordering is defined as placing multiple test orders for a single patient within a 5-minute interval.
-To ensure the robustness of our analysis, sensitivity tests were conducted on this cutoff point, affirming that our results are consistent and reliable, irrespective of minor variations in the batching definition.
+We define ``batching" in line with standard emergency medicine practices.
+It involves a physician ordering a comprehensive set of diagnostic tests at once, typically covering a broad range of potential diagnoses.
+This contrasts with non-batching, where tests are ordered more selectively based on the information available at the time, with additional tests potentially ordered later as needed.
 
-Our study identifies and differentiates between two distinct strategies of batching:
+We operationalize batching as the ordering of multiple diagnostic tests within a 5-minute window.
+Sensitivity analyses on this cutoff point showed that our results are robust to this definition.
+This operational definition aligns with the concept of batching as a single comprehensive effort.
+For our analysis, all lab-based testing ordered in this timeframe are categorized as one distinct type of testing (lab).
+In contrast, each imaging test (e.g., X-ray, CT scan) are considered a separate distinct test.
+Therefore, a batch in our study is defined as consisting of two or more distinct diagnostic tests, which could be a combination of lab and imaging tests or multiple imaging tests.
 
-* Lab + Image Batch: This category of batch order includes instances where a patient’s batched diagnostic tests comprise a combination of at least one imaging test (such as a contrasted CT scan, non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
-This mix is fascinating as it suggests a comprehensive diagnostic approach, possibly catering to complex medical cases.
-
-* Image + Image Batch: This category encompasses batch orders where a patient’s diagnostic tests consist of two or more distinct imaging tests.
-The reliance solely on imaging in these batches may indicate specific diagnostic pathways or a focus on conditions that primarily require visual diagnostic methods.
-
-It is important to note the differing implications of these batching strategies.
-While lab tests often can be processed concurrently with imaging tests, causing minimal additional waiting time, a batch of multiple imaging tests requires the patient’s presence for each test, potentially leading to significant increases in waiting time.
-Thus, the composition of a batch order can provide valuable insights into the physician’s decision-making process, particularly in balancing between patient waiting times and diagnostic uncertainty.
-This balance can be especially crucial in the context of varying complexity levels of patient encounters.
-
-Figure 1 illustrates the occurrence rates of each batching type across the top ten most frequently reported chief complaints, providing a visual representation of how batching practices vary with different patient presentations.
-
-\begin{figure}[h]
-  \centering
-  \caption{Batch Frequency by Chief Complaint}
-  \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
-\begin{tablenotes}
-\tiny
-\item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
-\end{tablenotes}  
-\end{figure}
+This approach reflects the realistic diagnostic strategies in an ED setting and allows for a clear distinction between batching and non-batching behaviors.
+The differentiation between lab tests as one collective unit and imaging tests as individual units is based on their operational impact in the ED, particularly regarding patient waiting time and processing queues.
 
@@
     \label{tab:wald_test}
-    \begin{tabular}{p{11cm}cc}
+    \begin{tabular}{p{10cm}cccc}
         \toprule
-        Chief Complaints & F-Statistic & $Pr(> F)$ \\
+        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
         \midrule
-        Abdominal Complaints & 1.37 & 0.106 \\
-        Back or Flank Pain & 1.00 & 0.451 \\
-        Chest Pain & 0.98 & 0.476 \\
-        Extremity Complaints & 0.97 & 0.495 \\
-        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
-        Gastrointestinal Issues & 0.98 & 0.480 \\
-        Neurological Issue & 0.75 & 0.793 \\
-        Shortness of Breath & 1.23 & 0.199 \\
-        Skin Complaints & 1.05 & 0.388 \\
-        Upper Respiratory Symptoms & 1.21 & 0.218 \\
+        Abdominal Complaints & 6232 & 2.587 & 0.108 \\ 
+        Back or Flank Pain & 2552 & 1.637 & 0.201 \\ 
+        Chest Pain & 3525 & 0.407 & 0.524 \\ 
+        Extremity Complaints & 5265 & 1.847 & 0.174 \\ 
+        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2381 & 0.023 & 0.880 \\ 
+        Gastrointestinal Issues & 3323 & 0.105 & 0.746 \\ 
+        Neurological Issue & 3495 & 0.135 & 0.713 \\ 
+        Shortness of Breath & 2966 & 1.324 & 0.250 \\ 
+        Skin Complaints & 2178 & 0.383 & 0.536 \\ 
+        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
         \midrule
-        Emergency Severity & F-Statistic & $Pr(> F)$ \\
+        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
         \midrule
-        ESI 1 or 2 & 1.09 & 0.346 \\
-        ESI 3, 4, or 5 & 1.24 & 0.196 \\
+        ESI 1 or 2 & 13914 & 0.011 & 0.915 \\ 
+        ESI 3, 4, or 5 & 29386 & 0.010 & 0.921 \\ 
         \bottomrule
@@
 
-In step two, the leniency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:
+In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:
 
@@
 The F-statistic is 50 when all controls and fixed effects are included.
-The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
+The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
 
@@
 Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination.
-
 Assignment to a physician in the top batching decile (relative to one in the bottom decile) is associated with a 0.2 unit increase in the number of diagnostic tests being ordered per patient encounter, without any additional change in the likelihood of a 72-hour return.
-With an average of 1,805 patient encounters per physician in our study period, this translates to an additional 361 diagnostic tests with no perceived benefit in terms of reducing returns or length of stay.
 
 The effects of testing inclination are also striking.
-We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 0.96 unit increase in the number of diagnostic tests being ordered per patient encounter, and -0.03 pp change in the probability of a patient returning within 72 hours.
+We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 9.6 excess diagnostic tests being ordered per 100 patient encounters, and -0.03 pp change in the probability of a patient returning within 72 hours.
 
@@
 To that end, we restrict attention to ED visits for complaints where batching occurs no more than 10 percent of the time (recall that our baseline sample only includes complaints with a >10 percent batching rate).
-We estimate a reduced-form regression of each main outcome on physician prescribing leniency for the subsample, following equation (3).
+We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3).
 The results of this exercise are displayed in Table \ref{tab:placebo_check}.
@@
 
-## Investigating Mechanisms: Excess Testing
+## Robustness
 
-In the previous section we find a particularly pronounced impact of batching on resource utilization.
-We reported a significant coefficient of 0.135 in the number of tests ordered due to batching, translating to 13.5 excess tests per 100 batch orders.
-Here, an ``excess test" is defined as one that would not have been ordered if the physician had chosen to order tests sequentially.
-This suggests that the information gain from sequential testing, where the need for additional tests is informed by the results of earlier tests, could lead to more efficient and targeted use of diagnostic resources.
+The 2SLS results presented in the previous sections are robust to several alternative specifications probing the key batching definition.
+Our primary definition operationalizes batching as the ordering of multiple diagnostic tests within a 5-minute window.
+This approach is grounded in standard emergency medicine practices and offers a clear distinction between batching and non-batching behaviors.
+To ensure the robustness of our findings, we performed several sensitivity analyses.
 
-To further understand this phenomenon, we probe into high-volume complaint areas in the ED.
-By focusing on these areas, we can discern which specific complaints are more likely to lead to excess testing under a batch ordering system.
-This analysis is crucial as it sheds light on the causal effect of batching in different clinical scenarios.
+(1) Variation in the Time Window for Batching: Firstly, we varied the time window for what constitutes a batch.
+The original 5-minute window was extended and contracted to understand its impact on the study's outcomes.
+Specifically, we considered scenarios where a batch is defined as two or more tests ordered within windows of 1 minute to 10 minutes.
+This range allowed us to capture a broader spectrum of batching behaviors and test the sensitivity of our results to different batching definitions.
 
-We consider the top 5 most common broad complaints in the ED, which account for 82% of all visits.
-Appendix table [...] .
-These include Cardiac/Chest-Related Issues, Gastrointestinal/Abdominal Issues, Respiratory-Related Issues, Neurological/Syncope Issues, and Musculoskeletal/Extremity Issues.
-We then examine the impact of batching on the number of tests ordered for each of these complaints.
+(2) Refinement of Batching Definition: Secondly, we refined our batching definition based on the sequence and timing of test orders.
+In the initial analysis, if two tests were ordered upfront and another test ordered later, it was counted as a batch.
+We modified this definition to consider a set of tests as a batch only if all tests ordered during a patient encounter came from that initial batch order.
+This adjustment ensures that our batching definition more accurately reflects a comprehensive diagnostic effort at the outset of patient care, rather than incremental decision-making.
 
-[...]
+The key finding from these robustness checks is the consistent impact of batching on key outcomes across all variations.
+As seen in Table \ref{tab:robustness_batching} and \ref{tab:robustness_batching2}, altering the time window for batching, whether narrowing it to as little as 1 minute or expanding it to 10 minutes, did not qualitatively change the results.
+Similarly, refining the definition of batching to consider only those tests ordered in the initial batch also had negligible impact on the study's main findings.
+These results lend further credibility to our initial findings, demonstrating that our conclusions about the impact of batching on patient outcomes in the ED are not sensitive to the specific operational definition of batching.
 
@@
 \begin{tablenotes}
-\small
+\tiny
 \item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample.
@@
 
-\begin{longtable}{|p{5cm}|p{12cm}|}
-\caption{Condensed Complaints} \\
+\newpage
+
+\begin{table}[!htbp] \centering 
+  \caption{2SLS Results: Robustness to Batching Definition (1-minute and 10-minute windows)} 
+  \label{tab:robustness_batching}
+
+\begin{tabular}{@{\extracolsep{10pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+ & \multicolumn{3}{c}{\textbf{1-minute window}} \\
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
+  & (0.203) & (0.018) & (0.009) \\ 
+  & & & \\ 
+ Batch & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
+  & (0.496) & (0.054) & (0.023) \\ 
+  & & & \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
+Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
+Residual Std. Error & 0.492 & 0.836 & 0.184 \\ 
 \hline 
-\textbf{Condensed Complaint Area} & \textbf{Complaints} \\
+\\[-1.8ex] & \multicolumn{3}{c}{\textbf{10-minute window}} \\
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.650$^{***}$ & 0.957$^{***}$ & $-$0.027$^{**}$ \\ 
+  & (0.241) & (0.022) & (0.011) \\ 
+  & & & \\ 
+ Batch & $-$0.205 & 0.141$^{***}$ & 0.017 \\ 
+  & (0.438) & (0.045) & (0.018) \\ 
+  & & & \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.146 & 0.315 & 0.009 \\ 
+Adjusted R$^{2}$ & 0.141 & 0.311 & 0.003 \\ 
+Residual Std. Error & 0.512 & 0.814 & 0.184 \\ 
 \hline 
-Gastrointestinal/Abdominal Issues & Abdominal Cramping, Abdominal Distention, Dyspepsia, Abdominal Pain, Ascites, Hernia, Abdominal Aortic Aneurysm, Abdominal Injury, Pancreatitis, Umbilical Hernia, Anal Fissure, Black or Bloody Stool, Constipation, GERD, Anal Fistula, Diarrhea, Dysphagia, Fecal Impaction, Fistula Follow Up, GIbleeding, GI Problem, Hemorrhoids, Morning Sickness, Nausea, Ostomy Care, Rectal Bleeding, Rectal Pain, Vomiting, Vomiting Blood, Vomiting During Pregnancy, GI Bleeding, Fecal Incontinence, Bloated, Hematochezia, Urine Leakage, Heartburn, Rectal Discharge, Urolithiasis, Ulcerative Colitis, Irritable Bowel Syndrome, Rectal Prolapse, Fistula Evaluation, Rectal Problems, Perianal Abscess, Fisula Evaluation, Stoma Dysfunction \\
+\hline \\[-1.8ex] 
+\end{tabular}
+\begin{tablenotes}
+\tiny
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
+
+\newpage
+
+\begin{table}[!htbp] \centering 
+  \caption{2SLS Results: Robustness to Batching Definition (all tests at once)} 
+  \label{tab:robustness_batching2}
+
+\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Batch & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
+  & (0.203) & (0.018) & (0.009) \\ 
+  & & & \\ 
+ `any.batch(fit)` & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
+  & (0.496) & (0.054) & (0.023) \\ 
+  & & & \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
+Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
+Residual Std. Error (df = 41667) & 0.492 & 0.836 & 0.184 \\ 
 \hline 
-Cardiac/Chest-Related Issues & Atrial Fibrillation, Atrial Flutter, Cardiac Valve Problem, Bradycardia, Irregular Heart Beat, Palpitations, POTS, Ventricular Tachycardia, Rapid Heart Rate, Heart Problem, Cardiac Arrest, Congestive Heart Failure, Circulatory Problem, Transient Ischemic Attack, Ventricular Tachycardia, Chest Injury, Chest Pain, Chest Wall Pain, Angina, Collarbone Injury, Rib Injury, Heart Pain \\
-\hline
-Musculoskeletal/Extremity Issues & Ankle Injury, Ankle Pain, Arm Injury, Arm Pain, Cold Extremity, Arm Swelling, Arthritis, Elbow Injury, Elbow Pain, Pseudogout, Extremity Pain, Extremity Weakness, Finger Injury, Hip Injury, Extremity Weakness, Finger Injury, Finger Pain, Dislocation, Foot Infection, Foot Injury, Foot Numbness, Foot Pain, Foot Swelling, Foot Ulcer, Foot Wound Check, Hand Injury, Hand Pain, Back Pain, Back Problem, Flank Pain, Sciatica, Back Injury, Disc Disorder, Assault Victim, Concussion, Facial Injury, Fall, Nasal Trauma, Head Injury, Head Laceration, Motor Vehicle Crash, Puncture Wound, Sexual Assault, Trauma, Domestic Violence, Gun Shot Wound, Work Related Injury, Motorcycle Crash, Injury, Bicycle Accident, Near Drowning, Lip Laceration \\
-\hline
-Neurological/Syncope Issues & Altered Mental Status, Cognitive Concerns, Facial Droop, Pre Syncope, Focal Weakness, Headache, Memory Loss, Migraine, Dementia, Dysphasia, Neuro Problem, Numbness, Paralysis, Seizures, Slurred Speech, Spasms, Stroke Like Symptoms, Tingling, Tremors, Trigeminal Neuralgia, Unable to Speak, Seizure Disorder, Insomnia, Parkinson's Disease, Loss of Consciousness, Neuropathy, Ataxia, Unable to speak, Peripheral Neuropathy, Stroke, Cerebrovascular Accident, Speech Problem, Acute Neurological Problem, Flashes, Light, Unresponsive, Multiple Sclerosis, Parkinson's Disease, Febrile Seizure, Paresthesia, Peripheral Neuropathy, Hydrocephalus, Spasticity, Neuroendocrine Tumor, Dizziness, Near Syncope, Syncope, Vertigo, Spells, Hypotension, Paroxysmal Positional Vertigo, Paroxysmal Positional Vertigo \\
-\hline
-Respiratory-Related Issues & Airway Obstruction, Aspiration, Pain With Breathing, Near Drowning, Respiratory Distress, Shortness of Breath, Wheezing, Increased Work Of Breathing, Difficulty Breathing, Choking, Oxygen Dependence, Hyperventilating, Orthopnea, Congestion, Cough, Coughing Up Blood, Flu Symptoms, Enlarged Tonsils, Peritonsillar Abscess, Nasal Congestion, Sinus Symptoms, Sinusitis, Sore Throat, Hoarseness, Throat Problem, Upper Respiratory Infection, Influenza, Laryngitis, Respiratory Arrest, Pneumonia, Pleural Effusion, Asthma, Croup, URI, Peritonsillar Abscess\\
-\hline
-\end{longtable}
+\hline \\[-1.8ex] 
+\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
+\end{tabular} 
+\end{table}
 
+\newpage
 :::

### 2024-01-25 manuscript/manuscript.tex
manuscript updates

@@
 
-\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the
-Emergency Department and the Impact on Care Delivery}
-
-\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the
-Emergency Department and the Impact on Care Delivery}
+\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variabilityin the ED}
+
+\TITLE{To Batch or Not to Batch: Test-Ordering Variabilityin the ED}
 
@@
 tests \citet{naseim2015}. A variety of diagnostic tests are used in the
-ED, including laboratory tests, imaging studies, and specialized tests
-such as electrocardiograms (ECGs) and point-of-care (POC) testing. These
-tests can provide valuable information about a patient's condition and
-help to guide treatment decisions.
+ED, including laboratory tests, imaging studies, and specialized tests.
+These tests can provide valuable information about a patient's condition
+and help to guide treatment decisions.
 
@@
 treated by a provider in the top decile of the tendency distribution,
-compared to being treated by someone in the bottom decile {[}INSERT
-RESULTS{]}.
+compared to being treated by someone in the bottom decile results in an
+additional 2 diagnostic test per 100 patient encounters after
+controlling for the physician's underlying propensity to test.
+
+Physician care takes on multiple dimensions besides the decision to
+batch order diagnostic tests. Employing a placebo exercise, we find
+evidence that that excess testing occurs after having seen a
+high-batching physician is due to batching from this physician, as
+opposed to other differences across physician in care that correlate
+with batching tendency: we find no effects of physician batch tendency
+on key outcomes for a ``placebo sample'' of patients who visit the ED
+for conditions that rarely result in batched orders.
+
+With the caveat that other unobserved dimensions of physician care may
+impact patient outcomes, we employ physician batch tendency as an
+instrumental variable for having tests batch ordered in the ED to
+quantify the effect of batching directly. We find that being treated by
+a top decile batching physician, compared to a bottom decile batching
+physician, results in an additional 13.5 tests per 100 patient
+encounters. We find no evidence that batch-ordering reduces ED length of
+stay or 72-hour return rates. These results suggest that the benefits of
+batch-ordering are limited, and that the practice may be associated with
+significant costs. Our findings are robust to analyses probing the key
+assumptions of our batching definition.
 
@@
 
-Our analytical lens is focused on the Emergency Department at the Mayo
-Clinic of Arizona, a distinguished tertiary care establishment. During
-the timeframe of our study, the ED recorded an annual visitation of
-approximately 45,000 patients. The department is singularly staffed by
-board-eligible or board-certified emergency physicians, abstaining from
-the services of nurse practitioners or physician assistants. A notable
-observation was that residents in rotation oversaw a low fraction,
-roughly 10\%, of the patient volume. Comprehensive patient data,
-encompassing demographics, chief complaints, vital signs, emergency
-severity, length of stay, and resource utilization metrics, were
-meticulously logged during the study period.
+Our study was conducted in the Emergency Department (ED) of the Mayo
+Clinic of Arizona, a tertiary care hospital without obstetrical
+services, an inpatient pediatrics unit, or a trauma designation. During
+the study period, the ED recorded approximately 43,000 visits per year,
+managed across 26 treatment rooms and up to 9 hallway spaces. The
+department is exclusively staffed by board-eligible or board-certified
+emergency physicians (EPs), with rotating residents overseeing about
+10\% of patient volume. Physicians operate in a unique workflow that
+includes staggered 8.5-hour shifts and a rotational patient assignment
+system, minimizing potential selection bias in patient encounters.
+
+We conducted a retrospective review of comprehensive ED operational data
+from 10/6/2018 through 12/31/2019, coinciding with the initiation of a
+new electronic medical record. The dataset includes detailed patient
+demographics, chief complaints, vital signs, emergency severity index
+(ESI), length of stay (LOS), and resource utilization metrics. This
+period was chosen to provide a robust data set while excluding the
+influence of the coronavirus pandemic.
 
@@
 Our research design focuses on adults who visit the Mayo Clinic of
-Arizona ED. We observe approximately \(48,000\) such visits until during
-the study period. To improve power, we drop encounters with rare chief
+Arizona ED. We observe approximately \(48,000\) such visits during the
+study period. To improve power, we drop encounters with rare chief
 complaints (\(<1000\) total encounters of this kind) and complaints
@@
 their ED encounter. While the patient could decide not to undergo the
-tests ordered by the physician, this is rare in practice, where we
-observe that approximately 97(?) percent of all tests ordered are
-conducted. Below, we detail our primary outcomes: (a) ED length of stay
-(LOS), (b) resource utilization, and (c) 72-hour return.
+tests ordered by the physician, this is rare in practice. Below, we
+detail our primary outcomes: (a) ED length of stay (LOS), (b) resource
+utilization, and (c) 72-hour return.
 
@@
 defined as the duration from a patient's arrival to the ED until their
-departure, whether by discharge or admission to the hospital. Although
-the literature provides a range for long ED-LOS, we operationalize long
-ED-LOS as stays exceeding the 75th percentile in our dataset for this
-study. This approach allows us to contextualize the definition of
-extended stays according to the specific demographics and case mix of
-the ED under study.
+departure, whether by discharge or admission to the hospital. The
+hypothesis is that batch testing may lead to shorter ED-LOS, potentially
+improving patient throughput and reducing crowding.
 
@@
 patient receives. In this study, we quantify resource utilization by the
-total number of diagnostic tests ordered per patient during their ED
-stay. This encompasses both initial and subsequent tests. The hypothesis
-is that batch testing may lead to variations in the number of tests
-ordered, potentially influencing the overall healthcare expenditure and
-efficiency.
+total number of distinct diagnostic tests ordered per patient during
+their ED stay. This encompasses both initial and subsequent tests. The
+hypothesis is that batch testing may lead to variations in the number of
+tests ordered, potentially influencing the overall healthcare
+expenditure and efficiency.
 
@@
 this outcome as the proportion of patients with a subsequent ED visit
-within 72 hours post-initial discharge.
+within 72 hours post-initial discharge. The hypothesis is that batch
+testing may lead to variations in the 72-hour return rate, potentially
+influencing the quality of care received.
 
@@
 
-In the context of our study, batching in diagnostic test ordering is
-defined as placing multiple test orders for a single patient within a
-5-minute interval. To ensure the robustness of our analysis, sensitivity
-tests were conducted on this cutoff point, affirming that our results
-are consistent and reliable, irrespective of minor variations in the
-batching definition.
-
-Our study identifies and differentiates between two distinct strategies
-of batching:
-
-\begin{itemize}
-\item
-  Lab + Image Batch: This category of batch order includes instances
-  where a patient's batched diagnostic tests comprise a combination of
-  at least one imaging test (such as a contrasted CT scan,
-  non-contrasted CT scan, X-ray, or ultrasound) and one laboratory test.
-  This mix is fascinating as it suggests a comprehensive diagnostic
-  approach, possibly catering to complex medical cases.
-\item
-  Image + Image Batch: This category encompasses batch orders where a
-  patient's diagnostic tests consist of two or more distinct imaging
-  tests. The reliance solely on imaging in these batches may indicate
-  specific diagnostic pathways or a focus on conditions that primarily
-  require visual diagnostic methods.
-\end{itemize}
-
-It is important to note the differing implications of these batching
-strategies. While lab tests often can be processed concurrently with
-imaging tests, causing minimal additional waiting time, a batch of
-multiple imaging tests requires the patient's presence for each test,
-potentially leading to significant increases in waiting time. Thus, the
-composition of a batch order can provide valuable insights into the
-physician's decision-making process, particularly in balancing between
-patient waiting times and diagnostic uncertainty. This balance can be
-especially crucial in the context of varying complexity levels of
-patient encounters.
-
-Figure 1 illustrates the occurrence rates of each batching type across
-the top ten most frequently reported chief complaints, providing a
-visual representation of how batching practices vary with different
-patient presentations.
-
-\begin{figure}[h]
-  \centering
-  \caption{Batch Frequency by Chief Complaint}
-  \includegraphics[width=1\textwidth]{../outputs/figures/Lab and Image Batch Rates.png}
-\begin{tablenotes}
-\tiny
-\item \textit{Figure 1 displays two statistics for the ten most common major chief complaint categories observed in our ED visit sample: the unadjusted batching rate for lab-image batching and image-image batching.}
-\end{tablenotes}  
-\end{figure}
+We define ``batching'' in line with standard emergency medicine
+practices. It involves a physician ordering a comprehensive set of
+diagnostic tests at once, typically covering a broad range of potential
+diagnoses. This contrasts with non-batching, where tests are ordered
+more selectively based on the information available at the time, with
+additional tests potentially ordered later as needed.
+
+We operationalize batching as the ordering of multiple diagnostic tests
+within a 5-minute window. Sensitivity analyses on this cutoff point
+showed that our results are robust to this definition. This operational
+definition aligns with the concept of batching as a single comprehensive
+effort. For our analysis, all lab-based testing ordered in this
+timeframe are categorized as one distinct type of testing (lab). In
+contrast, each imaging test (e.g., X-ray, CT scan) are considered a
+separate distinct test. Therefore, a batch in our study is defined as
+consisting of two or more distinct diagnostic tests, which could be a
+combination of lab and imaging tests or multiple imaging tests.
+
+This approach reflects the realistic diagnostic strategies in an ED
+setting and allows for a clear distinction between batching and
+non-batching behaviors. The differentiation between lab tests as one
+collective unit and imaging tests as individual units is based on their
+operational impact in the ED, particularly regarding patient waiting
+time and processing queues.
 
@@
     \label{tab:wald_test}
-    \begin{tabular}{p{11cm}cc}
+    \begin{tabular}{p{10cm}cccc}
         \toprule
-        Chief Complaints & F-Statistic & $Pr(> F)$ \\
-        \midrule
-        Abdominal Complaints & 1.37 & 0.106 \\
-        Back or Flank Pain & 1.00 & 0.451 \\
-        Chest Pain & 0.98 & 0.476 \\
-        Extremity Complaints & 0.97 & 0.495 \\
-        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.73 & 0.812 \\
-        Gastrointestinal Issues & 0.98 & 0.480 \\
-        Neurological Issue & 0.75 & 0.793 \\
-        Shortness of Breath & 1.23 & 0.199 \\
-        Skin Complaints & 1.05 & 0.388 \\
-        Upper Respiratory Symptoms & 1.21 & 0.218 \\
-        \midrule
-        Emergency Severity & F-Statistic & $Pr(> F)$ \\
-        \midrule
-        ESI 1 or 2 & 1.09 & 0.346 \\
-        ESI 3, 4, or 5 & 1.24 & 0.196 \\
+        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
+        \midrule
+        Abdominal Complaints & 6232 & 2.587 & 0.108 \\ 
+        Back or Flank Pain & 2552 & 1.637 & 0.201 \\ 
+        Chest Pain & 3525 & 0.407 & 0.524 \\ 
+        Extremity Complaints & 5265 & 1.847 & 0.174 \\ 
+        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2381 & 0.023 & 0.880 \\ 
+        Gastrointestinal Issues & 3323 & 0.105 & 0.746 \\ 
+        Neurological Issue & 3495 & 0.135 & 0.713 \\ 
+        Shortness of Breath & 2966 & 1.324 & 0.250 \\ 
+        Skin Complaints & 2178 & 0.383 & 0.536 \\ 
+        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
+        \midrule
+        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
+        \midrule
+        ESI 1 or 2 & 13914 & 0.011 & 0.915 \\ 
+        ESI 3, 4, or 5 & 29386 & 0.010 & 0.921 \\ 
         \bottomrule
@@
 
-In step two, the leniency measure for patient \(i\) seen by physician
+In step two, the tendency measure for patient \(i\) seen by physician
 \(j\) is computed as the average residual across all other patients seen
@@
 controls and fixed effects are included. The coefficient is greater than
-one because all emergency visits are used to construct the leniency
+one because all emergency visits are used to construct the tendency
 instrument, while the first stage is calculated using the baseline
@@
 ninetieth to tenth percentile in physician testing inclination.
-
 Assignment to a physician in the top batching decile (relative to one in
@@
 of diagnostic tests being ordered per patient encounter, without any
-additional change in the likelihood of a 72-hour return. With an average
-of 1,805 patient encounters per physician in our study period, this
-translates to an additional 361 diagnostic tests with no perceived
-benefit in terms of reducing returns or length of stay.
+additional change in the likelihood of a 72-hour return.
 
@@
 the bottom decile) is associated with 0.6\% change in the length of stay
-in the emergency department, a 0.96 unit increase in the number of
-diagnostic tests being ordered per patient encounter, and -0.03 pp
-change in the probability of a patient returning within 72 hours.
+in the emergency department, a 9.6 excess diagnostic tests being ordered
+per 100 patient encounters, and -0.03 pp change in the probability of a
+patient returning within 72 hours.
 
@@
 batching rate). We estimate a reduced-form regression of each main
-outcome on physician prescribing leniency for the subsample, following
+outcome on physician batch tendency for the subsample, following
 equation (3). The results of this exercise are displayed in Table
@@
 
-\hypertarget{investigating-mechanisms-excess-testing}{%
-\subsection{Investigating Mechanisms: Excess
-Testing}\label{investigating-mechanisms-excess-testing}}
-
-In the previous section we find a particularly pronounced impact of
-batching on resource utilization. We reported a significant coefficient
-of 0.135 in the number of tests ordered due to batching, translating to
-13.5 excess tests per 100 batch orders. Here, an ``excess test'' is
-defined as one that would not have been ordered if the physician had
-chosen to order tests sequentially. This suggests that the information
-gain from sequential testing, where the need for additional tests is
-informed by the results of earlier tests, could lead to more efficient
-and targeted use of diagnostic resources.
-
-To further understand this phenomenon, we probe into high-volume
-complaint areas in the ED. By focusing on these areas, we can discern
-which specific complaints are more likely to lead to excess testing
-under a batch ordering system. This analysis is crucial as it sheds
-light on the causal effect of batching in different clinical scenarios.
-
-We consider the top 5 most common broad complaints in the ED, which
-account for 82\% of all visits. Appendix table {[}\ldots{]} . These
-include Cardiac/Chest-Related Issues, Gastrointestinal/Abdominal Issues,
-Respiratory-Related Issues, Neurological/Syncope Issues, and
-Musculoskeletal/Extremity Issues. We then examine the impact of batching
-on the number of tests ordered for each of these complaints.
-
-{[}\ldots{]}
+\hypertarget{robustness}{%
+\subsection{Robustness}\label{robustness}}
+
+The 2SLS results presented in the previous sections are robust to
+several alternative specifications probing the key batching definition.
+Our primary definition operationalizes batching as the ordering of
+multiple diagnostic tests within a 5-minute window. This approach is
+grounded in standard emergency medicine practices and offers a clear
+distinction between batching and non-batching behaviors. To ensure the
+robustness of our findings, we performed several sensitivity analyses.
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\item
+  Variation in the Time Window for Batching: Firstly, we varied the time
+  window for what constitutes a batch. The original 5-minute window was
+  extended and contracted to understand its impact on the study's
+  outcomes. Specifically, we considered scenarios where a batch is
+  defined as two or more tests ordered within windows of 1 minute to 10
+  minutes. This range allowed us to capture a broader spectrum of
+  batching behaviors and test the sensitivity of our results to
+  different batching definitions.
+\item
+  Refinement of Batching Definition: Secondly, we refined our batching
+  definition based on the sequence and timing of test orders. In the
+  initial analysis, if two tests were ordered upfront and another test
+  ordered later, it was counted as a batch. We modified this definition
+  to consider a set of tests as a batch only if all tests ordered during
+  a patient encounter came from that initial batch order. This
+  adjustment ensures that our batching definition more accurately
+  reflects a comprehensive diagnostic effort at the outset of patient
+  care, rather than incremental decision-making.
+\end{enumerate}
+
+The key finding from these robustness checks is the consistent impact of
+batching on key outcomes across all variations. As seen in Table
+\ref{tab:robustness_batching} and \ref{tab:robustness_batching2},
+altering the time window for batching, whether narrowing it to as little
+as 1 minute or expanding it to 10 minutes, did not qualitatively change
+the results. Similarly, refining the definition of batching to consider
+only those tests ordered in the initial batch also had negligible impact
+on the study's main findings. These results lend further credibility to
+our initial findings, demonstrating that our conclusions about the
+impact of batching on patient outcomes in the ED are not sensitive to
+the specific operational definition of batching.
 
@@
 
-\begin{longtable}{|p{5cm}|p{12cm}|}
-\caption{Condensed Complaints} \\
-\hline
-\textbf{Condensed Complaint Area} & \textbf{Complaints} \\
-\hline
-Gastrointestinal/Abdominal Issues & Abdominal Cramping, Abdominal Distention, Dyspepsia, Abdominal Pain, Ascites, Hernia, Abdominal Aortic Aneurysm, Abdominal Injury, Pancreatitis, Umbilical Hernia, Anal Fissure, Black or Bloody Stool, Constipation, GERD, Anal Fistula, Diarrhea, Dysphagia, Fecal Impaction, Fistula Follow Up, GIbleeding, GI Problem, Hemorrhoids, Morning Sickness, Nausea, Ostomy Care, Rectal Bleeding, Rectal Pain, Vomiting, Vomiting Blood, Vomiting During Pregnancy, GI Bleeding, Fecal Incontinence, Bloated, Hematochezia, Urine Leakage, Heartburn, Rectal Discharge, Urolithiasis, Ulcerative Colitis, Irritable Bowel Syndrome, Rectal Prolapse, Fistula Evaluation, Rectal Problems, Perianal Abscess, Fisula Evaluation, Stoma Dysfunction \\
-\hline
-Cardiac/Chest-Related Issues & Atrial Fibrillation, Atrial Flutter, Cardiac Valve Problem, Bradycardia, Irregular Heart Beat, Palpitations, POTS, Ventricular Tachycardia, Rapid Heart Rate, Heart Problem, Cardiac Arrest, Congestive Heart Failure, Circulatory Problem, Transient Ischemic Attack, Ventricular Tachycardia, Chest Injury, Chest Pain, Chest Wall Pain, Angina, Collarbone Injury, Rib Injury, Heart Pain \\
-\hline
-Musculoskeletal/Extremity Issues & Ankle Injury, Ankle Pain, Arm Injury, Arm Pain, Cold Extremity, Arm Swelling, Arthritis, Elbow Injury, Elbow Pain, Pseudogout, Extremity Pain, Extremity Weakness, Finger Injury, Hip Injury, Extremity Weakness, Finger Injury, Finger Pain, Dislocation, Foot Infection, Foot Injury, Foot Numbness, Foot Pain, Foot Swelling, Foot Ulcer, Foot Wound Check, Hand Injury, Hand Pain, Back Pain, Back Problem, Flank Pain, Sciatica, Back Injury, Disc Disorder, Assault Victim, Concussion, Facial Injury, Fall, Nasal Trauma, Head Injury, Head Laceration, Motor Vehicle Crash, Puncture Wound, Sexual Assault, Trauma, Domestic Violence, Gun Shot Wound, Work Related Injury, Motorcycle Crash, Injury, Bicycle Accident, Near Drowning, Lip Laceration \\
-\hline
-Neurological/Syncope Issues & Altered Mental Status, Cognitive Concerns, Facial Droop, Pre Syncope, Focal Weakness, Headache, Memory Loss, Migraine, Dementia, Dysphasia, Neuro Problem, Numbness, Paralysis, Seizures, Slurred Speech, Spasms, Stroke Like Symptoms, Tingling, Tremors, Trigeminal Neuralgia, Unable to Speak, Seizure Disorder, Insomnia, Parkinson's Disease, Loss of Consciousness, Neuropathy, Ataxia, Unable to speak, Peripheral Neuropathy, Stroke, Cerebrovascular Accident, Speech Problem, Acute Neurological Problem, Flashes, Light, Unresponsive, Multiple Sclerosis, Parkinson's Disease, Febrile Seizure, Paresthesia, Peripheral Neuropathy, Hydrocephalus, Spasticity, Neuroendocrine Tumor, Dizziness, Near Syncope, Syncope, Vertigo, Spells, Hypotension, Paroxysmal Positional Vertigo, Paroxysmal Positional Vertigo \\
-\hline
-Respiratory-Related Issues & Airway Obstruction, Aspiration, Pain With Breathing, Near Drowning, Respiratory Distress, Shortness of Breath, Wheezing, Increased Work Of Breathing, Difficulty Breathing, Choking, Oxygen Dependence, Hyperventilating, Orthopnea, Congestion, Cough, Coughing Up Blood, Flu Symptoms, Enlarged Tonsils, Peritonsillar Abscess, Nasal Congestion, Sinus Symptoms, Sinusitis, Sore Throat, Hoarseness, Throat Problem, Upper Respiratory Infection, Influenza, Laryngitis, Respiratory Arrest, Pneumonia, Pleural Effusion, Asthma, Croup, URI, Peritonsillar Abscess\\
-\hline
-\end{longtable}
+\newpage
+
+\begin{table}[!htbp] \centering 
+  \caption{2SLS Results: Robustness to Batching Definition (1-minute and 10-minute windows)} 
+  \label{tab:robustness_batching}
+
+\begin{tabular}{@{\extracolsep{10pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+ & \multicolumn{3}{c}{\textbf{1-minute window}} \\
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
+  & (0.203) & (0.018) & (0.009) \\ 
+  & & & \\ 
+ Batch & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
+  & (0.496) & (0.054) & (0.023) \\ 
+  & & & \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
+Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
+Residual Std. Error & 0.492 & 0.836 & 0.184 \\ 
+\hline 
+\\[-1.8ex] & \multicolumn{3}{c}{\textbf{10-minute window}} \\
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Testing Inclination & 0.650$^{***}$ & 0.957$^{***}$ & $-$0.027$^{**}$ \\ 
+  & (0.241) & (0.022) & (0.011) \\ 
+  & & & \\ 
+ Batch & $-$0.205 & 0.141$^{***}$ & 0.017 \\ 
+  & (0.438) & (0.045) & (0.018) \\ 
+  & & & \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.146 & 0.315 & 0.009 \\ 
+Adjusted R$^{2}$ & 0.141 & 0.311 & 0.003 \\ 
+Residual Std. Error & 0.512 & 0.814 & 0.184 \\ 
+\hline 
+\hline \\[-1.8ex] 
+\end{tabular}
+\begin{tablenotes}
+\tiny
+\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\end{tablenotes}
+\end{table}
+
+\newpage
+
+\begin{table}[!htbp] \centering 
+  \caption{2SLS Results: Robustness to Batching Definition (all tests at once)} 
+  \label{tab:robustness_batching2}
+
+\begin{tabular}{@{\extracolsep{5pt}}lccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & Log LOS & Number of Tests & 72 Hour Return \\ 
+\\[-1.8ex] & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Batch & 0.592$^{***}$ & 0.975$^{***}$ & $-$0.025$^{***}$ \\ 
+  & (0.203) & (0.018) & (0.009) \\ 
+  & & & \\ 
+ `any.batch(fit)` & $-$0.101 & 0.159$^{***}$ & 0.021 \\ 
+  & (0.496) & (0.054) & (0.023) \\ 
+  & & & \\ 
+\textit{N} & 41,929 & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.209 & 0.278 & 0.009 \\ 
+Adjusted R$^{2}$ & 0.204 & 0.274 & 0.003 \\ 
+Residual Std. Error (df = 41667) & 0.492 & 0.836 & 0.184 \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
+\end{tabular} 
+\end{table}
+
+\newpage
 

### 2024-01-25 src/IV Validation.R

@@ -513,6 +513,32 @@ for(complaint in unique_complaints) {
 
 sink()
 
+unique_complaints[1]
+
+summary(felm(nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+       (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+       data = subset(data, GROUPED_COMPLAINT == unique_complaints[1])))
+
+summary(felm(nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+             data = subset(data, GROUPED_COMPLAINT == unique_complaints[2])))
+
+summary(felm(nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+             data = subset(data, GROUPED_COMPLAINT == unique_complaints[3])))
+
+summary(felm(nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+             data = subset(data, GROUPED_COMPLAINT == unique_complaints[4])))
+
+summary(felm(nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+             data = subset(data, GROUPED_COMPLAINT == unique_complaints[5])))
+
+summary(felm(nEDTests ~ test.inclination | dayofweekt + month_of_year + complaint_esi | 
+               (any.batch ~ batch.tendency + test.inclination)| ED_PROVIDER, 
+             data = subset(data, GROUPED_COMPLAINT == unique_complaints[6])))
+
 #=========================================================================
 # Specific Types of Test Batching: Lab+Image Batch
 #=========================================================================
@@ -633,3 +659,49 @@ stargazer(list(model.IV.LOS,
           style = 'QJE')
 
 sink()
+
+#=========================================================================
+# Robustness, Change definition of batching
+#=========================================================================
+
+data$any.batch[!is.na(data$sequenced)] <- 0
+
[…]

### 2024-09-09 Batching Draft 8/Batching Draft 8.Rmd
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
  - name: Jacob Jameson
    department:  Interfaculty Initiative in Health Policy
    affiliation: Harvard University
    location: Cambridge, MA 
  - name: Soroush Saghafian
    department:  Harvard Kennedy School of Government
    affiliation: Harvard University
    location: Cambridge, MA 
  - name: Robert Huckman
    department:  Harvard Business School
    affiliation: Harvard University
    location: Boston, MA 
  - name: Nicole Hodgson
    department:  Emergency Department
    affiliation: Mayo Clinic
    location: Phoenix, AZ
abstract: |

keywords:

bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
  - \usepackage{adjustbox}
  - \usepackage{multirow}
---

# Introduction

Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization. Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events. Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality [@bernstein2009].

One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests [@naseim2015]. A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests. These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.

A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially. This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing. Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results [@koch2018]. Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier. Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged. Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements [@saghafian2015], however, research surrounding test ordering strategies to improve the patient flow processes remains limited.

In this paper, we use data from two large EDs to quantify the benefits and consequences of batching versus sequentially ordering advanced imaging tests on patient length of stay, re-admission, and total imaging volume. Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests. When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side. Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests. We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period. The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.

With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly. Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm. In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings. Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines. Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.

In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes. This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization. It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.

# Literature Review

# Theoretical Background and Hypotheses

Emergency departments (EDs) are complex, time-sensitive environments where physicians must make rapid decisions under uncertainty [@asplin2003model; @jarvis2016ed]. Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.

## Diagnostic Decision-Making and Test Ordering in the ED

In the ED, physicians face the challenge of diagnosing and treating patients efficiently while maintaining high-quality care [@croskerry2013cognitive; @johnson2019decision]. Upon a patient's arrival, the physician must decide on a diagnostic strategy, which often involves ordering imaging tests. This decision-making process is influenced by various factors, including the physician's experience, the patient's presenting symptoms, and the ED's current state (e.g., crowding levels) [@cooke2013overview; @bernstein2009effect; @morley2018emergency].

The key distinction in our study is between batch ordering and sequential ordering of imaging tests. In batch ordering, a physician orders multiple tests simultaneously, while in sequential ordering, tests are ordered one at a time, with each subsequent test decision informed by the results of the previous tests. It's crucial to understand that batch ordering does not mean tests are performed simultaneously. Instead, it means that patients enter the queues for multiple tests at once. This distinction is important when considering the implications of batch ordering on ED operations and patient flow [@green2006managing; @batt2019impact].

## Queuing Theory and ED Operations

When a physician batch orders tests, the patient enters multiple queues simultaneously. This can be modeled using parallel queuing systems in operations research [@dai2010load; @zychlinski2019discrete]. While this might seem to expedite the process, it can lead to inefficiencies in resource allocation, as multiple departments (e.g., radiology, laboratory) must allocate resources to a single patient simultaneously, potentially leading to suboptimal resource utilization [@shi2016models; @wang2019data].

Moreover, batch ordering creates an information lag where results from one test cannot inform the decision to proceed with subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis]. This approach can also create blocking effects, where a patient waiting for multiple tests may block other patients who need only one of those tests, potentially increasing overall waiting times [@armony2015heterogeneous; @song2020managing].

## Reasons for Batch Ordering and Theoretical Implications

Physicians may choose to batch order for several reasons. They may believe it will speed up the diagnostic process [@kocher2012effect; @rosen2018quality], or they may be practicing defensive medicine to avoid missing critical diagnoses [@studdert2005defensive; @kanzaria2015emergency]. Batch ordering may also reduce the cognitive load required to make multiple sequential decisions [@croskerry2013cognitive; @nibbelink2018emergency], which can be particularly appealing in busy EDs where physicians are under time pressure [@bernstein2009effect; @morley2018emergency].

From a theoretical perspective, batch ordering presents several implications. It forgoes the opportunity for value of information considerations, where each test result could inform subsequent decisions [@smith2013impact; @akhlaghpour2020value]. Sequential ordering preserves the option to stop testing once a diagnosis is reached, aligning with real options theory in decision-making under uncertainty [@dixit1994investment; @garg2018value]. While batch ordering might seem efficient for individual patients, it could lead to systemic inefficiencies in resource utilization and patient flow [@green2006managing; @batt2019impact].

## Hypotheses Development

Based on this theoretical framework, we develop the following hypotheses:

### Impact on Resource Utilization

*Hypothesis 1: Batch ordering of diagnostic tests will lead to increased overall test utilization compared to sequential ordering.* This hypothesis is grounded in the idea that batch ordering bypasses the opportunity to use information from initial tests to inform the necessity of subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].

### Effect on Length of Stay

*Hypothesis 2: Batch ordering will have a significant impact on patient length of stay (LOS) in the ED, but the direction of this effect is ambiguous.* While batch ordering might reduce LOS by eliminating waiting times between test decisions [@kocher2012effect; @rosen2018quality], it could also increase LOS due to potential queuing inefficiencies and the time required to perform and interpret additional tests [@forster2003effect; @armony2015heterogeneous; @song2020managing].

### Quality of Care and Return Visits

*Hypothesis 3: Batch ordering will influence the rate of 72-hour return visits with admission, but the direction of this effect is uncertain.* Comprehensive initial testing might reduce missed diagnoses and thus return visits [@pines2010national; @sabbatini2016overzealous]. However, it could also lead to false positives and unnecessary follow-ups, potentially increasing return visits [@welch2011overdiagnosed; @o2018overdiagnosis].

### Hypothesis 4: Heterogeneity Across Presenting Complaints

*Hypothesis 4: The effects of batch ordering on resource utilization, length of stay, and quality of care will vary significantly across different presenting complaints.* This hypothesis is based on the understanding that different clinical presentations have varying levels of diagnostic uncertainty and risk [@cooke2013overview; @johnson2019decision], which may influence the appropriateness and impact of batch ordering.

By testing these hypotheses, we aim to provide a comprehensive understanding of the impacts of batch ordering in the ED, informing both clinical practice and operational policy. Our empirical strategy, detailed in the following sections, is designed to rigorously test these hypotheses and uncover the causal effects of batch ordering on ED performance and patient outcomes.

# Data and Methods

Our study uses data from two large U.S. emergency departments (EDs): the Mayo Clinic of Arizona and Massachusetts General Hospital (MGH). The MGH dataset, which includes 129,489 patient encounters from November 10, 2021 through December 10, 2022, provides a robust sample for validating the generalizability of our findings. However, our primary analysis focuses on the Mayo Clinic data due to its unique feature of random patient-physician assignment, which allows for stronger causal inference.

## Mayo Clinic Data

Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation. During the study period from October 6, 2018, through December 31, 2019, the ED recorded 48,854 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces. The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume. The data is summarized in Table \ref{tab:summary_stats}.

\begin{table}[ht]
\centering
\caption{Summary Statistics of Mayo Clinic Emergency Department Encounters}
\label{tab:summary_stats}
\begin{threeparttable}
\begin{tabular}{p{10cm}ccccc}
\toprule
Variable &  & {Mean} & {Q1} & {Median} & {Q3} \\
\midrule
\textbf{Emergency Department Characteristics} & & & & & \\
Total Patients & & {48,854} & {-} & {-} & {-} \\
Patients Admitted & & {18.7\%} & {-} & {-} & {-} \\
Patients Revisited within 72 Hours & & {3.81\%} & {-} & {-} & {-} \\
Patients with IV Fluids & & {35.7\%} & {-} & {-} & {-} \\
Patients with IV Meds & & {17.9\%} & {-} & {-} & {-} \\
Patients in ED & & {24.1} & {-} & {-} & {-} \\
Tachycardic & & {19.2\%} & {-} & {-} & {-} \\
Tachypneic & & {8.82\%} & {-} & {-} & {-} \\
Febrile & & {2.16\%} & {-} & {-} & {-} \\
Hypotensive & & {1.42\%} & {-} & {-} & {-} \\
ESI & & {2.81} & {-} & {-} & {-} \\
Time from Arrival to Triage (mins) & & {8.03} & {4} & {6} & {10} \\
Time from Triage to First Contact (mins) & & {80.2} & {11} & {29} & {61} \\
Average ED LOS (min) & & {246} & {-} & {-} & {-} \\
\midrule
\textbf{Patient Demographics} & & & & & \\
Percent Male & & {46.5\%} & {-} & {-} & {-} \\
Race: White & & {88.4\%} & {-} & {-} & {-} \\
Race: Black & & {4.17\%} & {-} & {-} & {-} \\
Race: Asian & & {3.05\%} & {-} & {-} & {-} \\
Gender: Female & & {53.5\%} & {-} & {-} & {-} \\
Arrival Age & & {57.7} & {43} & {61} & {74} \\
\midrule
\textbf{Diagnostic Tests and Outcomes} & & & & & \\
X-Rays Performed & & {43.3\%} & {-} & {-} & {-} \\
Ultrasounds Performed & & {11.3\%} & {-} & {-} & {-} \\
CTs Performed & & {35.5\%} & {-} & {-} & {-} \\
Labs Ordered & & {73.7\%} & {-} & {-} & {-} \\
Patients Discharged & & {66.8\%} & {-} & {-} & {-} \\
Patients Admitted & & {18.7\%} & {-} & {-} & {-} \\
Contrast CT Performed & & {17.7\%} & {-} & {-} & {-} \\
Time to Result: X-Ray (mins) & & {67.2} & {36} & {54} & {79} \\
Time to Result: Ultrasound (mins) & & {165} & {71} & {101} & {150} \\
Time to Result: Contrast CT (mins) & & {142} & {86} & {115} & {153} \\
Time to Result: Non-Contrast CT (mins) & & {89.7} & {50} & {70} & {102} \\
Time to Result: Lab (mins) & & {46.1} & {25} & {35} & {49} \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{threeparttable}
\end{table}

A key feature of the Mayo Clinic ED is its rotational patient assignment system, in which patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Balancing Test: Wald Test for Equality of Means}
    \label{tab:wald_test}
    \begin{tabular}{p{10cm}cccc}
        \toprule
        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 6,232 & 2.587 & 0.108 \\ 
        Back or Flank Pain & 2,552 & 1.637 & 0.201 \\ 
        Chest Pain & 3,525 & 0.407 & 0.524 \\ 
        Extremity Complaints & 5,265 & 1.847 & 0.174 \\ 
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2,381 & 0.023 & 0.880 \\ 
        Gastrointestinal Issues & 3,323 & 0.105 & 0.746 \\ 
        Neurological Issue & 3,495 & 0.135 & 0.713 \\ 
        Shortness of Breath & 2,966 & 1.324 & 0.250 \\ 
        Skin Complaints & 2,178 & 0.383 & 0.536 \\ 
        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
        \midrule
        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 13,914 & 0.011 & 0.915 \\ 
        ESI 3, 4, or 5 & 29,386 & 0.010 & 0.921 \\ 
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\small
\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{table}

We conducted a retrospective review of comprehensive ED operational data, coinciding with the initiation of a new electronic medical record. The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics. This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.

## Sample Construction

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe $48,854$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories as we do. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $15,821$ encounters involving full-time physicians who treat over 500 ED cases per year.

## Variable Definitions

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window. In Section [...], sensitivity analyses on this cutoff point showed that our results are robust to this definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study. Therefore, a batch in our study consists of two or more distinct imaging tests.

Our dependent variables are as follows:

*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings. It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.

*(b) Number of Unique Imaging Tests*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives. In this study, we quantify resource utilization by the total number of distinct imaging tests ordered per patient during their ED stay. This encompasses both initial and subsequent tests.

*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This outcome is used to assess the quality of care provided during the initial ED visit.

## Empirical Strategy

Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court. Similarly, we explore batching variation across physicians who work in the same emergency department. In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians. Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision. We employ both approaches and lay out their details in the next subsections.

### Batch Tendency Construction

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \alpha_{lab} + \varepsilon_{i,t}
\end{equation}

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision. Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case. As stated earlier, these controls are more than what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:

\begin{equation}
Tendency_{i,j}^{phys} =
\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{\varepsilon}_{i'}
\end{equation}

where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order imaging tests. Table \ref{table:first_stage} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $6$pp increase in the likelihood of having tests batch-ordered in the ED. The F-statistic is 94 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.

\begin{table}[!htbp] \centering 
  \caption{Comparison of First Stage Estimates}
  \label{table:first_stage}
  \begin{tabularx}{5.5in}{Xccc} % 'X' for the first column and 'c' for the others
  \toprule
   & \multicolumn{3}{c}{Coeffecient} \\
  \cmidrule{2-4}
   & (1) & (2) & (3) \\
  \midrule
  Batch Tendency & 1.92$^{***}$ & 1.91$^{***}$ & 1.76$^{***}$ \\ 
   & (0.07) & (0.06) & (1.76) \\ 
  \textit{Day of Week-Time of Day FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
  \textit{Month of Year FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
  \textit{Complaint/Severity FE} & & $\checkmark$ & $\checkmark$ \\
  \textit{Laboratory Tests Ordered} & & & $\checkmark$ \\
  \midrule
  F Statistic & 9.55 & 17.86 & 21.59 \\ 
  N & 16,361 & 16,361 & 16,361 \\ 
  \bottomrule
  \end{tabularx}
  \begin{tablenotes}
  \small
  \item Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
  \item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
  \end{tablenotes}
\end{table}

To estimate the reduced-form effects of being treated by a batch-preferring physician, we estimate the following equation:

\begin{equation}
Y_i = \mu_0 + \mu_1 Tendency_{i,j}^{phys} + \gamma X_i + \nu_i
\end{equation}

This reduced form will allow us to check that our instrument is a strong instrument. To study the effects of test batching in the ED on an outcome $Y_i$, we estimate the following 2SLS equations using our baseline sample:

\begin{equation}
Y_i = \beta_0 + \beta_1 Batched_i + \theta X_i + \varepsilon_i
\end{equation}

\begin{equation}
Batched_i = \delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i + \nu_i
\end{equation}

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour return with admission, and number of imaging tests ordered for the patient, and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, length of stay, and return with admission likelihood. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

## Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity. We reported a strong first stage (i.e., relevance) at the end of the previous Section. The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching. This is perhaps our strongest assumption and is at its core, untestable. However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings. First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher. Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest. The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis. Nevertheless, we take this assumption seriously and perform a placebo check in Section [...] as well as various robustness checks in Section [...].

Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects. It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician. The literature leveraging the judges design typically performs two informal tests for its implications. The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects). The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration). Appendix Table  [...] presents both of these tests in the two columns for various subsamples of interest.

# Results {#sec:4}

## Reduced-Form Results

In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department. We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination. Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.

To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency. Specifically, we calculate the physician's residualized leave-out average test rate, which serves as a proxy for their propensity to order tests. It is important to note the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination. This substantial correlation suggests that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies. Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.

Reduced-form regression results in Table \ref{table:reduced_form} reveal the effect the average total effect of batch tendency on resource utilization and LOS. However, it is important to note that the estimated total effect of batch tendency accounts for both direct and indirect (mediated) effects of batch tendency. We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered. The increased testing volume, in turn, is associated with an extended LOS.

\begin{table}[!htbp] 
\centering 
\caption{Reduced Form Estimates for Batch Tendency}
\label{table:reduced_form}
\small
\begin{threeparttable}
\begin{tabular}{lcccccc}
\toprule
 & 72Hr Admission & LOS & Imaging Tests & 72Hr Admission & LOS & Imaging Tests \\
\midrule
Batch Tendency & -0.033 & 2.038** & 4.655*** & -0.063*** & -0.829 & 1.137** \\
 & (0.028) & (0.994) & (0.554) & (0.055) & (2.393) & (0.464) \\
\textit{Baseline Controls} & $\checkmark$ & $\checkmark$ & $\checkmark$ &  $\checkmark$ & $\checkmark$ & $\checkmark$\\
\textit{Testing Inclination} &  &  &  &  $\checkmark$ & $\checkmark$ & $\checkmark$\\
\textit{Any Imaging Ordered} &  &  &  &  $\checkmark$ & $\checkmark$ & $\checkmark$\\
\midrule
N & 15,821 & 15,821 & 15,821 & 15,821 & 15,821 & 15,821 \\
R$^2$ & 0.006 & 0.303 & 0.199 & 0.006 & 0.364 & 0.591 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item Estimates of the reduced form described in the text. Robust standard errors are clustered at the physician level.
\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
\end{tablenotes}
\end{threeparttable}
\end{table}

## Placebo Check

In this section we investigate whether the reduced-form effects observed in Section [...] are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency. We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing. For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes. Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.

To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a >5 percent batching rate). We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3). The results of this exercise are displayed in Appendix Table [...]. They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.

## Instrumental Variables Results

To uncover the causal effect of test batching on key emergency department (ED) outcomes, we utilize a two-stage least squares (2SLS) approach, leveraging physician batch-ordering tendencies as an instrument. This strategy isolates the local average treatment effect (LATE) for patients whose diagnostic tests were batched due to their physician’s inclination to batch, rather than patient characteristics or other unobserved factors that may influence outcomes. The IV estimates capture the effect of test batching on ED length of stay, the number of imaging tests ordered, and the likelihood of a return visit within 72 hours with admission.

\begin{table}[!htbp] \centering 
  \caption{IV Results} 
  \label{} 
\begin{tabular}{@{\extracolsep{15pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & Log LOS & Imaging Tests & 72Hr Admission  \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Batched & $-$0.504 & 0.692$^{**}$ & $-$0.038 \\ 
  & (1.457) & (0.286) & (0.034) \\ 
  & & & \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular} 
\begin{tablenotes}
\small
\item Estimates of the reduced form described in the text. Robust standard errors are clustered at the physician level.
\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
\end{tablenotes}
\end{table}

*ED Length of Stay*—The IV results in column (1) of Table X indicate that test batching has a negative but statistically insignificant effect on ED length of stay. The coefficient of -0.504 suggests a potential reduction in time spent in the ED, but the confidence interval is wide, indicating that the effect may not be robust across different specifications. This finding implies that while batching may streamline certain diagnostic processes, other operational factors likely play a larger role in determining how long a patient remains in the ED. These results suggest that batching alone does not drastically impact the overall time patients spend in the ED, pointing to the need for further exploration of mechanisms that may moderate this relationship.

*Number of Imaging Tests Ordered*—In column (2), we find a statistically significant and positive effect of test batching on the number of imaging tests ordered. The coefficient of 0.692 (p < 0.05) indicates that batching leads to approximately 69% more imaging tests being conducted during the patient’s ED visit relative to if that patient's tests were sequenced. This result highlights a key consequence of batching: the grouping of diagnostic orders leads to a more intensive use of imaging resources. This increased resource utilization, while contributing to more comprehensive diagnostic evaluations, also raises concerns about the potential for overutilization and the associated costs and patient burden. This effect is particularly relevant in resource-constrained environments, where the trade-off between efficiency and resource use becomes crucial.

*Return Visits with Admission within 72 Hours*—The results in column (3) show that test batching has a negative effect on the likelihood of a return visit with admission within 72 hours, although this effect is not statistically significant. The point estimate of -0.038 suggests a small potential reduction in return visits for patients whose tests were batched, which may reflect more thorough initial diagnostics leading to fewer missed diagnoses or complications. However, the lack of statistical significance means we cannot draw firm conclusions from this result. This finding aligns with previous research suggesting that while batching may improve certain diagnostic processes, its effect on downstream outcomes such as return visits may be more complex and contingent on factors beyond test ordering patterns.

### Mechanism: Mediation Analysis

To further explore the mechanisms through which batching influences ED length of stay (LOS) and return visits with admission within 72 hours, we conducted a formal mediation analysis. Specifically, we assessed whether the number of imaging tests mediates the relationship between batching and these outcomes. This analysis helps to decompose the total effect of batching into direct and indirect effects, with the indirect effect capturing the mediation through imaging tests.

The mediation analysis was performed using a quasi-Bayesian approach with 1,000 simulations, adjusting for relevant covariates, including physician testing inclination, imaging use, day of the week, month of the year, patient complaints, and laboratory performance. The results for both outcomes are summarized in Table \ref{tab:med_results}, which presents the average causal mediation effect (ACME), average direct effect (ADE), and total effect for each outcome.

\begin{table}[h]
\centering
\caption{Mediation Analysis for ED Length of Stay and 72-Hour Return Visits}
\label{tab:med_results}
\begin{tabular}{lcccc}
\hline
Outcome      & Effect Type & Estimate & 95\% CI Lower & 95\% CI Upper \\ \hline
\multirow{4}{*}{\textbf{ED Length of Stay}} 
& & & & \\
    & ACME (Indirect Effect)  & 0.083   & 0.016  & 0.150  \\
    & ADE (Direct Effect)     & -0.550  & -3.686 & 2.300  \\
    & Total Effect            & -0.468  & -3.566 & 2.380  \\ \hline
\multirow{4}{*}{\textbf{72-Hour Return Visits}} 
& & & & \\
    & ACME (Indirect Effect)  & -0.0029  & -0.0059 & 0.000  \\
    & ADE (Direct Effect)     & -0.036   & -0.097  & 0.030  \\
    & Total Effect            & -0.039   & -0.099  & 0.030  \\ \hline
\end{tabular}
\end{table}

\textbf{Incorporating Contemporary Mediation Frameworks.}—In line with the classical literature on mediation (Baron and Kenny, 1986), mediation analysis traditionally proceeds in three steps: (1) testing the effect of the independent variable (batch order) on the dependent variable (LOS/72-hour return) without the mediator (imaging volume); (2) adding the mediator (imaging volume) to the model and testing the direct effect of the independent variable on the dependent variable; and (3) testing the mediator’s effect on the dependent variable in the presence of the independent variable.

However, recent literature has demonstrated that significant results in the zero-order effect (i.e., the total effect) are not required to justify moving forward with mediation analysis. Zhao et al. (2010) and Hayes (2022) argue that requiring significance in each step is an overly restrictive and outdated framework, which often overlooks indirect effects that can be meaningful even when direct effects are not. In our analysis, although we do not always find statistically significant direct effects for batching on LOS or return visits, the mediation pathway through imaging volume remains relevant for understanding the indirect effects of batching.

Thus, following the contemporary guidance in mediation analysis, we proceed with testing both direct and indirect effects simultaneously, allowing us to capture the more nuanced dynamics of how batching influences ED outcomes. This approach provides a richer understanding of the operational impacts of batching, particularly as it pertains to the trade-offs between diagnostic efficiency and resource utilization.

\textbf{Mediation Results for ED Length of Stay.}—The mediation analysis reveals that the indirect effect (ACME) of batching on ED length of stay through the number of imaging tests is $0.083 (p < 0.01)$, suggesting that the increased volume of diagnostic testing marginally extends patient stays in the ED. The direct effect (ADE) of batching on length of stay is negative but not statistically significant ($-0.550, p = 0.71$), and the total effect is also negative ($-0.468$) but insignificant. This indicates that while batching may reduce LOS, the effect is partially offset by the increased testing, leading to no substantial impact on overall ED throughput.

\textbf{Mediation Results for 72-Hour Return Visits.}—For return visits within 72 hours, the indirect effect through the number of imaging tests is negative ($-0.0029, p < 0.05$), indicating that the additional testing associated with batching may slightly reduce the likelihood of a return visit. However, the direct effect ($-0.036$) and total effect ($-0.039$) of batching on return visits are small and not statistically significant, suggesting that while comprehensive diagnostics may reduce return visits, the effect of batching on this outcome is generally weak.

\textbf{Conceptual Framework: Directed Acyclic Graph (DAG).}—Based on the findings from the mediation analysis, we propose a conceptual framework to illustrate the relationships uncovered by our empirical results. The Directed Acyclic Graph (DAG) shown in Figure \ref{fig:dag} provides a visual representation of the causal pathways between batch tendency (Z), batch ordering (X), imaging volume (M), and our outcomes (Y). The graph indicates that the indirect effect of batching operates through the volume of imaging tests, while the direct effect of batching on the outcomes bypasses this mediator.

\begin{figure}[h]
\centering
\includegraphics[width=3in]{model.png}
\caption{Directed Acyclic Graph Representing the Causal Pathways Between Batch Order, Imaging Volume, and ED Outcomes}
\label{fig:dag}
\end{figure}

In this DAG, batch tendency (Z) is used as an instrumental variable to predict batch order (X), which in turn affects both imaging volume (M) and our primary outcomes, including ED length of stay and 72-hour return with admission (Y). The unobserved confounding (U) is assumed to influence both imaging volume and the outcomes directly. The mediation analysis supports this structure, as it confirms that imaging volume mediates the relationship between batch ordering and ED outcomes, albeit modestly. By identifying this framework through our analysis, we provide a data-driven explanation for how batching influences ED operations.

\textbf{Interpretation of Mediation Results.}—These results show that the number of imaging tests serves as a partial mediator in the relationship between batching and ED outcomes. For length of stay, increased testing tends to prolong ED visits, but the overall effect of batching on LOS is negligible. For return visits, the increase in diagnostic testing slightly reduces the likelihood of a return within 72 hours, though the overall effect remains small and insignificant. These findings highlight the trade-offs involved in batching diagnostic orders: while batching may streamline certain processes, the associated increase in testing has modest effects on patient flow and outcomes.

### Time to Result

\newpage

# References

### 2024-09-09 Batching Draft 8/Batching-Draft-8.tex
getting going

--- 2024-09-09 Batching Draft 8/Batching Draft 8.Rmd
@@
----
-title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
-authors:
-  - name: Jacob Jameson
-    department:  Interfaculty Initiative in Health Policy
-    affiliation: Harvard University
-    location: Cambridge, MA 
-  - name: Soroush Saghafian
-    department:  Harvard Kennedy School of Government
-    affiliation: Harvard University
-    location: Cambridge, MA 
-  - name: Robert Huckman
-    department:  Harvard Business School
-    affiliation: Harvard University
-    location: Boston, MA 
-  - name: Nicole Hodgson
-    department:  Emergency Department
-    affiliation: Mayo Clinic
-    location: Phoenix, AZ
-abstract: |
+\documentclass{article}
 
-keywords:
+\usepackage{arxiv}
 
-bibliography: references.bib
-biblio-style: unsrt
-output: rticles::arxiv_article
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{tabularx}
-  - \usepackage{graphicx}
-  - \usepackage{tikz}
-  - \usepackage{siunitx}
-  - \usepackage{tablefootnote}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \usepackage{natbib}
-  - \usepackage{caption}
-  - \usepackage{adjustbox}
-  - \usepackage{multirow}
----
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{graphicx}
 
-# Introduction
+\title{To Batch or Not to Batch: Test-Ordering Variability in the ED}
 
-Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization.
-Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events.
-Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality [@bernstein2009].
+\author{
+    Jacob Jameson
+   \\
+    Interfaculty Initiative in Health Policy \\
+    Harvard University \\
+  Cambridge, MA \\
+  \texttt{} \\
+   \And
+    Soroush Saghafian
+   \\
+    Harvard Kennedy School of Government \\
+    Harvard University \\
+  Cambridge, MA \\
+  \texttt{} \\
+   \And
+    Robert Huckman
+   \\
+    Harvard Business School \\
+    Harvard University \\
+  Boston, MA \\
+  \texttt{} \\
+   \And
+    Nicole Hodgson
+   \\
+    Emergency Department \\
+    Mayo Clinic \\
+  Phoenix, AZ \\
+  \texttt{} \\
+  }
 
-One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests [@naseim2015].
-A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests.
-These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially.
-This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing.
-Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results [@koch2018].
-Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier.
-Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged.
-Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements [@saghafian2015], however, research surrounding test ordering strategies to improve the patient flow processes remains limited.
+% Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-In this paper, we use data from two large EDs to quantify the benefits and consequences of batching versus sequentially ordering advanced imaging tests on patient length of stay, re-admission, and total imaging volume.
-Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests.
-When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side.
-Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests.
-We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period.
-The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.
+\usepackage{booktabs}
+\usepackage{tabularx}
+\usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{siunitx}
+\usepackage{tablefootnote}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\usepackage{natbib}
+\usepackage{caption}
+\usepackage{adjustbox}
+\usepackage{multirow}
+\begin{document}
+\maketitle
 
-With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly.
-Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm.
-In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings.
-Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines.
-Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.
+\begin{abstract}
 
-In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes.
-This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization.
-It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.
+\end{abstract}
 
-# Literature Review
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-# Theoretical Background and Hypotheses
+Healthcare delivery, particularly in the emergency department (ED), is a
+delicate balance that involves ensuring optimal patient outcomes while
+optimizing resource utilization. Achieving these twin goals requires
+timely and accurate diagnosis, which in turn enables prompt and
+appropriate treatment, consequently improving patient prognosis and
+reducing the likelihood of adverse events. Furthermore, efficient
+patient discharge from the ED can help alleviate overcrowding, a severe
+issue with potential consequences including higher complication rates
+and increased mortality (S. Bernstein et al. 2009).
 
-Emergency departments (EDs) are complex, time-sensitive environments where physicians must make rapid decisions under uncertainty [@asplin2003model; @jarvis2016ed].
-Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.
+One important factor that can impact the speed and effectiveness of
+diagnosis in the ED is the availability and performance of diagnostic
+tests (Balogh, Miller, and Ball 2015). A variety of diagnostic tests are
+used in the ED, including laboratory tests, imaging studies, and
+specialized tests. These tests can provide valuable information about a
+patient's condition and help to guide treatment decisions.
 
-## Diagnostic Decision-Making and Test Ordering in the ED
+A critical question in this context pertains to whether physicians in
+the ED should batch order diagnostic tests or order them sequentially.
+This decision essentially represents a tradeoff between reducing patient
+length of stay and risk of over-testing. Over-testing, or performing
+unnecessary tests, can lead to increased costs, unnecessary patient
+anxiety, and potential harm from follow-up of false-positive results
+(Koch et al. 2018). Conversely, keeping a patient for an extended time
+to perform all possible tests could lead to ED overcrowding, an issue
+associated with severe consequences, as mentioned earlier. Instead, what
+is needed is a reasonable balance between the number of diagnostic tests
+performed and the total time the patient is kept in the ED before either
+being admitted or discharged. Several studies have demonstrated that
+optimizing the ED patient flow process can result in significant
+improvements (Saghafian, Austin, and Traub 2015), however, research
+surrounding test ordering strategies to improve the patient flow
+processes remains limited.
 
-In the ED, physicians face the challenge of diagnosing and treating patients efficiently while maintaining high-quality care [@croskerry2013cognitive; @johnson2019decision].
-Upon a patient's arrival, the physician must decide on a diagnostic strategy, which often involves ordering imaging tests.
-This decision-making process is influenced by various factors, including the physician's experience, the patient's presenting symptoms, and the ED's current state (e.g., crowding levels) [@cooke2013overview; @bernstein2009effect; @morley2018emergency].
+In this paper, we use data from two large EDs to quantify the benefits
+and consequences of batching versus sequentially ordering advanced
+imaging tests on patient length of stay, re-admission, and total imaging
+volume. Our empirical strategy exploits random assignment of patients to
+ED physicians who differ in their propensity to batch-order diagnostic
+tests. When patients arrive at the ED, they are assigned to a physician
+based on availability, with no discretion on either side. Thus, patients
+who arrive at the ED at similar times are randomly assigned to
+physicians who vary in their willingness to batch order diagnostic
+tests. We measure physician tendency to batch using a leave-out,
+residualized measure based on all other patients the physician has seen
+in the ED in the study period. The tendency measure strongly predicts
+the ED test batch outcome but is uncorrelated with patient and ED visit
+characteristics.
 
-The key distinction in our study is between batch ordering and sequential ordering of imaging tests.
-In batch ordering, a physician orders multiple tests simultaneously, while in sequential ordering, tests are ordered one at a time, with each subsequent test decision informed by the results of the previous tests.
-It's crucial to understand that batch ordering does not mean tests are performed simultaneously.
-Instead, it means that patients enter the queues for multiple tests at once.
-This distinction is important when considering the implications of batch ordering on ED operations and patient flow [@green2006managing; @batt2019impact].
+With the caveat that other unobserved dimensions of physician care may
+impact patient outcomes, we employ physician batch tendency as an
+instrumental variable for having tests batch ordered in the ED to
+quantify the effect of batching directly. Because of the institutional
+features of the ED, our research design closely approximates an RCT that
+assigns patients to batch-ordering or sequential-ordering arm. In the
+ED, patients have no discretion over choosing providers, and in our
+specific ED, physicians have discretion over choosing patients,
+alleviating major selection issues present in other health care
+settings. Furthermore, physicians exhibit wide variation in practice
+behavior in batch-ordering, even within the same hospital, while
+following the same guidelines. Finally, patient-physician interactions
+in the ED are typically well documented, short, and one-off,
+constraining physician decision-making to a more limited,
+better-observed choice set than present in settings such as specialty or
+primary care.
 
-## Queuing Theory and ED Operations
+In sum, exploiting practice variation in ED settings shuts down other
+(but not all) potential channels besides test batching that are present
+in other settings, determine length of stay, and impact patient
+outcomes. This approach allows us to move closer to identifying the
+causal impact of batch-ordering diagnostic tests on patient outcomes and
+resource utilization. It is important to note that this paper studies
+the impact of batch-ordering through a batching decision requiring
+clinical judgment (within practice norms) rather than through specific
+hospital policies, differences in adherence to clinical practice
+guidelines, or substandard care.
 
-When a physician batch orders tests, the patient enters multiple queues simultaneously.
-This can be modeled using parallel queuing systems in operations research [@dai2010load; @zychlinski2019discrete].
-While this might seem to expedite the process, it can lead to inefficiencies in resource allocation, as multiple departments (e.g., radiology, laboratory) must allocate resources to a single patient simultaneously, potentially leading to suboptimal resource utilization [@shi2016models; @wang2019data].
+\hypertarget{literature-review}{%
+\section{Literature Review}\label{literature-review}}
 
-Moreover, batch ordering creates an information lag where results from one test cannot inform the decision to proceed with subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].
-This approach can also create blocking effects, where a patient waiting for multiple tests may block other patients who need only one of those tests, potentially increasing overall waiting times [@armony2015heterogeneous; @song2020managing].
+\hypertarget{theoretical-background-and-hypotheses}{%
+\section{Theoretical Background and
+Hypotheses}\label{theoretical-background-and-hypotheses}}
 
-## Reasons for Batch Ordering and Theoretical Implications
+Emergency departments (EDs) are complex, time-sensitive environments
+where physicians must make rapid decisions under uncertainty (Asplin et
+al. 2003; Jarvis 2016). Our theoretical model aims to capture the
+nuanced dynamics of diagnostic decision-making in this high-stakes
+setting, with a particular focus on the practice of batch ordering
+imaging tests.
 
-Physicians may choose to batch order for several reasons.
-They may believe it will speed up the diagnostic process [@kocher2012effect; @rosen2018quality], or they may be practicing defensive medicine to avoid missing critical diagnoses [@studdert2005defensive; @kanzaria2015emergency].
-Batch ordering may also reduce the cognitive load required to make multiple sequential decisions [@croskerry2013cognitive; @nibbelink2018emergency], which can be particularly appealing in busy EDs where physicians are under time pressure [@bernstein2009effect; @morley2018emergency].
+\hypertarget{diagnostic-decision-making-and-test-ordering-in-the-ed}{%
+\subsection{Diagnostic Decision-Making and Test Ordering in the
+ED}\label{diagnostic-decision-making-and-test-ordering-in-the-ed}}
 
-From a theoretical perspective, batch ordering presents several implications.
-It forgoes the opportunity for value of information considerations, where each test result could inform subsequent decisions [@smith2013impact; @akhlaghpour2020value].
-Sequential ordering preserves the option to stop testing once a diagnosis is reached, aligning with real options theory in decision-making under uncertainty [@dixit1994investment; @garg2018value].
-While batch ordering might seem efficient for individual patients, it could lead to systemic inefficiencies in resource utilization and patient flow [@green2006managing; @batt2019impact].
+In the ED, physicians face the challenge of diagnosing and treating
+patients efficiently while maintaining high-quality care (Croskerry
+2013; Johnson 2019). Upon a patient's arrival, the physician must decide
+on a diagnostic strategy, which often involves ordering imaging tests.
+This decision-making process is influenced by various factors, including
+the physician's experience, the patient's presenting symptoms, and the
+ED's current state (e.g., crowding levels) (Cooke et al. 2013; S. L.
+Bernstein et al. 2009; Morley et al. 2018).
 
-## Hypotheses Development
+The key distinction in our study is between batch ordering and
+sequential ordering of imaging tests. In batch ordering, a physician
+orders multiple tests simultaneously, while in sequential ordering,
+tests are ordered one at a time, with each subsequent test decision
+informed by the results of the previous tests. It's crucial to
+understand that batch ordering does not mean tests are performed
+simultaneously. Instead, it means that patients enter the queues for
+multiple tests at once. This distinction is important when considering
+the implications of batch ordering on ED operations and patient flow
+(Green 2006; Batt and Terwiesch 2019).
 
-Based on this theoretical framework, we develop the following hypotheses:
+\hypertarget{queuing-theory-and-ed-operations}{%
+\subsection{Queuing Theory and ED
+Operations}\label{queuing-theory-and-ed-operations}}
 
-### Impact on Resource Utilization
+When a physician batch orders tests, the patient enters multiple queues
+simultaneously. This can be modeled using parallel queuing systems in
+operations research (Dai and He 2010; Zychlinski et al. 2019). While
+this might seem to expedite the process, it can lead to inefficiencies
+in resource allocation, as multiple departments (e.g., radiology,
+laboratory) must allocate resources to a single patient simultaneously,
+potentially leading to suboptimal resource utilization (Shi et al. 2016;
+Wang et al. 2019).
 
-*Hypothesis 1: Batch ordering of diagnostic tests will lead to increased overall test utilization compared to sequential ordering.* This hypothesis is grounded in the idea that batch ordering bypasses the opportunity to use information from initial tests to inform the necessity of subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].
+Moreover, batch ordering creates an information lag where results from
+one test cannot inform the decision to proceed with subsequent tests,
+potentially leading to unnecessary testing (Welch, Schwartz, and
+Woloshin 2011; O'Connor and Nichol 2018). This approach can also create
+blocking effects, where a patient waiting for multiple tests may block
+other patients who need only one of those tests, potentially increasing
+overall waiting times (Armony et al. 2015; Song et al. 2020).
 
-### Effect on Length of Stay
+\hypertarget{reasons-for-batch-ordering-and-theoretical-implications}{%
+\subsection{Reasons for Batch Ordering and Theoretical
+Implications}\label{reasons-for-batch-ordering-and-theoretical-implications}}
 
-*Hypothesis 2: Batch ordering will have a significant impact on patient length of stay (LOS) in the ED, but the direction of this effect is ambiguous.* While batch ordering might reduce LOS by eliminating waiting times between test decisions [@kocher2012effect; @rosen2018quality], it could also increase LOS due to potential queuing inefficiencies and the time required to perform and interpret additional tests [@forster2003effect; @armony2015heterogeneous; @song2020managing].
+Physicians may choose to batch order for several reasons. They may
+believe it will speed up the diagnostic process (Kocher et al. 2012;
+Rosen and Seda 2018), or they may be practicing defensive medicine to
+avoid missing critical diagnoses (Studdert et al. 2005; Kanzaria et al.
+2015). Batch ordering may also reduce the cognitive load required to
+make multiple sequential decisions (Croskerry 2013; Nibbelink and Brewer
+2018), which can be particularly appealing in busy EDs where physicians
+are under time pressure (S. L. Bernstein et al. 2009; Morley et al.
+2018).
 
-### Quality of Care and Return Visits
+From a theoretical perspective, batch ordering presents several
+implications. It forgoes the opportunity for value of information
+considerations, where each test result could inform subsequent decisions
+(Smith et al. 2013; Akhlaghpour 2020). Sequential ordering preserves the
+option to stop testing once a diagnosis is reached, aligning with real
+options theory in decision-making under uncertainty (Dixit and Pindyck
+1994; Garg, Wang, and Valerdi 2018). While batch ordering might seem
+efficient for individual patients, it could lead to systemic
+inefficiencies in resource utilization and patient flow (Green 2006;
+Batt and Terwiesch 2019).
 
-*Hypothesis 3: Batch ordering will influence the rate of 72-hour return visits with admission, but the direction of this effect is uncertain.* Comprehensive initial testing might reduce missed diagnoses and thus return visits [@pines2010national; @sabbatini2016overzealous].
-However, it could also lead to false positives and unnecessary follow-ups, potentially increasing return visits [@welch2011overdiagnosed; @o2018overdiagnosis].
+\hypertarget{hypotheses-development}{%
+\subsection{Hypotheses Development}\label{hypotheses-development}}
 
-### Hypothesis 4: Heterogeneity Across Presenting Complaints
+Based on this theoretical framework, we develop the following
+hypotheses:
 
-*Hypothesis 4: The effects of batch ordering on resource utilization, length of stay, and quality of care will vary significantly across different presenting complaints.* This hypothesis is based on the understanding that different clinical presentations have varying levels of diagnostic uncertainty and risk [@cooke2013overview; @johnson2019decision], which may influence the appropriateness and impact of batch ordering.
+\hypertarget{impact-on-resource-utilization}{%
+\subsubsection{Impact on Resource
+Utilization}\label{impact-on-resource-utilization}}
 
-By testing these hypotheses, we aim to provide a comprehensive understanding of the impacts of batch ordering in the ED, informing both clinical practice and operational policy.
-Our empirical strategy, detailed in the following sections, is designed to rigorously test these hypotheses and uncover the causal effects of batch ordering on ED performance and patient outcomes.
+\emph{Hypothesis 1: Batch ordering of diagnostic tests will lead to
+increased overall test utilization compared to sequential ordering.}
+This hypothesis is grounded in the idea that batch ordering bypasses the
+opportunity to use information from initial tests to inform the
+necessity of subsequent tests, potentially leading to unnecessary
+testing (Welch, Schwartz, and Woloshin 2011; O'Connor and Nichol 2018).
 
-# Data and Methods
+\hypertarget{effect-on-length-of-stay}{%
+\subsubsection{Effect on Length of
+Stay}\label{effect-on-length-of-stay}}
 
-Our study uses data from two large U.S.
-emergency departments (EDs): the Mayo Clinic of Arizona and Massachusetts General Hospital (MGH).
-The MGH dataset, which includes 129,489 patient encounters from November 10, 2021 through December 10, 2022, provides a robust sample for validating the generalizability of our findings.
-However, our primary analysis focuses on the Mayo Clinic data due to its unique feature of random patient-physician assignment, which allows for stronger causal inference.
+\emph{Hypothesis 2: Batch ordering will have a significant impact on
+patient length of stay (LOS) in the ED, but the direction of this effect
+is ambiguous.} While batch ordering might reduce LOS by eliminating
+waiting times between test decisions (Kocher et al. 2012; Rosen and Seda
+2018), it could also increase LOS due to potential queuing
+inefficiencies and the time required to perform and interpret additional
+tests (Forster et al. 2003; Armony et al. 2015; Song et al. 2020).
 
-## Mayo Clinic Data
+\hypertarget{quality-of-care-and-return-visits}{%
+\subsubsection{Quality of Care and Return
+Visits}\label{quality-of-care-and-return-visits}}
 
-Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation.
-During the study period from October 6, 2018, through December 31, 2019, the ED recorded 48,854 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces.
-The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume.
-The data is summarized in Table \ref{tab:summary_stats}.
+\emph{Hypothesis 3: Batch ordering will influence the rate of 72-hour
+return visits with admission, but the direction of this effect is
+uncertain.} Comprehensive initial testing might reduce missed diagnoses
+and thus return visits (Pines et al. 2010; Sabbatini et al. 2016).
+However, it could also lead to false positives and unnecessary
+follow-ups, potentially increasing return visits (Welch, Schwartz, and
+Woloshin 2011; O'Connor and Nichol 2018).
+
+\hypertarget{hypothesis-4-heterogeneity-across-presenting-complaints}{%
+\subsubsection{Hypothesis 4: Heterogeneity Across Presenting
+Complaints}\label{hypothesis-4-heterogeneity-across-presenting-complaints}}
+
+\emph{Hypothesis 4: The effects of batch ordering on resource
+utilization, length of stay, and quality of care will vary significantly
+across different presenting complaints.} This hypothesis is based on the
+understanding that different clinical presentations have varying levels
+of diagnostic uncertainty and risk (Cooke et al. 2013; Johnson 2019),
+which may influence the appropriateness and impact of batch ordering.
+
+By testing these hypotheses, we aim to provide a comprehensive
+understanding of the impacts of batch ordering in the ED, informing both
+clinical practice and operational policy. Our empirical strategy,
+detailed in the following sections, is designed to rigorously test these
+hypotheses and uncover the causal effects of batch ordering on ED
+performance and patient outcomes.
+
+\hypertarget{data-and-methods}{%
+\section{Data and Methods}\label{data-and-methods}}
+
+Our study uses data from two large U.S. emergency departments (EDs): the
+Mayo Clinic of Arizona and Massachusetts General Hospital (MGH). The MGH
+dataset, which includes 129,489 patient encounters from November 10,
+2021 through December 10, 2022, provides a robust sample for validating
+the generalizability of our findings. However, our primary analysis
+focuses on the Mayo Clinic data due to its unique feature of random
+patient-physician assignment, which allows for stronger causal
+inference.
+
+\hypertarget{mayo-clinic-data}{%
+\subsection{Mayo Clinic Data}\label{mayo-clinic-data}}
+
+Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a
+tertiary care hospital without obstetrical services, an inpatient
+pediatrics unit, or a trauma designation. During the study period from
+October 6, 2018, through December 31, 2019, the ED recorded 48,854
+visits per year, managed across 26 treatment rooms and up to 9 hallway
+spaces. The department is exclusively staffed by board-eligible or
+board-certified emergency physicians (EPs), with rotating residents
+overseeing about 10\% of patient volume. The data is summarized in Table
+\ref{tab:summary_stats}.
 
@@
 
-A key feature of the Mayo Clinic ED is its rotational patient assignment system, in which patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses.
-In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+A key feature of the Mayo Clinic ED is its rotational patient assignment
+system, in which patients arriving at the Mayo Clinic ED are randomly
+assigned to physicians via a rotational patient assignment algorithm
+Traub et al. (2016), which removes potential selection bias concerns for
+our analyses. In essence, barring arrival time and shift-level
+variation, the physician-to-patient matching can be deemed random. Table
+\ref{tab:wald_test} displays that patient encounters (regarding chief
+complaints and emergency severity) are equitably distributed across
+physicians within our study's cohort.
 
@@
 
-We conducted a retrospective review of comprehensive ED operational data, coinciding with the initiation of a new electronic medical record.
-The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics.
-This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.
+We conducted a retrospective review of comprehensive ED operational
+data, coinciding with the initiation of a new electronic medical record.
+The dataset includes detailed patient demographics, chief complaints,
+vital signs, emergency severity index (ESI), length of stay (LOS), and
+resource utilization metrics. This period was chosen to provide a robust
+data set while excluding the influence of the coronavirus pandemic.
 
-## Sample Construction
+\hypertarget{sample-construction}{%
+\subsection{Sample Construction}\label{sample-construction}}
 
-Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
-We observe $48,854$ such visits during the study period.
-To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5$ percent of the time.
-Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them.
-Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints.
-Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories as we do.
-In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $15,821$ encounters involving full-time physicians who treat over 500 ED cases per year.
+Our research design focuses on adults who visit the Mayo Clinic of
+Arizona ED. We observe \(48,854\) such visits during the study period.
+To improve power, we drop encounters with rare chief complaints
+(\(<1000\) total encounters of this kind) and complaints where a batch
+order occurs less than \(5\) percent of the time. Since batch orders are
+rare for these cases, our physician batch tendency instrument could
+suffer from a weak instrument problem if we included them. Examples of
+complaints dropped include Upper Respiratory Symptoms and Urinary
+Complaints. Excluding these conditions does not introduce selection bias
+unless physician test batching tendency is orthogonal to physician
+diagnosing behavior. While this assumption may be violated if we were to
+use a very detailed level of chief complaint information upon which to
+base our exclusion criterion, it is plausibly satisfied when using broad
+complaint categories as we do. In order to estimate a precise measure of
+physician-level batch tendency, we further restrict our sample to the
+\(15,821\) encounters involving full-time physicians who treat over 500
+ED cases per year.
 
-## Variable Definitions
+\hypertarget{variable-definitions}{%
+\subsection{Variable Definitions}\label{variable-definitions}}
 
-Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter.
-While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice.
-We define "batching" in line with standard emergency medicine practices.
-Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses.
-This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.
+Our explanatory variable in the IV analysis, \(Batched_i\), is an
+indicator for whether patient \(i\) has their tests batch-ordered at
+their ED encounter. While the patient could decide not to undergo the
+tests ordered by the physician, this is rare in practice. We define
+``batching'' in line with standard emergency medicine practices.
+Batching occurs when a physician simultaneously orders a comprehensive
+set of diagnostic tests, typically covering a broad range of potential
+diagnoses. This contrasts with sequential ordering, where tests are
+ordered sequentially based on the information obtained from each test as
+needed.
 
-We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window.
-In Section [...], sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study.
-Therefore, a batch in our study consists of two or more distinct imaging tests.
+We operationalize batching as occurring when multiple diagnostic imaging
+tests are ordered within a 5-minute window. In Section {[}\ldots{]},
+sensitivity analyses on this cutoff point showed that our results are
+robust to this definition. Each imaging test (e.g., X-ray, CT scan,
+Ultrasound) is considered a separate, distinct test for our study.
+Therefore, a batch in our study consists of two or more distinct imaging
+tests.
 
@@
 
-*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
-It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
+\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
+efficiency and patient throughput in emergency care settings. It is
+defined as the duration from a patient's arrival to the ED until their
+departure, whether by discharge or admission to the hospital.
 
-*(b) Number of Unique Imaging Tests*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives.
-In this study, we quantify resource utilization by the total number of distinct imaging tests ordered per patient during their ED stay.
-This encompasses both initial and subsequent tests.
+\emph{(b) Number of Unique Imaging Tests}-- Resource utilization in the
+ED typically refers to the extent of medical services and interventions
+a patient receives. In this study, we quantify resource utilization by
+the total number of distinct imaging tests ordered per patient during
+their ED stay. This encompasses both initial and subsequent tests.
 
-*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
-This outcome is used to assess the quality of care provided during the initial ED visit.
+\emph{(c) 72 Hour Return with Admission}-- This is a binary variable
+indicating whether the patient returns to the ED within 72 hours of
+their initial visit and is admitted to the hospital. This outcome is
+used to assess the quality of care provided during the initial ED visit.
 
-## Empirical Strategy
+\hypertarget{empirical-strategy}{%
+\subsection{Empirical Strategy}\label{empirical-strategy}}
 
-Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court.
-Similarly, we explore batching variation across physicians who work in the same emergency department.
-In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians.
-Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision.
-We employ both approaches and lay out their details in the next subsections.
+Our empirical strategy closely follows the literature that relies on
+quasi-random assignment of agents to cases, often referred to as the
+``judges design.'' Papers in this literature typically exploit variation
+in the sentencing leniency of judges who work in the same court.
+Similarly, we explore batching variation across physicians who work in
+the same emergency department. In its reduced form, under the assumption
+of quasi-random assignment, this approach allows researchers to identify
+the causal effect of being assigned to different types of physicians.
+Under additional assumptions, an instrumental variable approach
+identifies the causal effect of a given medical decision. We employ both
+approaches and lay out their details in the next subsections.
 
-### Batch Tendency Construction
+\hypertarget{batch-tendency-construction}{%
+\subsubsection{Batch Tendency
+Construction}\label{batch-tendency-construction}}
 
-To measure physician batch tendency, we use the physician's residualized leave-out average batch rate.
-This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways.
-First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.
+To measure physician batch tendency, we use the physician's residualized
+leave-out average batch rate. This measure is derived from two steps
+following the approaches taken by Doyle et al. (2015), Dobbie, Goldin,
+and Yang (2018), and Eichmeyer and Zhang (2022). First, we obtain
+residuals from a regression model, which includes all ED encounters in
+our sample period.
 
@@
 
-Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$.
-Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g.
-initiatives to eliminate excess testing) or seasonality in ED visits.
-We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$.
-Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision.
-Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case.
-As stated earlier, these controls are more than what is required for our quasi-random assignment assumption.
-Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.
+Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
+\(i\) had their imaging tests batch ordered on encounter that took place
+on date \(t\). Fixed effects include year-month fixed effects,
+\(\alpha_{ym}\), to control for time and seasonal variation in batching,
+such as hospital-specific policies (e.g.~initiatives to eliminate excess
+testing) or seasonality in ED visits. We also control for
+``shift-level'' variations that include both physician scheduling and
+patient arrival with day of week-time of day fixed effects,
+\(\alpha_{dt}\). Chief complaint by severity fixed effects,
+\(\alpha_{complaint}\), were also included to increase precision.
+Finally, a binary variable for whether or not laboratory tests were
+ordered, \(\alpha_{lab}\), was included to account for the complexity of
+the case. As stated earlier, these controls are more than what is
+required for our quasi-random assignment assumption. Under the
+assumption that we have captured the observables under which
+quasi-random assignment occurs in the ED, the unexplained variation--
+the physician's contribution-- resides in the error term,
+\(\varepsilon_{i,t}\).
 
-In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:
+In step two, the tendency measure for patient \(i\) seen by physician
+\(j\) is computed as the average residual across all other patients seen
+by the physician that year:
 
@@
 
-where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$.
-This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument.
-The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.
+where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
+residual from equation (1); \(J\) is the set of all ED encounters
+treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
+number of cases that physician has seen that year, excluding patient
+\(i\). This leave-out mean eliminates the mechanical bias that stems
+from patient \(i\)'s own case entering into the instrument. The measure
+is interpreted as the average (leave-out) batch rate of patient \(i\)'s
+physician, relative to other physicians in that hospital-year-month,
+hospital-day of week-time of day.
 
-We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order imaging tests.
-Table \ref{table:first_stage} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $6$pp increase in the likelihood of having tests batch-ordered in the ED.
-The F-statistic is 94 when all controls and fixed effects are included.
-The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
+We document that the Mayo Clinic ED physicians exhibit wide, systematic
+variation in their propensity to batch order imaging tests. Table
+\ref{table:first_stage} presents the ``first stage'' in a regression
+table: being assigned to a \(10\)pp higher batch-tendency physician is
+associated with a \(6\)pp increase in the likelihood of having tests
+batch-ordered in the ED. The F-statistic is 94 when all controls and
+fixed effects are included. The coefficient is greater than one because
+all emergency visits are used to construct the tendency instrument,
+while the first stage is calculated using the baseline sample only,
+which excludes the rare complaints.
 
@@
 
-To estimate the reduced-form effects of being treated by a batch-preferring physician, we estimate the following equation:
+To estimate the reduced-form effects of being treated by a
+batch-preferring physician, we estimate the following equation:
 
@@
 
-Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour return with admission, and number of imaging tests ordered for the patient, and $X_i$ is the same as in the reduced-form approach.
-$Batched_i$ variable suffers from potential endogeneity concerns.
-For example, injury severity may be unobserved and correlated with need to run multiple tests, length of stay, and return with admission likelihood.
-Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$.
-We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.
+Where \(Y_i\) represents our main outcomes of interest: length of stay,
+72 hour return with admission, and number of imaging tests ordered for
+the patient, and \(X_i\) is the same as in the reduced-form approach.
+\(Batched_i\) variable suffers from potential endogeneity concerns. For
+example, injury severity may be unobserved and correlated with need to
+run multiple tests, length of stay, and return with admission
+likelihood. Hence, we instrument \(Batched_i\) with the assigned
+physician \(j\)'s underlying tendency to batch,
+\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
+physician level to account for the assignment process of patients to
+physicians.
 
-## Identifying Assumptions
+\hypertarget{identifying-assumptions}{%
+\subsection{Identifying Assumptions}\label{identifying-assumptions}}
 
-The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence").
-The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.
+The reduced-form approach delivers an unbiased estimate of the causal
+effect of being treated by a higher tendency to batch physician, since
+assignment of patients to ED physicians is random, conditional on
+seasonality and shift (``conditional independence''). The
+residualization in equation (1) controls for more controls than required
+to achieve quasi-random assignment; they are included for statistical
+precision in measuring physician tendency to batch.
 
-Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity.
-We reported a strong first stage (i.e., relevance) at the end of the previous Section.
-The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching.
-This is perhaps our strongest assumption and is at its core, untestable.
-However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings.
-First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher.
-Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest.
-The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis.
-Nevertheless, we take this assumption seriously and perform a placebo check in Section [...] as well as various robustness checks in Section [...].
+Our instrumental variable approach, which aims to recover the causal
+effect of having diagnostic tests batch ordered, relies on three
+additional assumptions: relevance, exclusion, and monotonicity. We
+reported a strong first stage (i.e., relevance) at the end of the
+previous Section. The exclusion restriction requires that the instrument
+must influence the outcome of interest only through its effect on test
+batching. This is perhaps our strongest assumption and is at its core,
+untestable. However, several features of the ED setting suggest that
+such violation may likely only have a small impact and may be less
+concerning than in other health care settings. First, unlike in primary
+care settings, where the patient and primary care provider have many
+repeat encounters, the scope of what the emergency physician can do to
+impact medium-term outcomes is limited and well-observed by the
+researcher. Second, any violation of the exclusion restriction needs to
+directly affect the specific outcome of interest. The channel by which
+ED physicians can influence length of stay relative outcomes is likely
+through testing and diagnosis. Nevertheless, we take this assumption
+seriously and perform a placebo check in Section {[}\ldots{]} as well as
+various robustness checks in Section {[}\ldots{]}.
 
-Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects.
-It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician.
-The literature leveraging the judges design typically performs two informal tests for its implications.
-The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects).
-The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration).
-Appendix Table  [...] presents both of these tests in the two columns for various subsamples of interest.
+Finally, the monotonicity assumption is necessary for interpreting the
+coefficient estimates obtained from the IV approach as Local Average
+Treatment Effects (LATEs) if there are heterogeneous treatment effects.
+It requires that any patient who is (not) batched by a sequencer
+(batcher) would also (not) be batched by a batcher (sequencer)
+physician. The literature leveraging the judges design typically
+performs two informal tests for its implications. The first one provides
+that the first stage should be weakly positive for all subsamples
+(Dobbie, Goldin, and Yang (2018)). The second implication asserts that
+the instrument constructed by leaving out a particular subsample has
+predictive power over that same left-out subsample (Bhuller et al.
+(2020)). Appendix Table {[}\ldots{]} presents both of these tests in the
+two columns for various subsamples of interest.
 
-# Results {#sec:4}
+\hypertarget{sec:4}{%
+\section{Results}\label{sec:4}}
 
-## Reduced-Form Results
+\hypertarget{reduced-form-results}{%
+\subsection{Reduced-Form Results}\label{reduced-form-results}}
 
-In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department.
-We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination.
-Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.
+In this section, we explore the causal influence of physician batch
+tendency on patient outcomes and resource utilization in the emergency
+department. We posit that while batch tendency directly influences the
+practice of batch ordering tests, both batch tendency and batch ordering
+are concurrently influenced by a physician's testing inclination. Given
+that testing inclination directly affects primary outcomes, we include
+it as a control variable in our regression models to mitigate its
+confounding effects.
 
-To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency.
-Specifically, we calculate the physician's residualized leave-out average test rate, which serves as a proxy for their propensity to order tests.
-It is important to note the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
-This substantial correlation suggests that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies.
-Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.
+To quantify a physician's testing inclination, we employ a similar
+approach to that used in measuring physician batch tendency.
+Specifically, we calculate the physician's residualized leave-out
+average test rate, which serves as a proxy for their propensity to order
+tests. It is important to note the strong positive correlation
+(\(r = 0.79\)) between batch tendency and testing inclination. This
+substantial correlation suggests that physicians with a higher
+propensity to test may also exhibit a higher tendency to batch orders,
+potentially as a consequence of their testing strategies. Neglecting to
+account for testing inclination could lead to overestimated effects of
+batch tendency due to omitted variable bias.
 
-Reduced-form regression results in Table \ref{table:reduced_form} reveal the effect the average total effect of batch tendency on resource utilization and LOS.
-However, it is important to note that the estimated total effect of batch tendency accounts for both direct and indirect (mediated) effects of batch tendency.
-We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered.
-The increased testing volume, in turn, is associated with an extended LOS.
+Reduced-form regression results in Table \ref{table:reduced_form} reveal
+the effect the average total effect of batch tendency on resource
+utilization and LOS. However, it is important to note that the estimated
+total effect of batch tendency accounts for both direct and indirect
+(mediated) effects of batch tendency. We hypothesize that while batch
+ordering may streamline the testing process, resulting in a quicker
+completion of a given number of tests, it simultaneously appears to lead
+to an increase in the total number of tests ordered. The increased
+testing volume, in turn, is associated with an extended LOS.
 
@@
 
-## Placebo Check
+\hypertarget{placebo-check}{%
+\subsection{Placebo Check}\label{placebo-check}}
 
-In this section we investigate whether the reduced-form effects observed in Section [...] are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency.
-We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
-For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
-Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.
+In this section we investigate whether the reduced-form effects observed
+in Section {[}\ldots{]} are due to differences in batch rates across
+providers or due to other provider differences correlated with batch
+tendency. We start by studying reduced-form effects among patients with
+complaints that are never batched, as a ``placebo/falsification check.''
+By way of example, consider a patient who arrives at the ED with a
+urinary tact infection---a condition for which patients rarely undergo
+imaging testing. For such patients, we should expect to see no impact of
+batch tendency only if high batching and low batching physicians do not
+systematically differ in other dimensions of care relevant to patient
+outcomes. Conversely, if we do find a reduced-form effect for these
+patients, then high batch tendency physicians must systematically differ
+from low batch tendency physicians in other dimensions of care, beyond
+batching.
 
-To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a >5 percent batching rate).
-We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3).
-The results of this exercise are displayed in Appendix Table [...].
-They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
+To that end, we restrict attention to ED visits for complaints where
+batching occurs no more than 5 percent of the time (recall that our
+baseline sample only includes complaints with a \textgreater5 percent
+batching rate). We estimate a reduced-form regression of each main
+outcome on physician batch tendency for the subsample, following
+equation (3). The results of this exercise are displayed in Appendix
+Table {[}\ldots{]}. They show that in contrast to results for our main
+sample, the association between physician tendency to batch and a given
+outcome is statistically indistinguishable from zero and much smaller in
+magnitude for the samples of patients who visit the ED with health
+conditions that are rarely batched.
 
-## Instrumental Variables Results
+\hypertarget{instrumental-variables-results}{%
+\subsection{Instrumental Variables
+Results}\label{instrumental-variables-results}}
 
-To uncover the causal effect of test batching on key emergency department (ED) outcomes, we utilize a two-stage least squares (2SLS) approach, leveraging physician batch-ordering tendencies as an instrument.
-This strategy isolates the local average treatment effect (LATE) for patients whose diagnostic tests were batched due to their physician’s inclination to batch, rather than patient characteristics or other unobserved factors that may influence outcomes.
-The IV estimates capture the effect of test batching on ED length of stay, the number of imaging tests ordered, and the likelihood of a return visit within 72 hours with admission.
+To uncover the causal effect of test batching on key emergency
+department (ED) outcomes, we utilize a two-stage least squares (2SLS)
+approach, leveraging physician batch-ordering tendencies as an
+instrument. This strategy isolates the local average treatment effect
+(LATE) for patients whose diagnostic tests were batched due to their
+physician's inclination to batch, rather than patient characteristics or
+other unobserved factors that may influence outcomes. The IV estimates
+capture the effect of test batching on ED length of stay, the number of
+imaging tests ordered, and the likelihood of a return visit within 72
+hours with admission.
 
@@
 
-*ED Length of Stay*—The IV results in column (1) of Table X indicate that test batching has a negative but statistically insignificant effect on ED length of stay.
-The coefficient of -0.504 suggests a potential reduction in time spent in the ED, but the confidence interval is wide, indicating that the effect may not be robust across different specifications.
-This finding implies that while batching may streamline certain diagnostic processes, other operational factors likely play a larger role in determining how long a patient remains in the ED.
-These results suggest that batching alone does not drastically impact the overall time patients spend in the ED, pointing to the need for further exploration of mechanisms that may moderate this relationship.
+\emph{ED Length of Stay}---The IV results in column (1) of Table X
+indicate that test batching has a negative but statistically
+insignificant effect on ED length of stay. The coefficient of -0.504
+suggests a potential reduction in time spent in the ED, but the
+confidence interval is wide, indicating that the effect may not be
+robust across different specifications. This finding implies that while
+batching may streamline certain diagnostic processes, other operational
+factors likely play a larger role in determining how long a patient
+remains in the ED. These results suggest that batching alone does not
+drastically impact the overall time patients spend in the ED, pointing
+to the need for further exploration of mechanisms that may moderate this
+relationship.
 
-*Number of Imaging Tests Ordered*—In column (2), we find a statistically significant and positive effect of test batching on the number of imaging tests ordered.
-The coefficient of 0.692 (p < 0.05) indicates that batching leads to approximately 69% more imaging tests being conducted during the patient’s ED visit relative to if that patient's tests were sequenced.
-This result highlights a key consequence of batching: the grouping of diagnostic orders leads to a more intensive use of imaging resources.
-This increased resource utilization, while contributing to more comprehensive diagnostic evaluations, also raises concerns about the potential for overutilization and the associated costs and patient burden.
-This effect is particularly relevant in resource-constrained environments, where the trade-off between efficiency and resource use becomes crucial.
+\emph{Number of Imaging Tests Ordered}---In column (2), we find a
+statistically significant and positive effect of test batching on the
+number of imaging tests ordered. The coefficient of 0.692 (p \textless{}
+0.05) indicates that batching leads to approximately 69\% more imaging
+tests being conducted during the patient's ED visit relative to if that
+patient's tests were sequenced. This result highlights a key consequence
+of batching: the grouping of diagnostic orders leads to a more intensive
+use of imaging resources. This increased resource utilization, while
+contributing to more comprehensive diagnostic evaluations, also raises
+concerns about the potential for overutilization and the associated
+costs and patient burden. This effect is particularly relevant in
+resource-constrained environments, where the trade-off between
+efficiency and resource use becomes crucial.
 
-*Return Visits with Admission within 72 Hours*—The results in column (3) show that test batching has a negative effect on the likelihood of a return visit with admission within 72 hours, although this effect is not statistically significant.
-The point estimate of -0.038 suggests a small potential reduction in return visits for patients whose tests were batched, which may reflect more thorough initial diagnostics leading to fewer missed diagnoses or complications.
-However, the lack of statistical significance means we cannot draw firm conclusions from this result.
-This finding aligns with previous research suggesting that while batching may improve certain diagnostic processes, its effect on downstream outcomes such as return visits may be more complex and contingent on factors beyond test ordering patterns.
+\emph{Return Visits with Admission within 72 Hours}---The results in
+column (3) show that test batching has a negative effect on the
+likelihood of a return visit with admission within 72 hours, although
+this effect is not statistically significant. The point estimate of
+-0.038 suggests a small potential reduction in return visits for
+patients whose tests were batched, which may reflect more thorough
+initial diagnostics leading to fewer missed diagnoses or complications.
+However, the lack of statistical significance means we cannot draw firm
+conclusions from this result. This finding aligns with previous research
+suggesting that while batching may improve certain diagnostic processes,
+its effect on downstream outcomes such as return visits may be more
+complex and contingent on factors beyond test ordering patterns.
 
-### Mechanism: Mediation Analysis
+\hypertarget{mechanism-mediation-analysis}{%
+\subsubsection{Mechanism: Mediation
+Analysis}\label{mechanism-mediation-analysis}}
 
-To further explore the mechanisms through which batching influences ED length of stay (LOS) and return visits with admission within 72 hours, we conducted a formal mediation analysis.
-Specifically, we assessed whether the number of imaging tests mediates the relationship between batching and these outcomes.
-This analysis helps to decompose the total effect of batching into direct and indirect effects, with the indirect effect capturing the mediation through imaging tests.
+To further explore the mechanisms through which batching influences ED
+length of stay (LOS) and return visits with admission within 72 hours,
+we conducted a formal mediation analysis. Specifically, we assessed
+whether the number of imaging tests mediates the relationship between
+batching and these outcomes. This analysis helps to decompose the total
+effect of batching into direct and indirect effects, with the indirect
+effect capturing the mediation through imaging tests.
 
-The mediation analysis was performed using a quasi-Bayesian approach with 1,000 simulations, adjusting for relevant covariates, including physician testing inclination, imaging use, day of the week, month of the year, patient complaints, and laboratory performance.
-The results for both outcomes are summarized in Table \ref{tab:med_results}, which presents the average causal mediation effect (ACME), average direct effect (ADE), and total effect for each outcome.
+The mediation analysis was performed using a quasi-Bayesian approach
+with 1,000 simulations, adjusting for relevant covariates, including
+physician testing inclination, imaging use, day of the week, month of
+the year, patient complaints, and laboratory performance. The results
+for both outcomes are summarized in Table \ref{tab:med_results}, which
+presents the average causal mediation effect (ACME), average direct
+effect (ADE), and total effect for each outcome.
 
@@
 
-\textbf{Incorporating Contemporary Mediation Frameworks.}—In line with the classical literature on mediation (Baron and Kenny, 1986), mediation analysis traditionally proceeds in three steps: (1) testing the effect of the independent variable (batch order) on the dependent variable (LOS/72-hour return) without the mediator (imaging volume); (2) adding the mediator (imaging volume) to the model and testing the direct effect of the independent variable on the dependent variable; and (3) testing the mediator’s effect on the dependent variable in the presence of the independent variable.
+\textbf{Incorporating Contemporary Mediation Frameworks.}---In line with
+the classical literature on mediation (Baron and Kenny, 1986), mediation
+analysis traditionally proceeds in three steps: (1) testing the effect
+of the independent variable (batch order) on the dependent variable
+(LOS/72-hour return) without the mediator (imaging volume); (2) adding
+the mediator (imaging volume) to the model and testing the direct effect
+of the independent variable on the dependent variable; and (3) testing
+the mediator's effect on the dependent variable in the presence of the
+independent variable.
 
-However, recent literature has demonstrated that significant results in the zero-order effect (i.e., the total effect) are not required to justify moving forward with mediation analysis.
-Zhao et al.
-(2010) and Hayes (2022) argue that requiring significance in each step is an overly restrictive and outdated framework, which often overlooks indirect effects that can be meaningful even when direct effects are not.
-In our analysis, although we do not always find statistically significant direct effects for batching on LOS or return visits, the mediation pathway through imaging volume remains relevant for understanding the indirect effects of batching.
+However, recent literature has demonstrated that significant results in
+the zero-order effect (i.e., the total effect) are not required to
+justify moving forward with mediation analysis. Zhao et al.~(2010) and
+Hayes (2022) argue that requiring significance in each step is an overly
+restrictive and outdated framework, which often overlooks indirect
+effects that can be meaningful even when direct effects are not. In our
+analysis, although we do not always find statistically significant
+direct effects for batching on LOS or return visits, the mediation
+pathway through imaging volume remains relevant for understanding the
+indirect effects of batching.
 
-Thus, following the contemporary guidance in mediation analysis, we proceed with testing both direct and indirect effects simultaneously, allowing us to capture the more nuanced dynamics of how batching influences ED outcomes.
-This approach provides a richer understanding of the operational impacts of batching, particularly as it pertains to the trade-offs between diagnostic efficiency and resource utilization.
+Thus, following the contemporary guidance in mediation analysis, we
+proceed with testing both direct and indirect effects simultaneously,
+allowing us to capture the more nuanced dynamics of how batching
+influences ED outcomes. This approach provides a richer understanding of
+the operational impacts of batching, particularly as it pertains to the
+trade-offs between diagnostic efficiency and resource utilization.
 
-\textbf{Mediation Results for ED Length of Stay.}—The mediation analysis reveals that the indirect effect (ACME) of batching on ED length of stay through the number of imaging tests is $0.083 (p < 0.01)$, suggesting that the increased volume of diagnostic testing marginally extends patient stays in the ED.
-The direct effect (ADE) of batching on length of stay is negative but not statistically significant ($-0.550, p = 0.71$), and the total effect is also negative ($-0.468$) but insignificant.
-This indicates that while batching may reduce LOS, the effect is partially offset by the increased testing, leading to no substantial impact on overall ED throughput.
+\textbf{Mediation Results for ED Length of Stay.}---The mediation
+analysis reveals that the indirect effect (ACME) of batching on ED
+length of stay through the number of imaging tests is
+\(0.083 (p < 0.01)\), suggesting that the increased volume of diagnostic
+testing marginally extends patient stays in the ED. The direct effect
+(ADE) of batching on length of stay is negative but not statistically
+significant (\(-0.550, p = 0.71\)), and the total effect is also
+negative (\(-0.468\)) but insignificant. This indicates that while
+batching may reduce LOS, the effect is partially offset by the increased
+testing, leading to no substantial impact on overall ED throughput.
 
-\textbf{Mediation Results for 72-Hour Return Visits.}—For return visits within 72 hours, the indirect effect through the number of imaging tests is negative ($-0.0029, p < 0.05$), indicating that the additional testing associated with batching may slightly reduce the likelihood of a return visit.
-However, the direct effect ($-0.036$) and total effect ($-0.039$) of batching on return visits are small and not statistically significant, suggesting that while comprehensive diagnostics may reduce return visits, the effect of batching on this outcome is generally weak.
+\textbf{Mediation Results for 72-Hour Return Visits.}---For return
+visits within 72 hours, the indirect effect through the number of
+imaging tests is negative (\(-0.0029, p < 0.05\)), indicating that the
+additional testing associated with batching may slightly reduce the
+likelihood of a return visit. However, the direct effect (\(-0.036\))
+and total effect (\(-0.039\)) of batching on return visits are small and
+not statistically significant, suggesting that while comprehensive
+diagnostics may reduce return visits, the effect of batching on this
+outcome is generally weak.
 
-\textbf{Conceptual Framework: Directed Acyclic Graph (DAG).}—Based on the findings from the mediation analysis, we propose a conceptual framework to illustrate the relationships uncovered by our empirical results.
-The Directed Acyclic Graph (DAG) shown in Figure \ref{fig:dag} provides a visual representation of the causal pathways between batch tendency (Z), batch ordering (X), imaging volume (M), and our outcomes (Y).
-The graph indicates that the indirect effect of batching operates through the volume of imaging tests, while the direct effect of batching on the outcomes bypasses this mediator.
+\textbf{Conceptual Framework: Directed Acyclic Graph (DAG).}---Based on
+the findings from the mediation analysis, we propose a conceptual
+framework to illustrate the relationships uncovered by our empirical
+results. The Directed Acyclic Graph (DAG) shown in Figure \ref{fig:dag}
+provides a visual representation of the causal pathways between batch
+tendency (Z), batch ordering (X), imaging volume (M), and our outcomes
+(Y). The graph indicates that the indirect effect of batching operates
+through the volume of imaging tests, while the direct effect of batching
+on the outcomes bypasses this mediator.
 
@@
 
-In this DAG, batch tendency (Z) is used as an instrumental variable to predict batch order (X), which in turn affects both imaging volume (M) and our primary outcomes, including ED length of stay and 72-hour return with admission (Y).
-The unobserved confounding (U) is assumed to influence both imaging volume and the outcomes directly.
-The mediation analysis supports this structure, as it confirms that imaging volume mediates the relationship between batch ordering and ED outcomes, albeit modestly.
-By identifying this framework through our analysis, we provide a data-driven explanation for how batching influences ED operations.
+In this DAG, batch tendency (Z) is used as an instrumental variable to
+predict batch order (X), which in turn affects both imaging volume (M)
+and our primary outcomes, including ED length of stay and 72-hour return
+with admission (Y). The unobserved confounding (U) is assumed to
+influence both imaging volume and the outcomes directly. The mediation
+analysis supports this structure, as it confirms that imaging volume
+mediates the relationship between batch ordering and ED outcomes, albeit
+modestly. By identifying this framework through our analysis, we provide
+a data-driven explanation for how batching influences ED operations.
 
-\textbf{Interpretation of Mediation Results.}—These results show that the number of imaging tests serves as a partial mediator in the relationship between batching and ED outcomes.
-For length of stay, increased testing tends to prolong ED visits, but the overall effect of batching on LOS is negligible.
-For return visits, the increase in diagnostic testing slightly reduces the likelihood of a return within 72 hours, though the overall effect remains small and insignificant.
-These findings highlight the trade-offs involved in batching diagnostic orders: while batching may streamline certain processes, the associated increase in testing has modest effects on patient flow and outcomes.
-
-### Time to Result
+\textbf{Interpretation of Mediation Results.}---These results show that
+the number of imaging tests serves as a partial mediator in the
+relationship between batching and ED outcomes. For length of stay,
+increased testing tends to prolong ED visits, but the overall effect of
+batching on LOS is negligible. For return visits, the increase in
+diagnostic testing slightly reduces the likelihood of a return within 72
+hours, though the overall effect remains small and insignificant. These
+findings highlight the trade-offs involved in batching diagnostic
+orders: while batching may streamline certain processes, the associated
+increase in testing has modest effects on patient flow and outcomes.
 
@@
 
-# References
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-akhlaghpour2020value}{}}%
+Akhlaghpour, Hessameddin. 2020. {``The Value of Information in
+Sequential Decision Making.''} \emph{Proceedings of the National Academy
+of Sciences} 117 (48): 30925--31.
+
+\leavevmode\vadjust pre{\hypertarget{ref-armony2015heterogeneous}{}}%
+Armony, Mor, Shlomo Israelit, Avishai Mandelbaum, Yariv N Marmor, Yulia
+Tseytlin, and Galit B Yom-Tov. 2015. {``On Patient Flow in Hospitals: A
+Data-Based Queueing-Science Perspective.''} \emph{Stochastic Systems} 5
+(1): 146--94.
+
+\leavevmode\vadjust pre{\hypertarget{ref-asplin2003model}{}}%
+Asplin, Brent R, David J Magid, Karin V Rhodes, Leif I Solberg, Nicole
+Lurie, and Carlos A Camargo Jr. 2003. {``A Conceptual Model of Emergency
+Department Crowding.''} \emph{Annals of Emergency Medicine} 42 (2):
+173--80.
+
+\leavevmode\vadjust pre{\hypertarget{ref-naseim2015}{}}%
+Balogh, EP, BT Miller, and JR Ball. 2015. \emph{Improving Diagnosis in
+Health Care}. Edited by Committee on Diagnostic Error in Health Care,
+Board on Health Care Services, Institute of Medicine, The National
+Academies of Sciences Engineering, and Medicine. Washington (DC):
+National Academies Press (US).
+
+\leavevmode\vadjust pre{\hypertarget{ref-batt2019impact}{}}%
+Batt, Robert J, and Christian Terwiesch. 2019. {``The Impact of Process
+Design on ED Performance: A Discreteevent Simulation Study.''}
+\emph{Production and Operations Management} 28 (6): 1493--1510.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bernstein2009}{}}%
+Bernstein, SL, D Aronsky, R Duseja, S Epstein, D Handel, U Hwang, M
+McCarthy, et al. 2009. {``The Effect of Emergency Department Crowding on
+Clinically Oriented Outcomes.''} \emph{Academic Emergency Medicine} 16
+(1): 1--10.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bernstein2009effect}{}}%
+Bernstein, Steven L, Dominik Aronsky, Reena Duseja, Stephen Epstein, Dan
+Handel, Ula Hwang, Melissa McCarthy, et al. 2009. {``The Effect of
+Emergency Department Crowding on Clinically Oriented Outcomes.''}
+\emph{Academic Emergency Medicine} 16 (1): 1--10.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bhuller2020incarceration}{}}%
+Bhuller, Manudeep, Gordon B Dahl, Katrine V Løken, and Magne Mogstad.
+2020. {``Incarceration, Recidivism, and Employment.''} \emph{Journal of
+Political Economy} 128 (4): 1269--1324.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cooke2013overview}{}}%
+Cooke, Georga, Amanda Tapley, Elizabeth Holliday, Simon Morgan, Kim
+Henderson, Jean Ball, Mieke van Driel, Neil Spike, Rohan Kerr, and
+Parker Magin. 2013. {``An Overview of Clinical Decision Making: Biases
+and Guidelines.''} \emph{BMJ} 346: f1239.
+
+\leavevmode\vadjust pre{\hypertarget{ref-croskerry2013cognitive}{}}%
+Croskerry, Pat. 2013. {``From Mindless to Mindful Practice---Cognitive
+Bias and Clinical Decision Making.''} \emph{New England Journal of
+Medicine} 368 (26): 2445--48.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dai2010load}{}}%
+Dai, Jim, and Shuangchi He. 2010. {``Load Balancing and Capacity
+Planning in Queueing Networks with Abandonment.''} \emph{Operations
+Research} 58 (2): 429--39.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dixit1994investment}{}}%
+Dixit, Avinash K, and Robert S Pindyck. 1994. \emph{Investment Under
+Uncertainty}. Princeton university press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dobbie2018effects}{}}%
+Dobbie, Will, Jacob Goldin, and Crystal S. Yang. 2018. {``The Effects of
+Pretrial Detention on Conviction, Future Crime, and Employment: Evidence
+from Randomly Assigned Judges.''} \emph{American Economic Review} 108
+(2): 201--40.
+
+\leavevmode\vadjust pre{\hypertarget{ref-doyle2015measuring}{}}%
+Doyle, Joseph, John Graves, Jonathan Gruber, and Samuel Kleiner. 2015.
+{``Measuring Returns to Hospital Care: Evidence from Ambulance Referral
+Patterns.''} \emph{Journal of Political Economy} 123 (1): 170--214.
+\url{https://doi.org/10.1086/677756}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eichmeyer2022pathways}{}}%
+Eichmeyer, Sarah, and Jonathan Zhang. 2022. {``Pathways into Opioid
+Dependence: Evidence from Practice Variation in Emergency
+Departments.''} \emph{American Economic Journal: Applied Economics} 14
+(4): 271--300. \url{https://doi.org/10.1257/app.20210048}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-forster2003effect}{}}%
+Forster, Alan J, Ian Stiell, George Wells, Alexander J Lee, and Carl Van
+Walraven. 2003. {``The Effect of Hospital Occupancy on Emergency
+Department Length of Stay and Patient Disposition.''} \emph{Academic
+Emergency Medicine} 10 (2): 127--33.
+
+\leavevmode\vadjust pre{\hypertarget{ref-garg2018value}{}}%
+Garg, Ashish, Yingjie Wang, and Ricardo Valerdi. 2018. {``The Value of
+Options in Emergency Response Management.''} \emph{Systems Engineering}
+21 (4): 344--59.
+
+\leavevmode\vadjust pre{\hypertarget{ref-green2006managing}{}}%
+Green, Linda V. 2006. {``Managing Patient Demand in Healthcare
+Systems.''} \emph{Encyclopedia of Health Economics} 2: 33--39.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jarvis2016ed}{}}%
+Jarvis, Paul RK. 2016. {``ED Overcrowding and the Full Capacity Protocol
+Cross over Study: What Happens When a Hospital Is over Capacity?''}
+\emph{Emergency Medicine Journal} 33 (2): 115--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-johnson2019decision}{}}%
+Johnson, Gordon. 2019. {``Decision Making in Emergency Medicine:
+Rational, Improvisational or Both?''} \emph{Emergency Medicine
+Australasia} 31 (4): 525--26.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kanzaria2015emergency}{}}%
+Kanzaria, Hemal K, Jerome R Hoffman, Marc A Probst, John P Caloyeras,
+Sandra H Berry, and Robert H Brook. 2015. {``Emergency Physician
+Perceptions of Medically Unnecessary Advanced Diagnostic Imaging.''}
+\emph{Academic Emergency Medicine} 22 (4): 390--98.
+
+\leavevmode\vadjust pre{\hypertarget{ref-koch2018}{}}%
+Koch, C, K Roberts, C Petruccelli, and DJ Morgan. 2018. {``The Frequency
+of Unnecessary Testing in Hospitalized Patients.''} \emph{Am J Med} 131
+(5): 500--503. \url{https://doi.org/10.1016/j.amjmed.2017.11.025}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kocher2012effect}{}}%
+Kocher, Keith E, William J Meurer, Jeffrey S Desmond, and Brahmajee K
+Nallamothu. 2012. {``Effect of Testing and Treatment on Emergency
+Department Length of Stay Using a National Database.''} \emph{Academic
+Emergency Medicine} 19 (5): 525--34.
+
+\leavevmode\vadjust pre{\hypertarget{ref-morley2018emergency}{}}%
+Morley, Claire, Maria Unwin, Gregory M Peterson, Jim Stankovich, and
+Leigh Kinsman. 2018. {``Emergency Department Crowding: A Systematic
+Review of Causes, Consequences and Solutions.''} \emph{PloS One} 13 (8):
+e0203316.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nibbelink2018emergency}{}}%
+Nibbelink, Christine W, and Barbara B Brewer. 2018. {``Emergency Nursing
+Decision Making: Experience, Expertise, and Cognitive Strategies.''}
+\emph{Journal of Emergency Nursing} 44 (4): 360--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-o2018overdiagnosis}{}}%
+O'Connor, Raymond E, and Graham Nichol. 2018. {``Overdiagnosis and
+Overtreatment: Generalists---It's Time for a Grassroots Revolution.''}
+\emph{BMJ} 362: k2820.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pines2010national}{}}%
+Pines, Jesse M, Peter M Mullins, Joshua K Cooper, Lisa B Feng, and Karin
+E Roth. 2010. {``National Trends in Emergency Department Use, Care
+Patterns, and Quality of Care of Older Adults in the United States.''}
+\emph{Journal of the American Geriatrics Society} 58 (12): 2281--87.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rosen2018quality}{}}%
+Rosen, Charlotte L, and Gilbert Seda. 2018. {``Quality Improvement in
+Acute Care: How to Improve Your Emergency Department Right Now.''}
+\emph{Emergency Medicine Clinics} 36 (4): 835--51.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sabbatini2016overzealous}{}}%
+Sabbatini, Amber K, Luciano M Prevedello, Ali S Raja, Michael F Fialkow,
+Aaron D Sodickson, and Ramin Khorasani. 2016. {``Overzealous Use of the
+CT Scanner: How an ED Medical Director Can Use Data to Change Physician
+Behavior.''} \emph{Emergency Radiology} 23 (2): 173--79.
+
+\leavevmode\vadjust pre{\hypertarget{ref-saghafian2015}{}}%
+Saghafian, S, G Austin, and SJ Traub. 2015. {``Operations
+Research/Management Contributions to Emergency Department Patient Flow
+Optimization: Review and Research Prospects.''} \emph{IIE Transactions
+on Healthcare Systems Engineering} 5 (2): 101--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shi2016models}{}}%
+Shi, Pengyi, Mabel C Chou, Jim G Dai, Dingfan Ding, and Jing Sim. 2016.
+{``Models and Insights for Hospital Inpatient Operations: Time-Dependent
+ED Boarding Time.''} \emph{Management Science} 62 (1): 1--28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-smith2013impact}{}}%
+Smith, Michael, Craig Feied, Jonathan Handler, Michael Gillam, and
+William Chunn. 2013. {``The Impact of Using Emergency Physicians in the
+ED.''} \emph{The American Journal of Emergency Medicine} 31 (1):
+154--60.
+
+\leavevmode\vadjust pre{\hypertarget{ref-song2020managing}{}}%
+Song, Hummy, Anita L Tucker, Rachel Graue, Sarah Moravick, and Jillian J
+Yang. 2020. {``Managing Hospital Inpatient Bed Capacity Through
+Partitioning Care into Focused Wings.''} \emph{Manufacturing \& Service
+Operations Management} 22 (1): 70--87.
+
+\leavevmode\vadjust pre{\hypertarget{ref-studdert2005defensive}{}}%
+Studdert, David M, Michelle M Mello, William M Sage, Catherine M
+DesRoches, Jordon Peugh, Kinga Zapert, and Troyen A Brennan. 2005.
+{``Defensive Medicine Among High-Risk Specialist Physicians in a
+Volatile Malpractice Environment.''} \emph{Jama} 293 (21): 2609--17.
+
+\leavevmode\vadjust pre{\hypertarget{ref-traub2016emergency}{}}%
+Traub, Stephen J, Cameron F Stewart, Reza Didehban, Adam C Bartley,
+Soroush Saghafian, Vernon D Smith, Scott M Silvers, Ryan LeCheminant,
+and Christopher A Lipinski. 2016. {``Emergency Department Rotational
+Patient Assignment.''} \emph{Annals of Emergency Medicine} 67 (2):
+206--15. \url{https://doi.org/10.1016/j.annemergmed.2015.07.008}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wang2019data}{}}%
+Wang, Jingui, Yanfeng Ouyang, Xiaobo Qu, and Jingshan Li. 2019.
+{``Data-Driven Patient Scheduling in Emergency Departments: A Hybrid
+Robust-Stochastic Approach.''} \emph{Flexible Services and Manufacturing
+Journal} 31 (4): 1174--1202.
+
+\leavevmode\vadjust pre{\hypertarget{ref-welch2011overdiagnosed}{}}%
+Welch, H Gilbert, Lisa Schwartz, and Steve Woloshin. 2011.
+\emph{Overdiagnosed: Making People Sick in the Pursuit of Health}.
+Beacon Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zychlinski2019discrete}{}}%
+Zychlinski, Natalia, Avishai Mandelbaum, Petar Momcilovic, and Izack
+Cohen. 2019. {``A Discrete-Event Simulation Model for Managing Urgent
+Care Units.''} \emph{Operations Research for Health Care} 23: 100202.
+
+\end{CSLReferences}
+
+\bibliographystyle{unsrt}
+\bibliography{references.bib}
+
+\end{document}

### 2024-09-09 MgmtSci/MgmtSci.Rmd
getting going

--- 2024-09-09 Batching Draft 8/Batching Draft 8.Rmd
@@
 ---
-title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
+journal: mnsc
+title: "Image Batching: Batch Ordering Advanced Imaging Tests in the Emergency Department"
+shorttitle: "Image Batching"
 authors:
-  - name: Jacob Jameson
-    department:  Interfaculty Initiative in Health Policy
-    affiliation: Harvard University
-    location: Cambridge, MA 
-  - name: Soroush Saghafian
-    department:  Harvard Kennedy School of Government
-    affiliation: Harvard University
-    location: Cambridge, MA 
-  - name: Robert Huckman
-    department:  Harvard Business School
-    affiliation: Harvard University
-    location: Boston, MA 
-  - name: Nicole Hodgson
-    department:  Emergency Department
-    affiliation: Mayo Clinic
-    location: Phoenix, AZ
+- familyname: Jameson
+  othernames: Jacob
+  affiliation: Harvard Kennedy School, Harvard University
+  email: jacobjameson@g.harvard.edu
+- familyname: Soroush
+  othernames: Saghafian
+  affiliation: Harvard Kennedy School, Harvard University
+- familyname: Huckman
+  othernames:  Robert
+  affiliation: Harvard Business School, Harvard University
+- familyname: Hodgson
+  othernames:  Nicole
+  affiliation: Mayo Clinic, Emergency Department
+keywords:
+  - 3 to 6 keywords
+  - that do not appear in the title
 abstract: |
-
-keywords:
-
+  Abstract goes here
+acknowledgment: |
+  The authors would like to thank ...
 bibliography: references.bib
-biblio-style: unsrt
-output: rticles::arxiv_article
+blind: false
+output: rticles::informs_article
 header-includes:
@@
 
-Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization.
-Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events.
-Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality [@bernstein2009].
+Emergency departments (EDs) are high-pressure environments which globally face escalating challenges such as overcrowding, resource limitations, and increasing patient volumes [@mostafa2024strategies, @Sorup2013evaluation].
+These challenges can often be exacerbated by the need to provide timely and accurate diagnoses to patients, which often requires the use of advanced imaging tests such as computed tomography (CT) scans, ultrasound, X-ray, and magnetic resonance imaging (MRI) [@waheed2022impact].
+The management of diagnostic testing is a critical component of operational performance in EDs, where timely and accurate diagnoses can significantly impact patient outcomes [@committee2015improving].
+In many EDs, however, physicians have considerable discretion over how they order these tests, particularly with regard to advanced imaging [@valtchinov2019use].
+Yet little is known about the factors driving physicians' decisions to exercise such discretion and how test ordering should be managed when this discretion exists.
+In this paper, we explore the causal effects of batch ordering advanced imaging tests, a practice in which physicians order multiple tests simultaneously, on operational performance and patient outcomes in the ED.
 
-One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests [@naseim2015].
-A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests.
-These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.
+Worker discretion can improve system performance, but can sometimes enable workers to "choose the 'wrong' task (operationally)" [@vanDonselaar2010ordering].
+We consider the physician's decision to order imaging tests for their patient as an optimization problem, where the physician must balance the trade-offs between the benefits of ordering multiple tests simultaneously at the start of the patient encounter, which may expedite the diagnostic process if the tests are necessary for diagnosis and disposition, and the costs of ordering multiple tests simultaneously, which may increase the total time spent in the ED and the resource utilization of the department.
+Batching stands in contrast to the more common practice of ordering tests sequentially, where physicians order one test at a time, review the results, and then decide whether to order additional tests.
+The decision to batch order imaging tests is a form of discretion that physicians can exercise in the ED.
 
-A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially.
-This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing.
-Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results [@koch2018].
-Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier.
-Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged.
-Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements [@saghafian2015], however, research surrounding test ordering strategies to improve the patient flow processes remains limited.
+We address two research questions.
+First, what are the drivers of physician variation in the decision to batch order imaging tests?
+Second, what are the performance implications of batch ordering imaging tests?
+To identify the drivers of physician variation in the ordering of imaging tests, we consider the circumstances under which physicians are more likely to batch order imaging tests.
+We posit that the ability of physicians to identify an alternative test ordering strategy that they perceive as superior to the standard strategy will depend on the characteristics of the physician as well as the characteristics of the patient.
+With respect to the physician, we examine the role of physician experience.
+As for patient characteristics, we examine whether a patient has a complex chief complaint or a high acuity level, which may require additional diagnostic testing.
 
-In this paper, we use data from two large EDs to quantify the benefits and consequences of batching versus sequentially ordering advanced imaging tests on patient length of stay, re-admission, and total imaging volume.
-Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests.
-When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side.
-Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests.
-We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period.
-The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.
+We investigate these questions using operational data from two large EDs in the United States.
+We focus on imaging tests, which are commonly ordered in the ED and are associated with significant resource utilization.
+We use a unique dataset that contains detailed information on the timing of test orders, the timing of test results, and the timing of patient disposition.
+We exploit the random assignment of patients to physicians to identify the causal effects of batch ordering imaging tests on operational performance and patient outcomes.
+We find that physicians deviate from the standard test ordering strategy 42% of the time.
+We show that physicians are more likely to batch order imaging tests when they are more experienced, when there is an opportunity to follow a superior test ordering strategy by deviating, and when there is an opportunity to batch by deviating.
+When physicians deviate from the standard test ordering strategy, their average time to disposition increases by 13%.
+Other performance dimensions, including resource utilization and patient outcomes, are mostly unaffected.
+Our calculations suggest that forgoing deviations would have led to faster time to disposition, which could have saved 2,494 hours per year, increasing annual profits by 3%.
 
-With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly.
-Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm.
-In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings.
-Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines.
-Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.
+We also find that the
 
-In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes.
-This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization.
-It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.
+# Related Literature
 
-# Literature Review
+# Batch Ordering Advanced Imaging
 
-# Theoretical Background and Hypotheses
+## Drivers of Image Batching
 
-Emergency departments (EDs) are complex, time-sensitive environments where physicians must make rapid decisions under uncertainty [@asplin2003model; @jarvis2016ed].
-Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.
+## Implications for Emergency Department Operations and Patient Outcomes
 
-## Diagnostic Decision-Making and Test Ordering in the ED
+## Hetereogeneity in Orderging Strategy
 
-In the ED, physicians face the challenge of diagnosing and treating patients efficiently while maintaining high-quality care [@croskerry2013cognitive; @johnson2019decision].
-Upon a patient's arrival, the physician must decide on a diagnostic strategy, which often involves ordering imaging tests.
-This decision-making process is influenced by various factors, including the physician's experience, the patient's presenting symptoms, and the ED's current state (e.g., crowding levels) [@cooke2013overview; @bernstein2009effect; @morley2018emergency].
+# Setting, Data, and Models
 
-The key distinction in our study is between batch ordering and sequential ordering of imaging tests.
-In batch ordering, a physician orders multiple tests simultaneously, while in sequential ordering, tests are ordered one at a time, with each subsequent test decision informed by the results of the previous tests.
-It's crucial to understand that batch ordering does not mean tests are performed simultaneously.
-Instead, it means that patients enter the queues for multiple tests at once.
-This distinction is important when considering the implications of batch ordering on ED operations and patient flow [@green2006managing; @batt2019impact].
-
-## Queuing Theory and ED Operations
-
-When a physician batch orders tests, the patient enters multiple queues simultaneously.
-This can be modeled using parallel queuing systems in operations research [@dai2010load; @zychlinski2019discrete].
-While this might seem to expedite the process, it can lead to inefficiencies in resource allocation, as multiple departments (e.g., radiology, laboratory) must allocate resources to a single patient simultaneously, potentially leading to suboptimal resource utilization [@shi2016models; @wang2019data].
-
-Moreover, batch ordering creates an information lag where results from one test cannot inform the decision to proceed with subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].
-This approach can also create blocking effects, where a patient waiting for multiple tests may block other patients who need only one of those tests, potentially increasing overall waiting times [@armony2015heterogeneous; @song2020managing].
-
-## Reasons for Batch Ordering and Theoretical Implications
-
-Physicians may choose to batch order for several reasons.
-They may believe it will speed up the diagnostic process [@kocher2012effect; @rosen2018quality], or they may be practicing defensive medicine to avoid missing critical diagnoses [@studdert2005defensive; @kanzaria2015emergency].
-Batch ordering may also reduce the cognitive load required to make multiple sequential decisions [@croskerry2013cognitive; @nibbelink2018emergency], which can be particularly appealing in busy EDs where physicians are under time pressure [@bernstein2009effect; @morley2018emergency].
-
-From a theoretical perspective, batch ordering presents several implications.
-It forgoes the opportunity for value of information considerations, where each test result could inform subsequent decisions [@smith2013impact; @akhlaghpour2020value].
-Sequential ordering preserves the option to stop testing once a diagnosis is reached, aligning with real options theory in decision-making under uncertainty [@dixit1994investment; @garg2018value].
-While batch ordering might seem efficient for individual patients, it could lead to systemic inefficiencies in resource utilization and patient flow [@green2006managing; @batt2019impact].
-
-## Hypotheses Development
-
-Based on this theoretical framework, we develop the following hypotheses:
-
-### Impact on Resource Utilization
-
-*Hypothesis 1: Batch ordering of diagnostic tests will lead to increased overall test utilization compared to sequential ordering.* This hypothesis is grounded in the idea that batch ordering bypasses the opportunity to use information from initial tests to inform the necessity of subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].
-
-### Effect on Length of Stay
-
-*Hypothesis 2: Batch ordering will have a significant impact on patient length of stay (LOS) in the ED, but the direction of this effect is ambiguous.* While batch ordering might reduce LOS by eliminating waiting times between test decisions [@kocher2012effect; @rosen2018quality], it could also increase LOS due to potential queuing inefficiencies and the time required to perform and interpret additional tests [@forster2003effect; @armony2015heterogeneous; @song2020managing].
-
-### Quality of Care and Return Visits
-
-*Hypothesis 3: Batch ordering will influence the rate of 72-hour return visits with admission, but the direction of this effect is uncertain.* Comprehensive initial testing might reduce missed diagnoses and thus return visits [@pines2010national; @sabbatini2016overzealous].
-However, it could also lead to false positives and unnecessary follow-ups, potentially increasing return visits [@welch2011overdiagnosed; @o2018overdiagnosis].
-
-### Hypothesis 4: Heterogeneity Across Presenting Complaints
-
-*Hypothesis 4: The effects of batch ordering on resource utilization, length of stay, and quality of care will vary significantly across different presenting complaints.* This hypothesis is based on the understanding that different clinical presentations have varying levels of diagnostic uncertainty and risk [@cooke2013overview; @johnson2019decision], which may influence the appropriateness and impact of batch ordering.
-
-By testing these hypotheses, we aim to provide a comprehensive understanding of the impacts of batch ordering in the ED, informing both clinical practice and operational policy.
-Our empirical strategy, detailed in the following sections, is designed to rigorously test these hypotheses and uncover the causal effects of batch ordering on ED performance and patient outcomes.
-
-# Data and Methods
+## Empirical Setting
 
@@
 However, our primary analysis focuses on the Mayo Clinic data due to its unique feature of random patient-physician assignment, which allows for stronger causal inference.
+The data contain information on the timing of test orders, the timing of test results, and the timing of patient disposition, among various other important triage metrics and demographic features.
+We focus on imaging tests (x-rays, contrast CT, non-contrast CT, ultrasound) because unlike laboratory tests, imaging tests cannot be run simultaneously.
+Therefore, the operational implications of batch ordering imaging tests are more pronounced.
 
-## Mayo Clinic Data
+## Data
 
@@
 \begin{threeparttable}
-\begin{tabular}{p{10cm}ccccc}
+\begin{tabular}{lccccc}
 \toprule
@@
 
-\begin{table}[htbp]
+\begin{table}[ht]
     \centering
@@
     \label{tab:wald_test}
-    \begin{tabular}{p{10cm}cccc}
+    \begin{threeparttable}
+    \begin{tabular}{lcccc}
         \toprule
@@
 \begin{tablenotes}
-\small
 \item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
@@
 \end{tablenotes}
+\end{threeparttable}
 \end{table}
@@
 
-## Sample Construction
-
 Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
 We observe $48,854$ such visits during the study period.
-To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5$ percent of the time.
+To improve power, we drop encounters with rare "reasons for visit" (RVF) ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5\%$ of the time.
 Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them.
-Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints.
+Examples of complaints dropped include Skin Complaints and Urinary Complaints, as well as other complaints where multiple modalities of imaging is unlikely to occur.
 Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
@@
 
-## Variable Definitions
+### Instrumental Variable
 
-Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter.
-While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice.
+\
+Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch ordered during their ED encounter.
 We define "batching" in line with standard emergency medicine practices.
@@
 We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window.
-In Section [...], sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study.
+In Section 5.6, sensitivity analyses on this cutoff point showed that our results are robust to this definition.
+Each imaging test (e.g., X-ray, Contrast CT scan, Non-Contrast CT scan,  Ultrasound) is considered a separate, distinct test for our study.
 Therefore, a batch in our study consists of two or more distinct imaging tests.
 
-Our dependent variables are as follows:
+### Dependent Variables
 
-*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
-It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
+\
+\textit{Logarithm of Emergency Department Length of Stay (LOS)}: ED LOS is a key performance metric for ED operations.
+It is defined as the time from patient arrival to the ED to the time of discharge from the ED.
+We use this metric to evaluate the impact of batch ordering on ED efficiency.
+This variable is measured in minutes and in our sample is right-skewed.
+We log-transform this variable to normalize its distribution to meet the assumptions of linear regression.
 
-*(b) Number of Unique Imaging Tests*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives.
-In this study, we quantify resource utilization by the total number of distinct imaging tests ordered per patient during their ED stay.
-This encompasses both initial and subsequent tests.
+\textit{Logarithm of Time to Last Result}: Time to result is defined as the time from when a diagnostic test is ordered to when the result is available in the electronic medical record.
+The time stamps are at the minute level; that is, for each case, we know the minute in which a test is ordered by the physician and the minute in which the result is available.
+In order to fully capture the impact of batch ordering on the efficiency of diagnostic testing, we define time to last result (TIME TO RESULT) as the time from when the first image test is ordered to the the time when the last image test result is available.
+Time to result as a dependent variable is an understudied metric in the literature, and we believe it may capture a less noisy signal of the efficiency of diagnostic testing than LOS in certain circumstances.
+This variable is also right-skewed in our sample, so we log-transform it to normalize its distribution.
 
-*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
-This outcome is used to assess the quality of care provided during the initial ED visit.
+\textit{Total Imaging Tests Ordered}: Total imaging tests ordered (NUM_IMAGES) is a count of the number of distinct imaging tests ordered for a patient during the ED encounter.
+We use this variable to evaluate the impact of batch ordering on the number of imaging tests ordered.
 
-## Empirical Strategy
+\textit{72 Hour Return with Admission}: 72 Hour Return with Admission (72HRA) is a binary variable indicating whether a patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
+This variable is used to evaluate the impact of batch ordering on patient outcomes.
+
+## Identification Strategy
 
@@
 Similarly, we explore batching variation across physicians who work in the same emergency department.
-In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians.
-Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision.
-We employ both approaches and lay out their details in the next subsections.
-
-### Batch Tendency Construction
+In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians (i.e.
+physiancs with a higher or lower tendency to batch order imaging).
+Under additional assumptions, an instrumental variable approach identifies the causal effect of a given clincal decision.
 
@@
 
-Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour return with admission, and number of imaging tests ordered for the patient, and $X_i$ is the same as in the reduced-form approach.
+Where $Y_i$ represents our main outcomes of interest: ED LOS, TIME TO RESULT, NUM IMAGES, 72HRA, and $X_i$ is the same as in the reduced-form approach.
 $Batched_i$ variable suffers from potential endogeneity concerns.
@@
 
-## Identifying Assumptions
+### Identifying Assumptions
 
+\
 The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence").
@@
 
-# Results {#sec:4}
+## Econometric Models
 
-## Reduced-Form Results
+# Results and Discussion
+## Determenants of Image Batching
 
-In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department.
-We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination.
-Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.
+To investigate the drivers of batching behavior, we estimate the following regression model:
 
-To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency.
-Specifically, we calculate the physician's residualized leave-out average test rate, which serves as a proxy for their propensity to order tests.
-It is important to note the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
-This substantial correlation suggests that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies.
-Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.
+\small
+\begin{equation}
+Batched_{i,j} = \beta_0 + \beta_1 EXPERIENCE_{i,j} + \beta_2 SEXM_{i,j} + \beta_3 PATIENTS\_TBS + \beta_4 OCCUPANCY + X_{i,j} + \epsilon_i
+\end{equation}
 
-Reduced-form regression results in Table \ref{table:reduced_form} reveal the effect the average total effect of batch tendency on resource utilization and LOS.
-However, it is important to note that the estimated total effect of batch tendency accounts for both direct and indirect (mediated) effects of batch tendency.
-We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered.
-The increased testing volume, in turn, is associated with an extended LOS.
-
-\begin{table}[!htbp] 
-\centering 
-\caption{Reduced Form Estimates for Batch Tendency}
-\label{table:reduced_form}
-\small
-\begin{threeparttable}
-\begin{tabular}{lcccccc}
-\toprule
- & 72Hr Admission & LOS & Imaging Tests & 72Hr Admission & LOS & Imaging Tests \\
-\midrule
-Batch Tendency & -0.033 & 2.038** & 4.655*** & -0.063*** & -0.829 & 1.137** \\
- & (0.028) & (0.994) & (0.554) & (0.055) & (2.393) & (0.464) \\
-\textit{Baseline Controls} & $\checkmark$ & $\checkmark$ & $\checkmark$ &  $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\textit{Testing Inclination} &  &  &  &  $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\textit{Any Imaging Ordered} &  &  &  &  $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\midrule
-N & 15,821 & 15,821 & 15,821 & 15,821 & 15,821 & 15,821 \\
-R$^2$ & 0.006 & 0.303 & 0.199 & 0.006 & 0.364 & 0.591 \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}
-\small
-\item Estimates of the reduced form described in the text. Robust standard errors are clustered at the physician level.
-\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
-\end{tablenotes}
-\end{threeparttable}
-\end{table}
-
-## Placebo Check
-
-In this section we investigate whether the reduced-form effects observed in Section [...] are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency.
-We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
-For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
-Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.
-
-To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a >5 percent batching rate).
-We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3).
-The results of this exercise are displayed in Appendix Table [...].
-They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
-
-## Instrumental Variables Results
-
-To uncover the causal effect of test batching on key emergency department (ED) outcomes, we utilize a two-stage least squares (2SLS) approach, leveraging physician batch-ordering tendencies as an instrument.
-This strategy isolates the local average treatment effect (LATE) for patients whose diagnostic tests were batched due to their physician’s inclination to batch, rather than patient characteristics or other unobserved factors that may influence outcomes.
-The IV estimates capture the effect of test batching on ED length of stay, the number of imaging tests ordered, and the likelihood of a return visit within 72 hours with admission.
+\normalsize
+Where $Batched_{i,j}$ is a binary measure of whether physician $j$ batched tests for patient $i$.
+$EXPERIENCE_{i,j}$ is the number of years since the physician's residency graduation.
+$SEXM_{i,j}$ is a binary variable indicating the sex of the physician is male.
+$PATIENTS\_TBS$ is the number of remaining patients to be seen by the physician, excluding patient $i$, at the time of the patient's visit.
+$OCCUPANCY$ is the number of patients in the ED at the time of the arrival of the patient's visit.
+$X_{i,j}$ is the vectors of patient covariates described in the previous section.
+We cluster robust standard errors at the physician level.
 
 \begin{table}[!htbp] \centering 
-  \caption{IV Results} 
+  \caption{} 
   \label{} 
-\begin{tabular}{@{\extracolsep{15pt}}lccc} 
+\begin{tabular}{@{\extracolsep{5pt}}lcc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-\\[-1.8ex] & Log LOS & Imaging Tests & 72Hr Admission  \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
+ & Model 1 & Model 2 \\ 
+\\[-1.8ex] & (1) & (2)\\ 
 \hline \\[-1.8ex] 
- Batched & $-$0.504 & 0.692$^{**}$ & $-$0.038 \\ 
-  & (1.457) & (0.286) & (0.034) \\ 
-  & & & \\ 
+ EXPERIENCE & 0.001 (0.001) & 0.001 (0.001) \\ 
+  SEXM & 0.010 (0.028) & 0.010 (0.026) \\ 
+  PATIENTS\_TBS & $-$0.0002 (0.001) & 0.0001 (0.001) \\ 
+  OCCUPANCY & $-$0.0005 (0.0004) & $-$0.001 (0.0003) \\ 
+ & &   \\
+\textit{Day of Week-Time of Day FE} & $\checkmark$ & $\checkmark$  \\
+\textit{Month of Year FE} & $\checkmark$ & $\checkmark$   \\
+\textit{Complaint/Severity FE} & & $\checkmark$ \\
+\textit{Laboratory Tests Ordered} & & $\checkmark$  \\
+\hline \\[-1.8ex]
+ \textit{N} & 12,454 & 12,454 \\ 
+R$^{2}$ & 0.004 & 0.060 \\ 
+Adjusted R$^{2}$ & 0.0004 & 0.055 \\ 
+Residual Std. Error & 0.335 (df = 12411) & 0.326 (df = 12385) \\ 
 \hline 
 \hline \\[-1.8ex] 
-\end{tabular} 
-\begin{tablenotes}
-\small
-\item Estimates of the reduced form described in the text. Robust standard errors are clustered at the physician level.
-\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
-\end{tablenotes}
-\end{table}
-
-*ED Length of Stay*—The IV results in column (1) of Table X indicate that test batching has a negative but statistically insignificant effect on ED length of stay.
-The coefficient of -0.504 suggests a potential reduction in time spent in the ED, but the confidence interval is wide, indicating that the effect may not be robust across different specifications.
-This finding implies that while batching may streamline certain diagnostic processes, other operational factors likely play a larger role in determining how long a patient remains in the ED.
-These results suggest that batching alone does not drastically impact the overall time patients spend in the ED, pointing to the need for further exploration of mechanisms that may moderate this relationship.
-
-*Number of Imaging Tests Ordered*—In column (2), we find a statistically significant and positive effect of test batching on the number of imaging tests ordered.
-The coefficient of 0.692 (p < 0.05) indicates that batching leads to approximately 69% more imaging tests being conducted during the patient’s ED visit relative to if that patient's tests were sequenced.
-This result highlights a key consequence of batching: the grouping of diagnostic orders leads to a more intensive use of imaging resources.
-This increased resource utilization, while contributing to more comprehensive diagnostic evaluations, also raises concerns about the potential for overutilization and the associated costs and patient burden.
-This effect is particularly relevant in resource-constrained environments, where the trade-off between efficiency and resource use becomes crucial.
-
-*Return Visits with Admission within 72 Hours*—The results in column (3) show that test batching has a negative effect on the likelihood of a return visit with admission within 72 hours, although this effect is not statistically significant.
-The point estimate of -0.038 suggests a small potential reduction in return visits for patients whose tests were batched, which may reflect more thorough initial diagnostics leading to fewer missed diagnoses or complications.
-However, the lack of statistical significance means we cannot draw firm conclusions from this result.
-This finding aligns with previous research suggesting that while batching may improve certain diagnostic processes, its effect on downstream outcomes such as return visits may be more complex and contingent on factors beyond test ordering patterns.
-
-### Mechanism: Mediation Analysis
-
-To further explore the mechanisms through which batching influences ED length of stay (LOS) and return visits with admission within 72 hours, we conducted a formal mediation analysis.
-Specifically, we assessed whether the number of imaging tests mediates the relationship between batching and these outcomes.
-This analysis helps to decompose the total effect of batching into direct and indirect effects, with the indirect effect capturing the mediation through imaging tests.
-
-The mediation analysis was performed using a quasi-Bayesian approach with 1,000 simulations, adjusting for relevant covariates, including physician testing inclination, imaging use, day of the week, month of the year, patient complaints, and laboratory performance.
-The results for both outcomes are summarized in Table \ref{tab:med_results}, which presents the average causal mediation effect (ACME), average direct effect (ADE), and total effect for each outcome.
-
-\begin{table}[h]
-\centering
-\caption{Mediation Analysis for ED Length of Stay and 72-Hour Return Visits}
-\label{tab:med_results}
-\begin{tabular}{lcccc}
-\hline
-Outcome      & Effect Type & Estimate & 95\% CI Lower & 95\% CI Upper \\ \hline
-\multirow{4}{*}{\textbf{ED Length of Stay}} 
-& & & & \\
-    & ACME (Indirect Effect)  & 0.083   & 0.016  & 0.150  \\
-    & ADE (Direct Effect)     & -0.550  & -3.686 & 2.300  \\
-    & Total Effect            & -0.468  & -3.566 & 2.380  \\ \hline
-\multirow{4}{*}{\textbf{72-Hour Return Visits}} 
-& & & & \\
-    & ACME (Indirect Effect)  & -0.0029  & -0.0059 & 0.000  \\
-    & ADE (Direct Effect)     & -0.036   & -0.097  & 0.030  \\
-    & Total Effect            & -0.039   & -0.099  & 0.030  \\ \hline
+\textit{Notes:} & \multicolumn{2}{r}{$^{***}$Significant at the 1 percent level.} \\ 
+ & \multicolumn{2}{r}{$^{**}$Significant at the 5 percent level.} \\ 
+ & \multicolumn{2}{r}{$^{*}$Significant at the 10 percent level.} \\ 
 \end{tabular} 
@@
 
-\textbf{Incorporating Contemporary Mediation Frameworks.}—In line with the classical literature on mediation (Baron and Kenny, 1986), mediation analysis traditionally proceeds in three steps: (1) testing the effect of the independent variable (batch order) on the dependent variable (LOS/72-hour return) without the mediator (imaging volume); (2) adding the mediator (imaging volume) to the model and testing the direct effect of the independent variable on the dependent variable; and (3) testing the mediator’s effect on the dependent variable in the presence of the independent variable.
+In the baseline model (Model 1), we include only covaraites neccessary to achieve for quasi-random assignment.
+We find that the number of years since the physician's residency graduation, sex of the physician, number of remaining patients to be seen by the physician over the course of their shift, and the number of patients in the ED at the time of the arrival of the patient's visit are not significantly associated with the likelihood of batching tests.
+In Model 2, we include additional covariates to control for potential confounding factors such as the severity of the .
+We find that the number of years since the physician's residency graduation
 
-However, recent literature has demonstrated that significant results in the zero-order effect (i.e., the total effect) are not required to justify moving forward with mediation analysis.
-Zhao et al.
-(2010) and Hayes (2022) argue that requiring significance in each step is an overly restrictive and outdated framework, which often overlooks indirect effects that can be meaningful even when direct effects are not.
-In our analysis, although we do not always find statistically significant direct effects for batching on LOS or return visits, the mediation pathway through imaging volume remains relevant for understanding the indirect effects of batching.
+## Impact on Emergency Department Operations and Patient Outcomes
 
-Thus, following the contemporary guidance in mediation analysis, we proceed with testing both direct and indirect effects simultaneously, allowing us to capture the more nuanced dynamics of how batching influences ED outcomes.
-This approach provides a richer understanding of the operational impacts of batching, particularly as it pertains to the trade-offs between diagnostic efficiency and resource utilization.
+## Heterogeneity in Ordering Strategy
 
-\textbf{Mediation Results for ED Length of Stay.}—The mediation analysis reveals that the indirect effect (ACME) of batching on ED length of stay through the number of imaging tests is $0.083 (p < 0.01)$, suggesting that the increased volume of diagnostic testing marginally extends patient stays in the ED.
-The direct effect (ADE) of batching on length of stay is negative but not statistically significant ($-0.550, p = 0.71$), and the total effect is also negative ($-0.468$) but insignificant.
-This indicates that while batching may reduce LOS, the effect is partially offset by the increased testing, leading to no substantial impact on overall ED throughput.
+## Generealizability of Results
 
-\textbf{Mediation Results for 72-Hour Return Visits.}—For return visits within 72 hours, the indirect effect through the number of imaging tests is negative ($-0.0029, p < 0.05$), indicating that the additional testing associated with batching may slightly reduce the likelihood of a return visit.
-However, the direct effect ($-0.036$) and total effect ($-0.039$) of batching on return visits are small and not statistically significant, suggesting that while comprehensive diagnostics may reduce return visits, the effect of batching on this outcome is generally weak.
+## Managerial Implications
 
-\textbf{Conceptual Framework: Directed Acyclic Graph (DAG).}—Based on the findings from the mediation analysis, we propose a conceptual framework to illustrate the relationships uncovered by our empirical results.
-The Directed Acyclic Graph (DAG) shown in Figure \ref{fig:dag} provides a visual representation of the causal pathways between batch tendency (Z), batch ordering (X), imaging volume (M), and our outcomes (Y).
-The graph indicates that the indirect effect of batching operates through the volume of imaging tests, while the direct effect of batching on the outcomes bypasses this mediator.
+## Robustness Checks
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=3in]{model.png}
-\caption{Directed Acyclic Graph Representing the Causal Pathways Between Batch Order, Imaging Volume, and ED Outcomes}
-\label{fig:dag}
-\end{figure}
+# Conclusion
 
-In this DAG, batch tendency (Z) is used as an instrumental variable to predict batch order (X), which in turn affects both imaging volume (M) and our primary outcomes, including ED length of stay and 72-hour return with admission (Y).
-The unobserved confounding (U) is assumed to influence both imaging volume and the outcomes directly.
-The mediation analysis supports this structure, as it confirms that imaging volume mediates the relationship between batch ordering and ED outcomes, albeit modestly.
-By identifying this framework through our analysis, we provide a data-driven explanation for how batching influences ED operations.
+## Contributions
 
-\textbf{Interpretation of Mediation Results.}—These results show that the number of imaging tests serves as a partial mediator in the relationship between batching and ED outcomes.
-For length of stay, increased testing tends to prolong ED visits, but the overall effect of batching on LOS is negligible.
-For return visits, the increase in diagnostic testing slightly reduces the likelihood of a return within 72 hours, though the overall effect remains small and insignificant.
-These findings highlight the trade-offs involved in batching diagnostic orders: while batching may streamline certain processes, the associated increase in testing has modest effects on patient flow and outcomes.
-
-### Time to Result
+## Conclusion
 
 \newpage
-
-# References

### 2024-09-09 MgmtSci/MgmtSci.tex
getting going

--- 2024-09-09 MgmtSci/MgmtSci.Rmd
@@
----
-journal: mnsc
-title: "Image Batching: Batch Ordering Advanced Imaging Tests in the Emergency Department"
-shorttitle: "Image Batching"
-authors:
-- familyname: Jameson
-  othernames: Jacob
-  affiliation: Harvard Kennedy School, Harvard University
-  email: jacobjameson@g.harvard.edu
-- familyname: Soroush
-  othernames: Saghafian
-  affiliation: Harvard Kennedy School, Harvard University
-- familyname: Huckman
-  othernames:  Robert
-  affiliation: Harvard Business School, Harvard University
-- familyname: Hodgson
-  othernames:  Nicole
-  affiliation: Mayo Clinic, Emergency Department
-keywords:
-  - 3 to 6 keywords
-  - that do not appear in the title
-abstract: |
-  Abstract goes here
-acknowledgment: |
-  The authors would like to thank ...
-bibliography: references.bib
-blind: false
-output: rticles::informs_article
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{tabularx}
-  - \usepackage{graphicx}
-  - \usepackage{tikz}
-  - \usepackage{siunitx}
-  - \usepackage{tablefootnote}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \usepackage{natbib}
-  - \usepackage{caption}
-  - \usepackage{adjustbox}
-  - \usepackage{multirow}
----
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
+%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
+%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
+%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[,mnsc,nonblindrev]{informs}
 
-# Introduction
+\OneAndAHalfSpacedXI
+%%\OneAndAHalfSpacedXII % Current default line spacing
+%%\DoubleSpacedXII
+%%\DoubleSpacedXI
 
-Emergency departments (EDs) are high-pressure environments which globally face escalating challenges such as overcrowding, resource limitations, and increasing patient volumes [@mostafa2024strategies, @Sorup2013evaluation].
-These challenges can often be exacerbated by the need to provide timely and accurate diagnoses to patients, which often requires the use of advanced imaging tests such as computed tomography (CT) scans, ultrasound, X-ray, and magnetic resonance imaging (MRI) [@waheed2022impact].
-The management of diagnostic testing is a critical component of operational performance in EDs, where timely and accurate diagnoses can significantly impact patient outcomes [@committee2015improving].
-In many EDs, however, physicians have considerable discretion over how they order these tests, particularly with regard to advanced imaging [@valtchinov2019use].
-Yet little is known about the factors driving physicians' decisions to exercise such discretion and how test ordering should be managed when this discretion exists.
-In this paper, we explore the causal effects of batch ordering advanced imaging tests, a practice in which physicians order multiple tests simultaneously, on operational performance and patient outcomes in the ED.
+%% BEGIN MY ADDITIONS %%
+\usepackage{hyperref}
 
-Worker discretion can improve system performance, but can sometimes enable workers to "choose the 'wrong' task (operationally)" [@vanDonselaar2010ordering].
-We consider the physician's decision to order imaging tests for their patient as an optimization problem, where the physician must balance the trade-offs between the benefits of ordering multiple tests simultaneously at the start of the patient encounter, which may expedite the diagnostic process if the tests are necessary for diagnosis and disposition, and the costs of ordering multiple tests simultaneously, which may increase the total time spent in the ED and the resource utilization of the department.
-Batching stands in contrast to the more common practice of ordering tests sequentially, where physicians order one test at a time, review the results, and then decide whether to order additional tests.
-The decision to batch order imaging tests is a form of discretion that physicians can exercise in the ED.
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-We address two research questions.
-First, what are the drivers of physician variation in the decision to batch order imaging tests?
-Second, what are the performance implications of batch ordering imaging tests?
-To identify the drivers of physician variation in the ordering of imaging tests, we consider the circumstances under which physicians are more likely to batch order imaging tests.
-We posit that the ability of physicians to identify an alternative test ordering strategy that they perceive as superior to the standard strategy will depend on the characteristics of the physician as well as the characteristics of the patient.
-With respect to the physician, we examine the role of physician experience.
-As for patient characteristics, we examine whether a patient has a complex chief complaint or a high acuity level, which may require additional diagnostic testing.
+\usepackage{booktabs}
+\usepackage{tabularx}
+\usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{siunitx}
+\usepackage{tablefootnote}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\usepackage{natbib}
+\usepackage{caption}
+\usepackage{adjustbox}
+\usepackage{multirow}
 
-We investigate these questions using operational data from two large EDs in the United States.
-We focus on imaging tests, which are commonly ordered in the ED and are associated with significant resource utilization.
-We use a unique dataset that contains detailed information on the timing of test orders, the timing of test results, and the timing of patient disposition.
-We exploit the random assignment of patients to physicians to identify the causal effects of batch ordering imaging tests on operational performance and patient outcomes.
-We find that physicians deviate from the standard test ordering strategy 42% of the time.
-We show that physicians are more likely to batch order imaging tests when they are more experienced, when there is an opportunity to follow a superior test ordering strategy by deviating, and when there is an opportunity to batch by deviating.
-When physicians deviate from the standard test ordering strategy, their average time to disposition increases by 13%.
-Other performance dimensions, including resource utilization and patient outcomes, are mostly unaffected.
-Our calculations suggest that forgoing deviations would have led to faster time to disposition, which could have saved 2,494 hours per year, increasing annual profits by 3%.
+%% END MY ADDITIONS %%
+
+% Natbib setup for author-year style
+\usepackage{natbib}
+ \bibpunct[, ]{(}{)}{,}{a}{}{,}%
+ \def\bibfont{\small}%
+ \def\bibsep{\smallskipamount}%
+ \def\bibhang{24pt}%
+ \def\newblock{\ }%
+ \def\BIBand{and}%
+
+%% Setup of theorem styles. Outcomment only one.
+%% Preferred default is the first option.
+\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
+%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
+\ECRepeatTheorems
+
+%% Setup of the equation numbering system. Outcomment only one.
+%% Preferred default is the first option.
+\EquationsNumberedThrough    % Default: (1), (2), ...
+%\EquationsNumberedBySection % (1.1), (1.2), ...
+
+% For new submissions, leave this number blank.
+% For revisions, input the manuscript number assigned by the on-line
+% system along with a suffix ".Rx" where x is the revision number.
+\MANUSCRIPTNO{}
+
+%%%%%%%%%%%%%%%%
+\begin{document}
+%%%%%%%%%%%%%%%%
+
+% Outcomment only when entries are known. Otherwise leave as is and
+%   default values will be used.
+%\setcounter{page}{1}
+%\VOLUME{00}%
+%\NO{0}%
+%\MONTH{Xxxxx}% (month or a similar seasonal id)
+%\YEAR{0000}% e.g., 2005
+%\FIRSTPAGE{000}%
+%\LASTPAGE{000}%
+%\SHORTYEAR{00}% shortened year (two-digit)
+%\ISSUE{0000} %
+%\LONGFIRSTPAGE{0001} %
+%\DOI{10.1287/xxxx.0000.0000}%
+
+% Author's names for the running heads
+% Sample depending on the number of authors;
+\RUNAUTHOR{%
+Jameson, Soroush, Huckman
+ and Hodgson
+}
+% \RUNAUTHOR{Jones and Wilson}
+% \RUNAUTHOR{Jones, Miller, and Wilson}
+% \RUNAUTHOR{Jones et al.} % for four or more authors
+% Enter authors following the given pattern:
+%\RUNAUTHOR{}
+
+\RUNTITLE{Image Batching}
+
+\TITLE{Image Batching: Batch Ordering Advanced Imaging Tests in the
+Emergency Department}
+
+\ARTICLEAUTHORS{%
+\AUTHOR{Jacob Jameson}
+\AFF{Harvard Kennedy School, Harvard
+University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}
+
+\AUTHOR{Saghafian Soroush}
+\AFF{Harvard Kennedy School, Harvard University, \EMAIL{}}
+
+\AUTHOR{Robert Huckman}
+\AFF{Harvard Business School, Harvard University, \EMAIL{}}
+
+\AUTHOR{Nicole Hodgson}
+\AFF{Mayo Clinic, Emergency Department, \EMAIL{}}
+
+%
+}
+
+\ABSTRACT{Abstract goes here}
+
+\KEYWORDS{3 to 6 keywords; that do not appear in the title}
+
+\maketitle
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+Emergency departments (EDs) are high-pressure environments which
+globally face escalating challenges such as overcrowding, resource
+limitations, and increasing patient volumes
+\citep[\citet{Sorup2013evaluation}]{mostafa2024strategies}. These
+challenges can often be exacerbated by the need to provide timely and
+accurate diagnoses to patients, which often requires the use of advanced
+imaging tests such as computed tomography (CT) scans, ultrasound, X-ray,
+and magnetic resonance imaging (MRI) \citep{waheed2022impact}. The
+management of diagnostic testing is a critical component of operational
+performance in EDs, where timely and accurate diagnoses can
+significantly impact patient outcomes \citep{committee2015improving}. In
+many EDs, however, physicians have considerable discretion over how they
+order these tests, particularly with regard to advanced imaging
+\citep{valtchinov2019use}. Yet little is known about the factors driving
+physicians' decisions to exercise such discretion and how test ordering
+should be managed when this discretion exists. In this paper, we explore
+the causal effects of batch ordering advanced imaging tests, a practice
+in which physicians order multiple tests simultaneously, on operational
+performance and patient outcomes in the ED.
+
+Worker discretion can improve system performance, but can sometimes
+enable workers to ``choose the `wrong' task (operationally)''
+\citep{vanDonselaar2010ordering}. We consider the physician's decision
+to order imaging tests for their patient as an optimization problem,
+where the physician must balance the trade-offs between the benefits of
+ordering multiple tests simultaneously at the start of the patient
+encounter, which may expedite the diagnostic process if the tests are
+necessary for diagnosis and disposition, and the costs of ordering
+multiple tests simultaneously, which may increase the total time spent
+in the ED and the resource utilization of the department. Batching
+stands in contrast to the more common practice of ordering tests
+sequentially, where physicians order one test at a time, review the
+results, and then decide whether to order additional tests. The decision
+to batch order imaging tests is a form of discretion that physicians can
+exercise in the ED.
+
+We address two research questions. First, what are the drivers of
+physician variation in the decision to batch order imaging tests?
+Second, what are the performance implications of batch ordering imaging
+tests? To identify the drivers of physician variation in the ordering of
+imaging tests, we consider the circumstances under which physicians are
+more likely to batch order imaging tests. We posit that the ability of
+physicians to identify an alternative test ordering strategy that they
+perceive as superior to the standard strategy will depend on the
+characteristics of the physician as well as the characteristics of the
+patient. With respect to the physician, we examine the role of physician
+experience. As for patient characteristics, we examine whether a patient
+has a complex chief complaint or a high acuity level, which may require
+additional diagnostic testing.
+
+We investigate these questions using operational data from two large EDs
+in the United States. We focus on imaging tests, which are commonly
+ordered in the ED and are associated with significant resource
+utilization. We use a unique dataset that contains detailed information
+on the timing of test orders, the timing of test results, and the timing
+of patient disposition. We exploit the random assignment of patients to
+physicians to identify the causal effects of batch ordering imaging
+tests on operational performance and patient outcomes. We find that
+physicians deviate from the standard test ordering strategy 42\% of the
+time. We show that physicians are more likely to batch order imaging
+tests when they are more experienced, when there is an opportunity to
+follow a superior test ordering strategy by deviating, and when there is
+an opportunity to batch by deviating. When physicians deviate from the
+standard test ordering strategy, their average time to disposition
+increases by 13\%. Other performance dimensions, including resource
+utilization and patient outcomes, are mostly unaffected. Our
+calculations suggest that forgoing deviations would have led to faster
+time to disposition, which could have saved 2,494 hours per year,
+increasing annual profits by 3\%.
 
@@
 
-# Related Literature
+\hypertarget{related-literature}{%
+\section{Related Literature}\label{related-literature}}
 
-# Batch Ordering Advanced Imaging
+\hypertarget{batch-ordering-advanced-imaging}{%
+\section{Batch Ordering Advanced
+Imaging}\label{batch-ordering-advanced-imaging}}
 
-## Drivers of Image Batching
+\hypertarget{drivers-of-image-batching}{%
+\subsection{Drivers of Image Batching}\label{drivers-of-image-batching}}
 
-## Implications for Emergency Department Operations and Patient Outcomes
+\hypertarget{implications-for-emergency-department-operations-and-patient-outcomes}{%
+\subsection{Implications for Emergency Department Operations and Patient
+Outcomes}\label{implications-for-emergency-department-operations-and-patient-outcomes}}
 
-## Hetereogeneity in Orderging Strategy
+\hypertarget{hetereogeneity-in-orderging-strategy}{%
+\subsection{Hetereogeneity in Orderging
+Strategy}\label{hetereogeneity-in-orderging-strategy}}
 
-# Setting, Data, and Models
+\hypertarget{setting-data-and-models}{%
+\section{Setting, Data, and Models}\label{setting-data-and-models}}
 
-## Empirical Setting
+\hypertarget{empirical-setting}{%
+\subsection{Empirical Setting}\label{empirical-setting}}
 
-Our study uses data from two large U.S.
-emergency departments (EDs): the Mayo Clinic of Arizona and Massachusetts General Hospital (MGH).
-The MGH dataset, which includes 129,489 patient encounters from November 10, 2021 through December 10, 2022, provides a robust sample for validating the generalizability of our findings.
-However, our primary analysis focuses on the Mayo Clinic data due to its unique feature of random patient-physician assignment, which allows for stronger causal inference.
-The data contain information on the timing of test orders, the timing of test results, and the timing of patient disposition, among various other important triage metrics and demographic features.
-We focus on imaging tests (x-rays, contrast CT, non-contrast CT, ultrasound) because unlike laboratory tests, imaging tests cannot be run simultaneously.
-Therefore, the operational implications of batch ordering imaging tests are more pronounced.
+Our study uses data from two large U.S. emergency departments (EDs): the
+Mayo Clinic of Arizona and Massachusetts General Hospital (MGH). The MGH
+dataset, which includes 129,489 patient encounters from November 10,
+2021 through December 10, 2022, provides a robust sample for validating
+the generalizability of our findings. However, our primary analysis
+focuses on the Mayo Clinic data due to its unique feature of random
+patient-physician assignment, which allows for stronger causal
+inference. The data contain information on the timing of test orders,
+the timing of test results, and the timing of patient disposition, among
+various other important triage metrics and demographic features. We
+focus on imaging tests (x-rays, contrast CT, non-contrast CT,
+ultrasound) because unlike laboratory tests, imaging tests cannot be run
+simultaneously. Therefore, the operational implications of batch
+ordering imaging tests are more pronounced.
 
-## Data
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
 
-Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation.
-During the study period from October 6, 2018, through December 31, 2019, the ED recorded 48,854 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces.
-The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume.
-The data is summarized in Table \ref{tab:summary_stats}.
+Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a
+tertiary care hospital without obstetrical services, an inpatient
+pediatrics unit, or a trauma designation. During the study period from
+October 6, 2018, through December 31, 2019, the ED recorded 48,854
+visits per year, managed across 26 treatment rooms and up to 9 hallway
+spaces. The department is exclusively staffed by board-eligible or
+board-certified emergency physicians (EPs), with rotating residents
+overseeing about 10\% of patient volume. The data is summarized in Table
+\ref{tab:summary_stats}.
 
@@
 
-A key feature of the Mayo Clinic ED is its rotational patient assignment system, in which patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses.
-In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+A key feature of the Mayo Clinic ED is its rotational patient assignment
+system, in which patients arriving at the Mayo Clinic ED are randomly
+assigned to physicians via a rotational patient assignment algorithm
+\citet{traub2016emergency}, which removes potential selection bias
+concerns for our analyses. In essence, barring arrival time and
+shift-level variation, the physician-to-patient matching can be deemed
+random. Table \ref{tab:wald_test} displays that patient encounters
+(regarding chief complaints and emergency severity) are equitably
+distributed across physicians within our study's cohort.
 
@@
 
-We conducted a retrospective review of comprehensive ED operational data, coinciding with the initiation of a new electronic medical record.
-The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics.
-This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.
+We conducted a retrospective review of comprehensive ED operational
+data, coinciding with the initiation of a new electronic medical record.
+The dataset includes detailed patient demographics, chief complaints,
+vital signs, emergency severity index (ESI), length of stay (LOS), and
+resource utilization metrics. This period was chosen to provide a robust
+data set while excluding the influence of the coronavirus pandemic.
 
-Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
-We observe $48,854$ such visits during the study period.
-To improve power, we drop encounters with rare "reasons for visit" (RVF) ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5\%$ of the time.
-Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them.
-Examples of complaints dropped include Skin Complaints and Urinary Complaints, as well as other complaints where multiple modalities of imaging is unlikely to occur.
-Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories as we do.
-In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $15,821$ encounters involving full-time physicians who treat over 500 ED cases per year.
+Our research design focuses on adults who visit the Mayo Clinic of
+Arizona ED. We observe \(48,854\) such visits during the study period.
+To improve power, we drop encounters with rare ``reasons for visit''
+(RVF) (\(<1000\) total encounters of this kind) and complaints where a
+batch order occurs less than \(5\%\) of the time. Since batch orders are
+rare for these cases, our physician batch tendency instrument could
+suffer from a weak instrument problem if we included them. Examples of
+complaints dropped include Skin Complaints and Urinary Complaints, as
+well as other complaints where multiple modalities of imaging is
+unlikely to occur. Excluding these conditions does not introduce
+selection bias unless physician test batching tendency is orthogonal to
+physician diagnosing behavior. While this assumption may be violated if
+we were to use a very detailed level of chief complaint information upon
+which to base our exclusion criterion, it is plausibly satisfied when
+using broad complaint categories as we do. In order to estimate a
+precise measure of physician-level batch tendency, we further restrict
+our sample to the \(15,821\) encounters involving full-time physicians
+who treat over 500 ED cases per year.
 
-### Instrumental Variable
+\hypertarget{instrumental-variable}{%
+\subsubsection{Instrumental Variable}\label{instrumental-variable}}
 
-\
-Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch ordered during their ED encounter.
-We define "batching" in line with standard emergency medicine practices.
-Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses.
-This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.
+\hfill\break
+Our explanatory variable in the IV analysis, \(Batched_i\), is an
+indicator for whether patient \(i\) has their tests batch ordered during
+their ED encounter. We define ``batching'' in line with standard
+emergency medicine practices. Batching occurs when a physician
+simultaneously orders a comprehensive set of diagnostic tests, typically
+covering a broad range of potential diagnoses. This contrasts with
+sequential ordering, where tests are ordered sequentially based on the
+information obtained from each test as needed.
 
-We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window.
-In Section 5.6, sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-Each imaging test (e.g., X-ray, Contrast CT scan, Non-Contrast CT scan,  Ultrasound) is considered a separate, distinct test for our study.
-Therefore, a batch in our study consists of two or more distinct imaging tests.
+We operationalize batching as occurring when multiple diagnostic imaging
+tests are ordered within a 5-minute window. In Section 5.6, sensitivity
+analyses on this cutoff point showed that our results are robust to this
+definition. Each imaging test (e.g., X-ray, Contrast CT scan,
+Non-Contrast CT scan, Ultrasound) is considered a separate, distinct
+test for our study. Therefore, a batch in our study consists of two or
+more distinct imaging tests.
 
-### Dependent Variables
+\hypertarget{dependent-variables}{%
+\subsubsection{Dependent Variables}\label{dependent-variables}}
 
-\
-\textit{Logarithm of Emergency Department Length of Stay (LOS)}: ED LOS is a key performance metric for ED operations.
-It is defined as the time from patient arrival to the ED to the time of discharge from the ED.
-We use this metric to evaluate the impact of batch ordering on ED efficiency.
-This variable is measured in minutes and in our sample is right-skewed.
-We log-transform this variable to normalize its distribution to meet the assumptions of linear regression.
+\hfill\break
+\textit{Logarithm of Emergency Department Length of Stay (LOS)}: ED LOS
+is a key performance metric for ED operations. It is defined as the time
+from patient arrival to the ED to the time of discharge from the ED. We
+use this metric to evaluate the impact of batch ordering on ED
+efficiency. This variable is measured in minutes and in our sample is
+right-skewed. We log-transform this variable to normalize its
+distribution to meet the assumptions of linear regression.
 
-\textit{Logarithm of Time to Last Result}: Time to result is defined as the time from when a diagnostic test is ordered to when the result is available in the electronic medical record.
-The time stamps are at the minute level; that is, for each case, we know the minute in which a test is ordered by the physician and the minute in which the result is available.
-In order to fully capture the impact of batch ordering on the efficiency of diagnostic testing, we define time to last result (TIME TO RESULT) as the time from when the first image test is ordered to the the time when the last image test result is available.
-Time to result as a dependent variable is an understudied metric in the literature, and we believe it may capture a less noisy signal of the efficiency of diagnostic testing than LOS in certain circumstances.
-This variable is also right-skewed in our sample, so we log-transform it to normalize its distribution.
+\textit{Logarithm of Time to Last Result}: Time to result is defined as
+the time from when a diagnostic test is ordered to when the result is
+available in the electronic medical record. The time stamps are at the
+minute level; that is, for each case, we know the minute in which a test
+is ordered by the physician and the minute in which the result is
+available. In order to fully capture the impact of batch ordering on the
+efficiency of diagnostic testing, we define time to last result (TIME TO
+RESULT) as the time from when the first image test is ordered to the the
+time when the last image test result is available. Time to result as a
+dependent variable is an understudied metric in the literature, and we
+believe it may capture a less noisy signal of the efficiency of
+diagnostic testing than LOS in certain circumstances. This variable is
+also right-skewed in our sample, so we log-transform it to normalize its
+distribution.
 
-\textit{Total Imaging Tests Ordered}: Total imaging tests ordered (NUM_IMAGES) is a count of the number of distinct imaging tests ordered for a patient during the ED encounter.
-We use this variable to evaluate the impact of batch ordering on the number of imaging tests ordered.
+\textit{Total Imaging Tests Ordered}: Total imaging tests ordered
+(NUM\_IMAGES) is a count of the number of distinct imaging tests ordered
+for a patient during the ED encounter. We use this variable to evaluate
+the impact of batch ordering on the number of imaging tests ordered.
 
-\textit{72 Hour Return with Admission}: 72 Hour Return with Admission (72HRA) is a binary variable indicating whether a patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
-This variable is used to evaluate the impact of batch ordering on patient outcomes.
+\textit{72 Hour Return with Admission}: 72 Hour Return with Admission
+(72HRA) is a binary variable indicating whether a patient returns to the
+ED within 72 hours of their initial visit and is admitted to the
+hospital. This variable is used to evaluate the impact of batch ordering
+on patient outcomes.
 
-## Identification Strategy
+\hypertarget{identification-strategy}{%
+\subsection{Identification Strategy}\label{identification-strategy}}
 
-Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court.
-Similarly, we explore batching variation across physicians who work in the same emergency department.
-In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians (i.e.
-physiancs with a higher or lower tendency to batch order imaging).
-Under additional assumptions, an instrumental variable approach identifies the causal effect of a given clincal decision.
+Our empirical strategy closely follows the literature that relies on
+quasi-random assignment of agents to cases, often referred to as the
+``judges design.'' Papers in this literature typically exploit variation
+in the sentencing leniency of judges who work in the same court.
+Similarly, we explore batching variation across physicians who work in
+the same emergency department. In its reduced form, under the assumption
+of quasi-random assignment, this approach allows researchers to identify
+the causal effect of being assigned to different types of physicians
+(i.e.~physiancs with a higher or lower tendency to batch order imaging).
+Under additional assumptions, an instrumental variable approach
+identifies the causal effect of a given clincal decision.
 
-To measure physician batch tendency, we use the physician's residualized leave-out average batch rate.
-This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways.
-First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.
+To measure physician batch tendency, we use the physician's residualized
+leave-out average batch rate. This measure is derived from two steps
+following the approaches taken by \citet{doyle2015measuring},
+\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
+obtain residuals from a regression model, which includes all ED
+encounters in our sample period.
 
@@
 
-Where $Y_i$ represents our main outcomes of interest: ED LOS, TIME TO RESULT, NUM IMAGES, 72HRA, and $X_i$ is the same as in the reduced-form approach.
-$Batched_i$ variable suffers from potential endogeneity concerns.
-For example, injury severity may be unobserved and correlated with need to run multiple tests, length of stay, and return with admission likelihood.
-Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$.
-We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.
+Where \(Y_i\) represents our main outcomes of interest: ED LOS, TIME TO
+RESULT, NUM IMAGES, 72HRA, and \(X_i\) is the same as in the
+reduced-form approach. \(Batched_i\) variable suffers from potential
+endogeneity concerns. For example, injury severity may be unobserved and
+correlated with need to run multiple tests, length of stay, and return
+with admission likelihood. Hence, we instrument \(Batched_i\) with the
+assigned physician \(j\)'s underlying tendency to batch,
+\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
+physician level to account for the assignment process of patients to
+physicians.
 
-### Identifying Assumptions
+\hypertarget{identifying-assumptions}{%
+\subsubsection{Identifying Assumptions}\label{identifying-assumptions}}
 
-\
-The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence").
-The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.
+\hfill\break
+The reduced-form approach delivers an unbiased estimate of the causal
+effect of being treated by a higher tendency to batch physician, since
+assignment of patients to ED physicians is random, conditional on
+seasonality and shift (``conditional independence''). The
+residualization in equation (1) controls for more controls than required
+to achieve quasi-random assignment; they are included for statistical
+precision in measuring physician tendency to batch.
 
-Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity.
-We reported a strong first stage (i.e., relevance) at the end of the previous Section.
-The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching.
-This is perhaps our strongest assumption and is at its core, untestable.
-However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings.
-First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher.
-Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest.
-The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis.
-Nevertheless, we take this assumption seriously and perform a placebo check in Section [...] as well as various robustness checks in Section [...].
+Our instrumental variable approach, which aims to recover the causal
+effect of having diagnostic tests batch ordered, relies on three
+additional assumptions: relevance, exclusion, and monotonicity. We
+reported a strong first stage (i.e., relevance) at the end of the
+previous Section. The exclusion restriction requires that the instrument
+must influence the outcome of interest only through its effect on test
+batching. This is perhaps our strongest assumption and is at its core,
+untestable. However, several features of the ED setting suggest that
+such violation may likely only have a small impact and may be less
+concerning than in other health care settings. First, unlike in primary
+care settings, where the patient and primary care provider have many
+repeat encounters, the scope of what the emergency physician can do to
+impact medium-term outcomes is limited and well-observed by the
+researcher. Second, any violation of the exclusion restriction needs to
+directly affect the specific outcome of interest. The channel by which
+ED physicians can influence length of stay relative outcomes is likely
+through testing and diagnosis. Nevertheless, we take this assumption
+seriously and perform a placebo check in Section {[}\ldots{]} as well as
+various robustness checks in Section {[}\ldots{]}.
 
-Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects.
-It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician.
-The literature leveraging the judges design typically performs two informal tests for its implications.
-The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects).
-The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration).
-Appendix Table  [...] presents both of these tests in the two columns for various subsamples of interest.
+Finally, the monotonicity assumption is necessary for interpreting the
+coefficient estimates obtained from the IV approach as Local Average
+Treatment Effects (LATEs) if there are heterogeneous treatment effects.
+It requires that any patient who is (not) batched by a sequencer
+(batcher) would also (not) be batched by a batcher (sequencer)
+physician. The literature leveraging the judges design typically
+performs two informal tests for its implications. The first one provides
+that the first stage should be weakly positive for all subsamples
+(\citet{dobbie2018effects}). The second implication asserts that the
+instrument constructed by leaving out a particular subsample has
+predictive power over that same left-out subsample
+(\citet{bhuller2020incarceration}). Appendix Table {[}\ldots{]} presents
+both of these tests in the two columns for various subsamples of
+interest.
 
-## Econometric Models
+\hypertarget{econometric-models}{%
+\subsection{Econometric Models}\label{econometric-models}}
 
-# Results and Discussion
-## Determenants of Image Batching
+\hypertarget{results-and-discussion}{%
+\section{Results and Discussion}\label{results-and-discussion}}
 
-To investigate the drivers of batching behavior, we estimate the following regression model:
+\hypertarget{determenants-of-image-batching}{%
+\subsection{Determenants of Image
+Batching}\label{determenants-of-image-batching}}
+
+To investigate the drivers of batching behavior, we estimate the
+following regression model:
 
 \small
+
 \begin{equation}
@@
 \normalsize
-Where $Batched_{i,j}$ is a binary measure of whether physician $j$ batched tests for patient $i$.
-$EXPERIENCE_{i,j}$ is the number of years since the physician's residency graduation.
-$SEXM_{i,j}$ is a binary variable indicating the sex of the physician is male.
-$PATIENTS\_TBS$ is the number of remaining patients to be seen by the physician, excluding patient $i$, at the time of the patient's visit.
-$OCCUPANCY$ is the number of patients in the ED at the time of the arrival of the patient's visit.
-$X_{i,j}$ is the vectors of patient covariates described in the previous section.
-We cluster robust standard errors at the physician level.
+
+Where \(Batched_{i,j}\) is a binary measure of whether physician \(j\)
+batched tests for patient \(i\). \(EXPERIENCE_{i,j}\) is the number of
+years since the physician's residency graduation. \(SEXM_{i,j}\) is a
+binary variable indicating the sex of the physician is male.
+\(PATIENTS\_TBS\) is the number of remaining patients to be seen by the
+physician, excluding patient \(i\), at the time of the patient's visit.
+\(OCCUPANCY\) is the number of patients in the ED at the time of the
+arrival of the patient's visit. \(X_{i,j}\) is the vectors of patient
+covariates described in the previous section. We cluster robust standard
+errors at the physician level.
 
@@
 
-In the baseline model (Model 1), we include only covaraites neccessary to achieve for quasi-random assignment.
-We find that the number of years since the physician's residency graduation, sex of the physician, number of remaining patients to be seen by the physician over the course of their shift, and the number of patients in the ED at the time of the arrival of the patient's visit are not significantly associated with the likelihood of batching tests.
-In Model 2, we include additional covariates to control for potential confounding factors such as the severity of the .
-We find that the number of years since the physician's residency graduation
+In the baseline model (Model 1), we include only covaraites neccessary
+to achieve for quasi-random assignment. We find that the number of years
+since the physician's residency graduation, sex of the physician, number
+of remaining patients to be seen by the physician over the course of
+their shift, and the number of patients in the ED at the time of the
+arrival of the patient's visit are not significantly associated with the
+likelihood of batching tests. In Model 2, we include additional
+covariates to control for potential confounding factors such as the
+severity of the . We find that the number of years since the physician's
+residency graduation
 
-## Impact on Emergency Department Operations and Patient Outcomes
+\hypertarget{impact-on-emergency-department-operations-and-patient-outcomes}{%
+\subsection{Impact on Emergency Department Operations and Patient
+Outcomes}\label{impact-on-emergency-department-operations-and-patient-outcomes}}
 
-## Heterogeneity in Ordering Strategy
+\hypertarget{heterogeneity-in-ordering-strategy}{%
+\subsection{Heterogeneity in Ordering
+Strategy}\label{heterogeneity-in-ordering-strategy}}
 
-## Generealizability of Results
+\hypertarget{generealizability-of-results}{%
+\subsection{Generealizability of
+Results}\label{generealizability-of-results}}
 
-## Managerial Implications
+\hypertarget{managerial-implications}{%
+\subsection{Managerial Implications}\label{managerial-implications}}
 
-## Robustness Checks
+\hypertarget{robustness-checks}{%
+\subsection{Robustness Checks}\label{robustness-checks}}
 
-# Conclusion
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
-## Contributions
+\hypertarget{contributions}{%
+\subsection{Contributions}\label{contributions}}
 
-## Conclusion
+\hypertarget{conclusion-1}{%
+\subsection{Conclusion}\label{conclusion-1}}
 
 \newpage
+
+% Appendix here
+% Options are (1) APPENDIX (with or without general title) or
+%             (2) APPENDICES (if it has more than one unrelated sections)
+% Outcomment the appropriate case if necessary
+%
+% \begin{APPENDIX}{<Title of the Appendix>}
+% \end{APPENDIX}
+%
+%   or
+%
+% \begin{APPENDICES}
+% \section{<Title of Section A>}
+% \section{<Title of Section B>}
+% etc
+% \end{APPENDICES}
+
+% Acknowledgments here
+\ACKNOWLEDGMENT{The authors would like to thank \ldots{}}
+
+\bibliographystyle{informs2014}
+\bibliography{references.bib}
+
+\end{document}

### 2024-09-09 manuscript/CopyOfmanuscript.Rmd
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits...
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---

# Introduction {#sec:I}

Healthcare delivery, particularly in the emergency department (ED), requires a delicate balance between ensuring optimal patient outcomes and optimizing resource utilization. Achieving these twin goals necessitates timely and accurate diagnosis, which enables prompt and appropriate treatment, improving patient prognosis and reducing the likelihood of adverse events. Efficient patient discharge from the ED is crucial for alleviating overcrowding, a severe issue associated with higher complication rates and increased mortality @bernstein2009. In this complex environment, the availability and performance of diagnostic tests, especially imaging studies, play a pivotal role in shaping both patient care and operational efficiency @naseim2015.

A critical question in ED management pertains to whether physicians should batch order diagnostic imaging tests or order them sequentially. This decision represents a fundamental tradeoff between potentially reducing patient length of stay and risking over-testing. Batch ordering, where multiple imaging tests are ordered simultaneously, may expedite the diagnostic process but can lead to unnecessary tests, increased costs, patient anxiety, and potential harm from follow-up of false-positive results @koch2018. Conversely, a more sequential approach might reduce unnecessary testing but could extend a patient's time in the ED, potentially contributing to overcrowding. Striking the right balance between comprehensive evaluation and efficient resource use is crucial, yet research on optimal test ordering strategies in the ED remains limited @saghafian2015.

In this paper, we leverage data from over 45,000 ED patient visits to quantify the benefits and consequences of batching versus sequentially ordering diagnostic imaging tests. Our focus on imaging tests is particularly relevant given their significant impact on patient flow, resource utilization, and potential health risks from radiation exposure. We employ an empirical strategy that exploits the random assignment of patients to ED physicians who differ in their propensity to batch-order imaging tests. This quasi-experimental design allows us to overcome potential selection bias and move closer to identifying the causal impact of batch ordering on key outcomes such as patient length of stay, resource utilization, and the likelihood of return visits.

To capture physician tendencies towards batch ordering, we introduce a novel measure using a leave-out, residualized approach based on each physician's historical ordering patterns for all other patients seen during the study period. This tendency measure strongly predicts actual batch ordering behavior while remaining uncorrelated with patient and ED visit characteristics, providing a robust instrument for our analysis. Our findings reveal significant practice variation among physicians and demonstrate substantial impacts on ED operations and patient care trajectories.

We begin by evaluating the reduced-form effects of ED provider batch-ordering tendency on downstream patient outcomes and ED turnaround time. Our results indicate that practice variation, as captured by physician batch-ordering tendency, has large and significant consequences. For instance, being treated by a provider in the top decile of the batching tendency distribution, compared to one in the bottom decile, results in additional diagnostic imaging tests per 100 patient encounters, even after controlling for the physician's underlying propensity to test. This variation in practice patterns raises important questions about the optimal use of diagnostic resources in emergency care settings.

To address potential confounding factors, we employ a placebo exercise to demonstrate that the observed effects are primarily due to batching behavior rather than other aspects of physician care that might correlate with batching tendency. We find no significant effects of physician batch tendency on key outcomes for a "placebo sample" of patients visiting the ED for conditions that rarely result in batched orders. This strengthens our confidence in the specificity of the batching effect on patient outcomes and resource utilization.

Our research design closely approximates a randomized controlled trial that assigns patients to batch-ordering or sequential-ordering arms, leveraging the unique features of the ED setting. In this environment, patients have no discretion over provider selection, and in our specific ED, the random assignment of patients to physicians further mitigates selection issues. The wide variation in batch-ordering practices among physicians within the same hospital, despite following identical guidelines, provides the necessary variation for our analysis. Moreover, the typically short, one-off nature of ED patient-physician interactions constrains decision-making to a more limited, observable set of choices compared to other healthcare settings.

By exploiting this practice variation in ED settings, we aim to isolate the impact of test batching from other potential channels that might influence length of stay and patient outcomes in different healthcare contexts. This approach allows us to more accurately identify the causal impact of batch-ordering diagnostic imaging tests on patient outcomes and resource utilization. Importantly, our study examines batching decisions made within the bounds of normal clinical judgment and practice norms, rather than focusing on deviations from clinical guidelines or instances of substandard care.

The implications of this research extend beyond academic interest, offering potential insights for ED management, healthcare policy, and the design of clinical decision support systems. Understanding the true impacts of batch ordering practices could inform more efficient resource allocation strategies, improve patient flow, and ultimately enhance the quality of emergency care delivery. The remainder of this paper is structured as follows: The next section describes our data source and outlines the baseline sample. Section II describes the theoretical and conceptual model. Section IV details our empirical strategy and its accompanying identifying assumptions. Section V presents our results, including both reduced-form effects and instrumental variable analyses. Section V also discusses the implications of our findings for batch-ordering policies and ED management. The final section concludes with a summary of our contributions and directions for future research.

# Theoretical Model

Emergency departments (EDs) are complex environments where critical decisions about patient care must be made rapidly and under conditions of uncertainty. Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.

In a typical ED scenario, patients arrive with varying levels of acuity and a wide range of presenting symptoms. Upon arrival, each patient $i$ is assigned to a physician $j$, often based on availability rather than specific matching criteria. The physician's primary task is to diagnose and treat the patient efficiently and effectively, a process that frequently involves ordering diagnostic tests, including imaging studies.

Central to our model is the concept of the patient's underlying condition, denoted by $\tau_i$. This represents the true state of the patient's health, which is initially unknown to the treating physician. The physician's goal is to uncover $\tau_i$ as accurately and quickly as possible to provide appropriate care. However, the path to this discovery is not straightforward and involves critical decisions about diagnostic testing.

The physician's approach to ordering tests is influenced by various factors, including their individual tendencies and experiences. We represent this as $Z_j$, the physician's propensity to batch order tests. This tendency can be thought of as a stable characteristic of the physician, shaped by their training, past experiences, and personal practice style. The decision to batch order tests for a given patient, represented by $X_{ij}$, is a function of both the physician's tendency $Z_j$ and the specifics of the patient's presentation. Batch ordering involves requesting multiple imaging tests simultaneously, rather than in a sequential manner. This decision has cascading effects on the patient's journey through the ED.

\begin{figure}[h]
  \centering
  \caption{Conceptual Model}
  \label{fig:dag}
  \includegraphics[width=4.25in]{model.png}
  \begin{tablenotes}
    \small
    \item \textit{Figure \ref{fig:dag} illustrates the causal relationships between physician tendency, batch ordering decision, test volume, and patient outcomes.}
  \end{tablenotes}
\end{figure}

As illustrated in Figure \ref{fig:dag}, when a physician chooses to batch order ($X_{ij} = 1$), it directly influences the total volume of imaging tests performed, $M_{ij}$. Our model posits two key mechanisms for this relationship:

\begin{enumerate}
  \item Precautionary ordering: In the face of diagnostic uncertainty, physicians may opt for a comprehensive initial workup, potentially including tests that might not have been ordered in a sequential approach.
  \item Reduced information gain: By ordering tests simultaneously, physicians forego the opportunity to use the results of initial tests to inform the necessity of subsequent ones. This can lead to a higher overall test volume compared to a sequential approach where each test order is informed by preceding results.
\end{enumerate}

The total volume of tests, $M_{ij}$, in turn, impacts key patient outcomes, $Y_{ij}$, particularly length of stay (LOS) and the probability of a 72-hour return visit with admission. The relationship between test volume and these outcomes is complex. On one hand, more tests generally require more time, potentially increasing LOS. However, if the physician's initial diagnostic uncertainty is high, ordering more tests upfront might expedite the diagnostic process and lead to quicker treatment decisions.

Importantly, our model also allows for a direct effect of the batch ordering decision on outcomes, independent of its effect through test volume. This direct effect could represent, for example, the efficiency gains from parallel processing of multiple tests, which might reduce LOS even as the total number of tests increases. The model acknowledges the presence of unobserved confounding factors, $U_{ij}$, which might influence both the decision to batch order and patient outcomes. These could include factors like the physician's assessment of the patient's condition severity or subtle clinical cues not captured in standard data.

To formalize these relationships, we express them mathematically as follows:

\begin{equation}
p(X_{ij}|Z_j) = f(Z_j, U_{ij})
\end{equation}

\begin{equation}
M_{ij} = g(X_{ij}, U_{ij})
\end{equation}

\begin{equation}
Y_{ij} = h(X_{ij}, M_{ij}, U_{ij})
\end{equation}

To address potential endogeneity, our empirical strategy leverages the physician's general tendency to batch order, $Z_j$, as an instrument for the actual batching decision, $X_{ij}$. This approach allows us to estimate the causal effects of batch ordering on test volume and patient outcomes, even in the presence of unobserved confounders.

Our empirical analysis focuses on estimating these key relationships:

\begin{enumerate}
  \item The effect of physician tendency on batching decisions: $\frac{\partial X_{ij}}{\partial Z_j}$
  \item The effect of batching on test volume: $\frac{\partial M_{ij}}{\partial X_{ij}}$
  \item The total effect of batching on patient outcomes: $\frac{\partial Y_{ij}}{\partial X_{ij}}$
\end{enumerate}

In summary, our theoretical model provides a framework for understanding the complex decision-making process in ED diagnostic testing. It highlights the potential trade-offs involved in batch ordering – between comprehensive initial assessment and efficient resource use, between rapid testing and informed sequential decision-making. By formalizing these relationships, we set the stage for our empirical analysis, which aims to quantify these effects and provide insights for optimizing ED operations and patient care.

# Data and Definitions {#sec:II}

Our study was conducted in the Emergency Department (ED) of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation. During the study period, the ED recorded approximately 43,000 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces. The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume. Physicians operate in a unique workflow that includes staggered 8.5-hour shifts and a rotational patient assignment system, minimizing potential selection bias in patient encounters.

We conducted a retrospective review of comprehensive ED operational data from 10/6/2018 through 12/31/2019, coinciding with the initiation of a new electronic medical record. The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics. This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.

## Sample Construction and Summary Statistics

Our research design focuses on adult visits to the Mayo Clinic of Arizona ED. From an initial dataset of approximately 48,000 visits during the study period, we applied several exclusion criteria to improve statistical power and ensure the validity of our physician batch tendency instrument. We excluded encounters with rare chief complaints (fewer than 1000 total encounters) and complaints where batch ordering occurs in less than 5% of cases. Examples of excluded complaints include Upper Respiratory Symptoms and Urinary Complaints. This exclusion strategy is unlikely to introduce selection bias as long as physician test batching tendency is not strongly correlated with diagnosing behavior across broad complaint categories. To estimate a precise measure of physician-level batch tendency, we further restricted our sample to encounters involving full-time physicians who treat over 500 ED cases per year. These criteria resulted in a final analytical sample of 16,361 patient encounters.

Table \ref{tab:summary_statistics} provides a comprehensive overview of ED characteristics, patient demographics, and medical tests in our study site. The ED manages an average volume of 24.3 patients, indicating a significant yet manageable workload. The patient population exhibits a range of critical conditions, with 18.8% presenting as tachycardic, 9.51% as tachypneic, 2.43% with fever, and 1.57% as hypotensive. Diagnostic procedures in our sample heavily rely on laboratory tests, used in 78.5% of encounters. Imaging plays a crucial role, with X-rays ordered in 48.1% of visits, non-contrast CT scans in 23.1%, and contrast CT scans in 20.5%. Ultrasounds are used in 12.5% of encounters. Notably, 6.32% of visits involve batch ordering of imaging tests, a key focus of our study.

\begin{table}[ht]
\centering
\caption{Summary Statistics of Emergency Department Encounters}
\label{tab:summary_statistics}
\begin{tabular}{p{10.5cm}cccc}
\toprule
\textbf{} & \textbf{Mean} & \textbf{Q1} & \textbf{Median} & \textbf{Q3} \\
\midrule
\multicolumn{5}{l}{\textbf{Emergency Department Characteristics}} \\
Patients in ED & 24.3 & 16 & 25 & 33 \\
\\
Tachycardic & 0.188 & & & \\
Tachypneic & 0.095 & & & \\
Febrile & 0.024 & & & \\
Hypotensive & 0.016 & & & \\
\\
ESI & 2.74 & 2 & 3 & 3 \\
\\
Complaint: Abdominal Complaints & 0.169 & & & \\
Complaint: Extremity Complaints & 0.143 & & & \\
Complaint: Chest Pain & 0.096 & & & \\
Complaint: Neurological Issue & 0.095 & & & \\
Complaint: Gastrointestinal Issues & 0.090 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
Arrival Age & 58.3 & 44 & 61 & 75 \\
\\
Race: White & 0.886 & & & \\
Race: Black & 0.041 & & & \\
Race: Asian & 0.030 & & & \\
\\
Gender: Female & 0.545 & & & \\
\midrule
\multicolumn{5}{l}{\textbf{Tests}} \\
X-Ray & 0.481 & & & \\
Ultrasound & 0.125 & & & \\
Non-Contrast CT & 0.231 & & & \\
Contrast CT & 0.205 & & & \\
Lab & 0.785 & & & \\
\\
Imaging Tests were Batch Ordered & 0.063 & & & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text. Vital signs were categorized as follows: tachycardia (pulse more significant than $100$), tachypnea (respiratory rate greater than $20$), fever (temperature greater than $38^\circ C$), and hypotension (systolic blood pressure less than $90$).}
\end{tablenotes}
\end{table}

This sample composition and these descriptive statistics provide essential context for our subsequent analyses of batch ordering practices and their impacts on ED efficiency and patient outcomes. The diversity in patient characteristics, presenting complaints, and diagnostic approaches underscores the complexity of ED operations and the potential significance of test ordering strategies in this setting.

## Variable Definitions

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. Below, we detail our primary outcomes: (a) ED length of stay (LOS), and (b) total imaging volume, and (c) 72-hour return with admission.

*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings. It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital. The hypothesis is that batch testing may lead to shorter ED-LOS, potentially improving patient throughput and reducing crowding.

*(b) Resource Utilization*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives. In this study, we quantify resource utilization by the total number of distinct diagnostic tests ordered per patient during their ED stay. This encompasses both initial and subsequent tests. The hypothesis is that batch testing may lead to variations in the number of tests ordered, potentially influencing the overall healthcare expenditure and efficiency.

*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This outcome is used to assess the quality of care provided during the initial ED visit. The hypothesis is that batch testing may lead to variations in the quality of care provided, potentially influencing the likelihood of a return visit.

### Batching

We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window. In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study. Therefore, a batch in our study consists of two or more distinct imaging tests.

# Empirical Strategy {#sec:3}

Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court. Similarly, we explore batching variation across physicians who work in the same emergency department. In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians. Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision. We employ both approaches and lay out their details in the next subsections.

## Institutional Details on Patient-Physician Assignment

Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED. In most EDs, however, physicians have discretion in picking their patients. In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

\begin{table}[htbp]
    \centering
    \caption{Balancing Test: Wald Test for Equality of Means}
    \label{tab:wald_test}
    \begin{tabular}{p{10cm}cccc}
        \toprule
        Chief Complaints & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        Abdominal Complaints & 6232 & 2.587 & 0.108 \\ 
        Back or Flank Pain & 2552 & 1.637 & 0.201 \\ 
        Chest Pain & 3525 & 0.407 & 0.524 \\ 
        Extremity Complaints & 5265 & 1.847 & 0.174 \\ 
        Falls, Motor Vehicle Crashes, Assaults, and Trauma & 2381 & 0.023 & 0.880 \\ 
        Gastrointestinal Issues & 3323 & 0.105 & 0.746 \\ 
        Neurological Issue & 3495 & 0.135 & 0.713 \\ 
        Shortness of Breath & 2966 & 1.324 & 0.250 \\ 
        Skin Complaints & 2178 & 0.383 & 0.536 \\ 
        Upper Respiratory Symptoms & 1917 & 0.017 & 0.896 \\ 
        \midrule
        Emergency Severity & Frequency & F-Statistic & $Pr(> F)$ \\
        \midrule
        ESI 1 or 2 & 13914 & 0.011 & 0.915 \\ 
        ESI 3, 4, or 5 & 29386 & 0.010 & 0.921 \\ 
        \bottomrule
    \end{tabular}
\begin{tablenotes}
\small
\item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset. A balanced distribution implies that complaints and severity are evenly distributed across providers, which we expect to be the case due to randomization. The Wald F-statistic and p-value are reported. Robust standard errors (type HC1) were used to account for potential heteroscedasticity in the data.}
\end{tablenotes}
\end{table}

## Batch Tendency Construction

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \alpha_{lab} + \varepsilon_{i,t}
\end{equation}

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision. Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case. As stated earlier, these controls are more than what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:

\begin{equation}
Tendency_{i,j}^{phys} =
\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{\varepsilon}_{i'}
\end{equation}

where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure \ref{fig:physician_batching} graphs the frequency with which physicians batch order across different chief complaints, highlighting that the variation in batching differs starkly. We note that there is high correlation between

\begin{figure}[h]
  \centering
  \caption{Variation in Physician Batching}
  \label{fig:physician_batching}
  \includegraphics[width=4.25in]{../outputs/figures/Figure 1.png}
\begin{tablenotes}
\small
\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.}
\end{tablenotes}  
\end{figure}

\hyperref[table:first_stage]{Table } illustrates the "first stage" results, emphasizing the robust relevance of the instrument across various controls. Being assigned to a $10$pp higher batch-tendency physician is associated with a $17.6$pp increase in the probability of having tests batch-ordered in the ED. The F-statistic is 21.6 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only.

\begin{table}[!htbp] \centering 
  \caption{Comparison of First Stage Estimates}
  \label{table:first_stage}
  \begin{tabularx}{5.5in}{Xccc} % 'X' for the first column and 'c' for the others
  \toprule
   & \multicolumn{3}{c}{Coeffecient} \\
  \cmidrule{2-4}
   & (1) & (2) & (3) \\
  \midrule
  Batch Tendency & 1.92$^{***}$ & 1.91$^{***}$ & 1.76$^{***}$ \\ 
   & (0.07) & (0.06) & (1.76) \\ 
  \textit{Day of Week-Time of Day FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
  \textit{Month of Year FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
  \textit{Complaint/Severity FE} & & $\checkmark$ & $\checkmark$ \\
  \textit{Laboratory Tests Ordered} & & & $\checkmark$ \\
  \midrule
  F Statistic & 9.55 & 17.86 & 21.59 \\ 
  N & 16,361 & 16,361 & 16,361 \\ 
  \bottomrule
  \end{tabularx}
  \begin{tablenotes}
  \small
  \item Estimates of the first stage for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Column 3 corresponds to the baseline controls. Robust standard errors are clustered at the physician level.
  \item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
  \end{tablenotes}
\end{table}

To estimate the reduced-form effects of being treated by a batch-preferring physician, we estimate the following equation:

\begin{equation}
Y_i = \mu_0 + \mu_1 Tendency_{i,j}^{phys} + \gamma X_i + \nu_i
\end{equation}

This reduced form allows us to check the strength of our instrument. To study the effects of test batching in the ED, we employ a multi-stage approach using our baseline sample:

\begin{equation}
Tests_i = \alpha_0 + \alpha_1 Batched_i + \delta X_i + \eta_i
\end{equation}

\begin{equation}
Y_i = \beta_0 + \beta_1 Batched_i + \beta_2 Tests_i + \theta X_i + \varepsilon_i
\end{equation}

Where $Tests_i$ is the number of imaging tests ordered, and $Y_i$ represents our patient outcomes of interest: length of stay and 72-hour return with admission. $X_i$ is the same set of control variables as in the reduced-form approach. The $Batched_i$ variable suffers from potential endogeneity concerns. For example, unobserved injury severity may correlate with both the need for multiple tests and patient outcomes. To address this, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We use a causal mediation analysis to decompose the total effect of batching into its direct effect on outcomes and its indirect effect through the number of tests ordered. This approach employs bootstrapping to estimate confidence intervals for the direct and indirect effects. We cluster standard errors at the physician level to account for potential correlation in outcomes among patients treated by the same physician, and to properly adjust for the shared physician-level instrument across patients.

## Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (4) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity. We reported a strong first stage (i.e., relevance) at the end of the previous Section. The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching. This is perhaps our strongest assumption and is at its core, untestable. However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings. First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher. Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest. The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis. Nevertheless, we take this assumption seriously and perform a placebo check in Section 5.2 as well as various robustness checks in Section 6.

Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects. It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician. The literature leveraging the judges design typically performs two informal tests for its implications. The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects). The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration). Appendix Table  \ref{tab:monotonicity} presents both of these tests in the two columns for various subsamples of interest.

# Results {#sec:4}

## Reduced-Form Results

In this section, we explore the causal influence of physician batch tendency on total imaging volumne in the emergency department. We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination. Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.

To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency. Specifically, we calculate the physician's residualized leave-out average imaging test order rate, which serves as a proxy for their propensity to order tests. It is important to note the strong positive correlation between batch tendency and testing inclination, suggesting that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies. Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.

Reduced-form regression results in Table \ref{tab:reducedform} reveal the effect the average total effect of batch tendency on the number of distinct imaging tests ordered for our final sample. We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered. The increased testing volume, in turn, is associated with an extended LOS and possibly lower rate of return with admission.

\begin{table}[!htbp] \centering 
  \caption{Reduced Form Results: Effect of Batch Tendency on the Number of Unique Imaging Tests Ordered}
  \label{table:reduced_form}
  \begin{tabularx}{5.5in}{Xccc} % 'X' for the first column and 'c' for the others
  \toprule
   & \multicolumn{3}{c}{Coeffecient} \\
  \cmidrule{2-4}
   & (1) & (2) & (3) \\
  \midrule
  Batch Tendency  & 4.543$^{***}$ & 1.11$^{***}$ & 1.11$^{*}$ \\ 
   & (0.532) & (0.124) & (0.447) \\
  \textit{Baseline Controls} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Testing Inclination}  & & $\checkmark$ & $\checkmark$ \\
  \textit{Any Imaging Ordered} & & & $\checkmark$ \\ % Assumed to be included in (4)
  \midrule
  $R^2$ & 0.203 & 0.207 & 0.599 \\ 
  N & 16,361 & 16,361 & 16,361 \\ 
  \bottomrule
  \end{tabularx}
  \begin{tablenotes}
  \small
  \item Estimates of the reduced form for the baseline sample and baseline controls described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Robust standard errors are clustered at the physician level.
  \item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
  \end{tablenotes}
\end{table}

## Placebo Check

In this section we investigate whether the reduced-form effects observed in Section 5.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency. We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing. For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes. Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.

To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a $>5$ percent batching rate). We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (6). The results of this exercise are displayed in Appendix Table 1. They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.

## Instrumental Variables Results

This section presents our causal analysis of batch ordering imaging tests in the ED, focusing on its effects on resource utilization and patient outcomes. Our analysis reveals nuanced impacts of batch ordering, with significant implications for ED management and patient care.

### Effect of Batch Ordering on Imaging Tests and Patient Outcomes

Table \ref{tab:combined_results} presents the results of our two-stage least squares (2SLS) analysis and subsequent causal mediation analysis for our analytical sample.

\begin{table}[!htbp]
\centering
\caption{Combined 2SLS and Causal Mediation Analysis Results}
\label{tab:combined_results}
\begin{threeparttable}
\begin{tabularx}{\textwidth}{Xcccc}
\toprule
& Number of Imaging Tests & Log LOS & 72-hr Return with Admission \\ \cmidrule(l){2-2} \cmidrule(l){3-3} \cmidrule(l){4-5}
\midrule
\multicolumn{5}{l}{\textit{2SLS Results}} \\
\\
Batched & 0.699** & -0.539 & -0.036 & \\
& (0.286) & (1.534) & (0.034) & \\
\\
  \textit{Baseline Controls} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Testing Inclination} &  $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
  \textit{Imaging Ordered} & $\checkmark$ & $\checkmark$ & $\checkmark$ \\ 
\midrule
\multicolumn{5}{l}{\textit{Causal Mediation Analysis}} \\
\\
ACME & & 0.083*  [0.017, 0.15] & -0.003**  [-0.005, 0.00] \\
ADE &  & -0.565  [-3.608, 2.56] & -0.034  [-0.101, 0.03] \\
Total Effect &  & -0.481  [-3.549, 2.60] & -0.037  [-0.104, 0.03] \\
\bottomrule
\end{tabularx}
\end{threeparttable}
\end{table}

Our 2SLS estimates reveal a significant positive effect of batch ordering on the number of imaging tests performed. Specifically, batch ordering leads to an increase of 0.699 imaging tests (95% CI: [0.138, 1.260], p < 0.05) per patient encounter. This finding confirms our hypothesis that batch ordering results in a higher volume of diagnostic imaging compared to a sequential approach. The magnitude of this effect suggests that for every 100 batch orders, approximately 70 additional imaging tests are performed that would not have occurred under a sequential ordering strategy.

This excess in imaging tests can be attributed to the fundamental difference between batch and sequential ordering approaches. In a sequential approach, physicians can incorporate the information gained from initial tests to inform decisions about subsequent testing, potentially reducing unnecessary examinations. Batch ordering, by contrast, commits to multiple tests upfront, without the benefit of this stepwise information gain.

### Mediation Analysis: Imaging Volume as a Mediator

Our causal mediation analysis provides crucial insights into how the increased imaging volume affects key patient outcomes. For length of stay (LOS), we find a significant Average Causal Mediation Effect (ACME) of 0.0834 (95% CI: [0.0171, 0.15], p < 0.1). This positive ACME indicates that batch ordering indirectly extends ED stays through increased testing. Notably, while the Average Direct Effect (ADE) of batch ordering on LOS is negative (-0.565), it is not statistically significant. These findings challenge the conventional wisdom that batch ordering is a time-saving measure. While there may be some efficiency gains in the testing process itself (as suggested by the negative, albeit non-significant, ADE), these are more than offset by the additional time required to perform and interpret the excess tests generated by batch ordering. The net result is a tendency towards longer patient stays in the ED.

Interestingly, our analysis of 72-hour return rates with admission reveals a contrasting effect. We observe a small but statistically significant negative ACME of -0.00274 (95% CI: [-0.00550, 0.00], p < 0.05). This suggests that the increased imaging from batch ordering may slightly reduce the probability of patients returning to the ED within 72 hours and requiring admission. This finding indicates that the "shotgun" approach of comprehensive initial testing may have some benefits in terms of thorough patient evaluation, potentially catching issues that might otherwise lead to return visits. These results paint a nuanced picture of the impacts of batch ordering in the ED. On one hand, it leads to increased resource utilization and longer patient stays, which could strain ED capacity and potentially contribute to overcrowding. On the other hand, it may result in more comprehensive initial evaluations that slightly reduce short-term return visits.

The complexity of these findings underscores the need for careful consideration in ED test ordering strategies. While batch ordering may offer some benefits in terms of comprehensive initial evaluations, it also carries significant costs in terms of resource utilization and patient flow. ED managers and policymakers must weigh these trade-offs carefully when developing guidelines for diagnostic testing. These results also highlight the importance of considering both direct and indirect effects when evaluating ED processes. The apparent efficiency of batch ordering in the testing process itself may be misleading if one does not account for the downstream consequences of increased test volume. Given the varying impacts of batch ordering across different outcomes, it is likely that the appropriateness of this practice may depend on specific clinical scenarios or patient characteristics. To further explore this possibility and provide more targeted insights for ED management, we next conduct a subgroup analysis examining the effects of batch ordering across different presenting complaints.

### Subgroup Analysis by Chief Complaint

Our subgroup analysis reveals important differences in the effects of batch ordering across various presenting complaints, offering valuable insights for optimizing diagnostic strategies in the ED. Table \ref{tab:subgroup_analysis} presents these results.

\begin{table}[!htbp]
\centering
\caption{Subgroup Analysis Results by Chief Complaint}
\label{tab:subgroup_analysis}
\tiny
\begin{tabularx}{\textwidth}{Xcccccc}
\toprule
Chief Complaint & Batched Coefficient & ACME LOS & ADE LOS & ACME 72hr Return & ADE 72hr Return \\
\midrule
Extremity Complaints & 0.709* & 0.133* & 0.268 & -0.003 & -0.090 \\
 & (0.411) & [-0.030, 0.290] & [-3.640, 3.870] & [-0.012, 0.000] & [-0.214, 0.040] \\
 \\
Falls/MVCs/Trauma & 0.268 & 0.040 & -0.019 & -0.002 & 0.037 \\
 & (0.422) & [-0.103, 0.180] & [-3.109, 2.960] & [-0.009, 0.000] & [-0.055, 0.130] \\
 \\
Dizziness/Syncope & 1.572** & 0.193** & -1.353 & -0.011** & -0.023 \\
 & (0.701) & [0.015, 0.390] & [-4.043, 1.370] & [-0.025, 0.000] & [-0.270, 0.210] \\
 \\
Neurological Issue & 1.652 & 0.139 & -7.353 & 0.003 & -0.220 \\
 & (1.983) & [-0.242, 0.520] & [-16.602, 2.500] & [-0.020, 0.030] & [-0.987, 0.560] \\
 \\
Fatigue and Weakness & 0.793 & 0.072 & -0.286 & -0.002 & 0.010 \\
 & (0.686) & [-0.059, 0.220] & [-1.422, 0.940] & [-0.013, 0.000] & [-0.102, 0.120] \\
 \\
Fevers/Sweats/Chills & -0.181 & -0.025 & -1.086 & 0.002 & 0.061 \\
 & (0.858) & [-0.250, 0.200] & [-3.079, 1.000] & [-0.020, 0.030] & [-0.336, 0.430] \\
\bottomrule
\end{tabularx}
\end{table}

For extremity complaints, which range from minor injuries to potentially severe conditions like deep vein thrombosis, batch ordering significantly increases imaging volume (coefficient = 0.709, p < 0.1) and indirectly extends ED length of stay (ACME = 0.133, 95% CI: [-0.030, 0.290], p < 0.1). However, this approach also shows a trend towards reduced 72-hour returns (ACME = -0.003, 95% CI: [-0.012, 0.000]). These findings suggest that while batch ordering for extremity complaints leads to more resource utilization and longer stays, it may be preventing missed diagnoses of serious conditions like fractures or vascular issues. For extremity complaints, a targeted batch ordering approach may be beneficial. Clinicians should consider batch ordering for patients presenting with symptoms suggestive of more serious conditions (e.g., suspected fractures, vascular compromises) or for those with complex medical histories. For minor injuries or low-risk presentations, a sequential approach might be more appropriate to balance resource use and patient flow.

Dizziness and syncope complaints demonstrate even stronger effects of batch ordering. The significant increase in imaging volume (coefficient = 1.572, p < 0.05) translates to longer ED stays (ACME = 0.193, 95% CI: [0.015, 0.390], p < 0.05) but also significantly reduces 72-hour returns (ACME = -0.011, 95% CI: [-0.025, 0.000], p < 0.05). This pattern likely reflects the complex nature of these complaints, which can stem from benign causes or indicate serious underlying conditions like stroke or cardiac issues. For dizziness and syncope, the reduction in return visits justifies a more liberal use of batch ordering. The potential to identify serious underlying conditions outweighs the increased length of stay. However, clinicians should still use clinical judgment to identify low-risk patients (e.g., young patients with clear peripheral vertigo) who might benefit from a more sequential approach.

For other complaint categories, including falls/trauma, neurological issues, and general symptoms like fatigue or fever, the benefits of batch ordering are less clear-cut. In these cases, the lack of significant effects on return visits suggests that a more selective, sequential approach to test ordering may be more appropriate. For these complaints, we recommend a more individualized approach. Clinicians should rely on thorough initial assessments and risk stratification to determine the need for comprehensive imaging. Decision support tools incorporating factors like patient age, comorbidities, and specific symptoms could guide this process.

It is important to consider several key tradeoffs when implementing batch ordering in the ED:

1. Resource Utilization vs. Patient Safety: While batch ordering increases resource use and ED length of stay, the potential reduction in return visits for certain complaints suggests improved diagnostic accuracy. EDs must weigh the upfront costs against the long-term benefits of preventing missed diagnoses and return visits.

2. ED Flow vs. Comprehensive Care: Longer ED stays due to batch ordering could impact overall ED flow. However, for complaints like dizziness/syncope, this tradeoff appears justified by the reduced return rates. EDs should consider implementing parallel processes (e.g., rapid results reporting, streamlined discharge planning) to mitigate the impact on patient flow.

3. Cost Considerations: While our study doesn't directly address costs, the increase in imaging tests from batch ordering likely increases immediate healthcare costs. However, this may be offset by reduced costs from prevented return visits and earlier diagnosis of serious conditions. A comprehensive cost-benefit analysis would be valuable for future research.

## Robustness

### Deifinition of Batching

The 2SLS results presented in the previous sections are robust to several alternative specifications probing the key batching definition. Our primary definition operationalizes batching as the ordering of multiple diagnostic tests within a 5-minute window. This approach is grounded in standard emergency medicine practices and offers a clear distinction between batching and non-batching behaviors. To ensure the robustness of our findings, we performed several sensitivity analyses.

(1) Variation in the Time Window for Batching: Firstly, we varied the time window for what constitutes a batch. The original 5-minute window was extended and contracted to understand its impact on the study's outcomes. Specifically, we considered scenarios where a batch is defined as two or more tests ordered within windows of 1 minute to 10 minutes. This range allowed us to capture a broader spectrum of batching behaviors and test the sensitivity of our results to different batching definitions.

(2) Refinement of Batching Definition: Secondly, we refined our batching definition based on the sequence and timing of test orders. In the initial analysis, if two tests were ordered upfront and another test ordered later, it was counted as a batch. We modified this definition to consider a set of tests as a batch only if all tests ordered during a patient encounter came from that initial batch order. This adjustment ensures that our batching definition more accurately reflects a comprehensive diagnostic effort at the outset of patient care, rather than incremental decision-making.

The key finding from these robustness checks is the consistent impact of batching on key outcomes across all variations. As seen in Appendix Table 2 and 3, altering the time window for batching, whether narrowing it to as little as 1 minute or expanding it to 10 minutes, did not qualitatively change the results. Similarly, refining the definition of batching to consider only those tests ordered in the initial batch also had negligible impact on the study's main findings. These results lend further credibility to our initial findings, demonstrating that our conclusions about the impact of batching on patient outcomes in the ED are not sensitive to the specific operational definition of batching.

### External Validity

To assess the generalizability of our findings and ensure their robustness, we conducted an additional analysis using data from the Emergency Department of Massachusetts General Hospital (MGH). This high-volume ED sees approximately 110,000 patient encounters annually, providing a diverse and extensive dataset for comparison. Our analysis utilized 12 months of ED visit data from MGH, encompassing a wide range of patient presentations, diagnoses, and outcomes. This dataset provides a valuable contrast to our primary study site, offering insights into how batch ordering practices and their effects may vary in a different clinical environment.

It is important to note that the MGH ED does not employ the same quasi-random assignment of patients to physicians as our primary study site. This difference in patient allocation necessitated an alternative analytical approach. To address potential selection bias and create a fair comparison between batched and non-batched cases, we employed a propensity score matching technique, which involves:

- Score Calculation: We calculated propensity scores for each patient encounter based on key variables including: Chief complaint, Severity (as measured by the Emergency Severity Index or an equivalent triage score), Time of day, Day of week, and Patient demographics (age, gender).

- Matching: Using these propensity scores, we matched patients who received batch-ordered imaging tests with those who did not. We employed a nearest-neighbor matching algorithm with a caliper width of 0.2 standard deviations of the logit of the propensity score, which is widely accepted in the literature as a balance between matching quality and number of matched pairs.

- Balance Checking: After matching, we conducted balance checks to ensure that the matched groups were comparable across all covariates used in the propensity score calculation. This included visual inspection of covariate distributions and statistical tests (e.g., standardized mean differences) to confirm adequate balance.

- Outcome Comparison: With our matched sample, we compared outcomes between the batched and non-batched groups, focusing on the same key metrics as our primary analysis: length of stay, number of imaging tests performed, and 72-hour return rates with admission.

This approach allows us to estimate the effect of batch ordering in a setting where randomization is not feasible, by creating comparable groups of batched and non-batched cases. While this method cannot fully replicate the strengths of our primary quasi-experimental design, it provides a valuable check on the external validity of our findings. By comparing the results from this analysis to our primary findings, we can assess whether the effects of batch ordering on ED outcomes are consistent across different hospital settings and patient populations. This comparison enhances the generalizability of our conclusions and provides additional context for our recommendations on ED imaging practices. The results of this external validation analysis are presented in Appendix Table XX.

# Conclusion

This study provides empirical evidence on the impacts of batch ordering imaging tests in emergency departments, illuminating the trade-offs between comprehensive initial assessments and efficient resource utilization. Our findings, derived from a quasi-experimental design leveraging physician practice variation, align closely with our theoretical model of ED decision-making. We demonstrate that batch ordering, while leading to more thorough initial evaluations, results in increased resource use and longer ED stays.

The fundamental tension we uncover is between the information gain approach of sequential testing and the comprehensive but potentially excessive nature of batch ordering. Our results indicate that batch ordering leads to approximately 70 additional imaging tests per 100 patient encounters compared to a sequential approach. This substantial increase in testing volume translates to longer ED stays but also slightly reduced 72-hour return rates with admission. These findings underscore the complex decision-making environment in EDs, where physicians must balance the need for thorough diagnostics against operational efficiency and patient flow.

Importantly, our subgroup analysis reveals significant heterogeneity in the effects of batch ordering across different presenting complaints. For conditions with high diagnostic uncertainty or potential for serious underlying issues, such as dizziness or syncope, the benefits of comprehensive initial testing may outweigh the costs of increased resource utilization. Conversely, for more straightforward presentations, a sequential, information gain approach appears more appropriate. This variation in effects emphasizes the need for nuanced, complaint-specific approaches to test ordering in emergency settings.

Our results have significant implications for ED management and health policy. They suggest that a one-size-fits-all approach to diagnostic testing in the ED is suboptimal. Instead, ED managers and policymakers should consider implementing targeted batch ordering strategies that account for the specific nature of patient presentations and the relative risks of missed diagnoses versus overcrowding and resource strain. Future research should focus on developing and validating decision support tools that can guide clinicians in choosing between batch and sequential ordering based on patient-specific factors and presenting complaints.

In conclusion, this study contributes to our understanding of ED operations by quantifying the causal impacts of different diagnostic testing strategies. By highlighting the nuanced trade-offs involved in batch ordering, we provide a foundation for more informed decision-making in emergency care. As healthcare systems continue to grapple with the dual challenges of providing high-quality care and managing limited resources, insights from studies like this one will be crucial in shaping evidence-based policies and practices in emergency medicine.

\newpage

### 2024-09-09 manuscript/CopyOfmanuscript.tex
getting going

--- 2024-09-09 manuscript/CopyOfmanuscript.Rmd
@@
----
-title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
-authors:
-- familyname: Jameson
-  othernames: Jacob
-  affiliation: Interfaculty Initiative in Health Policy, Harvard University
-  email: jacobjameson@g.harvard.edu
-- familyname: Saghafian
-  othernames:  Soroush
-  affiliation: Kennedy School of Government, Harvard University 
-  email: soroush_saghafian@hks.harvard.edu
-- familyname: Huckman
-  othernames:  Robert
-  affiliation: Harvard Business School 
-  email: rhuckman@hbs.edu
-- familyname: Hodgson
-  othernames:  Nicole
-  affiliation: Mayo Clinic
-  email: Hodgson.Nicole@mayo.edu
-keywords:
-  - Emergency Department, Operational Effeciency, Diagnostic Testing
-abstract: |
-  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits...
-bibliography: references.bib
-blind: false
-output: rticles::informs_article
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{tabularx}
-  - \usepackage{graphicx}
-  - \usepackage{tikz}
-  - \usepackage{siunitx}
-  - \usepackage{tablefootnote}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \usepackage{natbib}
-  - \usepackage{caption}
----
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
+%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
+%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
+%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[,,nonblindrev]{informs}
 
-# Introduction {#sec:I}
+\OneAndAHalfSpacedXI
+%%\OneAndAHalfSpacedXII % Current default line spacing
+%%\DoubleSpacedXII
+%%\DoubleSpacedXI
 
-Healthcare delivery, particularly in the emergency department (ED), requires a delicate balance between ensuring optimal patient outcomes and optimizing resource utilization.
-Achieving these twin goals necessitates timely and accurate diagnosis, which enables prompt and appropriate treatment, improving patient prognosis and reducing the likelihood of adverse events.
-Efficient patient discharge from the ED is crucial for alleviating overcrowding, a severe issue associated with higher complication rates and increased mortality @bernstein2009.
-In this complex environment, the availability and performance of diagnostic tests, especially imaging studies, play a pivotal role in shaping both patient care and operational efficiency @naseim2015.
+%% BEGIN MY ADDITIONS %%
+\usepackage{hyperref}
 
-A critical question in ED management pertains to whether physicians should batch order diagnostic imaging tests or order them sequentially.
-This decision represents a fundamental tradeoff between potentially reducing patient length of stay and risking over-testing.
-Batch ordering, where multiple imaging tests are ordered simultaneously, may expedite the diagnostic process but can lead to unnecessary tests, increased costs, patient anxiety, and potential harm from follow-up of false-positive results @koch2018.
-Conversely, a more sequential approach might reduce unnecessary testing but could extend a patient's time in the ED, potentially contributing to overcrowding.
-Striking the right balance between comprehensive evaluation and efficient resource use is crucial, yet research on optimal test ordering strategies in the ED remains limited @saghafian2015.
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-In this paper, we leverage data from over 45,000 ED patient visits to quantify the benefits and consequences of batching versus sequentially ordering diagnostic imaging tests.
-Our focus on imaging tests is particularly relevant given their significant impact on patient flow, resource utilization, and potential health risks from radiation exposure.
-We employ an empirical strategy that exploits the random assignment of patients to ED physicians who differ in their propensity to batch-order imaging tests.
-This quasi-experimental design allows us to overcome potential selection bias and move closer to identifying the causal impact of batch ordering on key outcomes such as patient length of stay, resource utilization, and the likelihood of return visits.
+\usepackage{booktabs}
+\usepackage{tabularx}
+\usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{siunitx}
+\usepackage{tablefootnote}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\usepackage{natbib}
+\usepackage{caption}
 
-To capture physician tendencies towards batch ordering, we introduce a novel measure using a leave-out, residualized approach based on each physician's historical ordering patterns for all other patients seen during the study period.
-This tendency measure strongly predicts actual batch ordering behavior while remaining uncorrelated with patient and ED visit characteristics, providing a robust instrument for our analysis.
-Our findings reveal significant practice variation among physicians and demonstrate substantial impacts on ED operations and patient care trajectories.
+%% END MY ADDITIONS %%
 
-We begin by evaluating the reduced-form effects of ED provider batch-ordering tendency on downstream patient outcomes and ED turnaround time.
-Our results indicate that practice variation, as captured by physician batch-ordering tendency, has large and significant consequences.
-For instance, being treated by a provider in the top decile of the batching tendency distribution, compared to one in the bottom decile, results in additional diagnostic imaging tests per 100 patient encounters, even after controlling for the physician's underlying propensity to test.
-This variation in practice patterns raises important questions about the optimal use of diagnostic resources in emergency care settings.
+% Natbib setup for author-year style
+\usepackage{natbib}
+ \bibpunct[, ]{(}{)}{,}{a}{}{,}%
+ \def\bibfont{\small}%
+ \def\bibsep{\smallskipamount}%
+ \def\bibhang{24pt}%
+ \def\newblock{\ }%
+ \def\BIBand{and}%
 
-To address potential confounding factors, we employ a placebo exercise to demonstrate that the observed effects are primarily due to batching behavior rather than other aspects of physician care that might correlate with batching tendency.
-We find no significant effects of physician batch tendency on key outcomes for a "placebo sample" of patients visiting the ED for conditions that rarely result in batched orders.
-This strengthens our confidence in the specificity of the batching effect on patient outcomes and resource utilization.
+%% Setup of theorem styles. Outcomment only one.
+%% Preferred default is the first option.
+\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
+%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
+\ECRepeatTheorems
 
-Our research design closely approximates a randomized controlled trial that assigns patients to batch-ordering or sequential-ordering arms, leveraging the unique features of the ED setting.
-In this environment, patients have no discretion over provider selection, and in our specific ED, the random assignment of patients to physicians further mitigates selection issues.
-The wide variation in batch-ordering practices among physicians within the same hospital, despite following identical guidelines, provides the necessary variation for our analysis.
-Moreover, the typically short, one-off nature of ED patient-physician interactions constrains decision-making to a more limited, observable set of choices compared to other healthcare settings.
+%% Setup of the equation numbering system. Outcomment only one.
+%% Preferred default is the first option.
+\EquationsNumberedThrough    % Default: (1), (2), ...
+%\EquationsNumberedBySection % (1.1), (1.2), ...
 
-By exploiting this practice variation in ED settings, we aim to isolate the impact of test batching from other potential channels that might influence length of stay and patient outcomes in different healthcare contexts.
-This approach allows us to more accurately identify the causal impact of batch-ordering diagnostic imaging tests on patient outcomes and resource utilization.
-Importantly, our study examines batching decisions made within the bounds of normal clinical judgment and practice norms, rather than focusing on deviations from clinical guidelines or instances of substandard care.
+% For new submissions, leave this number blank.
+% For revisions, input the manuscript number assigned by the on-line
+% system along with a suffix ".Rx" where x is the revision number.
+\MANUSCRIPTNO{}
 
-The implications of this research extend beyond academic interest, offering potential insights for ED management, healthcare policy, and the design of clinical decision support systems.
-Understanding the true impacts of batch ordering practices could inform more efficient resource allocation strategies, improve patient flow, and ultimately enhance the quality of emergency care delivery.
-The remainder of this paper is structured as follows: The next section describes our data source and outlines the baseline sample.
-Section II describes the theoretical and conceptual model.
-Section IV details our empirical strategy and its accompanying identifying assumptions.
-Section V presents our results, including both reduced-form effects and instrumental variable analyses.
-Section V also discusses the implications of our findings for batch-ordering policies and ED management.
-The final section concludes with a summary of our contributions and directions for future research.
+%%%%%%%%%%%%%%%%
+\begin{document}
+%%%%%%%%%%%%%%%%
 
-# Theoretical Model
+% Outcomment only when entries are known. Otherwise leave as is and
+%   default values will be used.
+%\setcounter{page}{1}
+%\VOLUME{00}%
+%\NO{0}%
+%\MONTH{Xxxxx}% (month or a similar seasonal id)
+%\YEAR{0000}% e.g., 2005
+%\FIRSTPAGE{000}%
+%\LASTPAGE{000}%
+%\SHORTYEAR{00}% shortened year (two-digit)
+%\ISSUE{0000} %
+%\LONGFIRSTPAGE{0001} %
+%\DOI{10.1287/xxxx.0000.0000}%
 
-Emergency departments (EDs) are complex environments where critical decisions about patient care must be made rapidly and under conditions of uncertainty.
-Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.
+% Author's names for the running heads
+% Sample depending on the number of authors;
+\RUNAUTHOR{%
+Jameson, Saghafian, Huckman
+ and Hodgson
+}
+% \RUNAUTHOR{Jones and Wilson}
+% \RUNAUTHOR{Jones, Miller, and Wilson}
+% \RUNAUTHOR{Jones et al.} % for four or more authors
+% Enter authors following the given pattern:
+%\RUNAUTHOR{}
 
-In a typical ED scenario, patients arrive with varying levels of acuity and a wide range of presenting symptoms.
-Upon arrival, each patient $i$ is assigned to a physician $j$, often based on availability rather than specific matching criteria.
-The physician's primary task is to diagnose and treat the patient efficiently and effectively, a process that frequently involves ordering diagnostic tests, including imaging studies.
+\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}
 
-Central to our model is the concept of the patient's underlying condition, denoted by $\tau_i$.
-This represents the true state of the patient's health, which is initially unknown to the treating physician.
-The physician's goal is to uncover $\tau_i$ as accurately and quickly as possible to provide appropriate care.
-However, the path to this discovery is not straightforward and involves critical decisions about diagnostic testing.
+\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}
 
-The physician's approach to ordering tests is influenced by various factors, including their individual tendencies and experiences.
-We represent this as $Z_j$, the physician's propensity to batch order tests.
-This tendency can be thought of as a stable characteristic of the physician, shaped by their training, past experiences, and personal practice style.
-The decision to batch order tests for a given patient, represented by $X_{ij}$, is a function of both the physician's tendency $Z_j$ and the specifics of the patient's presentation.
-Batch ordering involves requesting multiple imaging tests simultaneously, rather than in a sequential manner.
-This decision has cascading effects on the patient's journey through the ED.
+\ARTICLEAUTHORS{%
+\AUTHOR{Jacob Jameson}
+\AFF{Interfaculty Initiative in Health Policy, Harvard
+University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}
+
+\AUTHOR{Soroush Saghafian}
+\AFF{Kennedy School of Government, Harvard
+University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}
+
+\AUTHOR{Robert Huckman}
+\AFF{Harvard Business
+School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}
+
+\AUTHOR{Nicole Hodgson}
+\AFF{Mayo
+Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}
+
+%
+}
+
+\ABSTRACT{We use practice variation across physicians to uncover the
+role of test-ordering on care delivery in the ED. Using records of over
+45,000 Mayo Clinic emergency department visits\ldots{}}
+
+\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
+Testing}
+
+\maketitle
+
+\hypertarget{sec:I}{%
+\section{Introduction}\label{sec:I}}
+
+Healthcare delivery, particularly in the emergency department (ED),
+requires a delicate balance between ensuring optimal patient outcomes
+and optimizing resource utilization. Achieving these twin goals
+necessitates timely and accurate diagnosis, which enables prompt and
+appropriate treatment, improving patient prognosis and reducing the
+likelihood of adverse events. Efficient patient discharge from the ED is
+crucial for alleviating overcrowding, a severe issue associated with
+higher complication rates and increased mortality \citet{bernstein2009}.
+In this complex environment, the availability and performance of
+diagnostic tests, especially imaging studies, play a pivotal role in
+shaping both patient care and operational efficiency \citet{naseim2015}.
+
+A critical question in ED management pertains to whether physicians
+should batch order diagnostic imaging tests or order them sequentially.
+This decision represents a fundamental tradeoff between potentially
+reducing patient length of stay and risking over-testing. Batch
+ordering, where multiple imaging tests are ordered simultaneously, may
+expedite the diagnostic process but can lead to unnecessary tests,
+increased costs, patient anxiety, and potential harm from follow-up of
+false-positive results \citet{koch2018}. Conversely, a more sequential
+approach might reduce unnecessary testing but could extend a patient's
+time in the ED, potentially contributing to overcrowding. Striking the
+right balance between comprehensive evaluation and efficient resource
+use is crucial, yet research on optimal test ordering strategies in the
+ED remains limited \citet{saghafian2015}.
+
+In this paper, we leverage data from over 45,000 ED patient visits to
+quantify the benefits and consequences of batching versus sequentially
+ordering diagnostic imaging tests. Our focus on imaging tests is
+particularly relevant given their significant impact on patient flow,
+resource utilization, and potential health risks from radiation
+exposure. We employ an empirical strategy that exploits the random
+assignment of patients to ED physicians who differ in their propensity
+to batch-order imaging tests. This quasi-experimental design allows us
+to overcome potential selection bias and move closer to identifying the
+causal impact of batch ordering on key outcomes such as patient length
+of stay, resource utilization, and the likelihood of return visits.
+
+To capture physician tendencies towards batch ordering, we introduce a
+novel measure using a leave-out, residualized approach based on each
+physician's historical ordering patterns for all other patients seen
+during the study period. This tendency measure strongly predicts actual
+batch ordering behavior while remaining uncorrelated with patient and ED
+visit characteristics, providing a robust instrument for our analysis.
+Our findings reveal significant practice variation among physicians and
+demonstrate substantial impacts on ED operations and patient care
+trajectories.
+
+We begin by evaluating the reduced-form effects of ED provider
+batch-ordering tendency on downstream patient outcomes and ED turnaround
+time. Our results indicate that practice variation, as captured by
+physician batch-ordering tendency, has large and significant
+consequences. For instance, being treated by a provider in the top
+decile of the batching tendency distribution, compared to one in the
+bottom decile, results in additional diagnostic imaging tests per 100
+patient encounters, even after controlling for the physician's
+underlying propensity to test. This variation in practice patterns
+raises important questions about the optimal use of diagnostic resources
+in emergency care settings.
+
+To address potential confounding factors, we employ a placebo exercise
+to demonstrate that the observed effects are primarily due to batching
+behavior rather than other aspects of physician care that might
+correlate with batching tendency. We find no significant effects of
+physician batch tendency on key outcomes for a ``placebo sample'' of
+patients visiting the ED for conditions that rarely result in batched
+orders. This strengthens our confidence in the specificity of the
+batching effect on patient outcomes and resource utilization.
+
+Our research design closely approximates a randomized controlled trial
+that assigns patients to batch-ordering or sequential-ordering arms,
+leveraging the unique features of the ED setting. In this environment,
+patients have no discretion over provider selection, and in our specific
+ED, the random assignment of patients to physicians further mitigates
+selection issues. The wide variation in batch-ordering practices among
+physicians within the same hospital, despite following identical
+guidelines, provides the necessary variation for our analysis. Moreover,
+the typically short, one-off nature of ED patient-physician interactions
+constrains decision-making to a more limited, observable set of choices
+compared to other healthcare settings.
+
+By exploiting this practice variation in ED settings, we aim to isolate
+the impact of test batching from other potential channels that might
+influence length of stay and patient outcomes in different healthcare
+contexts. This approach allows us to more accurately identify the causal
+impact of batch-ordering diagnostic imaging tests on patient outcomes
+and resource utilization. Importantly, our study examines batching
+decisions made within the bounds of normal clinical judgment and
+practice norms, rather than focusing on deviations from clinical
+guidelines or instances of substandard care.
+
+The implications of this research extend beyond academic interest,
+offering potential insights for ED management, healthcare policy, and
+the design of clinical decision support systems. Understanding the true
+impacts of batch ordering practices could inform more efficient resource
+allocation strategies, improve patient flow, and ultimately enhance the
+quality of emergency care delivery. The remainder of this paper is
+structured as follows: The next section describes our data source and
+outlines the baseline sample. Section II describes the theoretical and
+conceptual model. Section IV details our empirical strategy and its
+accompanying identifying assumptions. Section V presents our results,
+including both reduced-form effects and instrumental variable analyses.
+Section V also discusses the implications of our findings for
+batch-ordering policies and ED management. The final section concludes
+with a summary of our contributions and directions for future research.
+
+\hypertarget{theoretical-model}{%
+\section{Theoretical Model}\label{theoretical-model}}
+
+Emergency departments (EDs) are complex environments where critical
+decisions about patient care must be made rapidly and under conditions
+of uncertainty. Our theoretical model aims to capture the nuanced
+dynamics of diagnostic decision-making in this high-stakes setting, with
+a particular focus on the practice of batch ordering imaging tests.
+
+In a typical ED scenario, patients arrive with varying levels of acuity
+and a wide range of presenting symptoms. Upon arrival, each patient
+\(i\) is assigned to a physician \(j\), often based on availability
+rather than specific matching criteria. The physician's primary task is
+to diagnose and treat the patient efficiently and effectively, a process
+that frequently involves ordering diagnostic tests, including imaging
+studies.
+
+Central to our model is the concept of the patient's underlying
+condition, denoted by \(\tau_i\). This represents the true state of the
+patient's health, which is initially unknown to the treating physician.
+The physician's goal is to uncover \(\tau_i\) as accurately and quickly
+as possible to provide appropriate care. However, the path to this
+discovery is not straightforward and involves critical decisions about
+diagnostic testing.
+
+The physician's approach to ordering tests is influenced by various
+factors, including their individual tendencies and experiences. We
+represent this as \(Z_j\), the physician's propensity to batch order
+tests. This tendency can be thought of as a stable characteristic of the
+physician, shaped by their training, past experiences, and personal
+practice style. The decision to batch order tests for a given patient,
+represented by \(X_{ij}\), is a function of both the physician's
+tendency \(Z_j\) and the specifics of the patient's presentation. Batch
+ordering involves requesting multiple imaging tests simultaneously,
+rather than in a sequential manner. This decision has cascading effects
+on the patient's journey through the ED.
 
@@
 
-As illustrated in Figure \ref{fig:dag}, when a physician chooses to batch order ($X_{ij} = 1$), it directly influences the total volume of imaging tests performed, $M_{ij}$.
-Our model posits two key mechanisms for this relationship:
+As illustrated in Figure \ref{fig:dag}, when a physician chooses to
+batch order (\(X_{ij} = 1\)), it directly influences the total volume of
+imaging tests performed, \(M_{ij}\). Our model posits two key mechanisms
+for this relationship:
 
@@
 
-The total volume of tests, $M_{ij}$, in turn, impacts key patient outcomes, $Y_{ij}$, particularly length of stay (LOS) and the probability of a 72-hour return visit with admission.
-The relationship between test volume and these outcomes is complex.
-On one hand, more tests generally require more time, potentially increasing LOS.
-However, if the physician's initial diagnostic uncertainty is high, ordering more tests upfront might expedite the diagnostic process and lead to quicker treatment decisions.
+The total volume of tests, \(M_{ij}\), in turn, impacts key patient
+outcomes, \(Y_{ij}\), particularly length of stay (LOS) and the
+probability of a 72-hour return visit with admission. The relationship
+between test volume and these outcomes is complex. On one hand, more
+tests generally require more time, potentially increasing LOS. However,
+if the physician's initial diagnostic uncertainty is high, ordering more
+tests upfront might expedite the diagnostic process and lead to quicker
+treatment decisions.
 
-Importantly, our model also allows for a direct effect of the batch ordering decision on outcomes, independent of its effect through test volume.
-This direct effect could represent, for example, the efficiency gains from parallel processing of multiple tests, which might reduce LOS even as the total number of tests increases.
-The model acknowledges the presence of unobserved confounding factors, $U_{ij}$, which might influence both the decision to batch order and patient outcomes.
-These could include factors like the physician's assessment of the patient's condition severity or subtle clinical cues not captured in standard data.
+Importantly, our model also allows for a direct effect of the batch
+ordering decision on outcomes, independent of its effect through test
+volume. This direct effect could represent, for example, the efficiency
+gains from parallel processing of multiple tests, which might reduce LOS
+even as the total number of tests increases. The model acknowledges the
+presence of unobserved confounding factors, \(U_{ij}\), which might
+influence both the decision to batch order and patient outcomes. These
+could include factors like the physician's assessment of the patient's
+condition severity or subtle clinical cues not captured in standard
+data.
 
-To formalize these relationships, we express them mathematically as follows:
+To formalize these relationships, we express them mathematically as
+follows:
 
@@
 
-To address potential endogeneity, our empirical strategy leverages the physician's general tendency to batch order, $Z_j$, as an instrument for the actual batching decision, $X_{ij}$.
-This approach allows us to estimate the causal effects of batch ordering on test volume and patient outcomes, even in the presence of unobserved confounders.
+To address potential endogeneity, our empirical strategy leverages the
+physician's general tendency to batch order, \(Z_j\), as an instrument
+for the actual batching decision, \(X_{ij}\). This approach allows us to
+estimate the causal effects of batch ordering on test volume and patient
+outcomes, even in the presence of unobserved confounders.
 
@@
 
-In summary, our theoretical model provides a framework for understanding the complex decision-making process in ED diagnostic testing.
-It highlights the potential trade-offs involved in batch ordering – between comprehensive initial assessment and efficient resource use, between rapid testing and informed sequential decision-making.
-By formalizing these relationships, we set the stage for our empirical analysis, which aims to quantify these effects and provide insights for optimizing ED operations and patient care.
+In summary, our theoretical model provides a framework for understanding
+the complex decision-making process in ED diagnostic testing. It
+highlights the potential trade-offs involved in batch ordering --
+between comprehensive initial assessment and efficient resource use,
+between rapid testing and informed sequential decision-making. By
+formalizing these relationships, we set the stage for our empirical
+analysis, which aims to quantify these effects and provide insights for
+optimizing ED operations and patient care.
 
-# Data and Definitions {#sec:II}
+\hypertarget{sec:II}{%
+\section{Data and Definitions}\label{sec:II}}
 
-Our study was conducted in the Emergency Department (ED) of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation.
-During the study period, the ED recorded approximately 43,000 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces.
-The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume.
-Physicians operate in a unique workflow that includes staggered 8.5-hour shifts and a rotational patient assignment system, minimizing potential selection bias in patient encounters.
+Our study was conducted in the Emergency Department (ED) of the Mayo
+Clinic of Arizona, a tertiary care hospital without obstetrical
+services, an inpatient pediatrics unit, or a trauma designation. During
+the study period, the ED recorded approximately 43,000 visits per year,
+managed across 26 treatment rooms and up to 9 hallway spaces. The
+department is exclusively staffed by board-eligible or board-certified
+emergency physicians (EPs), with rotating residents overseeing about
+10\% of patient volume. Physicians operate in a unique workflow that
+includes staggered 8.5-hour shifts and a rotational patient assignment
+system, minimizing potential selection bias in patient encounters.
 
-We conducted a retrospective review of comprehensive ED operational data from 10/6/2018 through 12/31/2019, coinciding with the initiation of a new electronic medical record.
-The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics.
-This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.
+We conducted a retrospective review of comprehensive ED operational data
+from 10/6/2018 through 12/31/2019, coinciding with the initiation of a
+new electronic medical record. The dataset includes detailed patient
+demographics, chief complaints, vital signs, emergency severity index
+(ESI), length of stay (LOS), and resource utilization metrics. This
+period was chosen to provide a robust data set while excluding the
+influence of the coronavirus pandemic.
 
-## Sample Construction and Summary Statistics
+\hypertarget{sample-construction-and-summary-statistics}{%
+\subsection{Sample Construction and Summary
+Statistics}\label{sample-construction-and-summary-statistics}}
 
-Our research design focuses on adult visits to the Mayo Clinic of Arizona ED.
-From an initial dataset of approximately 48,000 visits during the study period, we applied several exclusion criteria to improve statistical power and ensure the validity of our physician batch tendency instrument.
-We excluded encounters with rare chief complaints (fewer than 1000 total encounters) and complaints where batch ordering occurs in less than 5% of cases.
-Examples of excluded complaints include Upper Respiratory Symptoms and Urinary Complaints.
-This exclusion strategy is unlikely to introduce selection bias as long as physician test batching tendency is not strongly correlated with diagnosing behavior across broad complaint categories.
-To estimate a precise measure of physician-level batch tendency, we further restricted our sample to encounters involving full-time physicians who treat over 500 ED cases per year.
-These criteria resulted in a final analytical sample of 16,361 patient encounters.
+Our research design focuses on adult visits to the Mayo Clinic of
+Arizona ED. From an initial dataset of approximately 48,000 visits
+during the study period, we applied several exclusion criteria to
+improve statistical power and ensure the validity of our physician batch
+tendency instrument. We excluded encounters with rare chief complaints
+(fewer than 1000 total encounters) and complaints where batch ordering
+occurs in less than 5\% of cases. Examples of excluded complaints
+include Upper Respiratory Symptoms and Urinary Complaints. This
+exclusion strategy is unlikely to introduce selection bias as long as
+physician test batching tendency is not strongly correlated with
+diagnosing behavior across broad complaint categories. To estimate a
+precise measure of physician-level batch tendency, we further restricted
+our sample to encounters involving full-time physicians who treat over
+500 ED cases per year. These criteria resulted in a final analytical
+sample of 16,361 patient encounters.
 
-Table \ref{tab:summary_statistics} provides a comprehensive overview of ED characteristics, patient demographics, and medical tests in our study site.
-The ED manages an average volume of 24.3 patients, indicating a significant yet manageable workload.
-The patient population exhibits a range of critical conditions, with 18.8% presenting as tachycardic, 9.51% as tachypneic, 2.43% with fever, and 1.57% as hypotensive.
-Diagnostic procedures in our sample heavily rely on laboratory tests, used in 78.5% of encounters.
-Imaging plays a crucial role, with X-rays ordered in 48.1% of visits, non-contrast CT scans in 23.1%, and contrast CT scans in 20.5%.
-Ultrasounds are used in 12.5% of encounters.
-Notably, 6.32% of visits involve batch ordering of imaging tests, a key focus of our study.
+Table \ref{tab:summary_statistics} provides a comprehensive overview of
+ED characteristics, patient demographics, and medical tests in our study
+site. The ED manages an average volume of 24.3 patients, indicating a
+significant yet manageable workload. The patient population exhibits a
+range of critical conditions, with 18.8\% presenting as tachycardic,
+9.51\% as tachypneic, 2.43\% with fever, and 1.57\% as hypotensive.
+Diagnostic procedures in our sample heavily rely on laboratory tests,
+used in 78.5\% of encounters. Imaging plays a crucial role, with X-rays
+ordered in 48.1\% of visits, non-contrast CT scans in 23.1\%, and
+contrast CT scans in 20.5\%. Ultrasounds are used in 12.5\% of
+encounters. Notably, 6.32\% of visits involve batch ordering of imaging
+tests, a key focus of our study.
 
@@
 
-This sample composition and these descriptive statistics provide essential context for our subsequent analyses of batch ordering practices and their impacts on ED efficiency and patient outcomes.
-The diversity in patient characteristics, presenting complaints, and diagnostic approaches underscores the complexity of ED operations and the potential significance of test ordering strategies in this setting.
+This sample composition and these descriptive statistics provide
+essential context for our subsequent analyses of batch ordering
+practices and their impacts on ED efficiency and patient outcomes. The
+diversity in patient characteristics, presenting complaints, and
+diagnostic approaches underscores the complexity of ED operations and
+the potential significance of test ordering strategies in this setting.
 
-## Variable Definitions
+\hypertarget{variable-definitions}{%
+\subsection{Variable Definitions}\label{variable-definitions}}
 
-Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter.
-While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice.
-Below, we detail our primary outcomes: (a) ED length of stay (LOS), and (b) total imaging volume, and (c) 72-hour return with admission.
+Our explanatory variable in the IV analysis, \(Batched_i\), is an
+indicator for whether patient \(i\) has their tests batch-ordered at
+their ED encounter. While the patient could decide not to undergo the
+tests ordered by the physician, this is rare in practice. Below, we
+detail our primary outcomes: (a) ED length of stay (LOS), and (b) total
+imaging volume, and (c) 72-hour return with admission.
 
-*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
-It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
-The hypothesis is that batch testing may lead to shorter ED-LOS, potentially improving patient throughput and reducing crowding.
+\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
+efficiency and patient throughput in emergency care settings. It is
+defined as the duration from a patient's arrival to the ED until their
+departure, whether by discharge or admission to the hospital. The
+hypothesis is that batch testing may lead to shorter ED-LOS, potentially
+improving patient throughput and reducing crowding.
 
-*(b) Resource Utilization*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives.
-In this study, we quantify resource utilization by the total number of distinct diagnostic tests ordered per patient during their ED stay.
-This encompasses both initial and subsequent tests.
-The hypothesis is that batch testing may lead to variations in the number of tests ordered, potentially influencing the overall healthcare expenditure and efficiency.
+\emph{(b) Resource Utilization}-- Resource utilization in the ED
+typically refers to the extent of medical services and interventions a
+patient receives. In this study, we quantify resource utilization by the
+total number of distinct diagnostic tests ordered per patient during
+their ED stay. This encompasses both initial and subsequent tests. The
+hypothesis is that batch testing may lead to variations in the number of
+tests ordered, potentially influencing the overall healthcare
+expenditure and efficiency.
 
-*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
-This outcome is used to assess the quality of care provided during the initial ED visit.
-The hypothesis is that batch testing may lead to variations in the quality of care provided, potentially influencing the likelihood of a return visit.
+\emph{(c) 72 Hour Return with Admission}-- This is a binary variable
+indicating whether the patient returns to the ED within 72 hours of
+their initial visit and is admitted to the hospital. This outcome is
+used to assess the quality of care provided during the initial ED visit.
+The hypothesis is that batch testing may lead to variations in the
+quality of care provided, potentially influencing the likelihood of a
+return visit.
 
-### Batching
+\hypertarget{batching}{%
+\subsubsection{Batching}\label{batching}}
 
-We define "batching" in line with standard emergency medicine practices.
-Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses.
-This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.
+We define ``batching'' in line with standard emergency medicine
+practices. Batching occurs when a physician simultaneously orders a
+comprehensive set of diagnostic tests, typically covering a broad range
+of potential diagnoses. This contrasts with sequential ordering, where
+tests are ordered sequentially based on the information obtained from
+each test as needed.
 
-We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window.
-In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study.
-Therefore, a batch in our study consists of two or more distinct imaging tests.
+We operationalize batching as occurring when multiple diagnostic imaging
+tests are ordered within a 5-minute window. In Section 4.4, sensitivity
+analyses on this cutoff point showed that our results are robust to this
+definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is
+considered a separate, distinct test for our study. Therefore, a batch
+in our study consists of two or more distinct imaging tests.
 
-# Empirical Strategy {#sec:3}
+\hypertarget{sec:3}{%
+\section{Empirical Strategy}\label{sec:3}}
 
-Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court.
-Similarly, we explore batching variation across physicians who work in the same emergency department.
-In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians.
-Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision.
-We employ both approaches and lay out their details in the next subsections.
+Our empirical strategy closely follows the literature that relies on
+quasi-random assignment of agents to cases, often referred to as the
+``judges design.'' Papers in this literature typically exploit variation
+in the sentencing leniency of judges who work in the same court.
+Similarly, we explore batching variation across physicians who work in
+the same emergency department. In its reduced form, under the assumption
+of quasi-random assignment, this approach allows researchers to identify
+the causal effect of being assigned to different types of physicians.
+Under additional assumptions, an instrumental variable approach
+identifies the causal effect of a given medical decision. We employ both
+approaches and lay out their details in the next subsections.
 
-## Institutional Details on Patient-Physician Assignment
+\hypertarget{institutional-details-on-patient-physician-assignment}{%
+\subsection{Institutional Details on Patient-Physician
+Assignment}\label{institutional-details-on-patient-physician-assignment}}
 
-Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED.
-In most EDs, however, physicians have discretion in picking their patients.
-In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses.
-In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+Contrary to most healthcare settings where patients exhibit choice, they
+are predominantly passive in their physician assignment in the ED. In
+most EDs, however, physicians have discretion in picking their patients.
+In contrast, patients arriving at the Mayo Clinic ED are randomly
+assigned to physicians via a rotational patient assignment algorithm
+\citet{traub2016emergency}, which removes potential selection bias
+concerns for our analyses. In essence, barring arrival time and
+shift-level variation, the physician-to-patient matching can be deemed
+random. Table \ref{tab:wald_test} displays that patient encounters
+(regarding chief complaints and emergency severity) are equitably
+distributed across physicians within our study's cohort.
 
@@
 
-## Batch Tendency Construction
+\hypertarget{batch-tendency-construction}{%
+\subsection{Batch Tendency
+Construction}\label{batch-tendency-construction}}
 
-To measure physician batch tendency, we use the physician's residualized leave-out average batch rate.
-This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways.
-First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.
+To measure physician batch tendency, we use the physician's residualized
+leave-out average batch rate. This measure is derived from two steps
+following the approaches taken by \citet{doyle2015measuring},
+\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
+obtain residuals from a regression model, which includes all ED
+encounters in our sample period.
 
@@
 
-where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$.
-This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument.
-The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.
+where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
+residual from equation (1); \(J\) is the set of all ED encounters
+treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
+number of cases that physician has seen that year, excluding patient
+\(i\). This leave-out mean eliminates the mechanical bias that stems
+from patient \(i\)'s own case entering into the instrument. The measure
+is interpreted as the average (leave-out) batch rate of patient \(i\)'s
+physician, relative to other physicians in that hospital-year-month,
+hospital-day of week-time of day.
 
-We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure \ref{fig:physician_batching} graphs the frequency with which physicians batch order across different chief complaints, highlighting that the variation in batching differs starkly.
-We note that there is high correlation between
+We document that the Mayo Clinic ED physicians exhibit wide, systematic
+variation in their propensity to batch order diagnostic tests. Figure
+\ref{fig:physician_batching} graphs the frequency with which physicians
+batch order across different chief complaints, highlighting that the
+variation in batching differs starkly. We note that there is high
+correlation between
 
@@
 
-\hyperref[table:first_stage]{Table } illustrates the "first stage" results, emphasizing the robust relevance of the instrument across various controls.
-Being assigned to a $10$pp higher batch-tendency physician is associated with a $17.6$pp increase in the probability of having tests batch-ordered in the ED.
-The F-statistic is 21.6 when all controls and fixed effects are included.
-The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only.
+\hyperref[table:first_stage]{Table } illustrates the ``first stage''
+results, emphasizing the robust relevance of the instrument across
+various controls. Being assigned to a \(10\)pp higher batch-tendency
+physician is associated with a \(17.6\)pp increase in the probability of
+having tests batch-ordered in the ED. The F-statistic is 21.6 when all
+controls and fixed effects are included. The coefficient is greater than
+one because all emergency visits are used to construct the tendency
+instrument, while the first stage is calculated using the baseline
+sample only.
 
@@
 
-This reduced form allows us to check the strength of our instrument. To study the effects of test batching in the ED, we employ a multi-stage approach using our baseline sample:
+This reduced form allows us to check the strength of our instrument. To
+study the effects of test batching in the ED, we employ a multi-stage
+approach using our baseline sample:
 
@@
 
-Where $Tests_i$ is the number of imaging tests ordered, and $Y_i$ represents our patient outcomes of interest: length of stay and 72-hour return with admission.
-$X_i$ is the same set of control variables as in the reduced-form approach.
-The $Batched_i$ variable suffers from potential endogeneity concerns.
-For example, unobserved injury severity may correlate with both the need for multiple tests and patient outcomes.
-To address this, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$.
-We use a causal mediation analysis to decompose the total effect of batching into its direct effect on outcomes and its indirect effect through the number of tests ordered.
-This approach employs bootstrapping to estimate confidence intervals for the direct and indirect effects.
-We cluster standard errors at the physician level to account for potential correlation in outcomes among patients treated by the same physician, and to properly adjust for the shared physician-level instrument across patients.
+Where \(Tests_i\) is the number of imaging tests ordered, and \(Y_i\)
+represents our patient outcomes of interest: length of stay and 72-hour
+return with admission. \(X_i\) is the same set of control variables as
+in the reduced-form approach. The \(Batched_i\) variable suffers from
+potential endogeneity concerns. For example, unobserved injury severity
+may correlate with both the need for multiple tests and patient
+outcomes. To address this, we instrument \(Batched_i\) with the assigned
+physician \(j\)'s underlying tendency to batch,
+\(Tendency_{i,j}^{phys}\). We use a causal mediation analysis to
+decompose the total effect of batching into its direct effect on
+outcomes and its indirect effect through the number of tests ordered.
+This approach employs bootstrapping to estimate confidence intervals for
+the direct and indirect effects. We cluster standard errors at the
+physician level to account for potential correlation in outcomes among
+patients treated by the same physician, and to properly adjust for the
+shared physician-level instrument across patients.
 
-## Identifying Assumptions
+\hypertarget{identifying-assumptions}{%
+\subsection{Identifying Assumptions}\label{identifying-assumptions}}
 
-The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence").
-The residualization in equation (4) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.
+The reduced-form approach delivers an unbiased estimate of the causal
+effect of being treated by a higher tendency to batch physician, since
+assignment of patients to ED physicians is random, conditional on
+seasonality and shift (``conditional independence''). The
+residualization in equation (4) controls for more controls than required
+to achieve quasi-random assignment; they are included for statistical
+precision in measuring physician tendency to batch.
 
-Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity.
-We reported a strong first stage (i.e., relevance) at the end of the previous Section.
-The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching.
-This is perhaps our strongest assumption and is at its core, untestable.
-However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings.
-First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher.
-Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest.
-The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis.
-Nevertheless, we take this assumption seriously and perform a placebo check in Section 5.2 as well as various robustness checks in Section 6.
+Our instrumental variable approach, which aims to recover the causal
+effect of having diagnostic tests batch ordered, relies on three
+additional assumptions: relevance, exclusion, and monotonicity. We
+reported a strong first stage (i.e., relevance) at the end of the
+previous Section. The exclusion restriction requires that the instrument
+must influence the outcome of interest only through its effect on test
+batching. This is perhaps our strongest assumption and is at its core,
+untestable. However, several features of the ED setting suggest that
+such violation may likely only have a small impact and may be less
+concerning than in other health care settings. First, unlike in primary
+care settings, where the patient and primary care provider have many
+repeat encounters, the scope of what the emergency physician can do to
+impact medium-term outcomes is limited and well-observed by the
+researcher. Second, any violation of the exclusion restriction needs to
+directly affect the specific outcome of interest. The channel by which
+ED physicians can influence length of stay relative outcomes is likely
+through testing and diagnosis. Nevertheless, we take this assumption
+seriously and perform a placebo check in Section 5.2 as well as various
+robustness checks in Section 6.
 
-Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects.
-It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician.
-The literature leveraging the judges design typically performs two informal tests for its implications.
-The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects).
-The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration).
-Appendix Table  \ref{tab:monotonicity} presents both of these tests in the two columns for various subsamples of interest.
+Finally, the monotonicity assumption is necessary for interpreting the
+coefficient estimates obtained from the IV approach as Local Average
+Treatment Effects (LATEs) if there are heterogeneous treatment effects.
+It requires that any patient who is (not) batched by a sequencer
+(batcher) would also (not) be batched by a batcher (sequencer)
+physician. The literature leveraging the judges design typically
+performs two informal tests for its implications. The first one provides
+that the first stage should be weakly positive for all subsamples
+(\citet{dobbie2018effects}). The second implication asserts that the
+instrument constructed by leaving out a particular subsample has
+predictive power over that same left-out subsample
+(\citet{bhuller2020incarceration}). Appendix Table
+\ref{tab:monotonicity} presents both of these tests in the two columns
+for various subsamples of interest.
 
-# Results {#sec:4}
+\hypertarget{sec:4}{%
+\section{Results}\label{sec:4}}
 
-## Reduced-Form Results
+\hypertarget{reduced-form-results}{%
+\subsection{Reduced-Form Results}\label{reduced-form-results}}
 
-In this section, we explore the causal influence of physician batch tendency on total imaging volumne in the emergency department.
-We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination.
-Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.
+In this section, we explore the causal influence of physician batch
+tendency on total imaging volumne in the emergency department. We posit
+that while batch tendency directly influences the practice of batch
+ordering tests, both batch tendency and batch ordering are concurrently
+influenced by a physician's testing inclination. Given that testing
+inclination directly affects primary outcomes, we include it as a
+control variable in our regression models to mitigate its confounding
+effects.
 
-To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency.
-Specifically, we calculate the physician's residualized leave-out average imaging test order rate, which serves as a proxy for their propensity to order tests.
-It is important to note the strong positive correlation between batch tendency and testing inclination, suggesting that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies.
-Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.
+To quantify a physician's testing inclination, we employ a similar
+approach to that used in measuring physician batch tendency.
+Specifically, we calculate the physician's residualized leave-out
+average imaging test order rate, which serves as a proxy for their
+propensity to order tests. It is important to note the strong positive
+correlation between batch tendency and testing inclination, suggesting
+that physicians with a higher propensity to test may also exhibit a
+higher tendency to batch orders, potentially as a consequence of their
+testing strategies. Neglecting to account for testing inclination could
+lead to overestimated effects of batch tendency due to omitted variable
+bias.
 
-Reduced-form regression results in Table \ref{tab:reducedform} reveal the effect the average total effect of batch tendency on the number of distinct imaging tests ordered for our final sample.
-We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered.
-The increased testing volume, in turn, is associated with an extended LOS and possibly lower rate of return with admission.
+Reduced-form regression results in Table \ref{tab:reducedform} reveal
+the effect the average total effect of batch tendency on the number of
+distinct imaging tests ordered for our final sample. We hypothesize that
+while batch ordering may streamline the testing process, resulting in a
+quicker completion of a given number of tests, it simultaneously appears
+to lead to an increase in the total number of tests ordered. The
+increased testing volume, in turn, is associated with an extended LOS
+and possibly lower rate of return with admission.
 
@@
 
-## Placebo Check
+\hypertarget{placebo-check}{%
+\subsection{Placebo Check}\label{placebo-check}}
 
-In this section we investigate whether the reduced-form effects observed in Section 5.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency.
-We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
-For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
-Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.
+In this section we investigate whether the reduced-form effects observed
+in Section 5.1 are due to differences in batch rates across providers or
+due to other provider differences correlated with batch tendency. We
+start by studying reduced-form effects among patients with complaints
+that are never batched, as a ``placebo/falsification check.'' By way of
+example, consider a patient who arrives at the ED with a urinary tact
+infection---a condition for which patients rarely undergo imaging
+testing. For such patients, we should expect to see no impact of batch
+tendency only if high batching and low batching physicians do not
+systematically differ in other dimensions of care relevant to patient
+outcomes. Conversely, if we do find a reduced-form effect for these
+patients, then high batch tendency physicians must systematically differ
+from low batch tendency physicians in other dimensions of care, beyond
+batching.
 
-To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a $>5$ percent batching rate).
-We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (6).
-The results of this exercise are displayed in Appendix Table 1.
-They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
+To that end, we restrict attention to ED visits for complaints where
+batching occurs no more than 5 percent of the time (recall that our
+baseline sample only includes complaints with a \(>5\) percent batching
+rate). We estimate a reduced-form regression of each main outcome on
+physician batch tendency for the subsample, following equation (6). The
+results of this exercise are displayed in Appendix Table 1. They show
+that in contrast to results for our main sample, the association between
+physician tendency to batch and a given outcome is statistically
+indistinguishable from zero and much smaller in magnitude for the
+samples of patients who visit the ED with health conditions that are
+rarely batched.
 
-## Instrumental Variables Results
+\hypertarget{instrumental-variables-results}{%
+\subsection{Instrumental Variables
+Results}\label{instrumental-variables-results}}
 
-This section presents our causal analysis of batch ordering imaging tests in the ED, focusing on its effects on resource utilization and patient outcomes.
-Our analysis reveals nuanced impacts of batch ordering, with significant implications for ED management and patient care.
+This section presents our causal analysis of batch ordering imaging
+tests in the ED, focusing on its effects on resource utilization and
+patient outcomes. Our analysis reveals nuanced impacts of batch
+ordering, with significant implications for ED management and patient
+care.
 
-### Effect of Batch Ordering on Imaging Tests and Patient Outcomes
+\hypertarget{effect-of-batch-ordering-on-imaging-tests-and-patient-outcomes}{%
+\subsubsection{Effect of Batch Ordering on Imaging Tests and Patient
+Outcomes}\label{effect-of-batch-ordering-on-imaging-tests-and-patient-outcomes}}
 
-Table \ref{tab:combined_results} presents the results of our two-stage least squares (2SLS) analysis and subsequent causal mediation analysis for our analytical sample.
+Table \ref{tab:combined_results} presents the results of our two-stage
+least squares (2SLS) analysis and subsequent causal mediation analysis
+for our analytical sample.
 
@@
 
-Our 2SLS estimates reveal a significant positive effect of batch ordering on the number of imaging tests performed.
-Specifically, batch ordering leads to an increase of 0.699 imaging tests (95% CI: [0.138, 1.260], p < 0.05) per patient encounter.
-This finding confirms our hypothesis that batch ordering results in a higher volume of diagnostic imaging compared to a sequential approach.
-The magnitude of this effect suggests that for every 100 batch orders, approximately 70 additional imaging tests are performed that would not have occurred under a sequential ordering strategy.
+Our 2SLS estimates reveal a significant positive effect of batch
+ordering on the number of imaging tests performed. Specifically, batch
+ordering leads to an increase of 0.699 imaging tests (95\% CI: {[}0.138,
+1.260{]}, p \textless{} 0.05) per patient encounter. This finding
+confirms our hypothesis that batch ordering results in a higher volume
+of diagnostic imaging compared to a sequential approach. The magnitude
+of this effect suggests that for every 100 batch orders, approximately
+70 additional imaging tests are performed that would not have occurred
+under a sequential ordering strategy.
 
-This excess in imaging tests can be attributed to the fundamental difference between batch and sequential ordering approaches.
-In a sequential approach, physicians can incorporate the information gained from initial tests to inform decisions about subsequent testing, potentially reducing unnecessary examinations.
-Batch ordering, by contrast, commits to multiple tests upfront, without the benefit of this stepwise information gain.
+This excess in imaging tests can be attributed to the fundamental
+difference between batch and sequential ordering approaches. In a
+sequential approach, physicians can incorporate the information gained
+from initial tests to inform decisions about subsequent testing,
+potentially reducing unnecessary examinations. Batch ordering, by
+contrast, commits to multiple tests upfront, without the benefit of this
+stepwise information gain.
 
-### Mediation Analysis: Imaging Volume as a Mediator
+\hypertarget{mediation-analysis-imaging-volume-as-a-mediator}{%
+\subsubsection{Mediation Analysis: Imaging Volume as a
+Mediator}\label{mediation-analysis-imaging-volume-as-a-mediator}}
 
-Our causal mediation analysis provides crucial insights into how the increased imaging volume affects key patient outcomes.
-For length of stay (LOS), we find a significant Average Causal Mediation Effect (ACME) of 0.0834 (95% CI: [0.0171, 0.15], p < 0.1).
-This positive ACME indicates that batch ordering indirectly extends ED stays through increased testing.
-Notably, while the Average Direct Effect (ADE) of batch ordering on LOS is negative (-0.565), it is not statistically significant.
-These findings challenge the conventional wisdom that batch ordering is a time-saving measure.
-While there may be some efficiency gains in the testing process itself (as suggested by the negative, albeit non-significant, ADE), these are more than offset by the additional time required to perform and interpret the excess tests generated by batch ordering.
-The net result is a tendency towards longer patient stays in the ED.
+Our causal mediation analysis provides crucial insights into how the
+increased imaging volume affects key patient outcomes. For length of
+stay (LOS), we find a significant Average Causal Mediation Effect (ACME)
+of 0.0834 (95\% CI: {[}0.0171, 0.15{]}, p \textless{} 0.1). This
+positive ACME indicates that batch ordering indirectly extends ED stays
+through increased testing. Notably, while the Average Direct Effect
+(ADE) of batch ordering on LOS is negative (-0.565), it is not
+statistically significant. These findings challenge the conventional
+wisdom that batch ordering is a time-saving measure. While there may be
+some efficiency gains in the testing process itself (as suggested by the
+negative, albeit non-significant, ADE), these are more than offset by
+the additional time required to perform and interpret the excess tests
+generated by batch ordering. The net result is a tendency towards longer
+patient stays in the ED.
 
-Interestingly, our analysis of 72-hour return rates with admission reveals a contrasting effect.
-We observe a small but statistically significant negative ACME of -0.00274 (95% CI: [-0.00550, 0.00], p < 0.05).
-This suggests that the increased imaging from batch ordering may slightly reduce the probability of patients returning to the ED within 72 hours and requiring admission.
-This finding indicates that the "shotgun" approach of comprehensive initial testing may have some benefits in terms of thorough patient evaluation, potentially catching issues that might otherwise lead to return visits.
-These results paint a nuanced picture of the impacts of batch ordering in the ED.
-On one hand, it leads to increased resource utilization and longer patient stays, which could strain ED capacity and potentially contribute to overcrowding.
-On the other hand, it may result in more comprehensive initial evaluations that slightly reduce short-term return visits.
+Interestingly, our analysis of 72-hour return rates with admission
+reveals a contrasting effect. We observe a small but statistically
+significant negative ACME of -0.00274 (95\% CI: {[}-0.00550, 0.00{]}, p
+\textless{} 0.05). This suggests that the increased imaging from batch
+ordering may slightly reduce the probability of patients returning to
+the ED within 72 hours and requiring admission. This finding indicates
+that the ``shotgun'' approach of comprehensive initial testing may have
+some benefits in terms of thorough patient evaluation, potentially
+catching issues that might otherwise lead to return visits. These
+results paint a nuanced picture of the impacts of batch ordering in the
+ED. On one hand, it leads to increased resource utilization and longer
+patient stays, which could strain ED capacity and potentially contribute
+to overcrowding. On the other hand, it may result in more comprehensive
+initial evaluations that slightly reduce short-term return visits.
 
-The complexity of these findings underscores the need for careful consideration in ED test ordering strategies.
-While batch ordering may offer some benefits in terms of comprehensive initial evaluations, it also carries significant costs in terms of resource utilization and patient flow.
-ED managers and policymakers must weigh these trade-offs carefully when developing guidelines for diagnostic testing.
-These results also highlight the importance of considering both direct and indirect effects when evaluating ED processes.
-The apparent efficiency of batch ordering in the testing process itself may be misleading if one does not account for the downstream consequences of increased test volume.
-Given the varying impacts of batch ordering across different outcomes, it is likely that the appropriateness of this practice may depend on specific clinical scenarios or patient characteristics.
-To further explore this possibility and provide more targeted insights for ED management, we next conduct a subgroup analysis examining the effects of batch ordering across different presenting complaints.
+The complexity of these findings underscores the need for careful
+consideration in ED test ordering strategies. While batch ordering may
+offer some benefits in terms of comprehensive initial evaluations, it
+also carries significant costs in terms of resource utilization and
+patient flow. ED managers and policymakers must weigh these trade-offs
+carefully when developing guidelines for diagnostic testing. These
+results also highlight the importance of considering both direct and
+indirect effects when evaluating ED processes. The apparent efficiency
+of batch ordering in the testing process itself may be misleading if one
+does not account for the downstream consequences of increased test
+volume. Given the varying impacts of batch ordering across different
+outcomes, it is likely that the appropriateness of this practice may
+depend on specific clinical scenarios or patient characteristics. To
+further explore this possibility and provide more targeted insights for
+ED management, we next conduct a subgroup analysis examining the effects
+of batch ordering across different presenting complaints.
 
-### Subgroup Analysis by Chief Complaint
+\hypertarget{subgroup-analysis-by-chief-complaint}{%
+\subsubsection{Subgroup Analysis by Chief
+Complaint}\label{subgroup-analysis-by-chief-complaint}}
 
-Our subgroup analysis reveals important differences in the effects of batch ordering across various presenting complaints, offering valuable insights for optimizing diagnostic strategies in the ED.
-Table \ref{tab:subgroup_analysis} presents these results.
+Our subgroup analysis reveals important differences in the effects of
+batch ordering across various presenting complaints, offering valuable
+insights for optimizing diagnostic strategies in the ED. Table
+\ref{tab:subgroup_analysis} presents these results.
 
@@
 
-For extremity complaints, which range from minor injuries to potentially severe conditions like deep vein thrombosis, batch ordering significantly increases imaging volume (coefficient = 0.709, p < 0.1) and indirectly extends ED length of stay (ACME = 0.133, 95% CI: [-0.030, 0.290], p < 0.1).
-However, this approach also shows a trend towards reduced 72-hour returns (ACME = -0.003, 95% CI: [-0.012, 0.000]).
-These findings suggest that while batch ordering for extremity complaints leads to more resource utilization and longer stays, it may be preventing missed diagnoses of serious conditions like fractures or vascular issues.
-For extremity complaints, a targeted batch ordering approach may be beneficial.
-Clinicians should consider batch ordering for patients presenting with symptoms suggestive of more serious conditions (e.g., suspected fractures, vascular compromises) or for those with complex medical histories.
-For minor injuries or low-risk presentations, a sequential approach might be more appropriate to balance resource use and patient flow.
+For extremity complaints, which range from minor injuries to potentially
+severe conditions like deep vein thrombosis, batch ordering
+significantly increases imaging volume (coefficient = 0.709, p
+\textless{} 0.1) and indirectly extends ED length of stay (ACME = 0.133,
+95\% CI: {[}-0.030, 0.290{]}, p \textless{} 0.1). However, this approach
+also shows a trend towards reduced 72-hour returns (ACME = -0.003, 95\%
+CI: {[}-0.012, 0.000{]}). These findings suggest that while batch
+ordering for extremity complaints leads to more resource utilization and
+longer stays, it may be preventing missed diagnoses of serious
+conditions like fractures or vascular issues. For extremity complaints,
+a targeted batch ordering approach may be beneficial. Clinicians should
+consider batch ordering for patients presenting with symptoms suggestive
+of more serious conditions (e.g., suspected fractures, vascular
+compromises) or for those with complex medical histories. For minor
+injuries or low-risk presentations, a sequential approach might be more
+appropriate to balance resource use and patient flow.
 
-Dizziness and syncope complaints demonstrate even stronger effects of batch ordering.
-The significant increase in imaging volume (coefficient = 1.572, p < 0.05) translates to longer ED stays (ACME = 0.193, 95% CI: [0.015, 0.390], p < 0.05) but also significantly reduces 72-hour returns (ACME = -0.011, 95% CI: [-0.025, 0.000], p < 0.05).
-This pattern likely reflects the complex nature of these complaints, which can stem from benign causes or indicate serious underlying conditions like stroke or cardiac issues.
-For dizziness and syncope, the reduction in return visits justifies a more liberal use of batch ordering.
-The potential to identify serious underlying conditions outweighs the increased length of stay.
-However, clinicians should still use clinical judgment to identify low-risk patients (e.g., young patients with clear peripheral vertigo) who might benefit from a more sequential approach.
+Dizziness and syncope complaints demonstrate even stronger effects of
+batch ordering. The significant increase in imaging volume (coefficient
+= 1.572, p \textless{} 0.05) translates to longer ED stays (ACME =
+0.193, 95\% CI: {[}0.015, 0.390{]}, p \textless{} 0.05) but also
+significantly reduces 72-hour returns (ACME = -0.011, 95\% CI:
+{[}-0.025, 0.000{]}, p \textless{} 0.05). This pattern likely reflects
+the complex nature of these complaints, which can stem from benign
+causes or indicate serious underlying conditions like stroke or cardiac
+issues. For dizziness and syncope, the reduction in return visits
+justifies a more liberal use of batch ordering. The potential to
+identify serious underlying conditions outweighs the increased length of
+stay. However, clinicians should still use clinical judgment to identify
+low-risk patients (e.g., young patients with clear peripheral vertigo)
+who might benefit from a more sequential approach.
 
-For other complaint categories, including falls/trauma, neurological issues, and general symptoms like fatigue or fever, the benefits of batch ordering are less clear-cut.
-In these cases, the lack of significant effects on return visits suggests that a more selective, sequential approach to test ordering may be more appropriate.
-For these complaints, we recommend a more individualized approach.
-Clinicians should rely on thorough initial assessments and risk stratification to determine the need for comprehensive imaging.
-Decision support tools incorporating factors like patient age, comorbidities, and specific symptoms could guide this process.
+For other complaint categories, including falls/trauma, neurological
+issues, and general symptoms like fatigue or fever, the benefits of
+batch ordering are less clear-cut. In these cases, the lack of
+significant effects on return visits suggests that a more selective,
+sequential approach to test ordering may be more appropriate. For these
+complaints, we recommend a more individualized approach. Clinicians
+should rely on thorough initial assessments and risk stratification to
+determine the need for comprehensive imaging. Decision support tools
+incorporating factors like patient age, comorbidities, and specific
+symptoms could guide this process.
 
-It is important to consider several key tradeoffs when implementing batch ordering in the ED:
+It is important to consider several key tradeoffs when implementing
+batch ordering in the ED:
 
-1.
-Resource Utilization vs.
-Patient Safety: While batch ordering increases resource use and ED length of stay, the potential reduction in return visits for certain complaints suggests improved diagnostic accuracy.
-EDs must weigh the upfront costs against the long-term benefits of preventing missed diagnoses and return visits.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Resource Utilization vs.~Patient Safety: While batch ordering
+  increases resource use and ED length of stay, the potential reduction
+  in return visits for certain complaints suggests improved diagnostic
+  accuracy. EDs must weigh the upfront costs against the long-term
+  benefits of preventing missed diagnoses and return visits.
+\item
+  ED Flow vs.~Comprehensive Care: Longer ED stays due to batch ordering
+  could impact overall ED flow. However, for complaints like
+  dizziness/syncope, this tradeoff appears justified by the reduced
+  return rates. EDs should consider implementing parallel processes
+  (e.g., rapid results reporting, streamlined discharge planning) to
+  mitigate the impact on patient flow.
+\item
+  Cost Considerations: While our study doesn't directly address costs,
+  the increase in imaging tests from batch ordering likely increases
+  immediate healthcare costs. However, this may be offset by reduced
+  costs from prevented return visits and earlier diagnosis of serious
+  conditions. A comprehensive cost-benefit analysis would be valuable
+  for future research.
+\end{enumerate}
 
-2.
-ED Flow vs.
-Comprehensive Care: Longer ED stays due to batch ordering could impact overall ED flow.
-However, for complaints like dizziness/syncope, this tradeoff appears justified by the reduced return rates.
-EDs should consider implementing parallel processes (e.g., rapid results reporting, streamlined discharge planning) to mitigate the impact on patient flow.
+\hypertarget{robustness}{%
+\subsection{Robustness}\label{robustness}}
 
-3.
-Cost Considerations: While our study doesn't directly address costs, the increase in imaging tests from batch ordering likely increases immediate healthcare costs.
-However, this may be offset by reduced costs from prevented return visits and earlier diagnosis of serious conditions.
-A comprehensive cost-benefit analysis would be valuable for future research.
+\hypertarget{deifinition-of-batching}{%
+\subsubsection{Deifinition of Batching}\label{deifinition-of-batching}}
 
-## Robustness
+The 2SLS results presented in the previous sections are robust to
+several alternative specifications probing the key batching definition.
+Our primary definition operationalizes batching as the ordering of
+multiple diagnostic tests within a 5-minute window. This approach is
+grounded in standard emergency medicine practices and offers a clear
+distinction between batching and non-batching behaviors. To ensure the
+robustness of our findings, we performed several sensitivity analyses.
 
-### Deifinition of Batching
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\item
+  Variation in the Time Window for Batching: Firstly, we varied the time
+  window for what constitutes a batch. The original 5-minute window was
+  extended and contracted to understand its impact on the study's
+  outcomes. Specifically, we considered scenarios where a batch is
+  defined as two or more tests ordered within windows of 1 minute to 10
+  minutes. This range allowed us to capture a broader spectrum of
+  batching behaviors and test the sensitivity of our results to
+  different batching definitions.
+\item
+  Refinement of Batching Definition: Secondly, we refined our batching
+  definition based on the sequence and timing of test orders. In the
+  initial analysis, if two tests were ordered upfront and another test
+  ordered later, it was counted as a batch. We modified this definition
+  to consider a set of tests as a batch only if all tests ordered during
+  a patient encounter came from that initial batch order. This
+  adjustment ensures that our batching definition more accurately
+  reflects a comprehensive diagnostic effort at the outset of patient
+  care, rather than incremental decision-making.
+\end{enumerate}
 
-The 2SLS results presented in the previous sections are robust to several alternative specifications probing the key batching definition.
-Our primary definition operationalizes batching as the ordering of multiple diagnostic tests within a 5-minute window.
-This approach is grounded in standard emergency medicine practices and offers a clear distinction between batching and non-batching behaviors.
-To ensure the robustness of our findings, we performed several sensitivity analyses.
+The key finding from these robustness checks is the consistent impact of
+batching on key outcomes across all variations. As seen in Appendix
+Table 2 and 3, altering the time window for batching, whether narrowing
+it to as little as 1 minute or expanding it to 10 minutes, did not
+qualitatively change the results. Similarly, refining the definition of
+batching to consider only those tests ordered in the initial batch also
+had negligible impact on the study's main findings. These results lend
+further credibility to our initial findings, demonstrating that our
+conclusions about the impact of batching on patient outcomes in the ED
+are not sensitive to the specific operational definition of batching.
 
-(1) Variation in the Time Window for Batching: Firstly, we varied the time window for what constitutes a batch.
-The original 5-minute window was extended and contracted to understand its impact on the study's outcomes.
-Specifically, we considered scenarios where a batch is defined as two or more tests ordered within windows of 1 minute to 10 minutes.
-This range allowed us to capture a broader spectrum of batching behaviors and test the sensitivity of our results to different batching definitions.
+\hypertarget{external-validity}{%
+\subsubsection{External Validity}\label{external-validity}}
 
-(2) Refinement of Batching Definition: Secondly, we refined our batching definition based on the sequence and timing of test orders.
-In the initial analysis, if two tests were ordered upfront and another test ordered later, it was counted as a batch.
-We modified this definition to consider a set of tests as a batch only if all tests ordered during a patient encounter came from that initial batch order.
-This adjustment ensures that our batching definition more accurately reflects a comprehensive diagnostic effort at the outset of patient care, rather than incremental decision-making.
+To assess the generalizability of our findings and ensure their
+robustness, we conducted an additional analysis using data from the
+Emergency Department of Massachusetts General Hospital (MGH). This
+high-volume ED sees approximately 110,000 patient encounters annually,
+providing a diverse and extensive dataset for comparison. Our analysis
+utilized 12 months of ED visit data from MGH, encompassing a wide range
+of patient presentations, diagnoses, and outcomes. This dataset provides
+a valuable contrast to our primary study site, offering insights into
+how batch ordering practices and their effects may vary in a different
+clinical environment.
 
-The key finding from these robustness checks is the consistent impact of batching on key outcomes across all variations.
-As seen in Appendix Table 2 and 3, altering the time window for batching, whether narrowing it to as little as 1 minute or expanding it to 10 minutes, did not qualitatively change the results.
-Similarly, refining the definition of batching to consider only those tests ordered in the initial batch also had negligible impact on the study's main findings.
-These results lend further credibility to our initial findings, demonstrating that our conclusions about the impact of batching on patient outcomes in the ED are not sensitive to the specific operational definition of batching.
+It is important to note that the MGH ED does not employ the same
+quasi-random assignment of patients to physicians as our primary study
+site. This difference in patient allocation necessitated an alternative
+analytical approach. To address potential selection bias and create a
+fair comparison between batched and non-batched cases, we employed a
+propensity score matching technique, which involves:
 
-### External Validity
+\begin{itemize}
+\item
+  Score Calculation: We calculated propensity scores for each patient
+  encounter based on key variables including: Chief complaint, Severity
+  (as measured by the Emergency Severity Index or an equivalent triage
+  score), Time of day, Day of week, and Patient demographics (age,
+  gender).
+\item
+  Matching: Using these propensity scores, we matched patients who
+  received batch-ordered imaging tests with those who did not. We
+  employed a nearest-neighbor matching algorithm with a caliper width of
+  0.2 standard deviations of the logit of the propensity score, which is
+  widely accepted in the literature as a balance between matching
+  quality and number of matched pairs.
+\item
+  Balance Checking: After matching, we conducted balance checks to
+  ensure that the matched groups were comparable across all covariates
+  used in the propensity score calculation. This included visual
+  inspection of covariate distributions and statistical tests (e.g.,
+  standardized mean differences) to confirm adequate balance.
+\item
+  Outcome Comparison: With our matched sample, we compared outcomes
+  between the batched and non-batched groups, focusing on the same key
+  metrics as our primary analysis: length of stay, number of imaging
+  tests performed, and 72-hour return rates with admission.
+\end{itemize}
 
-To assess the generalizability of our findings and ensure their robustness, we conducted an additional analysis using data from the Emergency Department of Massachusetts General Hospital (MGH).
-This high-volume ED sees approximately 110,000 patient encounters annually, providing a diverse and extensive dataset for comparison.
-Our analysis utilized 12 months of ED visit data from MGH, encompassing a wide range of patient presentations, diagnoses, and outcomes.
-This dataset provides a valuable contrast to our primary study site, offering insights into how batch ordering practices and their effects may vary in a different clinical environment.
+This approach allows us to estimate the effect of batch ordering in a
+setting where randomization is not feasible, by creating comparable
+groups of batched and non-batched cases. While this method cannot fully
+replicate the strengths of our primary quasi-experimental design, it
+provides a valuable check on the external validity of our findings. By
+comparing the results from this analysis to our primary findings, we can
+assess whether the effects of batch ordering on ED outcomes are
+consistent across different hospital settings and patient populations.
+This comparison enhances the generalizability of our conclusions and
+provides additional context for our recommendations on ED imaging
+practices. The results of this external validation analysis are
+presented in Appendix Table XX.
 
-It is important to note that the MGH ED does not employ the same quasi-random assignment of patients to physicians as our primary study site.
-This difference in patient allocation necessitated an alternative analytical approach.
-To address potential selection bias and create a fair comparison between batched and non-batched cases, we employed a propensity score matching technique, which involves:
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
-- Score Calculation: We calculated propensity scores for each patient encounter based on key variables including: Chief complaint, Severity (as measured by the Emergency Severity Index or an equivalent triage score), Time of day, Day of week, and Patient demographics (age, gender).
+This study provides empirical evidence on the impacts of batch ordering
+imaging tests in emergency departments, illuminating the trade-offs
+between comprehensive initial assessments and efficient resource
+utilization. Our findings, derived from a quasi-experimental design
+leveraging physician practice variation, align closely with our
+theoretical model of ED decision-making. We demonstrate that batch
+ordering, while leading to more thorough initial evaluations, results in
+increased resource use and longer ED stays.
 
-- Matching: Using these propensity scores, we matched patients who received batch-ordered imaging tests with those who did not.
-We employed a nearest-neighbor matching algorithm with a caliper width of 0.2 standard deviations of the logit of the propensity score, which is widely accepted in the literature as a balance between matching quality and number of matched pairs.
+The fundamental tension we uncover is between the information gain
+approach of sequential testing and the comprehensive but potentially
+excessive nature of batch ordering. Our results indicate that batch
+ordering leads to approximately 70 additional imaging tests per 100
+patient encounters compared to a sequential approach. This substantial
+increase in testing volume translates to longer ED stays but also
+slightly reduced 72-hour return rates with admission. These findings
+underscore the complex decision-making environment in EDs, where
+physicians must balance the need for thorough diagnostics against
+operational efficiency and patient flow.
 
-- Balance Checking: After matching, we conducted balance checks to ensure that the matched groups were comparable across all covariates used in the propensity score calculation.
-This included visual inspection of covariate distributions and statistical tests (e.g., standardized mean differences) to confirm adequate balance.
+Importantly, our subgroup analysis reveals significant heterogeneity in
+the effects of batch ordering across different presenting complaints.
+For conditions with high diagnostic uncertainty or potential for serious
+underlying issues, such as dizziness or syncope, the benefits of
+comprehensive initial testing may outweigh the costs of increased
+resource utilization. Conversely, for more straightforward
+presentations, a sequential, information gain approach appears more
+appropriate. This variation in effects emphasizes the need for nuanced,
+complaint-specific approaches to test ordering in emergency settings.
 
-- Outcome Comparison: With our matched sample, we compared outcomes between the batched and non-batched groups, focusing on the same key metrics as our primary analysis: length of stay, number of imaging tests performed, and 72-hour return rates with admission.
+Our results have significant implications for ED management and health
+policy. They suggest that a one-size-fits-all approach to diagnostic
+testing in the ED is suboptimal. Instead, ED managers and policymakers
+should consider implementing targeted batch ordering strategies that
+account for the specific nature of patient presentations and the
+relative risks of missed diagnoses versus overcrowding and resource
+strain. Future research should focus on developing and validating
+decision support tools that can guide clinicians in choosing between
+batch and sequential ordering based on patient-specific factors and
+presenting complaints.
 
-This approach allows us to estimate the effect of batch ordering in a setting where randomization is not feasible, by creating comparable groups of batched and non-batched cases.
-While this method cannot fully replicate the strengths of our primary quasi-experimental design, it provides a valuable check on the external validity of our findings.
-By comparing the results from this analysis to our primary findings, we can assess whether the effects of batch ordering on ED outcomes are consistent across different hospital settings and patient populations.
-This comparison enhances the generalizability of our conclusions and provides additional context for our recommendations on ED imaging practices.
-The results of this external validation analysis are presented in Appendix Table XX.
-
-# Conclusion
-
-This study provides empirical evidence on the impacts of batch ordering imaging tests in emergency departments, illuminating the trade-offs between comprehensive initial assessments and efficient resource utilization.
-Our findings, derived from a quasi-experimental design leveraging physician practice variation, align closely with our theoretical model of ED decision-making.
-We demonstrate that batch ordering, while leading to more thorough initial evaluations, results in increased resource use and longer ED stays.
-
-The fundamental tension we uncover is between the information gain approach of sequential testing and the comprehensive but potentially excessive nature of batch ordering.
-Our results indicate that batch ordering leads to approximately 70 additional imaging tests per 100 patient encounters compared to a sequential approach.
-This substantial increase in testing volume translates to longer ED stays but also slightly reduced 72-hour return rates with admission.
-These findings underscore the complex decision-making environment in EDs, where physicians must balance the need for thorough diagnostics against operational efficiency and patient flow.
-
-Importantly, our subgroup analysis reveals significant heterogeneity in the effects of batch ordering across different presenting complaints.
-For conditions with high diagnostic uncertainty or potential for serious underlying issues, such as dizziness or syncope, the benefits of comprehensive initial testing may outweigh the costs of increased resource utilization.
-Conversely, for more straightforward presentations, a sequential, information gain approach appears more appropriate.
-This variation in effects emphasizes the need for nuanced, complaint-specific approaches to test ordering in emergency settings.
-
-Our results have significant implications for ED management and health policy.
-They suggest that a one-size-fits-all approach to diagnostic testing in the ED is suboptimal.
-Instead, ED managers and policymakers should consider implementing targeted batch ordering strategies that account for the specific nature of patient presentations and the relative risks of missed diagnoses versus overcrowding and resource strain.
-Future research should focus on developing and validating decision support tools that can guide clinicians in choosing between batch and sequential ordering based on patient-specific factors and presenting complaints.
-
-In conclusion, this study contributes to our understanding of ED operations by quantifying the causal impacts of different diagnostic testing strategies.
-By highlighting the nuanced trade-offs involved in batch ordering, we provide a foundation for more informed decision-making in emergency care.
-As healthcare systems continue to grapple with the dual challenges of providing high-quality care and managing limited resources, insights from studies like this one will be crucial in shaping evidence-based policies and practices in emergency medicine.
+In conclusion, this study contributes to our understanding of ED
+operations by quantifying the causal impacts of different diagnostic
+testing strategies. By highlighting the nuanced trade-offs involved in
+batch ordering, we provide a foundation for more informed
+decision-making in emergency care. As healthcare systems continue to
+grapple with the dual challenges of providing high-quality care and
+managing limited resources, insights from studies like this one will be
+crucial in shaping evidence-based policies and practices in emergency
+medicine.
 
 \newpage
+
+% Appendix here
+% Options are (1) APPENDIX (with or without general title) or
+%             (2) APPENDICES (if it has more than one unrelated sections)
+% Outcomment the appropriate case if necessary
+%
+% \begin{APPENDIX}{<Title of the Appendix>}
+% \end{APPENDIX}
+%
+%   or
+%
+% \begin{APPENDICES}
+% \section{<Title of Section A>}
+% \section{<Title of Section B>}
+% etc
+% \end{APPENDICES}
+
+% Acknowledgments here
+\ACKNOWLEDGMENT{}
+
+\bibliographystyle{informs2014}
+\bibliography{references.bib}
+
+\end{document}

### 2024-09-09 manuscript/New Results.Rmd
getting going

---
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
---
## Batch Tendency Construction

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

\begin{equation}
Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \varepsilon_{i,t}
\end{equation}

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on date $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision. As stated earlier, these controls are what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:

\begin{equation}
Tendency_{i,j}^{phys} =
\frac{1}{N_{-i,j}} \sum_{i' \in \{J \backslash i\}}\hat{\varepsilon}_{i'}
\end{equation}

where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

\newpage

## First Stage

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Tables \ref{tab:linear_regression} and \ref{tab:probit_regression} present the ``first stage" results using both linear and probit regression models. In the linear model, being assigned to a physician with a 10 percentage point (pp) higher batch-tendency is associated with a 10.78 pp increase in the likelihood of having tests batch-ordered in the ED. The probit model suggests an even stronger relationship, with a 10 pp increase in batch-tendency associated with approximately a 35.9 pp increase in the probability of batching on average. Both models demonstrate a strong and statistically significant relationship between physician batch-tendency and the likelihood of test batching, supporting the relevance of our instrument.

\begin{table}[ht]
\centering
\caption{First Stage: Linear Regression of Batch Tendency on Test Batching}
\label{tab:linear_regression}
\begin{tabular}{lccc}
\hline
 & Model 1 & Model 2 & Model 3 \\
\hline
Batching Tendency & 1.096*** & 1.078*** & 1.078*** \\
 & (0.034) & (0.027) & (0.027) \\
N & 46,783 & 46,783 & 46,783 \\
R$^2$ & 0.012 & 0.064 & 0.064 \\
Adjusted R$^2$ & 0.011 & 0.061 & 0.061 \\
\hline
\end{tabular}
\begin{flushleft}
\textit{Linear regression estimates of the first stage. Robust standard errors are in parentheses. Model 1 includes only the batching tendency. Model 2 adds seasonality, shift fixed effects, and chief complaint. Model 3 further adds patient observables.} \\
*** p < 0.001, ** p < 0.01, * p < 0.05
\end{flushleft}
\end{table}

\begin{table}[ht]
\centering
\caption{First Stage: Probit Regression of Batch Tendency on Test Batching}
\label{tab:probit_regression}
\begin{tabular}{lccc}
\hline
 & Model 1 & Model 2 & Model 3 \\
\hline
Batching Tendency & 8.507*** & 8.989*** & 8.984*** \\
 & (0.393) & (0.415) & (0.415) \\
N & 46,783 & 46,783 & 46,783 \\
Log Likelihood & -9,565.239 & -8,617.761 & -8,614.765 \\
Akaike Inf. Crit. & 19,210.480 & 17,509.520 & 17,515.530 \\
\hline
\end{tabular}
\begin{flushleft}
\textit{Probit regression estimates of the first stage. Standard errors are in parentheses. Model specifications are the same as in Table \ref{tab:linear_regression}.} \\
*** p < 0.001, ** p < 0.01, * p < 0.05
\end{flushleft}
\end{table}

Both the linear and probit models demonstrate the strength of our instrument, with F-statistics well above conventional thresholds in the linear model and highly significant coefficients in both models. The probit model, while potentially more theoretically appropriate for our binary outcome, yields results consistent with the linear model in terms of the strong, positive relationship between physician batch-tendency and the likelihood of test batching. These results support the relevance of our instrument in the subsequent instrumental variable analysis.

\begin{figure}[h]
  \centering
  \includegraphics[width=4in]{../outputs/figures/roc.png}
\begin{tablenotes}
\small
\item \textit{}
\end{tablenotes}  
\end{figure}

\newpage

## Reduced Form

Not controlling for testing inclination:

\begin{table}[!htbp] \centering 
  \caption{Reduced Form} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & Return and Admit (72hr) & Log ED LOS & Imaging Tests \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 batch.tendency & $-$0.077$^{***}$ & 1.976$^{**}$ & 3.444$^{***}$ \\ 
  & (0.016) & (0.955) & (0.410) \\ 
  & & & \\ 
\textit{N} & 46,783 & 46,783 & 46,783 \\ 
R$^{2}$ & 0.005 & 0.302 & 0.257 \\ 
Adjusted R$^{2}$ & 0.002 & 0.301 & 0.255 \\ 
Residual Std. Error & 0.115 (df = 46645) & 0.472 (df = 46722) & 0.666 (df = 46645) \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
 & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
 & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
\end{tabular} 
\end{table}

Controlling for testing inclincation:

\begin{table}[!htbp] \centering 
  \caption{Reduced Form} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & Return and Admit (72hr) & Log ED LOS & Imaging Tests \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 batch.tendency & $-$0.091$^{***}$ & $-$0.804 & 0.201$^{**}$ \\ 
  & (0.035) & (2.339) & (0.097) \\ 
  & & & \\ 
 test.inclination & 0.004 & 0.850 & 0.972$^{***}$ \\ 
  & (0.011) & (0.596) & (0.019) \\ 
  & & & \\ 
\textit{N} & 46,783 & 46,783 & 46,783 \\ 
R$^{2}$ & 0.005 & 0.339 & 0.260 \\ 
Adjusted R$^{2}$ & 0.002 & 0.337 & 0.258 \\ 
Residual Std. Error (df = 46644) & 0.115 & 0.459 & 0.665 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{3}{r}{$^{***}$Significant at the 1 percent level.} \\ 
 & \multicolumn{3}{r}{$^{**}$Significant at the 5 percent level.} \\ 
 & \multicolumn{3}{r}{$^{*}$Significant at the 10 percent level.} \\ 
\end{tabular} 
\end{table}

\newpage

## Instrumental Variables Results

In this section, we examine the causal effects of batch ordering in the ED using a two-stage instrumental variables approach with a probit model in the first stage. This method, sometimes referred to as a probit-2SLS or probit-IV approach\footnote{Wooldridge, J. M. (2010). Econometric analysis of cross section and panel data. MIT press. (Chapter 15 discusses nonlinear models in IV estimation)}\footnote{Angrist, J. D., \& Pischke, J. S. (2008). Mostly harmless econometrics: An empiricist's companion. Princeton university press. (They discuss various IV approaches, including with binary endogenous variables)}, allows us to account for the binary nature of our endogenous variable (batch ordering) while addressing potential endogeneity concerns.

Our first stage employs a probit model to estimate the probability of batch ordering as a function of our instrument (physician batch-tendency) and other covariates. We then use the predicted probabilities from this probit model as an instrument in the second stage linear regression.

Table \ref{tab:iv_probit_results} presents the effects on our main outcomes. We begin by estimating the effects using this probit-2SLS approach. Following this, we extend our analysis to estimate the direct and indirect effects of batch ordering on LOS using a causal mediation framework, which incorporates our probit first-stage estimates.
To conduct the mediation analysis, we employ a custom bootstrapped approach that respects the clustered nature of our data. This method allows us to decompose the total effect of batch ordering on length of stay (LOS) into its direct effect and its indirect effect through the number of ED tests ordered. Our approach involves the following steps:

- We first use a probit model to predict the probability of batch ordering.

- We then estimate the effect of these predicted probabilities on the mediator (number of ED tests).

- Finally, we estimate the effects of both the predicted batching probabilities and the mediator on the outcome (LOS).

- We use clustered bootstrapping to estimate confidence intervals and p-values for the direct and indirect effects, ensuring that we account for the clustered structure of our data at the ED provider level.

This approach allows us to quantify not only the overall impact of batch ordering on LOS, but also to understand the mechanism through which this effect occurs, particularly the role played by the number of tests ordered. By using clustered bootstrapping, we ensure that our standard errors and confidence intervals accurately reflect the dependency structure within our data.

\begin{table}[ht]
\centering
\caption{IV Probit Regression Results}
\label{tab:iv_probit_results}
\begin{tabular}{lccc}
\hline
 & Log ED LOS & Imaging Tests & Return and Admit (72hr)  \\
\hline
Coefficient & -1.078257 & 0.6579455 & -0.008485203 \\
Std. Error & 0.3353349 & 0.2127876 & 0.01395781 \\
p-value & 0.001303237 & 0.001989105 & 0.543245 \\
R$^2$ & 0.3413561 & 0.5555298 & 0.005026105 \\
\hline
\end{tabular}
\begin{flushleft}
\textit{Notes:} Coefficients are from the second stage of an IV probit regression. Standard errors are clustered by ED\_PROVIDER. \\
*** p $<$ 0.01, ** p $<$ 0.05, * p $<$ 0.1.
\end{flushleft}
\end{table}

\begin{table}[htbp]
\centering
\caption{Mediation Analysis Results: Effect of Batch Ordering on Length of Stay}
\label{tab:mediation_results}
\begin{tabular}{lcccc}
\hline
Effect & Estimate & 95\% CI & Z-score & P-value \\
\hline
Indirect (via ED tests) & 0.134 & [0.067, 0.180] & 59.950 & <0.001 \\
Direct & -0.957 & [-1.485, 0.347] & -2.805 & 0.005 \\
Total & -0.823 & -- & -- & -- \\
\hline
\end{tabular}
\begin{tablenotes}
\small
\item Note: CI = Confidence Interval. Total effect is the sum of indirect and direct effects.
\end{tablenotes}
\end{table}

Our mediation analysis reveals a nuanced relationship between batch ordering and ED length of stay (LOS). Batch ordering has a significant positive indirect effect on LOS (0.134, p<0.001) mediated through the number of ED tests, suggesting that batching leads to more tests being ordered, which increases LOS. However, this is counterbalanced by a larger negative direct effect (-0.957, p=0.005), indicating that batch ordering also introduces efficiencies that reduce LOS. The total effect (-0.823) suggests that, overall, batch ordering decreases LOS. These findings imply that while batch ordering may increase the number of tests, it ultimately leads to shorter ED stays, likely due to improved process efficiency.

\begin{table}[htbp]
\centering
\caption{Mediation Analysis Results: Effect of Batch Ordering on 72-Hour Return and Admission}
\label{tab:mediation_results_72hr}
\begin{tabular}{lcccc}
\hline
Effect & Estimate & 95\% CI & Z-score & P-value \\
\hline
Indirect (via ED tests) & -0.00161 & [-0.00219, -0.00036] & -9.028 & <0.001 \\
Direct & 0.00011 & [-0.01740, 0.01826] & 0.006 & 0.995 \\
Total & -0.00150 & -- & -- & -- \\
\hline
\end{tabular}
\begin{tablenotes}
\small
\item Note: CI = Confidence Interval. Total effect is the sum of indirect and direct effects.
\end{tablenotes}
\end{table}

The mediation analysis for 72-hour return and admission rates reveals a small but significant indirect effect of batch ordering. Batch ordering has a negative indirect effect (-0.00161, p<0.001) on 72-hour returns and admissions, mediated through the number of ED tests. This suggests that the increased testing associated with batch ordering slightly reduces the likelihood of patient returns or admissions within 72 hours. The direct effect (0.00011, p=0.995) is not statistically significant, indicating no evidence of a direct impact of batch ordering on this outcome. The total effect (-0.00150) suggests a minor overall reduction in 72-hour returns and admissions associated with batch ordering, primarily driven by the indirect effect through increased testing.

### 2024-09-09 manuscript/manuscript v2.Rmd
getting going

--- 2024-01-25 manuscript/manuscript.Rmd
@@
 ---
-title: "To Batch or Not to Batch: Test-Ordering Variabilityin the ED"
+title: "Judicious Batching: The Impact of Batch Ordering Imaging Tests in the Emergency Department"
 authors:
@@
   - \usepackage{tabularx}
+  - \usepackage{graphicx}
+  - \usepackage{tikz}
   - \usepackage{siunitx}
@@
   - \usepackage{natbib}
+  - \usepackage{caption}
 ---
@@
 With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly.
-We find that being treated by a top decile batching physician, compared to a bottom decile batching physician, results in an additional 13.5 tests per 100 patient encounters.
-We find no evidence that batch-ordering reduces ED length of stay or 72-hour return rates.
-These results suggest that the benefits of batch-ordering are limited, and that the practice may be associated with significant costs.
-Our findings are robust to analyses probing the key assumptions of our batching definition.
-
 Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm.
@@
 We observe approximately $48,000$ such visits during the study period.
-To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $10$ percent of the time.
+To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $2$ percent of the time.
 Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them.
-Examples of complaints dropped include urinary complaints and allergic reactions.
+Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints.
 Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
-In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving physicians who treat over 500 ED cases per year.
+While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories.
+In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $36,765$ encounters involving full-time physicians who treat over 500 ED cases per year.
 
@@
 While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice.
-Below, we detail our primary outcomes: (a) ED length of stay (LOS), (b) resource utilization, and (c) 72-hour return.
+Below, we detail our primary outcomes: (a) ED length of stay (LOS), and (b) resource utilization.
 
@@
 
-  *(c) 72-hour return*-- The rate of patients returning to the ED within 72 hours of their initial visit is a proxy for the quality of care received.
-A high 72-hour return rate might indicate inadequate treatment or diagnostic oversight during the initial visit.
-We measure this outcome as the proportion of patients with a subsequent ED visit within 72 hours post-initial discharge.
-The hypothesis is that batch testing may lead to variations in the 72-hour return rate, potentially influencing the quality of care received.
+*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
+This outcome is used to assess the quality of care provided during the initial ED visit.
+The hypothesis is that batch testing may lead to variations in the quality of care provided, potentially influencing the likelihood of a return visit.
 
@@
 
-We define ``batching" in line with standard emergency medicine practices.
-It involves a physician ordering a comprehensive set of diagnostic tests at once, typically covering a broad range of potential diagnoses.
-This contrasts with non-batching, where tests are ordered more selectively based on the information available at the time, with additional tests potentially ordered later as needed.
+We define "batching" in line with standard emergency medicine practices.
+Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses.
+This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.
 
-We operationalize batching as the ordering of multiple diagnostic tests within a 5-minute window.
-Sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-This operational definition aligns with the concept of batching as a single comprehensive effort.
-For our analysis, all lab-based testing ordered in this timeframe are categorized as one distinct type of testing (lab).
-In contrast, each imaging test (e.g., X-ray, CT scan) are considered a separate distinct test.
-Therefore, a batch in our study is defined as consisting of two or more distinct diagnostic tests, which could be a combination of lab and imaging tests or multiple imaging tests.
-
-This approach reflects the realistic diagnostic strategies in an ED setting and allows for a clear distinction between batching and non-batching behaviors.
-The differentiation between lab tests as one collective unit and imaging tests as individual units is based on their operational impact in the ED, particularly regarding patient waiting time and processing queues.
+We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window.
+In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition.
+Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study.
+Therefore, a batch in our study consists of two or more distinct imaging tests.
 
@@
 
-\hyperref[tab:summary_statistics]{Table 1} presents a detailed overview of the emergency department (ED) characteristics, patient demographics, and medical tests for our baseline sample, derived from the data collected during the study period.
-On average, the emergency department manages a volume of approximately 24 patients (Mean = 24.18), indicating a significant but manageable patient load.
-Key physiological markers such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are prevalent among patients, albeit at varying degrees, highlighting critical aspects of emergency care.
-Notably, a small proportion of patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01) conditions, emphasizing the diversity of cases encountered in the ED setting.
-The Emergency Severity Index (ESI) averages at 2.78, suggesting that most cases fall within the moderate acuity level.
+hyperref[tab]{Table 1} provides a concise overview of emergency department (ED) characteristics, patient demographics, and medical tests based on our study data.
+The ED maintains an average volume of 24.3 patients, reflecting a significant yet manageable workload.
+Key physiological indicators include tachycardia (18.8%) and tachypnea (9.51%), showcasing the critical conditions often treated in emergency settings.
+Additionally, fever and hypotension are observed in 2.43% and 1.57% of patients, respectively, highlighting the variety of urgent care demands.
 
-Common chief complaints in the emergency department paint a vivid picture of patient needs and healthcare demands.
-The data reveals abdominal complaints (Mean = 0.14), extremity complaints (Mean = 0.12), and chest pain (Mean = 0.08) as the most frequently occurring, followed by neurological (Mean = 0.08) and gastrointestinal issues (Mean = 0.08).
-This information is crucial in understanding the primary reasons for ED visits and the required resources for patient care.
+Patient complaints commonly involve abdominal issues (16.9%), extremity problems (14.3%), and chest pain (9.58%), along with neurological (9.50%) and gastrointestinal complaints (9.04%).
+These figures emphasize the frequent health concerns that drive ED visits and the resources needed for effective response.
 
-The patient characteristics in our sample depict a diverse demographic landscape.
-The average arrival age in the ED is approximately 58 years (Mean = 58.33), indicating a predominantly middle-aged population.
-The racial composition is predominantly White (Mean = 0.89), with smaller proportions of Black (Mean = 0.04) and Asian (Mean = 0.03) patients.
-Notably, a slight majority of the patients are female (Mean = 0.54), offering insights into gender dynamics in healthcare utilization.
+The patient demographic profile shows an average age of 58.3 years, predominantly middle-aged and elderly individuals.
+Racial distribution is mainly White (88.6%), with smaller percentages of Black (4.12%) and Asian (2.97%) populations.
+A slight majority of patients are female (54.5%), indicating gender trends in healthcare utilization.
 
-In terms of diagnostic tests, the data reveals a high reliance on lab tests (Mean = 0.77), indicating their crucial role in patient diagnosis and management.
-X-Ray (Mean = 0.47) and CT scans, both non-contrast (Mean = 0.21) and contrast (Mean = 0.19), are also frequently employed, underscoring the importance of imaging in modern emergency medicine.
-Ultrasound usage (Mean = 0.11) and the practice of batch ordering tests (Mean = 0.38) further reflect the operational aspects of the ED and the strategies employed to manage patient care effectively.
+Diagnostic procedures heavily rely on lab tests (78.5%), essential for thorough patient evaluations.
+Imaging tests such as X-rays (48.1%) and CT scans—both non-contrast (23.1%) and contrast (20.5%)—play pivotal roles in diagnostics.
+The use of ultrasounds (12.5%) and the strategy of batch ordering tests (6.32%) are critical operational aspects of the ED, reflecting approaches to efficient patient management.
 
@@
 \centering
-\caption{Summary Statistics}
+\caption{Summary Statistics of Emergency Department Encounters}
 \label{tab:summary_statistics}
@@
 \midrule
-\multicolumn{5}{l}{\textbf{ED Characteristics}} \\
-ED Volume & 24.18 & 15 & 25 & 32 \\
-Tachycardic & 0.19 & & & \\
-Tachypneic & 0.09 & & & \\
-Febrile & 0.02 & & & \\
-Hypotensive & 0.01 & & & \\ 
-ESI & 2.78 & 2 & 3 & 3 \\
-Complaint: Abdominal & 0.14 & & & \\
-Complaint: Extremity & 0.12 & & & \\
-Complaint: Chest Pain & 0.08 & & & \\
-Complaint: Neurological & 0.08 & & & \\
-Complaint: Gastrointestinal & 0.08 & & & \\
+\multicolumn{5}{l}{\textbf{Emergency Department Characteristics}} \\
+Patients in ED & 24.3 & 16 & 25 & 33 \\
+\\
+Tachycardic & 0.188 & & & \\
+Tachypneic & 0.095 & & & \\
+Febrile & 0.024 & & & \\
+Hypotensive & 0.016 & & & \\
+\\
+ESI & 2.74 & 2 & 3 & 3 \\
+\\
+Complaint: Abdominal Complaints & 0.169 & & & \\
+Complaint: Extremity Complaints & 0.143 & & & \\
+Complaint: Chest Pain & 0.096 & & & \\
+Complaint: Neurological Issue & 0.095 & & & \\
+Complaint: Gastrointestinal Issues & 0.090 & & & \\
 \midrule
 \multicolumn{5}{l}{\textbf{Patient Characteristics}} \\
-Arrival Age & 58.33 & 44 & 62 & 75 \\
-Race: White & 0.89 & & & \\
-Race: Black & 0.04 & & & \\
-Race: Asian & 0.03 & & & \\
-Gender: Female & 0.54 & & & \\
+Arrival Age & 58.3 & 44 & 61 & 75 \\
+\\
+Race: White & 0.886 & & & \\
+Race: Black & 0.041 & & & \\
+Race: Asian & 0.030 & & & \\
+\\
+Gender: Female & 0.545 & & & \\
 \midrule
 \multicolumn{5}{l}{\textbf{Tests}} \\
-X-Ray & 0.47 & & & \\
-Ultrasound & 0.11 & & & \\
-Non-Contrast CT & 0.21 & & & \\
-Contrast CT & 0.19 & & & \\
-Lab & 0.77 & & & \\
-Tests were Batch Ordered & 0.38 & & & \\
+X-Ray & 0.481 & & & \\
+Ultrasound & 0.125 & & & \\
+Non-Contrast CT & 0.231 & & & \\
+Contrast CT & 0.205 & & & \\
+Lab & 0.785 & & & \\
+\\
+Imaging Tests were Batch Ordered & 0.063 & & & \\
 \bottomrule
@@
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{This table reports summary statistics for the baseline sample of emergency department visits during the study period described in the text.
@@
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Table \ref{tab:wald_test} reports the results of a Wald test which was conducted to assess the balance of chief complaints across providers in our dataset.
@@
 \begin{equation}
-Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint} + \varepsilon_{i,t}
+Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \alpha_{lab} + \varepsilon_{i,t}
 \end{equation}
 
-Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on date $t$.
+Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$.
 Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g.
@@
 Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision.
-As stated earlier, these controls are what is required for our quasi-random assignment assumption.
+Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case.
+As stated earlier, these controls are more than what is required for our quasi-random assignment assumption.
 Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.
@@
 We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure \ref{fig:physician_batching} graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically.
-\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $8$pp increase in the likelihood of having tests batch-ordered in the ED.
-The F-statistic is 50 when all controls and fixed effects are included.
-The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
+Figure \ref{fig:physician_batching} graphs the frequency with which physicians batch order across different chief complaints, highlighting that the variation in batching differs starkly.
+We note that there is high correlation between
 
@@
   \centering
-  \caption{Batch Frequency by Physician for High Volume Complaints}
+  \caption{Variation in Physician Batching}
   \label{fig:physician_batching}
-  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
+  \includegraphics[width=4.25in]{../outputs/figures/Figure 1.png}
 \begin{tablenotes}
-\tiny
-\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.
-Physicians are mapped on the x-axis, revealing those with a systematically heightened tendency to batch (in red) compared to their peers who batch less frequently (in blue).
-The quartile of batchers is calculated based on the batching rate across all complaint areas per physician.
-Complaints presented in this figure are from patient encounters that exhibit the highest variance in physician batching.
-Physicians who batch (not batch) in one category of complaints tend to also batch (not batch) in other categories.}
+\small
+\item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.}
 \end{tablenotes}  
@@
 
-\begin{table}[ht]
-\centering
-\caption{First Stage: Effect of Batch Tendency on Test Batching}
-\label{tab:regression}
-\begin{tabular}{p{9cm}ccc}
-\toprule
-\textbf{} & Model 1 & Model 2 & Model 3 \\
-\midrule
-Batching Tendency & 1.06 *** & 1.08 *** & 1.08 *** \\
- & (0.01) & (0.01) & (0.01) \\
-\midrule
-F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
-F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
-\midrule
-Num. obs. & 43327 & 43327 & 43327 \\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & No & Yes & Yes \\
-Patient observables? & No & No & Yes \\
-\bottomrule
+\hyperref[table:first_stage]{Table 1} illustrates the "first stage" results from both linear and probit models, emphasizing the robust relevance of the instrument across different statistical frameworks.
+In both modeling approaches, a 10 percentage point increase in the batch-tendency of physicians consistently results in a significant increase in the likelihood of batch ordering tests in the emergency department—1.15 percentage points in the linear model and 93.24 percentage points in the probit model.
+This consistent significance across models underscores the strength and relevance of batch tendency as an instrumental variable.
+Given the robustness to different model specifications, we move forward with a probit model in our first stage to estimate the effect of batch ordering on patient outcomes.
+
+\begin{table}[!htbp] \centering 
+  \caption{Comparison of First Stage Estimates: Linear vs. Probit Models}
+  \label{table:first_stage}
+\begin{tabular}{@{\extracolsep{5pt}}lcccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+ & \multicolumn{3}{c}{Linear Model} & \multicolumn{3}{c}{Probit Model} \\ 
+\cline{2-4} \cline{5-7}
+\\[-1.8ex] & (1) & (2) & (3) & (1) & (2) & (3)\\ 
+\hline \\[-1.8ex] 
+ Batch Tendency & 1.25$^{***}$ & 1.24$^{***}$ & 1.15$^{***}$ & 8.657$^{***}$ & 9.135$^{***}$ & 8.515$^{***}$ \\ 
+  & (0.04) & (0.03) & (0.02) & (0.659) & (0.709) & (0.789) \\ 
+  & & & & & & \\
+\textit{Day of Week-Time of Day FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
+\textit{Month of Year FE} & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
+\textit{Complaint/Severity FE} & & $\checkmark$ & $\checkmark$ & & $\checkmark$ & $\checkmark$ \\
+\textit{Laboratory Tests Ordered} & & & $\checkmark$ & & & $\checkmark$ \\
+\hline 
+\hline \\[-1.8ex] 
+F Statistic & 13.13 & 26.02 & 29.38 &  &  &  \\
+Akaike Inf. Crit. & & & & 16,957.210 & 15,657.200 & 15,349.280 \\ 
+\\
+\textit{N} & 36,765 & 36,765 & 36,765 & 36,765 & 36,765 & 36,765 \\
+
+\hline \\[-1.8ex] 
 \end{tabular} 
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Estimates of the first stage for the baseline sample described in the text.
@@
 Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
 Column 3 corresponds to the baseline controls.
@@
 
-This reduced form will allow us to check that our instrument is a strong instrument.
-To study the effects of test batching in the ED on an outcome $Y_i$, we estimate the following 2SLS equations using our baseline sample:
+This reduced form will allow us to check the strength of our instrument.
+To study the effects of test batching in the ED on an outcome $Y_i$, we employ a two-stage estimation approach using our baseline sample:
 
@@
 \begin{equation}
-Batched_i = \delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i + \nu_i
+P(Batched_i = 1) = \Phi(\delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i)
 \end{equation}
 
-Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization. 
-and $X_i$ is the same as in the reduced-form approach.
+Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization, and $X_i$ is the same as in the reduced-form approach.
 $Batched_i$ variable suffers from potential endogeneity concerns.
@@
 In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department.
-Our reduced form regression results, as presented in Model 1 of Table \ref{tab:reducedform}, reveal that batch tendency correlates with prolonged ED length of stay (LOS), increased test ordering, and does not materially alter the likelihood of a 72-hour return.
-Notably, these associations are nuanced by an underlying factor: the physician's testing inclination.
+We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination.
+Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.
 
-\begin{table}[htbp]
-\centering
-\caption{Reduced-Form Results: Length of Stay, Number of Tests, and 72-Hour Return}
-\label{tab:reducedform}
-\begin{tabular}{p{6.5cm}ccc}
-\toprule
-& \multicolumn{1}{c}{ln(LOS)}
-& \multicolumn{1}{c}{Number of Tests}
-& \multicolumn{1}{c}{72-Hour Return} \\
-& \multicolumn{1}{c}{(1)}
-& \multicolumn{1}{c}{(2)}
-& \multicolumn{1}{c}{(3)} \\
-\midrule
-\textit{Model 1} & & & \\
-Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
-                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
-\\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
-Hospital volume? & Yes & Yes & Yes \\
-\midrule
-\midrule
-\textit{Model 2} & & & \\
-Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
-                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
-                    &     &     &  \\
-Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
-                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
-\\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
-Hospital volume? & Yes & Yes & Yes \\
-\bottomrule
+To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency.
+Specifically, we calculate the physician's residualized leave-out average test rate, which serves as a proxy for their propensity to order tests.
+It is important to note the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
+This substantial correlation suggests that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies.
+Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.
+
+Reduced-form regression results in Table \ref{tab:reducedform} reveal the effect the average total effect of batch tendency on resource utilization and LOS.
+However, it is important to note that the estimated total effect of batch tendency accounts for both direct and indirect (mediated) effects of batch tendency.
+We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered.
+The increased testing volume, in turn, is associated with an extended LOS.
+
+\begin{table}[!htbp] \centering 
+\caption{Reduced-Form Results: Length of Stay and Number of Tests}
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lcc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & ln(LOS) & Number of Tests \\ 
+\\[-1.8ex] & (1) & (2)\\ 
+\hline \\[-1.8ex] 
+ Batch Tendency & $-$0.109 & 0.149$^{***}$ \\ 
+  & (0.464) & (0.048) \\ 
+  & & \\ 
+ Testing Inclination & 0.605$^{**}$ & 0.952$^{***}$ \\ 
+  & (0.240) & (0.023) \\ 
+  & & \\ 
+Seasonality and shift fixed effects? & Yes & Yes \\
+Chief Complaint? & Yes & Yes  \\
+\hline 
+\textit{N} & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.215 & 0.265 \\ 
+Adjusted R$^{2}$ & 0.213 & 0.263 \\ 
+Residual Std. Error (df = 41800) & 0.490 & 0.842 \\ 
+\hline 
+\hline \\[-1.8ex] 
 \end{tabular} 
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Estimates of the reduced form are for the baseline sample described in the text.
-Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability.
 Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
 Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
 Robust standard errors are clustered at the physician level.}
@@
 
-We construct a physician's testing inclination using the same method used to measure physician batch tendency, but instead we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test.
-A key consideration in our analysis is the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
-Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often as a result of ordering more tests.
-Failure to control for this inclination could result in biased estimates of the effect of batch tendency.
-
-The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG}) elucidates the interconnected pathways between batch tendency, batch ordering, and primary outcomes.
-We posit that while batch tendency directly influences the practice of batch ordering tests, batch tendency and batch ordering are concurrently influenced by testing inclination.
-This is critical as testing inclination has a direct effect on primary outcomes, necessitating its inclusion as a control variable to disentangle its confounding effects.
+We introduce a Directed Acyclic Graph (DAG) to illustrate the interconnected pathways between batch tendency, batch ordering, and primary outcomes.
+The DAG in Figure \ref{fig:DAG} shows that batch tendency directly influences batch ordering, which in turn affects the number of tests ordered.
+The number of tests ordered, in turn, influences the length of stay.
+The DAG also illustrates the effect of batch tendency on the length of stay  is mediated by the number of tests ordered.
 
@@
   \label{fig:DAG}
-  \includegraphics[width=4in]{../outputs/figures/DAG.png}
+  \includegraphics[width=6in]{../outputs/figures/dagitty.png}
 \begin{tablenotes}
@@
 
-In Model 2 of Table \ref{tab:reducedform}, we control for testing inclination and observe notable changes in the coefficients for batch tendency.
-The previously significant positive association with ln(LOS) diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination.
-Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination.
-Assignment to a physician in the top batching decile (relative to one in the bottom decile) is associated with a 0.2 unit increase in the number of diagnostic tests being ordered per patient encounter, without any additional change in the likelihood of a 72-hour return.
-
-The effects of testing inclination are also striking.
-We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 9.6 excess diagnostic tests being ordered per 100 patient encounters, and -0.03 pp change in the probability of a patient returning within 72 hours.
+To determine whether or not the effect of batch tendency on LOS is mediated by the number of tests ordered, we estimate a mediation analysis using the approach outlined by @imai2010general.
+Our analysis, scrutinizes the pathways through which physician batch tendency influences patient length of stay (LOS) in the emergency department.
+The analysis delineates both direct effects, which capture the influence of batch tendency on LOS without intermediaries, and indirect effects, which operate through the mediator of test ordering volume.
+Our findings underscore a significant Average Direct Effect (ADE) of batch tendency on LOS, estimated at -0.15  (p = 0.046).
 
@@
 Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
 Testing inclination? & Yes & Yes & Yes \\
@@
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{This table reports the estimated coefficients of a reduced-form regression of our main outcomes on physician batching tendency for samples based on the patient’s chief complaint field.
@@
 In this section we examine the causal effects of batch ordering in the ED.
-Mirroring our presentation of the reduced-form results, we turn to presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED LOS, number of tests, and 72-hour return.
+Mirroring our presentation of the reduced-form results, we turn to presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED LOS and number of tests ordered.
+We first start by estimating the effects of batch ordering on the number of tests ordered using the standard 2SLS approach.
+Next we estimate the direct and indirect effects of batch ordering on LOS using a causal mediation analysis.
 
-\begin{table}[!htbp] \centering 
-  \caption{2SLS Results: Length of Stay, Number of Tests, and 72-Hour Return} 
-  \label{tab:2SLS} 
-\begin{tabular}{p{7cm}ccc}
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & ln(LOS) & Number of Tests & 72 Hour Return \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
-\hline \\[-1.8ex] 
- Batch & $-$0.086 & 0.135$^{***}$ & 0.018 \\ 
-  & (0.419) & (0.046) & (0.018) \\ 
-  & & & \\ 
- Testing Inclination & 0.210$^{**}$ & 0.335$^{***}$ & $-$0.009$^{***}$ \\ 
-  & (0.082) & (0.008) & (0.003) \\ 
-  & & & \\ 
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
-Hospital volume? & Yes & Yes & Yes \\
+\begin{table}[!htbp]
+\centering
+\begin{threeparttable}
+\caption{Combined 2SLS and Causal Mediation Analysis Results}
+\label{tab:combined_results}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
+\toprule
+& \multicolumn{2}{c}{2SLS Results} \\
+\cmidrule{2-3}
+& ln(LOS) & Number of Tests \\
 \midrule
-\textit{N} & 41,929 & 41,929 & 41,929 \\ 
-R$^{2}$ & 0.199 & 0.312 & 0.008 \\ 
-Adjusted R$^{2}$ & 0.194 & 0.307 & 0.002 \\ 
-Residual Std. Error (df = 41667) & 0.496 & 0.817 & 0.184 \\ 
-\\
+Batch & -0.086 & 0.135$^{***}$ \\
+& (0.419) & (0.046) \\
+& & \\
+\addlinespace
+Seasonality and Shift Fixed Effects & Yes & Yes \\
+Chief Complaint and Severity & Yes & Yes \\
+\midrule
+Observations & 41,901 & 41,901 \\
+R$^{2}$ & 0.183 & 0.310 \\
 \bottomrule
-\end{tabular}
+\end{tabular*}
+
+\smallskip % Adds some space before the next part of the table
+
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
+\toprule
+& \multicolumn{2}{c}{Causal Mediation Analysis} \\
+\cmidrule{2-3}
+Effect & Estimate & 95\% CI \\
+\midrule
+ACME & 0.036 & [-0.009, 0.08] \\
+ADE & -0.136 & [-0.221, -0.04] \\
+Total Effect & -0.100 & [-0.195, 0.00] \\
+Proportion Mediated & -0.333 & [-2.831, 1.21] \\
+\bottomrule
+\end{tabular*}
+
 \begin{tablenotes}
-\tiny
-\item \textit{Estimates of the IV model are for the baseline sample described in the text.
-Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
-Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
-Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
-Robust standard errors are clustered at the physician level.}
-\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\small
+\item The 2SLS results are based on robust standard errors clustered at the physician level, with controls for seasonality, shift fixed effects, chief complaint, and severity.
+The causal mediation analysis examines whether the number of tests mediates the effect of batching on length of stay, using a quasi-Bayesian confidence interval approach with 1000 simulations.
+\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
 \end{tablenotes}
+\end{threeparttable}
 \end{table}
 
-*ED length of stay (LOS).*-- Our analysis reveals that batch ordering does not significantly reduce the length of stay in the ED.
-This finding holds even after accounting for various control variables, including ED volume and patient acuity.
-Contrary to common assumptions that batching tests could expedite patient throughput, our results suggest that the efficiency gains could be offset.
-A possible explanation lies in excess tests being ordered due to batching, which could lead to increased wait times for patients.
+*ED length of stay (LOS).*-- Our causal mediation analysis reveals that the Average Direct Effect (ADE) of test batching on the natural log of LOS is -0.13607, indicating that, when not considering the mediating effect of the number of tests, batching leads to a decrease in the expected LOS.
+Specifically, this translates to an approximate 12.7 percent reduction in LOS when tests are batched, all else being equal.
+This result, complemented by our 2SLS estimate of -0.086, suggests that the direct effect of batch ordering on LOS is negative.
+Our results also suggest a "washing out" effect, where the effeciency gains from batching are offset by the increased number of tests ordered.
 This is consistent with our finding that batching increases the number of tests ordered.
@@
 *Resource Utilization.*-- The impact of batching on resource utilization is pronounced.
-We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 2,206 additional tests during our study period.
-This is likely due to the information gain that occurs from sequential test ordering, where a physician may order additional tests after receiving the results of the first test.
-
-*72-hour Return*-- Our findings indicate that batch ordering does not significantly impact the likelihood of a 72-hour return.
-This suggests that despite the increased number of tests, there is no corresponding decrease in short-term readmission rates.
-This highlights the need for more targeted testing strategies that balance comprehensive assessment with resource efficiency.
+We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 13.5 excess tests per 100 batch orders, where excess means tests that would not have been ordered had the physician not batched.
+This result is consistent with the notion that batching leads to increased resource utilization in the ED.
 
 The instrumental variable analysis presented here underscores the complexities underlying batch ordering practices in the ED.
-While batching is often seen by physicians as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization without corresponding gains in patient throughput or reductions in short-term return rates.
+While batching is often seen by physicians as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization and that the corresponding gains in patient throughput are washed out.
 These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
@@
 This is particularly pronounced in high-volume complaint areas, where batching leads to a significant increase in the number of tests ordered.
-However, we find no significant impact of batching on ED length of stay or 72-hour return rates.
-This suggests that despite the increased number of tests, there is no corresponding decrease in short-term readmission rates.
-This highlights the need for more targeted testing strategies that balance comprehensive assessment with resource efficiency.
+However, we find no significant impact of batching on ED length of stay.
+Causal mediation analysis reveals that the direct effect of batching on length of stay is negative, but this effect is offset by the increased number of tests ordered.
+This suggests that the efficiency gains from batching are washed out by the increased resource utilization.
+
+Our findings have important implications for emergency care practice.
+While batching is often seen as a method to enhance efficiency, our results suggest that it may lead to increased resource utilization and that the corresponding gains in patient throughput are washed out.
+These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
+Future research should explore the impact of batching on other outcomes, such as patient satisfaction and clinical outcomes, to provide a more comprehensive understanding of the implications of batching in the emergency department.
 
@@
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Column 1 displays the first stage coefficient of batched on the baseline physician batch tendency instrument for the corresponding sub-sample.
@@
 \begin{tablenotes}
-\tiny
+\small
 \item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
@@
 \newpage
+
+\begin{table}[ht]
+\centering
+\caption{2SLS Results for the "Batch" Variable Across Different General Complaints: Outcome of Number of Tests}
+\begin{tabular}{lcc}
+\hline
+\textbf{Complaint} & \textbf{Batching Coefficient} & \textbf{Standard Error} \\
+\hline
+Musculoskeletal/Extremity & 0.476** & (0.235)  \\
+Respiratory-Related & 0.244 & (0.209)   \\
+General/Other Symptoms & 0.179 & (0.207)   \\
+Gastrointestinal/Abdominal & -0.092 & (0.128)   \\
+Cardiac/Chest-Related & 0.270 & (0.264)   \\
+Neurological/Syncope & -0.080 & (0.493)   \\
+\hline
+\end{tabular}
+\label{tab:batch_variable_outcome}
+\end{table}
+
 :::

### 2024-09-09 manuscript/manuscript-v2.tex
getting going

--- 2024-09-09 manuscript/manuscript v2.Rmd
@@
----
-title: "Judicious Batching: The Impact of Batch Ordering Imaging Tests in the Emergency Department"
-authors:
-- familyname: Jameson
-  othernames: Jacob
-  affiliation: Interfaculty Initiative in Health Policy, Harvard University
-  email: jacobjameson@g.harvard.edu
-- familyname: Saghafian
-  othernames:  Soroush
-  affiliation: Kennedy School of Government, Harvard University 
-  email: soroush_saghafian@hks.harvard.edu
-- familyname: Huckman
-  othernames:  Robert
-  affiliation: Harvard Business School 
-  email: rhuckman@hbs.edu
-- familyname: Hodgson
-  othernames:  Nicole
-  affiliation: Mayo Clinic
-  email: Hodgson.Nicole@mayo.edu
-keywords:
-  - Emergency Department, Operational Effeciency, Diagnostic Testing
-abstract: |
-  We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED.
-Using records of over 45,000 Mayo Clinic emergency department visits, we find that quasi-random assignment to a top (versus bottom) decile batching physician significantly increases excess diagnostic testing.
-Instrumental variable results show that batching leads to an additional 13.5 tests per 100 batches, resulting in excess tests that would not have been ordered under a sequential-ordering strategy.
-Contrary to expectations, batch ordering does not reduce ED length of stay or affect 72-hour return rates, challenging the perceived efficiency of batching and highlighting the necessity for more targeted testing strategies in emergency care.
-acknowledgment: |
-  The authors would like to thank ..
-bibliography: references.bib
-blind: false
-output: rticles::informs_article
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{tabularx}
-  - \usepackage{graphicx}
-  - \usepackage{tikz}
-  - \usepackage{siunitx}
-  - \usepackage{tablefootnote}
-  - \usepackage{longtable}
-  - \usepackage{threeparttable}
-  - \usepackage{natbib}
-  - \usepackage{caption}
----
-
-# Introduction {#sec:I}
-
-Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization.
-Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events.
-Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality @bernstein2009.
-
-One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests @naseim2015.
-A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests.
-These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.
-
-A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially.
-This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing.
-Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results @koch2018.
-Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier.
-Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged.
-Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements @saghafian2015, however, research surrounding test ordering strategies to improve the patient flow processes remains limited.
-
-In this paper, we use data from over 45,000 patient visits to the ED that occur during our study period to quantify the benefits and consequences of batching versus sequentially ordering diagnostic tests on patient length of stay, re-admission, and resource utilization.
-Our empirical strategy exploits random assignment of patients to ED physicians who differ in their propensity to batch-order diagnostic tests.
-When patients arrive at the ED, they are assigned to a physician based on availability, with no discretion on either side.
-Thus, patients who arrive at the ED at similar times are randomly assigned to physicians who vary in their willingness to batch order diagnostic tests.
-We measure physician tendency to batch using a leave-out, residualized measure based on all other patients the physician has seen in the ED in the study period.
-The tendency measure strongly predicts the ED test batch outcome but is uncorrelated with patient and ED visit characteristics.
-
-We start by evaluating the reduced-form effects of ED provider batch-ordering tendency on downstream patient outcomes and turnaround time.
-We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences.
-Being treated by a provider in the top decile of the tendency distribution, compared to being treated by someone in the bottom decile results in an additional 2 diagnostic test per 100 patient encounters after controlling for the physician's underlying propensity to test.
-
-Physician care takes on multiple dimensions besides the decision to batch order diagnostic tests.
-Employing a placebo exercise, we find evidence that that excess testing occurs after having seen a high-batching physician is due to batching from this physician, as opposed to other differences across physician in care that correlate with batching tendency: we find no effects of physician batch tendency on key outcomes for a ``placebo sample" of patients who visit the ED for conditions that rarely result in batched orders.
-
-With the caveat that other unobserved dimensions of physician care may impact patient outcomes, we employ physician batch tendency as an instrumental variable for having tests batch ordered in the ED to quantify the effect of batching directly.
-Because of the institutional features of the ED, our research design closely approximates an RCT that assigns patients to batch-ordering or sequential-ordering arm.
-In the ED, patients have no discretion over choosing providers, and in our specific ED, physicians have discretion over choosing patients, alleviating major selection issues present in other health care settings.
-Furthermore, physicians exhibit wide variation in practice behavior in batch-ordering, even within the same hospital, while following the same guidelines.
-Finally, patient-physician interactions in the ED are typically well documented, short, and one-off, constraining physician decision-making to a more limited, better-observed choice set than present in settings such as specialty or primary care.
-
-In sum, exploiting practice variation in ED settings shuts down other (but not all) potential channels besides test batching that are present in other settings, determine length of stay, and impact patient outcomes.
-This approach allows us to move closer to identifying the causal impact of batch-ordering diagnostic tests on patient outcomes and resource utilization.
-It is important to note that this paper studies the impact of batch-ordering through a batching decision requiring clinical judgment (within practice norms) rather than through specific hospital policies, differences in adherence to clinical practice guidelines, or substandard care.
-
-The remainder of this paper is structured as follows.
-The next section describes the data source and outlines our baseline sample.
-The empirical strategy and its accompanying identifying assumptions are laid out in Section III.
-Section IV presents the results.
-Section V draws implications for batch-ordering policies.
-The last section concludes.
-
-# Data and Definitions {#sec:II}
-
-Our study was conducted in the Emergency Department (ED) of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation.
-During the study period, the ED recorded approximately 43,000 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces.
-The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume.
-Physicians operate in a unique workflow that includes staggered 8.5-hour shifts and a rotational patient assignment system, minimizing potential selection bias in patient encounters.
-
-We conducted a retrospective review of comprehensive ED operational data from 10/6/2018 through 12/31/2019, coinciding with the initiation of a new electronic medical record.
-The dataset includes detailed patient demographics, chief complaints, vital signs, emergency severity index (ESI), length of stay (LOS), and resource utilization metrics.
-This period was chosen to provide a robust data set while excluding the influence of the coronavirus pandemic.
-
-## Sample Construction
-
-Our research design focuses on adults who visit the Mayo Clinic of Arizona ED.
-We observe approximately $48,000$ such visits during the study period.
-To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $2$ percent of the time.
-Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them.
-Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints.
-Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories.
-In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $36,765$ encounters involving full-time physicians who treat over 500 ED cases per year.
-
-## Variable Definitions
-
-Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter.
-While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice.
-Below, we detail our primary outcomes: (a) ED length of stay (LOS), and (b) resource utilization.
-
-*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings.
-It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.
-The hypothesis is that batch testing may lead to shorter ED-LOS, potentially improving patient throughput and reducing crowding.
-
-*(b) Resource Utilization*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives.
-In this study, we quantify resource utilization by the total number of distinct diagnostic tests ordered per patient during their ED stay.
-This encompasses both initial and subsequent tests.
-The hypothesis is that batch testing may lead to variations in the number of tests ordered, potentially influencing the overall healthcare expenditure and efficiency.
-
-*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital.
-This outcome is used to assess the quality of care provided during the initial ED visit.
-The hypothesis is that batch testing may lead to variations in the quality of care provided, potentially influencing the likelihood of a return visit.
-
-### Batching
-
-We define "batching" in line with standard emergency medicine practices.
-Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses.
-This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.
-
-We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window.
-In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition.
-Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study.
-Therefore, a batch in our study consists of two or more distinct imaging tests.
-
-## Summary Statistics
-
-hyperref[tab]{Table 1} provides a concise overview of emergency department (ED) characteristics, patient demographics, and medical tests based on our study data.
-The ED maintains an average volume of 24.3 patients, reflecting a significant yet manageable workload.
-Key physiological indicators include tachycardia (18.8%) and tachypnea (9.51%), showcasing the critical conditions often treated in emergency settings.
-Additionally, fever and hypotension are observed in 2.43% and 1.57% of patients, respectively, highlighting the variety of urgent care demands.
-
-Patient complaints commonly involve abdominal issues (16.9%), extremity problems (14.3%), and chest pain (9.58%), along with neurological (9.50%) and gastrointestinal complaints (9.04%).
-These figures emphasize the frequent health concerns that drive ED visits and the resources needed for effective response.
-
-The patient demographic profile shows an average age of 58.3 years, predominantly middle-aged and elderly individuals.
-Racial distribution is mainly White (88.6%), with smaller percentages of Black (4.12%) and Asian (2.97%) populations.
-A slight majority of patients are female (54.5%), indicating gender trends in healthcare utilization.
-
-Diagnostic procedures heavily rely on lab tests (78.5%), essential for thorough patient evaluations.
-Imaging tests such as X-rays (48.1%) and CT scans—both non-contrast (23.1%) and contrast (20.5%)—play pivotal roles in diagnostics.
-The use of ultrasounds (12.5%) and the strategy of batch ordering tests (6.32%) are critical operational aspects of the ED, reflecting approaches to efficient patient management.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
+%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
+%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
+%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[,,nonblindrev]{informs}
+
+\OneAndAHalfSpacedXI
+%%\OneAndAHalfSpacedXII % Current default line spacing
+%%\DoubleSpacedXII
+%%\DoubleSpacedXI
+
+%% BEGIN MY ADDITIONS %%
+\usepackage{hyperref}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{booktabs}
+\usepackage{tabularx}
+\usepackage{graphicx}
+\usepackage{tikz}
+\usepackage{siunitx}
+\usepackage{tablefootnote}
+\usepackage{longtable}
+\usepackage{threeparttable}
+\usepackage{natbib}
+\usepackage{caption}
+
+%% END MY ADDITIONS %%
+
+% Natbib setup for author-year style
+\usepackage{natbib}
+ \bibpunct[, ]{(}{)}{,}{a}{}{,}%
+ \def\bibfont{\small}%
+ \def\bibsep{\smallskipamount}%
+ \def\bibhang{24pt}%
+ \def\newblock{\ }%
+ \def\BIBand{and}%
+
+%% Setup of theorem styles. Outcomment only one.
+%% Preferred default is the first option.
+\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
+%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
+\ECRepeatTheorems
+
+%% Setup of the equation numbering system. Outcomment only one.
+%% Preferred default is the first option.
+\EquationsNumberedThrough    % Default: (1), (2), ...
+%\EquationsNumberedBySection % (1.1), (1.2), ...
+
+% For new submissions, leave this number blank.
+% For revisions, input the manuscript number assigned by the on-line
+% system along with a suffix ".Rx" where x is the revision number.
+\MANUSCRIPTNO{}
+
+%%%%%%%%%%%%%%%%
+\begin{document}
+%%%%%%%%%%%%%%%%
+
+% Outcomment only when entries are known. Otherwise leave as is and
+%   default values will be used.
+%\setcounter{page}{1}
+%\VOLUME{00}%
+%\NO{0}%
+%\MONTH{Xxxxx}% (month or a similar seasonal id)
+%\YEAR{0000}% e.g., 2005
+%\FIRSTPAGE{000}%
+%\LASTPAGE{000}%
+%\SHORTYEAR{00}% shortened year (two-digit)
+%\ISSUE{0000} %
+%\LONGFIRSTPAGE{0001} %
+%\DOI{10.1287/xxxx.0000.0000}%
+
+% Author's names for the running heads
+% Sample depending on the number of authors;
+\RUNAUTHOR{%
+Jameson, Saghafian, Huckman
+ and Hodgson
+}
+% \RUNAUTHOR{Jones and Wilson}
+% \RUNAUTHOR{Jones, Miller, and Wilson}
+% \RUNAUTHOR{Jones et al.} % for four or more authors
+% Enter authors following the given pattern:
+%\RUNAUTHOR{}
+
+\RUNTITLE{Judicious Batching: The Impact of Batch Ordering Imaging Tests
+in the Emergency Department}
+
+\TITLE{Judicious Batching: The Impact of Batch Ordering Imaging Tests in
+the Emergency Department}
+
+\ARTICLEAUTHORS{%
+\AUTHOR{Jacob Jameson}
+\AFF{Interfaculty Initiative in Health Policy, Harvard
+University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}
+
+\AUTHOR{Soroush Saghafian}
+\AFF{Kennedy School of Government, Harvard
+University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}
+
+\AUTHOR{Robert Huckman}
+\AFF{Harvard Business
+School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}
+
+\AUTHOR{Nicole Hodgson}
+\AFF{Mayo
+Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}
+
+%
+}
+
+\ABSTRACT{We use practice variation across physicians to uncover the
+role of test-ordering on care delivery in the ED. Using records of over
+45,000 Mayo Clinic emergency department visits, we find that
+quasi-random assignment to a top (versus bottom) decile batching
+physician significantly increases excess diagnostic testing.
+Instrumental variable results show that batching leads to an additional
+13.5 tests per 100 batches, resulting in excess tests that would not
+have been ordered under a sequential-ordering strategy. Contrary to
+expectations, batch ordering does not reduce ED length of stay or affect
+72-hour return rates, challenging the perceived efficiency of batching
+and highlighting the necessity for more targeted testing strategies in
+emergency care.}
+
+\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
+Testing}
+
+\maketitle
+
+\hypertarget{sec:I}{%
+\section{Introduction}\label{sec:I}}
+
+Healthcare delivery, particularly in the emergency department (ED), is a
+delicate balance that involves ensuring optimal patient outcomes while
+optimizing resource utilization. Achieving these twin goals requires
+timely and accurate diagnosis, which in turn enables prompt and
+appropriate treatment, consequently improving patient prognosis and
+reducing the likelihood of adverse events. Furthermore, efficient
+patient discharge from the ED can help alleviate overcrowding, a severe
+issue with potential consequences including higher complication rates
+and increased mortality \citet{bernstein2009}.
+
+One important factor that can impact the speed and effectiveness of
+diagnosis in the ED is the availability and performance of diagnostic
+tests \citet{naseim2015}. A variety of diagnostic tests are used in the
+ED, including laboratory tests, imaging studies, and specialized tests.
+These tests can provide valuable information about a patient's condition
+and help to guide treatment decisions.
+
+A critical question in this context pertains to whether physicians in
+the ED should batch order diagnostic tests or order them sequentially.
+This decision essentially represents a tradeoff between reducing patient
+length of stay and risk of over-testing. Over-testing, or performing
+unnecessary tests, can lead to increased costs, unnecessary patient
+anxiety, and potential harm from follow-up of false-positive results
+\citet{koch2018}. Conversely, keeping a patient for an extended time to
+perform all possible tests could lead to ED overcrowding, an issue
+associated with severe consequences, as mentioned earlier. Instead, what
+is needed is a reasonable balance between the number of diagnostic tests
+performed and the total time the patient is kept in the ED before either
+being admitted or discharged. Several studies have demonstrated that
+optimizing the ED patient flow process can result in significant
+improvements \citet{saghafian2015}, however, research surrounding test
+ordering strategies to improve the patient flow processes remains
+limited.
+
+In this paper, we use data from over 45,000 patient visits to the ED
+that occur during our study period to quantify the benefits and
+consequences of batching versus sequentially ordering diagnostic tests
+on patient length of stay, re-admission, and resource utilization. Our
+empirical strategy exploits random assignment of patients to ED
+physicians who differ in their propensity to batch-order diagnostic
+tests. When patients arrive at the ED, they are assigned to a physician
+based on availability, with no discretion on either side. Thus, patients
+who arrive at the ED at similar times are randomly assigned to
+physicians who vary in their willingness to batch order diagnostic
+tests. We measure physician tendency to batch using a leave-out,
+residualized measure based on all other patients the physician has seen
+in the ED in the study period. The tendency measure strongly predicts
+the ED test batch outcome but is uncorrelated with patient and ED visit
+characteristics.
+
+We start by evaluating the reduced-form effects of ED provider
+batch-ordering tendency on downstream patient outcomes and turnaround
+time. We find that practice variation as captured by physician
+batch-ordering tendency has large and significant consequences. Being
+treated by a provider in the top decile of the tendency distribution,
+compared to being treated by someone in the bottom decile results in an
+additional 2 diagnostic test per 100 patient encounters after
+controlling for the physician's underlying propensity to test.
+
+Physician care takes on multiple dimensions besides the decision to
+batch order diagnostic tests. Employing a placebo exercise, we find
+evidence that that excess testing occurs after having seen a
+high-batching physician is due to batching from this physician, as
+opposed to other differences across physician in care that correlate
+with batching tendency: we find no effects of physician batch tendency
+on key outcomes for a ``placebo sample'' of patients who visit the ED
+for conditions that rarely result in batched orders.
+
+With the caveat that other unobserved dimensions of physician care may
+impact patient outcomes, we employ physician batch tendency as an
+instrumental variable for having tests batch ordered in the ED to
+quantify the effect of batching directly. Because of the institutional
+features of the ED, our research design closely approximates an RCT that
+assigns patients to batch-ordering or sequential-ordering arm. In the
+ED, patients have no discretion over choosing providers, and in our
+specific ED, physicians have discretion over choosing patients,
+alleviating major selection issues present in other health care
+settings. Furthermore, physicians exhibit wide variation in practice
+behavior in batch-ordering, even within the same hospital, while
+following the same guidelines. Finally, patient-physician interactions
+in the ED are typically well documented, short, and one-off,
+constraining physician decision-making to a more limited,
+better-observed choice set than present in settings such as specialty or
+primary care.
+
+In sum, exploiting practice variation in ED settings shuts down other
+(but not all) potential channels besides test batching that are present
+in other settings, determine length of stay, and impact patient
+outcomes. This approach allows us to move closer to identifying the
+causal impact of batch-ordering diagnostic tests on patient outcomes and
+resource utilization. It is important to note that this paper studies
+the impact of batch-ordering through a batching decision requiring
+clinical judgment (within practice norms) rather than through specific
+hospital policies, differences in adherence to clinical practice
+guidelines, or substandard care.
+
+The remainder of this paper is structured as follows. The next section
+describes the data source and outlines our baseline sample. The
+empirical strategy and its accompanying identifying assumptions are laid
+out in Section III. Section IV presents the results. Section V draws
+implications for batch-ordering policies. The last section concludes.
+
+\hypertarget{sec:II}{%
+\section{Data and Definitions}\label{sec:II}}
+
+Our study was conducted in the Emergency Department (ED) of the Mayo
+Clinic of Arizona, a tertiary care hospital without obstetrical
+services, an inpatient pediatrics unit, or a trauma designation. During
+the study period, the ED recorded approximately 43,000 visits per year,
+managed across 26 treatment rooms and up to 9 hallway spaces. The
+department is exclusively staffed by board-eligible or board-certified
+emergency physicians (EPs), with rotating residents overseeing about
+10\% of patient volume. Physicians operate in a unique workflow that
+includes staggered 8.5-hour shifts and a rotational patient assignment
+system, minimizing potential selection bias in patient encounters.
+
+We conducted a retrospective review of comprehensive ED operational data
+from 10/6/2018 through 12/31/2019, coinciding with the initiation of a
+new electronic medical record. The dataset includes detailed patient
+demographics, chief complaints, vital signs, emergency severity index
+(ESI), length of stay (LOS), and resource utilization metrics. This
+period was chosen to provide a robust data set while excluding the
+influence of the coronavirus pandemic.
+
+\hypertarget{sample-construction}{%
+\subsection{Sample Construction}\label{sample-construction}}
+
+Our research design focuses on adults who visit the Mayo Clinic of
+Arizona ED. We observe approximately \(48,000\) such visits during the
+study period. To improve power, we drop encounters with rare chief
+complaints (\(<1000\) total encounters of this kind) and complaints
+where a batch order occurs less than \(2\) percent of the time. Since
+batch orders are rare for these cases, our physician batch tendency
+instrument could suffer from a weak instrument problem if we included
+them. Examples of complaints dropped include Upper Respiratory Symptoms
+and Urinary Complaints. Excluding these conditions does not introduce
+selection bias unless physician test batching tendency is orthogonal to
+physician diagnosing behavior. While this assumption may be violated if
+we were to use a very detailed level of chief complaint information upon
+which to base our exclusion criterion, it is plausibly satisfied when
+using broad complaint categories. In order to estimate a precise measure
+of physician-level batch tendency, we further restrict our sample to the
+\(36,765\) encounters involving full-time physicians who treat over 500
+ED cases per year.
+
+\hypertarget{variable-definitions}{%
+\subsection{Variable Definitions}\label{variable-definitions}}
+
+Our explanatory variable in the IV analysis, \(Batched_i\), is an
+indicator for whether patient \(i\) has their tests batch-ordered at
+their ED encounter. While the patient could decide not to undergo the
+tests ordered by the physician, this is rare in practice. Below, we
+detail our primary outcomes: (a) ED length of stay (LOS), and (b)
+resource utilization.
+
+\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
+efficiency and patient throughput in emergency care settings. It is
+defined as the duration from a patient's arrival to the ED until their
+departure, whether by discharge or admission to the hospital. The
+hypothesis is that batch testing may lead to shorter ED-LOS, potentially
+improving patient throughput and reducing crowding.
+
+\emph{(b) Resource Utilization}-- Resource utilization in the ED
+typically refers to the extent of medical services and interventions a
+patient receives. In this study, we quantify resource utilization by the
+total number of distinct diagnostic tests ordered per patient during
+their ED stay. This encompasses both initial and subsequent tests. The
+hypothesis is that batch testing may lead to variations in the number of
+tests ordered, potentially influencing the overall healthcare
+expenditure and efficiency.
+
+\emph{(c) 72 Hour Return with Admission}-- This is a binary variable
+indicating whether the patient returns to the ED within 72 hours of
+their initial visit and is admitted to the hospital. This outcome is
+used to assess the quality of care provided during the initial ED visit.
+The hypothesis is that batch testing may lead to variations in the
+quality of care provided, potentially influencing the likelihood of a
+return visit.
+
+\hypertarget{batching}{%
+\subsubsection{Batching}\label{batching}}
+
+We define ``batching'' in line with standard emergency medicine
+practices. Batching occurs when a physician simultaneously orders a
+comprehensive set of diagnostic tests, typically covering a broad range
+of potential diagnoses. This contrasts with sequential ordering, where
+tests are ordered sequentially based on the information obtained from
+each test as needed.
+
+We operationalize batching as occurring when multiple diagnostic imaging
+tests are ordered within a 5-minute window. In Section 4.4, sensitivity
+analyses on this cutoff point showed that our results are robust to this
+definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is
+considered a separate, distinct test for our study. Therefore, a batch
+in our study consists of two or more distinct imaging tests.
+
+\hypertarget{summary-statistics}{%
+\subsection{Summary Statistics}\label{summary-statistics}}
+
+hyperref{[}tab{]}\{Table 1\} provides a concise overview of emergency
+department (ED) characteristics, patient demographics, and medical tests
+based on our study data. The ED maintains an average volume of 24.3
+patients, reflecting a significant yet manageable workload. Key
+physiological indicators include tachycardia (18.8\%) and tachypnea
+(9.51\%), showcasing the critical conditions often treated in emergency
+settings. Additionally, fever and hypotension are observed in 2.43\% and
+1.57\% of patients, respectively, highlighting the variety of urgent
+care demands.
+
+Patient complaints commonly involve abdominal issues (16.9\%), extremity
+problems (14.3\%), and chest pain (9.58\%), along with neurological
+(9.50\%) and gastrointestinal complaints (9.04\%). These figures
+emphasize the frequent health concerns that drive ED visits and the
+resources needed for effective response.
+
+The patient demographic profile shows an average age of 58.3 years,
+predominantly middle-aged and elderly individuals. Racial distribution
+is mainly White (88.6\%), with smaller percentages of Black (4.12\%) and
+Asian (2.97\%) populations. A slight majority of patients are female
+(54.5\%), indicating gender trends in healthcare utilization.
+
+Diagnostic procedures heavily rely on lab tests (78.5\%), essential for
+thorough patient evaluations. Imaging tests such as X-rays (48.1\%) and
+CT scans---both non-contrast (23.1\%) and contrast (20.5\%)---play
+pivotal roles in diagnostics. The use of ultrasounds (12.5\%) and the
+strategy of batch ordering tests (6.32\%) are critical operational
+aspects of the ED, reflecting approaches to efficient patient
+management.
 
@@
 
-# Empirical Strategy {#sec:3}
-
-Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court.
-Similarly, we explore batching variation across physicians who work in the same emergency department.
-In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians.
-Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision.
-We employ both approaches and lay out their details in the next subsections.
-
-## Institutional Details on Patient-Physician Assignment
-
-Contrary to most healthcare settings where patients exhibit choice, they are predominantly passive in their physician assignment in the ED.
-In most EDs, however, physicians have discretion in picking their patients.
-In contrast, patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses.
-In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random.
-Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.
+\hypertarget{sec:3}{%
+\section{Empirical Strategy}\label{sec:3}}
+
+Our empirical strategy closely follows the literature that relies on
+quasi-random assignment of agents to cases, often referred to as the
+``judges design.'' Papers in this literature typically exploit variation
+in the sentencing leniency of judges who work in the same court.
+Similarly, we explore batching variation across physicians who work in
+the same emergency department. In its reduced form, under the assumption
+of quasi-random assignment, this approach allows researchers to identify
+the causal effect of being assigned to different types of physicians.
+Under additional assumptions, an instrumental variable approach
+identifies the causal effect of a given medical decision. We employ both
+approaches and lay out their details in the next subsections.
+
+\hypertarget{institutional-details-on-patient-physician-assignment}{%
+\subsection{Institutional Details on Patient-Physician
+Assignment}\label{institutional-details-on-patient-physician-assignment}}
+
+Contrary to most healthcare settings where patients exhibit choice, they
+are predominantly passive in their physician assignment in the ED. In
+most EDs, however, physicians have discretion in picking their patients.
+In contrast, patients arriving at the Mayo Clinic ED are randomly
+assigned to physicians via a rotational patient assignment algorithm
+\citet{traub2016emergency}, which removes potential selection bias
+concerns for our analyses. In essence, barring arrival time and
+shift-level variation, the physician-to-patient matching can be deemed
+random. Table \ref{tab:wald_test} displays that patient encounters
+(regarding chief complaints and emergency severity) are equitably
+distributed across physicians within our study's cohort.
 
@@
 
-## Batch Tendency Construction
-
-To measure physician batch tendency, we use the physician's residualized leave-out average batch rate.
-This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways.
-First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.
+\hypertarget{batch-tendency-construction}{%
+\subsection{Batch Tendency
+Construction}\label{batch-tendency-construction}}
+
+To measure physician batch tendency, we use the physician's residualized
+leave-out average batch rate. This measure is derived from two steps
+following the approaches taken by \citet{doyle2015measuring},
+\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
+obtain residuals from a regression model, which includes all ED
+encounters in our sample period.
 
@@
 
-Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$.
-Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g.
-initiatives to eliminate excess testing) or seasonality in ED visits.
-We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$.
-Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision.
-Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case.
-As stated earlier, these controls are more than what is required for our quasi-random assignment assumption.
-Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.
-
-In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:
+Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
+\(i\) had their imaging tests batch ordered on encounter that took place
+on date \(t\). Fixed effects include year-month fixed effects,
+\(\alpha_{ym}\), to control for time and seasonal variation in batching,
+such as hospital-specific policies (e.g.~initiatives to eliminate excess
+testing) or seasonality in ED visits. We also control for
+``shift-level'' variations that include both physician scheduling and
+patient arrival with day of week-time of day fixed effects,
+\(\alpha_{dt}\). Chief complaint by severity fixed effects,
+\(\alpha_{complaint}\), were also included to increase precision.
+Finally, a binary variable for whether or not laboratory tests were
+ordered, \(\alpha_{lab}\), was included to account for the complexity of
+the case. As stated earlier, these controls are more than what is
+required for our quasi-random assignment assumption. Under the
+assumption that we have captured the observables under which
+quasi-random assignment occurs in the ED, the unexplained variation--
+the physician's contribution-- resides in the error term,
+\(\varepsilon_{i,t}\).
+
+In step two, the tendency measure for patient \(i\) seen by physician
+\(j\) is computed as the average residual across all other patients seen
+by the physician that year:
 
@@
 
-where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$.
-This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument.
-The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.
-
-We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests.
-Figure \ref{fig:physician_batching} graphs the frequency with which physicians batch order across different chief complaints, highlighting that the variation in batching differs starkly.
-We note that there is high correlation between
+where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
+residual from equation (1); \(J\) is the set of all ED encounters
+treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
+number of cases that physician has seen that year, excluding patient
+\(i\). This leave-out mean eliminates the mechanical bias that stems
+from patient \(i\)'s own case entering into the instrument. The measure
+is interpreted as the average (leave-out) batch rate of patient \(i\)'s
+physician, relative to other physicians in that hospital-year-month,
+hospital-day of week-time of day.
+
+We document that the Mayo Clinic ED physicians exhibit wide, systematic
+variation in their propensity to batch order diagnostic tests. Figure
+\ref{fig:physician_batching} graphs the frequency with which physicians
+batch order across different chief complaints, highlighting that the
+variation in batching differs starkly. We note that there is high
+correlation between
 
@@
 
-\hyperref[table:first_stage]{Table 1} illustrates the "first stage" results from both linear and probit models, emphasizing the robust relevance of the instrument across different statistical frameworks.
-In both modeling approaches, a 10 percentage point increase in the batch-tendency of physicians consistently results in a significant increase in the likelihood of batch ordering tests in the emergency department—1.15 percentage points in the linear model and 93.24 percentage points in the probit model.
-This consistent significance across models underscores the strength and relevance of batch tendency as an instrumental variable.
-Given the robustness to different model specifications, we move forward with a probit model in our first stage to estimate the effect of batch ordering on patient outcomes.
+\hyperref[table:first_stage]{Table 1} illustrates the ``first stage''
+results from both linear and probit models, emphasizing the robust
+relevance of the instrument across different statistical frameworks. In
+both modeling approaches, a 10 percentage point increase in the
+batch-tendency of physicians consistently results in a significant
+increase in the likelihood of batch ordering tests in the emergency
+department---1.15 percentage points in the linear model and 93.24
+percentage points in the probit model. This consistent significance
+across models underscores the strength and relevance of batch tendency
+as an instrumental variable. Given the robustness to different model
+specifications, we move forward with a probit model in our first stage
+to estimate the effect of batch ordering on patient outcomes.
 
@@
 This reduced form will allow us to check the strength of our instrument.
-To study the effects of test batching in the ED on an outcome $Y_i$, we employ a two-stage estimation approach using our baseline sample:
+To study the effects of test batching in the ED on an outcome \(Y_i\),
+we employ a two-stage estimation approach using our baseline sample:
 
@@
 
-Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization, and $X_i$ is the same as in the reduced-form approach.
-$Batched_i$ variable suffers from potential endogeneity concerns.
-For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay.
-Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$.
-We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.
-
-## Identifying Assumptions
-
-The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence").
-The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.
-
-Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity.
-We reported a strong first stage (i.e., relevance) at the end of the previous Section.
-The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching.
-This is perhaps our strongest assumption and is at its core, untestable.
-However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings.
-First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher.
-Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest.
-The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis.
-Nevertheless, we take this assumption seriously and perform a placebo check in Section 4.2 as well as various robustness checks in Section 4.4.
-
-Finally, the monotonicity assumption is necessary for interpreting the coefficient estimates obtained from the IV approach as Local Average Treatment Effects (LATEs) if there are heterogeneous treatment effects.
-It requires that any patient who is (not) batched by a sequencer (batcher) would also (not) be batched by a batcher (sequencer) physician.
-The literature leveraging the judges design typically performs two informal tests for its implications.
-The first one provides that the first stage should be weakly positive for all subsamples (@dobbie2018effects).
-The second implication asserts that the instrument constructed by leaving out a particular subsample has predictive power over that same left-out subsample (@bhuller2020incarceration).
-Appendix Table  \ref{tab:monotonicity} presents both of these tests in the two columns for various subsamples of interest.
-
-# Results {#sec:4}
-
-## Reduced-Form Results
-
-In this section, we explore the causal influence of physician batch tendency on patient outcomes and resource utilization in the emergency department.
-We posit that while batch tendency directly influences the practice of batch ordering tests, both batch tendency and batch ordering are concurrently influenced by a physician's testing inclination.
-Given that testing inclination directly affects primary outcomes, we include it as a control variable in our regression models to mitigate its confounding effects.
-
-To quantify a physician's testing inclination, we employ a similar approach to that used in measuring physician batch tendency.
-Specifically, we calculate the physician's residualized leave-out average test rate, which serves as a proxy for their propensity to order tests.
-It is important to note the strong positive correlation ($r = 0.79$) between batch tendency and testing inclination.
-This substantial correlation suggests that physicians with a higher propensity to test may also exhibit a higher tendency to batch orders, potentially as a consequence of their testing strategies.
-Neglecting to account for testing inclination could lead to overestimated effects of batch tendency due to omitted variable bias.
-
-Reduced-form regression results in Table \ref{tab:reducedform} reveal the effect the average total effect of batch tendency on resource utilization and LOS.
-However, it is important to note that the estimated total effect of batch tendency accounts for both direct and indirect (mediated) effects of batch tendency.
-We hypothesize that while batch ordering may streamline the testing process, resulting in a quicker completion of a given number of tests, it simultaneously appears to lead to an increase in the total number of tests ordered.
-The increased testing volume, in turn, is associated with an extended LOS.
+Where \(Y_i\) represents our main outcomes of interest: length of stay,
+72 hour readmission, and resource utilization, and \(X_i\) is the same
+as in the reduced-form approach. \(Batched_i\) variable suffers from
+potential endogeneity concerns. For example, injury severity may be
+unobserved and correlated with need to run multiple tests, which in turn
+also affects length of stay. Hence, we instrument \(Batched_i\) with the
+assigned physician \(j\)'s underlying tendency to batch,
+\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
+physician level to account for the assignment process of patients to
+physicians.
+
+\hypertarget{identifying-assumptions}{%
+\subsection{Identifying Assumptions}\label{identifying-assumptions}}
+
+The reduced-form approach delivers an unbiased estimate of the causal
+effect of being treated by a higher tendency to batch physician, since
+assignment of patients to ED physicians is random, conditional on
+seasonality and shift (``conditional independence''). The
+residualization in equation (1) controls for more controls than required
+to achieve quasi-random assignment; they are included for statistical
+precision in measuring physician tendency to batch.
+
+Our instrumental variable approach, which aims to recover the causal
+effect of having diagnostic tests batch ordered, relies on three
+additional assumptions: relevance, exclusion, and monotonicity. We
+reported a strong first stage (i.e., relevance) at the end of the
+previous Section. The exclusion restriction requires that the instrument
+must influence the outcome of interest only through its effect on test
+batching. This is perhaps our strongest assumption and is at its core,
+untestable. However, several features of the ED setting suggest that
+such violation may likely only have a small impact and may be less
+concerning than in other health care settings. First, unlike in primary
+care settings, where the patient and primary care provider have many
+repeat encounters, the scope of what the emergency physician can do to
+impact medium-term outcomes is limited and well-observed by the
+researcher. Second, any violation of the exclusion restriction needs to
+directly affect the specific outcome of interest. The channel by which
+ED physicians can influence length of stay relative outcomes is likely
+through testing and diagnosis. Nevertheless, we take this assumption
+seriously and perform a placebo check in Section 4.2 as well as various
+robustness checks in Section 4.4.
+
+Finally, the monotonicity assumption is necessary for interpreting the
+coefficient estimates obtained from the IV approach as Local Average
+Treatment Effects (LATEs) if there are heterogeneous treatment effects.
+It requires that any patient who is (not) batched by a sequencer
+(batcher) would also (not) be batched by a batcher (sequencer)
+physician. The literature leveraging the judges design typically
+performs two informal tests for its implications. The first one provides
+that the first stage should be weakly positive for all subsamples
+(\citet{dobbie2018effects}). The second implication asserts that the
+instrument constructed by leaving out a particular subsample has
+predictive power over that same left-out subsample
+(\citet{bhuller2020incarceration}). Appendix Table
+\ref{tab:monotonicity} presents both of these tests in the two columns
+for various subsamples of interest.
+
+\hypertarget{sec:4}{%
+\section{Results}\label{sec:4}}
+
+\hypertarget{reduced-form-results}{%
+\subsection{Reduced-Form Results}\label{reduced-form-results}}
+
+In this section, we explore the causal influence of physician batch
+tendency on patient outcomes and resource utilization in the emergency
+department. We posit that while batch tendency directly influences the
+practice of batch ordering tests, both batch tendency and batch ordering
+are concurrently influenced by a physician's testing inclination. Given
+that testing inclination directly affects primary outcomes, we include
+it as a control variable in our regression models to mitigate its
+confounding effects.
+
+To quantify a physician's testing inclination, we employ a similar
+approach to that used in measuring physician batch tendency.
+Specifically, we calculate the physician's residualized leave-out
+average test rate, which serves as a proxy for their propensity to order
+tests. It is important to note the strong positive correlation
+(\(r = 0.79\)) between batch tendency and testing inclination. This
+substantial correlation suggests that physicians with a higher
+propensity to test may also exhibit a higher tendency to batch orders,
+potentially as a consequence of their testing strategies. Neglecting to
+account for testing inclination could lead to overestimated effects of
+batch tendency due to omitted variable bias.
+
+Reduced-form regression results in Table \ref{tab:reducedform} reveal
+the effect the average total effect of batch tendency on resource
+utilization and LOS. However, it is important to note that the estimated
+total effect of batch tendency accounts for both direct and indirect
+(mediated) effects of batch tendency. We hypothesize that while batch
+ordering may streamline the testing process, resulting in a quicker
+completion of a given number of tests, it simultaneously appears to lead
+to an increase in the total number of tests ordered. The increased
+testing volume, in turn, is associated with an extended LOS.
 
@@
 
-We introduce a Directed Acyclic Graph (DAG) to illustrate the interconnected pathways between batch tendency, batch ordering, and primary outcomes.
-The DAG in Figure \ref{fig:DAG} shows that batch tendency directly influences batch ordering, which in turn affects the number of tests ordered.
-The number of tests ordered, in turn, influences the length of stay.
-The DAG also illustrates the effect of batch tendency on the length of stay  is mediated by the number of tests ordered.
+We introduce a Directed Acyclic Graph (DAG) to illustrate the
+interconnected pathways between batch tendency, batch ordering, and
+primary outcomes. The DAG in Figure \ref{fig:DAG} shows that batch
+tendency directly influences batch ordering, which in turn affects the
+number of tests ordered. The number of tests ordered, in turn,
+influences the length of stay. The DAG also illustrates the effect of
+batch tendency on the length of stay is mediated by the number of tests
+ordered.
 
@@
 
-To determine whether or not the effect of batch tendency on LOS is mediated by the number of tests ordered, we estimate a mediation analysis using the approach outlined by @imai2010general.
-Our analysis, scrutinizes the pathways through which physician batch tendency influences patient length of stay (LOS) in the emergency department.
-The analysis delineates both direct effects, which capture the influence of batch tendency on LOS without intermediaries, and indirect effects, which operate through the mediator of test ordering volume.
-Our findings underscore a significant Average Direct Effect (ADE) of batch tendency on LOS, estimated at -0.15  (p = 0.046).
-
-## Placebo Check
-
-In this section we investigate whether the reduced-form effects observed in Section 4.1 are due to differences in batch rates across providers or due to other provider differences correlated with batch tendency.
-We start by studying reduced-form effects among patients with complaints that are never batched, as a ``placebo/falsification check.” By way of example, consider a patient who arrives at the ED with a urinary tact infection—a condition for which patients rarely undergo imaging testing.
-For such patients, we should expect to see no impact of batch tendency only if high batching and low batching physicians do not systematically differ in other dimensions of care relevant to patient outcomes.
-Conversely, if we do find a reduced-form effect for these patients, then high batch tendency physicians must systematically differ from low batch tendency physicians in other dimensions of care, beyond batching.
-
-To that end, we restrict attention to ED visits for complaints where batching occurs no more than 10 percent of the time (recall that our baseline sample only includes complaints with a >10 percent batching rate).
-We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3).
-The results of this exercise are displayed in Table \ref{tab:placebo_check}.
-They show that in contrast to results for our main sample, the association between physician tendency to batch and a given outcome is statistically indistinguishable from zero and much smaller in magnitude for the samples of patients who visit the ED with health conditions that are rarely batched.
+To determine whether or not the effect of batch tendency on LOS is
+mediated by the number of tests ordered, we estimate a mediation
+analysis using the approach outlined by \citet{imai2010general}. Our
+analysis, scrutinizes the pathways through which physician batch
+tendency influences patient length of stay (LOS) in the emergency
+department. The analysis delineates both direct effects, which capture
+the influence of batch tendency on LOS without intermediaries, and
+indirect effects, which operate through the mediator of test ordering
+volume. Our findings underscore a significant Average Direct Effect
+(ADE) of batch tendency on LOS, estimated at -0.15 (p = 0.046).
+
+\hypertarget{placebo-check}{%
+\subsection{Placebo Check}\label{placebo-check}}
+
+In this section we investigate whether the reduced-form effects observed
+in Section 4.1 are due to differences in batch rates across providers or
+due to other provider differences correlated with batch tendency. We
+start by studying reduced-form effects among patients with complaints
+that are never batched, as a ``placebo/falsification check.'' By way of
+example, consider a patient who arrives at the ED with a urinary tact
+infection---a condition for which patients rarely undergo imaging
+testing. For such patients, we should expect to see no impact of batch
+tendency only if high batching and low batching physicians do not
+systematically differ in other dimensions of care relevant to patient
+outcomes. Conversely, if we do find a reduced-form effect for these
+patients, then high batch tendency physicians must systematically differ
+from low batch tendency physicians in other dimensions of care, beyond
+batching.
+
+To that end, we restrict attention to ED visits for complaints where
+batching occurs no more than 10 percent of the time (recall that our
+baseline sample only includes complaints with a \textgreater10 percent
+batching rate). We estimate a reduced-form regression of each main
+outcome on physician batch tendency for the subsample, following
+equation (3). The results of this exercise are displayed in Table
+\ref{tab:placebo_check}. They show that in contrast to results for our
+main sample, the association between physician tendency to batch and a
+given outcome is statistically indistinguishable from zero and much
+smaller in magnitude for the samples of patients who visit the ED with
+health conditions that are rarely batched.
 
@@
 
-## Instrumental Variables Results
-
-In this section we examine the causal effects of batch ordering in the ED.
-Mirroring our presentation of the reduced-form results, we turn to presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED LOS and number of tests ordered.
-We first start by estimating the effects of batch ordering on the number of tests ordered using the standard 2SLS approach.
-Next we estimate the direct and indirect effects of batch ordering on LOS using a causal mediation analysis.
+\hypertarget{instrumental-variables-results}{%
+\subsection{Instrumental Variables
+Results}\label{instrumental-variables-results}}
+
+In this section we examine the causal effects of batch ordering in the
+ED. Mirroring our presentation of the reduced-form results, we turn to
+presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED
+LOS and number of tests ordered. We first start by estimating the
+effects of batch ordering on the number of tests ordered using the
+standard 2SLS approach. Next we estimate the direct and indirect effects
+of batch ordering on LOS using a causal mediation analysis.
 
@@
 
-*ED length of stay (LOS).*-- Our causal mediation analysis reveals that the Average Direct Effect (ADE) of test batching on the natural log of LOS is -0.13607, indicating that, when not considering the mediating effect of the number of tests, batching leads to a decrease in the expected LOS.
-Specifically, this translates to an approximate 12.7 percent reduction in LOS when tests are batched, all else being equal.
-This result, complemented by our 2SLS estimate of -0.086, suggests that the direct effect of batch ordering on LOS is negative.
-Our results also suggest a "washing out" effect, where the effeciency gains from batching are offset by the increased number of tests ordered.
-This is consistent with our finding that batching increases the number of tests ordered.
-
-*Resource Utilization.*-- The impact of batching on resource utilization is pronounced.
-We observe a 0.135 increase in the number of tests ordered due to batching, corresponding to approximately 13.5 excess tests per 100 batch orders, where excess means tests that would not have been ordered had the physician not batched.
-This result is consistent with the notion that batching leads to increased resource utilization in the ED.
-
-The instrumental variable analysis presented here underscores the complexities underlying batch ordering practices in the ED.
-While batching is often seen by physicians as a method to enhance efficiency, our findings suggest that it may lead to increased resource utilization and that the corresponding gains in patient throughput are washed out.
-These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
-
-## Robustness
-
-The 2SLS results presented in the previous sections are robust to several alternative specifications probing the key batching definition.
-Our primary definition operationalizes batching as the ordering of multiple diagnostic tests within a 5-minute window.
-This approach is grounded in standard emergency medicine practices and offers a clear distinction between batching and non-batching behaviors.
-To ensure the robustness of our findings, we performed several sensitivity analyses.
-
-(1) Variation in the Time Window for Batching: Firstly, we varied the time window for what constitutes a batch.
-The original 5-minute window was extended and contracted to understand its impact on the study's outcomes.
-Specifically, we considered scenarios where a batch is defined as two or more tests ordered within windows of 1 minute to 10 minutes.
-This range allowed us to capture a broader spectrum of batching behaviors and test the sensitivity of our results to different batching definitions.
-
-(2) Refinement of Batching Definition: Secondly, we refined our batching definition based on the sequence and timing of test orders.
-In the initial analysis, if two tests were ordered upfront and another test ordered later, it was counted as a batch.
-We modified this definition to consider a set of tests as a batch only if all tests ordered during a patient encounter came from that initial batch order.
-This adjustment ensures that our batching definition more accurately reflects a comprehensive diagnostic effort at the outset of patient care, rather than incremental decision-making.
-
-The key finding from these robustness checks is the consistent impact of batching on key outcomes across all variations.
-As seen in Table \ref{tab:robustness_batching} and \ref{tab:robustness_batching2}, altering the time window for batching, whether narrowing it to as little as 1 minute or expanding it to 10 minutes, did not qualitatively change the results.
-Similarly, refining the definition of batching to consider only those tests ordered in the initial batch also had negligible impact on the study's main findings.
-These results lend further credibility to our initial findings, demonstrating that our conclusions about the impact of batching on patient outcomes in the ED are not sensitive to the specific operational definition of batching.
-
-# Conclusion
-
-Our results highlight that changes in resource utilization and costs can arise as a consequence of small variations in physician test ordering strategies in a single medical encounter, at the emergency department.
-Our instrumental variable approach suggests that the causal effects of batching are substantial: 13.5 excess tests arise per 100 batched orders.
-This is particularly pronounced in high-volume complaint areas, where batching leads to a significant increase in the number of tests ordered.
-However, we find no significant impact of batching on ED length of stay.
-Causal mediation analysis reveals that the direct effect of batching on length of stay is negative, but this effect is offset by the increased number of tests ordered.
-This suggests that the efficiency gains from batching are washed out by the increased resource utilization.
+\emph{ED length of stay (LOS).}-- Our causal mediation analysis reveals
+that the Average Direct Effect (ADE) of test batching on the natural log
+of LOS is -0.13607, indicating that, when not considering the mediating
+effect of the number of tests, batching leads to a decrease in the
+expected LOS. Specifically, this translates to an approximate 12.7
+percent reduction in LOS when tests are batched, all else being equal.
+This result, complemented by our 2SLS estimate of -0.086, suggests that
+the direct effect of batch ordering on LOS is negative. Our results also
+suggest a ``washing out'' effect, where the effeciency gains from
+batching are offset by the increased number of tests ordered. This is
+consistent with our finding that batching increases the number of tests
+ordered.
+
+\emph{Resource Utilization.}-- The impact of batching on resource
+utilization is pronounced. We observe a 0.135 increase in the number of
+tests ordered due to batching, corresponding to approximately 13.5
+excess tests per 100 batch orders, where excess means tests that would
+not have been ordered had the physician not batched. This result is
+consistent with the notion that batching leads to increased resource
+utilization in the ED.
+
+The instrumental variable analysis presented here underscores the
+complexities underlying batch ordering practices in the ED. While
+batching is often seen by physicians as a method to enhance efficiency,
+our findings suggest that it may lead to increased resource utilization
+and that the corresponding gains in patient throughput are washed out.
+These insights point to the need for a more nuanced approach to test
+ordering in emergency care, one that carefully weighs the benefits of
+comprehensive assessment against the risks and costs of potential
+over-testing.
+
+\hypertarget{robustness}{%
+\subsection{Robustness}\label{robustness}}
+
+The 2SLS results presented in the previous sections are robust to
+several alternative specifications probing the key batching definition.
+Our primary definition operationalizes batching as the ordering of
+multiple diagnostic tests within a 5-minute window. This approach is
+grounded in standard emergency medicine practices and offers a clear
+distinction between batching and non-batching behaviors. To ensure the
+robustness of our findings, we performed several sensitivity analyses.
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\item
+  Variation in the Time Window for Batching: Firstly, we varied the time
+  window for what constitutes a batch. The original 5-minute window was
+  extended and contracted to understand its impact on the study's
+  outcomes. Specifically, we considered scenarios where a batch is
+  defined as two or more tests ordered within windows of 1 minute to 10
+  minutes. This range allowed us to capture a broader spectrum of
+  batching behaviors and test the sensitivity of our results to
+  different batching definitions.
+\item
+  Refinement of Batching Definition: Secondly, we refined our batching
+  definition based on the sequence and timing of test orders. In the
+  initial analysis, if two tests were ordered upfront and another test
+  ordered later, it was counted as a batch. We modified this definition
+  to consider a set of tests as a batch only if all tests ordered during
+  a patient encounter came from that initial batch order. This
+  adjustment ensures that our batching definition more accurately
+  reflects a comprehensive diagnostic effort at the outset of patient
+  care, rather than incremental decision-making.
+\end{enumerate}
+
+The key finding from these robustness checks is the consistent impact of
+batching on key outcomes across all variations. As seen in Table
+\ref{tab:robustness_batching} and \ref{tab:robustness_batching2},
+altering the time window for batching, whether narrowing it to as little
+as 1 minute or expanding it to 10 minutes, did not qualitatively change
+the results. Similarly, refining the definition of batching to consider
+only those tests ordered in the initial batch also had negligible impact
+on the study's main findings. These results lend further credibility to
+our initial findings, demonstrating that our conclusions about the
+impact of batching on patient outcomes in the ED are not sensitive to
+the specific operational definition of batching.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+Our results highlight that changes in resource utilization and costs can
+arise as a consequence of small variations in physician test ordering
+strategies in a single medical encounter, at the emergency department.
+Our instrumental variable approach suggests that the causal effects of
+batching are substantial: 13.5 excess tests arise per 100 batched
+orders. This is particularly pronounced in high-volume complaint areas,
+where batching leads to a significant increase in the number of tests
+ordered. However, we find no significant impact of batching on ED length
+of stay. Causal mediation analysis reveals that the direct effect of
+batching on length of stay is negative, but this effect is offset by the
+increased number of tests ordered. This suggests that the efficiency
+gains from batching are washed out by the increased resource
+utilization.
 
 Our findings have important implications for emergency care practice.
-While batching is often seen as a method to enhance efficiency, our results suggest that it may lead to increased resource utilization and that the corresponding gains in patient throughput are washed out.
-These insights point to the need for a more nuanced approach to test ordering in emergency care, one that carefully weighs the benefits of comprehensive assessment against the risks and costs of potential over-testing.
-Future research should explore the impact of batching on other outcomes, such as patient satisfaction and clinical outcomes, to provide a more comprehensive understanding of the implications of batching in the emergency department.
+While batching is often seen as a method to enhance efficiency, our
+results suggest that it may lead to increased resource utilization and
+that the corresponding gains in patient throughput are washed out. These
+insights point to the need for a more nuanced approach to test ordering
+in emergency care, one that carefully weighs the benefits of
+comprehensive assessment against the risks and costs of potential
+over-testing. Future research should explore the impact of batching on
+other outcomes, such as patient satisfaction and clinical outcomes, to
+provide a more comprehensive understanding of the implications of
+batching in the emergency department.
 
@@
 
-:::
+\end{APPENDIX}
+
+% Appendix here
+% Options are (1) APPENDIX (with or without general title) or
+%             (2) APPENDICES (if it has more than one unrelated sections)
+% Outcomment the appropriate case if necessary
+%
+% \begin{APPENDIX}{<Title of the Appendix>}
+% \end{APPENDIX}
+%
+%   or
+%
+% \begin{APPENDICES}
+% \section{<Title of Section A>}
+% \section{<Title of Section B>}
+% etc
+% \end{APPENDICES}
+
+% Acknowledgments here
+\ACKNOWLEDGMENT{The authors would like to thank ..}
+
+\bibliographystyle{informs2014}
+\bibliography{references.bib}
+
+\end{document}

### 2024-09-09 manuscript/manuscript-v32.tex
getting going

--- 2024-09-09 manuscript/manuscript-v2.tex
@@
 percentage points in the probit model. This consistent significance
-across models underscores the strength and relevance of batch tendency
-as an instrumental variable. Given the robustness to different model
-specifications, we move forward with a probit model in our first stage
-to estimate the effect of batch ordering on patient outcomes.
+across models underscores the strength and validity of the
+batch-tendency as an instrumental variable, confirming its utility in
+controlling for unobserved confounders and providing a reliable basis
+for further causal inference. Clustering standard errors at the provider
+level ensures robust inference, reinforcing the instrument's credibility
+irrespective of the analytical method employed.
 
@@
 
-This reduced form will allow us to check the strength of our instrument.
-To study the effects of test batching in the ED on an outcome \(Y_i\),
-we employ a two-stage estimation approach using our baseline sample:
+This reduced form will allow us to check that our instrument is a strong
+instrument. To study the effects of test batching in the ED on an
+outcome \(Y_i\), we estimate the following 2SLS equations using our
+baseline sample:
 
@@
 \begin{equation}
-P(Batched_i = 1) = \Phi(\delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i)
+Batched_i = \delta_0 + \delta_1 Tendency_{i,j}^{phys} + \delta_2 X_i + \nu_i
 \end{equation}
@@
 Where \(Y_i\) represents our main outcomes of interest: length of stay,
-72 hour readmission, and resource utilization, and \(X_i\) is the same
+72 hour readmission, and resource utilization. and \(X_i\) is the same
 as in the reduced-form approach. \(Batched_i\) variable suffers from

### 2024-09-09 manuscript/manuscript.Rmd
getting going

@@
 ---
-title: "To Batch or Not to Batch: Test-Ordering Variabilityin the ED"
+title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
 authors:
@@
 Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior.
-While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
-In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving physicians who treat over 500 ED cases per year.
+While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories.
+In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $42,000$ encounters involving full-time physicians who treat over 500 ED cases per year.
 
@@
 
-  *(c) 72-hour return*-- The rate of patients returning to the ED within 72 hours of their initial visit is a proxy for the quality of care received.
-A high 72-hour return rate might indicate inadequate treatment or diagnostic oversight during the initial visit.
-We measure this outcome as the proportion of patients with a subsequent ED visit within 72 hours post-initial discharge.
-The hypothesis is that batch testing may lead to variations in the 72-hour return rate, potentially influencing the quality of care received.
-
 ### Batching
@@
 We operationalize batching as the ordering of multiple diagnostic tests within a 5-minute window.
-Sensitivity analyses on this cutoff point showed that our results are robust to this definition.
+In Section 4.4, sensitivity analyses on this cutoff point showed that our results are robust to this definition.
 This operational definition aligns with the concept of batching as a single comprehensive effort.
@@
 \hyperref[tab:summary_statistics]{Table 1} presents a detailed overview of the emergency department (ED) characteristics, patient demographics, and medical tests for our baseline sample, derived from the data collected during the study period.
+Data is at the patient encounter level.
 On average, the emergency department manages a volume of approximately 24 patients (Mean = 24.18), indicating a significant but manageable patient load.
@@
 \begin{equation}
-Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint} + \varepsilon_{i,t}
+Batched_{i,t} = \alpha_0 + \alpha_{ym} + \alpha_{dt} + \alpha_{complaint\_esi} + \varepsilon_{i,t}
 \end{equation}
@@
 Figure \ref{fig:physician_batching} graphs the histogram of batch-ordering frequency by physician for popular chief complaints, highlighting that the variation in batching differs systematically.
-\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $8$pp increase in the likelihood of having tests batch-ordered in the ED.
-The F-statistic is 50 when all controls and fixed effects are included.
+\hyperref[tab:regression]{Table 3} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $6$pp increase in the likelihood of having tests batch-ordered in the ED.
+The F-statistic is 94 when all controls and fixed effects are included.
 The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.
@@
   \label{fig:physician_batching}
-  \includegraphics[width=1\textwidth]{../outputs/figures/Figure 2.png}
+  \includegraphics[width=1\textwidth]{../outputs/figures/batch_ordering.png}
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Figure \ref{fig:physician_batching} illuminates the marked differences among physicians in their propensity to batch-order diagnostic tests.
@@
 \midrule
-Batching Tendency & 1.06 *** & 1.08 *** & 1.08 *** \\
+Batching Tendency & 1.06 *** & 1.06 *** & 1.06 *** \\
  & (0.01) & (0.01) & (0.01) \\
 \midrule
-F statistic (full model) & 7.02 & 50.34 & 49.31  \\  
+F statistic (full model) & 23.87 & 94.41 & 90.44  \\  
 F (full model): p-value & $<0.001$ & $<0.001$ & $<0.001$ \\
 \midrule
-Num. obs. & 43327 & 43327 & 43327 \\
+Num. obs. & 41929 & 41929 & 41929 \\
 Seasonality and shift fixed effects? & Yes & Yes & Yes \\
@@
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Estimates of the first stage for the baseline sample described in the text.

### 2024-09-09 manuscript/manuscript.tex
getting going

@@
 \usepackage{tabularx}
+\usepackage{graphicx}
+\usepackage{tikz}
 \usepackage{siunitx}
@@
 \usepackage{natbib}
+\usepackage{caption}
 
@@
 instrumental variable for having tests batch ordered in the ED to
-quantify the effect of batching directly. We find that being treated by
-a top decile batching physician, compared to a bottom decile batching
-physician, results in an additional 13.5 tests per 100 patient
-encounters. We find no evidence that batch-ordering reduces ED length of
-stay or 72-hour return rates. These results suggest that the benefits of
-batch-ordering are limited, and that the practice may be associated with
-significant costs. Our findings are robust to analyses probing the key
-assumptions of our batching definition.
-
-Because of the institutional features of the ED, our research design
-closely approximates an RCT that assigns patients to batch-ordering or
-sequential-ordering arm. In the ED, patients have no discretion over
-choosing providers, and in our specific ED, physicians have discretion
-over choosing patients, alleviating major selection issues present in
-other health care settings. Furthermore, physicians exhibit wide
-variation in practice behavior in batch-ordering, even within the same
-hospital, while following the same guidelines. Finally,
-patient-physician interactions in the ED are typically well documented,
-short, and one-off, constraining physician decision-making to a more
-limited, better-observed choice set than present in settings such as
-specialty or primary care.
+quantify the effect of batching directly. Because of the institutional
+features of the ED, our research design closely approximates an RCT that
+assigns patients to batch-ordering or sequential-ordering arm. In the
+ED, patients have no discretion over choosing providers, and in our
+specific ED, physicians have discretion over choosing patients,
+alleviating major selection issues present in other health care
+settings. Furthermore, physicians exhibit wide variation in practice
+behavior in batch-ordering, even within the same hospital, while
+following the same guidelines. Finally, patient-physician interactions
+in the ED are typically well documented, short, and one-off,
+constraining physician decision-making to a more limited,
+better-observed choice set than present in settings such as specialty or
+primary care.
 
@@
 which to base our exclusion criterion, it is plausibly satisfied when
-using broad complaint
-categories\footnote{The idea being that, for a patient who presents with an ankle fracture, different physicians may choose different detailed diagnosis codes within the three-digit ICD-9 824 (``Fracture of ankle"), but are unlikely to disagree at the broader chief complaint category (``Extremetiy Complaint").}.
-In order to estimate a precise measure of physician-level batch
-tendency, we further restrict our sample to the \(42,000\) encounters
-involving physicians who treat over 500 ED cases per year.
+using broad complaint categories. In order to estimate a precise measure
+of physician-level batch tendency, we further restrict our sample to the
+\(42,000\) encounters involving full-time physicians who treat over 500
+ED cases per year.
 
@@
 tests ordered by the physician, this is rare in practice. Below, we
-detail our primary outcomes: (a) ED length of stay (LOS), (b) resource
-utilization, and (c) 72-hour return.
+detail our primary outcomes: (a) ED length of stay (LOS), and (b)
+resource utilization.
 
@@
 
-\emph{(c) 72-hour return}-- The rate of patients returning to the ED
-within 72 hours of their initial visit is a proxy for the quality of
-care received. A high 72-hour return rate might indicate inadequate
-treatment or diagnostic oversight during the initial visit. We measure
-this outcome as the proportion of patients with a subsequent ED visit
-within 72 hours post-initial discharge. The hypothesis is that batch
-testing may lead to variations in the 72-hour return rate, potentially
-influencing the quality of care received.
-
 \hypertarget{batching}{%
@@
 We operationalize batching as the ordering of multiple diagnostic tests
-within a 5-minute window. Sensitivity analyses on this cutoff point
-showed that our results are robust to this definition. This operational
-definition aligns with the concept of batching as a single comprehensive
-effort. For our analysis, all lab-based testing ordered in this
-timeframe are categorized as one distinct type of testing (lab). In
+within a 5-minute window. In Section 4.4, sensitivity analyses on this
+cutoff point showed that our results are robust to this definition. This
+operational definition aligns with the concept of batching as a single
+comprehensive effort. For our analysis, all lab-based testing ordered in
+this timeframe are categorized as one distinct type of testing (lab). In
 contrast, each imaging test (e.g., X-ray, CT scan) are considered a
@@
 and medical tests for our baseline sample, derived from the data
-collected during the study period. On average, the emergency department
-manages a volume of approximately 24 patients (Mean = 24.18), indicating
-a significant but manageable patient load. Key physiological markers
-such as tachycardia (Mean = 0.19) and tachypnea (Mean = 0.09) are
-prevalent among patients, albeit at varying degrees, highlighting
-critical aspects of emergency care. Notably, a small proportion of
-patients exhibit febrile (Mean = 0.02) and hypotensive (Mean = 0.01)
-conditions, emphasizing the diversity of cases encountered in the ED
-setting. The Emergency Severity Index (ESI) averages at 2.78, suggesting
-that most cases fall within the moderate acuity level.
+collected during the study period. Data is at the patient encounter
+level. On average, the emergency department manages a volume of
+approximately 24 patients (Mean = 24.18), indicating a significant but
+manageable patient load. Key physiological markers such as tachycardia
+(Mean = 0.19) and tachypnea (Mean = 0.09) are prevalent among patients,
+albeit at varying degrees, highlighting critical aspects of emergency
+care. Notably, a small proportion of patients exhibit febrile (Mean =
+0.02) and hypotensive (Mean = 0.01) conditions, emphasizing the
+diversity of cases encountered in the ED setting. The Emergency Severity
+Index (ESI) averages at 2.78, suggesting that most cases fall within the
+moderate acuity level.
 
@@
 regression table: being assigned to a \(10\)pp higher batch-tendency
-physician is associated with a \(8\)pp increase in the likelihood of
-having tests batch-ordered in the ED. The F-statistic is 50 when all
+physician is associated with a \(6\)pp increase in the likelihood of
+having tests batch-ordered in the ED. The F-statistic is 94 when all
 controls and fixed effects are included. The coefficient is greater than
@@
 tendency on patient outcomes and resource utilization in the emergency
-department. Our reduced form regression results, as presented in Model 1
-of Table \ref{tab:reducedform}, reveal that batch tendency correlates
-with prolonged ED length of stay (LOS), increased test ordering, and
-does not materially alter the likelihood of a 72-hour return. Notably,
-these associations are nuanced by an underlying factor: the physician's
-testing inclination.
-
-\begin{table}[htbp]
-\centering
-\caption{Reduced-Form Results: Length of Stay, Number of Tests, and 72-Hour Return}
-\label{tab:reducedform}
-\begin{tabular}{p{6.5cm}ccc}
-\toprule
-& \multicolumn{1}{c}{ln(LOS)}
-& \multicolumn{1}{c}{Number of Tests}
-& \multicolumn{1}{c}{72-Hour Return} \\
-& \multicolumn{1}{c}{(1)}
-& \multicolumn{1}{c}{(2)}
-& \multicolumn{1}{c}{(3)} \\
-\midrule
-\textit{Model 1} & & & \\
-Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
-                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
-\\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
-Hospital volume? & Yes & Yes & Yes \\
-\midrule
-\midrule
-\textit{Model 2} & & & \\
-Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
-                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
-                    &     &     &  \\
-Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
-                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
-\\
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
-Hospital volume? & Yes & Yes & Yes \\
-\bottomrule
+department. We posit that while batch tendency directly influences the
+practice of batch ordering tests, both batch tendency and batch ordering
+are concurrently influenced by a physician's testing inclination. Given
+that testing inclination directly affects primary outcomes, we include
+it as a control variable in our regression models to mitigate its
+confounding effects.
+
+To quantify a physician's testing inclination, we employ a similar
+approach to that used in measuring physician batch tendency.
+Specifically, we calculate the physician's residualized leave-out
+average test rate, which serves as a proxy for their propensity to order
+tests. It is important to note the strong positive correlation
+(\(r = 0.79\)) between batch tendency and testing inclination. This
+substantial correlation suggests that physicians with a higher
+propensity to test may also exhibit a higher tendency to batch orders,
+potentially as a consequence of their testing strategies. Neglecting to
+account for testing inclination could lead to overestimated effects of
+batch tendency due to omitted variable bias.
+
+Reduced-form regression results in Table \ref{tab:reducedform} reveal
+the effect the average total effect of batch tendency on resource
+utilization and LOS. However, it is important to note that the estimated
+total effect of batch tendency accounts for both direct and indirect
+(mediated) effects of batch tendency. We hypothesize that while batch
+ordering may streamline the testing process, resulting in a quicker
+completion of a given number of tests, it simultaneously appears to lead
+to an increase in the total number of tests ordered. The increased
+testing volume, in turn, is associated with an extended LOS.
+
+\begin{table}[!htbp] \centering 
+\caption{Reduced-Form Results: Length of Stay and Number of Tests}
+  \label{} 
+\begin{tabular}{@{\extracolsep{5pt}}lcc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & ln(LOS) & Number of Tests \\ 
+\\[-1.8ex] & (1) & (2)\\ 
+\hline \\[-1.8ex] 
+ Batch Tendency & $-$0.109 & 0.149$^{***}$ \\ 
+  & (0.464) & (0.048) \\ 
+  & & \\ 
+ Testing Inclination & 0.605$^{**}$ & 0.952$^{***}$ \\ 
+  & (0.240) & (0.023) \\ 
+  & & \\ 
+Seasonality and shift fixed effects? & Yes & Yes \\
+Chief Complaint? & Yes & Yes  \\
+\hline 
+\textit{N} & 41,929 & 41,929 \\ 
+R$^{2}$ & 0.215 & 0.265 \\ 
+Adjusted R$^{2}$ & 0.213 & 0.263 \\ 
+Residual Std. Error (df = 41800) & 0.490 & 0.842 \\ 
+\hline 
+\hline \\[-1.8ex] 
 \end{tabular} 
 \begin{tablenotes}
-\tiny
+\small
 \item \textit{Estimates of the reduced form are for the baseline sample described in the text.
-Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability.
 Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
 Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
 Robust standard errors are clustered at the physician level.}
@@
 
-We construct a physician's testing inclination using the same method
-used to measure physician batch tendency, but instead we use the
-physician's residualized leave-out average test rate to get a measure of
-a physicians propensity to test. A key consideration in our analysis is
-the strong positive correlation (\(r = 0.79\)) between batch tendency
-and testing inclination. Such a high degree of correlation raises the
-concern that physicians with a higher testing inclination may batch more
-often as a result of ordering more tests. Failure to control for this
-inclination could result in biased estimates of the effect of batch
-tendency.
-
-The introduction of a Directed Acyclic Graph (Figure \ref{fig:DAG})
-elucidates the interconnected pathways between batch tendency, batch
-ordering, and primary outcomes. We posit that while batch tendency
-directly influences the practice of batch ordering tests, batch tendency
-and batch ordering are concurrently influenced by testing inclination.
-This is critical as testing inclination has a direct effect on primary
-outcomes, necessitating its inclusion as a control variable to
-disentangle its confounding effects.
+We introduce a Directed Acyclic Graph (DAG) to illustrate the
+interconnected pathways between batch tendency, batch ordering, and
+primary outcomes. The DAG in Figure \ref{fig:DAG} shows that batch
+tendency directly influences batch ordering, which in turn affects the
+number of tests ordered. The number of tests ordered, in turn,
+influences the length of stay. The DAG also illustrates the effect of
+batch tendency on the length of stay is mediated by the number of tests
+ordered.
 
@@
 
-In Model 2 of Table \ref{tab:reducedform}, we control for testing
-inclination and observe notable changes in the coefficients for batch
-tendency. The previously significant positive association with ln(LOS)
-diminishes, suggesting that the initial estimate may have been capturing
-the combined effects of batching and testing inclination. Estimated
-coefficients for batch tendency are scaled by the difference in batch
-tendency going from the ninetieth to tenth percentile in physician batch
-tendency for interpretability, and coefficients for testing inclination
-are scaled by the difference in testing inclination going from the
-ninetieth to tenth percentile in physician testing inclination.
-Assignment to a physician in the top batching decile (relative to one in
-the bottom decile) is associated with a 0.2 unit increase in the number
-of diagnostic tests being ordered per patient encounter, without any
-additional change in the likelihood of a 72-hour return.
-
-The effects of testing inclination are also striking. We find that a
-movement from a physician in the top testing decile (relative to one in
-the bottom decile) is associated with 0.6\% change in the length of stay
-in the emergency department, a 9.6 excess diagnostic tests being ordered
-per 100 patient encounters, and -0.03 pp change in the probability of a
-patient returning within 72 hours.
+To determine whether or not the effect of batch tendency on LOS is
+mediated by the number of tests ordered, we estimate a mediation
+analysis using the approach outlined by \citet{imai2010general}. Our
+analysis, scrutinizes the pathways through which physician batch
+tendency influences patient length of stay (LOS) in the emergency
+department. The analysis delineates both direct effects, which capture
+the influence of batch tendency on LOS without intermediaries, and
+indirect effects, which operate through the mediator of test ordering
+volume. Our findings underscore a significant Average Direct Effect
+(ADE) of batch tendency on LOS, estimated at -0.15 (p = 0.046).
 
@@
 presenting effects in Table \ref{tab:2SLS} on our main outcomes of ED
-LOS, number of tests, and 72-hour return.
-
-\begin{table}[!htbp] \centering 
-  \caption{2SLS Results: Length of Stay, Number of Tests, and 72-Hour Return} 
-  \label{tab:2SLS} 
-\begin{tabular}{p{7cm}ccc}
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & ln(LOS) & Number of Tests & 72 Hour Return \\ 
-\\[-1.8ex] & (1) & (2) & (3)\\ 
-\hline \\[-1.8ex] 
- Batch & $-$0.086 & 0.135$^{***}$ & 0.018 \\ 
-  & (0.419) & (0.046) & (0.018) \\ 
-  & & & \\ 
- Testing Inclination & 0.210$^{**}$ & 0.335$^{***}$ & $-$0.009$^{***}$ \\ 
-  & (0.082) & (0.008) & (0.003) \\ 
-  & & & \\ 
-Seasonality and shift fixed effects? & Yes & Yes & Yes \\
-Chief Complaint? & Yes & Yes & Yes \\
-Patient observables? & Yes & Yes & Yes \\
-Hospital volume? & Yes & Yes & Yes \\
-\midrule
-\textit{N} & 41,929 & 41,929 & 41,929 \\ 
-R$^{2}$ & 0.199 & 0.312 & 0.008 \\ 
-Adjusted R$^{2}$ & 0.194 & 0.307 & 0.002 \\ 
-Residual Std. Error (df = 41667) & 0.496 & 0.817 & 0.184 \\ 
-\\
+LOS and number of tests ordered. We first start by estimating the
+effects of batch ordering on the number of tests ordered using the
+standard 2SLS approach. Next we estimate the direct and indirect effects
+of batch ordering on LOS using a causal mediation analysis.
+
+\begin{table}[!htbp]
+\centering
+\begin{threeparttable}
+\caption{Combined 2SLS and Causal Mediation Analysis Results}
+\label{tab:combined_results}
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
+\toprule
+& \multicolumn{2}{c}{2SLS Results} \\
+\cmidrule{2-3}
+& ln(LOS) & Number of Tests \\
+\midrule
+Batch & -0.086 & 0.135$^{***}$ \\
+& (0.419) & (0.046) \\
+& & \\
+\addlinespace
+Seasonality and Shift Fixed Effects & Yes & Yes \\
+Chief Complaint and Severity & Yes & Yes \\
+\midrule
+Observations & 41,901 & 41,901 \\
+R$^{2}$ & 0.183 & 0.310 \\
 \bottomrule
-\end{tabular}
+\end{tabular*}
+
+\smallskip % Adds some space before the next part of the table
+
+\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} lcc}
+\toprule
+& \multicolumn{2}{c}{Causal Mediation Analysis} \\
+\cmidrule{2-3}
+Effect & Estimate & 95\% CI \\
+\midrule
+ACME & 0.036 & [-0.009, 0.08] \\
+ADE & -0.136 & [-0.221, -0.04] \\
+Total Effect & -0.100 & [-0.195, 0.00] \\
+Proportion Mediated & -0.333 & [-2.831, 1.21] \\
+\bottomrule
+\end{tabular*}
+
 \begin{tablenotes}
-\tiny
-\item \textit{Estimates of the IV model are for the baseline sample described in the text.
-Coefficients on testing inclination are scaled by the difference between testing inclination of the ninetieth and tenth percentile physicians for interpretability.
-Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects.
-Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter.
-Patient observables include sex dummy, race/ethnicity, and age bins.
-Robust standard errors are clustered at the physician level.}
-\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
+\small
+\item The 2SLS results are based on robust standard errors clustered at the physician level, with controls for seasonality, shift fixed effects, chief complaint, and severity.
+The causal mediation analysis examines whether the number of tests mediates the effect of batching on length of stay, using a quasi-Bayesian confidence interval approach with 1000 simulations.
+\item $^{***} p < 0.001$, $^{**} p < 0.01$, $^{*} p < 0.05$.
 \end{tablenotes}
+\end{threeparttable}
 \end{table}
 
-\emph{ED length of stay (LOS).}-- Our analysis reveals that batch
-ordering does not significantly reduce the length of stay in the ED.
-This finding holds even after accounting for various control variables,
-including ED volume and patient acuity. Contrary to common assumptions
-that batching tests could expedite patient throughput, our results
-suggest that the efficiency gains could be offset. A possible
-explanation lies in excess tests being ordered due to batching, which
-could lead to increased wait times for patients. This is consistent with
-our finding that batching increases the number of tests ordered.
+\emph{ED length of stay (LOS).}-- Our causal mediation analysis reveals
+that the Average Direct Effect (ADE) of test batching on the natural log
+of LOS is -0.13607, indicating that, when not considering the mediating
+effect of the number of tests, batching leads to a decrease in the
+expected LOS. Specifically, this translates to an approximate 12.7
+percent reduction in LOS when tests are batched, all else being equal.
+This result, complemented by our 2SLS estimate of -0.086, suggests that
+the direct effect of batch ordering on LOS is negative. Our results also
+suggest a ``washing out'' effect, where the effeciency gains from
+batching are offset by the increased number of tests ordered. This is
+consistent with our finding that batching increases the number of tests
+ordered.
 
@@
 utilization is pronounced. We observe a 0.135 increase in the number of
-tests ordered due to batching, corresponding to approximately 2,206
-additional tests during our study period. This is likely due to the
-information gain that occurs from sequential test ordering, where a
-physician may order additional tests after receiving the results of the
-first test.
-
-\emph{72-hour Return}-- Our findings indicate that batch ordering does
-not significantly impact the likelihood of a 72-hour return. This
-suggests that despite the increased number of tests, there is no
-corresponding decrease in short-term readmission rates. This highlights
-the need for more targeted testing strategies that balance comprehensive
-assessment with resource efficiency.
+tests ordered due to batching, corresponding to approximately 13.5
+excess tests per 100 batch orders, where excess means tests that would
+not have been ordered had the physician not batched. This result is
+consistent with the notion that batching leads to increased resource
+utilization in the ED.
 
@@
 our findings suggest that it may lead to increased resource utilization
-without corresponding gains in patient throughput or reductions in
-short-term return rates. These insights point to the need for a more
-nuanced approach to test ordering in emergency care, one that carefully
-weighs the benefits of comprehensive assessment against the risks and
-costs of potential over-testing.
+and that the corresponding gains in patient throughput are washed out.
+These insights point to the need for a more nuanced approach to test
+ordering in emergency care, one that carefully weighs the benefits of
+comprehensive assessment against the risks and costs of potential
+over-testing.
 
@@
 ordered. However, we find no significant impact of batching on ED length
-of stay or 72-hour return rates. This suggests that despite the
-increased number of tests, there is no corresponding decrease in
-short-term readmission rates. This highlights the need for more targeted
-testing strategies that balance comprehensive assessment with resource
-efficiency.
+of stay. Causal mediation analysis reveals that the direct effect of
+batching on length of stay is negative, but this effect is offset by the
+increased number of tests ordered. This suggests that the efficiency
+gains from batching are washed out by the increased resource
+utilization.
+
+Our findings have important implications for emergency care practice.
+While batching is often seen as a method to enhance efficiency, our
+results suggest that it may lead to increased resource utilization and
+that the corresponding gains in patient throughput are washed out. These
+insights point to the need for a more nuanced approach to test ordering
+in emergency care, one that carefully weighs the benefits of
+comprehensive assessment against the risks and costs of potential
+over-testing. Future research should explore the impact of batching on
+other outcomes, such as patient satisfaction and clinical outcomes, to
+provide a more comprehensive understanding of the implications of
+batching in the emergency department.
 
@@
 
+\begin{table}[ht]
+\centering
+\caption{2SLS Results for the "Batch" Variable Across Different General Complaints: Outcome of Number of Tests}
+\begin{tabular}{lcc}
+\hline
+\textbf{Complaint} & \textbf{Batching Coefficient} & \textbf{Standard Error} \\
+\hline
+Musculoskeletal/Extremity & 0.476** & (0.235)  \\
+Respiratory-Related & 0.244 & (0.209)   \\
+General/Other Symptoms & 0.179 & (0.207)   \\
+Gastrointestinal/Abdominal & -0.092 & (0.128)   \\
+Cardiac/Chest-Related & 0.270 & (0.264)   \\
+Neurological/Syncope & -0.080 & (0.493)   \\
+\hline
+\end{tabular}
+\label{tab:batch_variable_outcome}
+\end{table}
+
 \end{APPENDIX}

### 2024-09-09 v2/manuscriptv2.Rmd
getting going

--- 2024-09-09 manuscript/CopyOfmanuscript.Rmd
@@
 ---
-title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
+title: "The Impact of Batch Ordering Imaging Tests in the Emergency Department"
 authors:
-- familyname: Jameson
-  othernames: Jacob
-  affiliation: Interfaculty Initiative in Health Policy, Harvard University
+  - name: Jacob Jameson
+    department: Interfaculty Initiative in Health Policy
+    affiliation: Harvard University
+    location: Cambridge, MA
     email: jacobjameson@g.harvard.edu
-- familyname: Saghafian
-  othernames:  Soroush
-  affiliation: Kennedy School of Government, Harvard University 
+  - name: Soroush Saghafian
+    department: Kennedy School of Government
+    affiliation: Harvard University
+    location: Cambridge, MA
     email: soroush_saghafian@hks.harvard.edu
-- familyname: Huckman
-  othernames:  Robert
-  affiliation: Harvard Business School 
+  - name: Robert Huckman
+    department: Harvard Business School
+    affiliation: Harvard University
+    location: Boston, MA
     email: rhuckman@hbs.edu
-- familyname: Hodgson
-  othernames:  Nicole
+  - name: Nicole Hodgson
+    department: Department of Emergency Medicine
     affiliation: Mayo Clinic
+    location: Phoenix, AZ
     email: Hodgson.Nicole@mayo.edu
-keywords:
-  - Emergency Department, Operational Effeciency, Diagnostic Testing
 abstract: |
   We use practice variation across physicians to uncover the role of test-ordering on care delivery in the ED. Using records of over 45,000 Mayo Clinic emergency department visits...
+keywords:
+  - Emergency Department
+  - Operational Efficiency
+  - Diagnostic Testing
+acknowledgment: |
+  The authors would like to thank ..
 bibliography: references.bib
-blind: false
-output: rticles::informs_article
+output: rticles::arxiv_article
 header-includes:
@@
 
+To determine whether or not the effect of batch tendency on LOS is mediated by the number of tests ordered, we estimate a mediation analysis using the approach outlined by @imai2010general.
+Our analysis, scrutinizes the pathways through which physician batch tendency influences patient length of stay (LOS) in the emergency department.
+The analysis delineates both direct effects, which capture the influence of batch tendency on LOS without intermediaries, and indirect effects, which operate through the mediator of test ordering volume.
+Our findings underscore a significant Average Direct Effect (ADE) of batch tendency on LOS, estimated at -0.15  (p = 0.046).
+
 ## Placebo Check
@@
 
-To that end, we restrict attention to ED visits for complaints where batching occurs no more than 5 percent of the time (recall that our baseline sample only includes complaints with a $>5$ percent batching rate).
-We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (6).
+To that end, we restrict attention to ED visits for complaints where batching occurs no more than 10 percent of the time (recall that our baseline sample only includes complaints with a >2 percent batching rate).
+We estimate a reduced-form regression of each main outcome on physician batch tendency for the subsample, following equation (3).
 The results of this exercise are displayed in Appendix Table 1.
@@
 \end{tabularx}
+\begin{tablenotes}
+\small
+\item Note: ** p < 0.05, * p < 0.1. Full controls included. Estimates from 2SLS and causal mediation analysis, presenting impacts on number of tests, length of stay, and 72-hour return rates.
+\end{tablenotes}
 \end{threeparttable}
@@
 \end{tabularx}
+\small Note: * p < 0.1, ** p < 0.05.
+ACME: Average Causal Mediation Effect, ADE: Average Direct Effect.
+LOS is measured in log-transformed hours.
+72hr Return is the probability of return within 72 hours with admission.
 \end{table}
@@
 # Conclusion
-
-This study provides empirical evidence on the impacts of batch ordering imaging tests in emergency departments, illuminating the trade-offs between comprehensive initial assessments and efficient resource utilization.
-Our findings, derived from a quasi-experimental design leveraging physician practice variation, align closely with our theoretical model of ED decision-making.
-We demonstrate that batch ordering, while leading to more thorough initial evaluations, results in increased resource use and longer ED stays.
-
-The fundamental tension we uncover is between the information gain approach of sequential testing and the comprehensive but potentially excessive nature of batch ordering.
-Our results indicate that batch ordering leads to approximately 70 additional imaging tests per 100 patient encounters compared to a sequential approach.
-This substantial increase in testing volume translates to longer ED stays but also slightly reduced 72-hour return rates with admission.
-These findings underscore the complex decision-making environment in EDs, where physicians must balance the need for thorough diagnostics against operational efficiency and patient flow.
-
-Importantly, our subgroup analysis reveals significant heterogeneity in the effects of batch ordering across different presenting complaints.
-For conditions with high diagnostic uncertainty or potential for serious underlying issues, such as dizziness or syncope, the benefits of comprehensive initial testing may outweigh the costs of increased resource utilization.
-Conversely, for more straightforward presentations, a sequential, information gain approach appears more appropriate.
-This variation in effects emphasizes the need for nuanced, complaint-specific approaches to test ordering in emergency settings.
-
-Our results have significant implications for ED management and health policy.
-They suggest that a one-size-fits-all approach to diagnostic testing in the ED is suboptimal.
-Instead, ED managers and policymakers should consider implementing targeted batch ordering strategies that account for the specific nature of patient presentations and the relative risks of missed diagnoses versus overcrowding and resource strain.
-Future research should focus on developing and validating decision support tools that can guide clinicians in choosing between batch and sequential ordering based on patient-specific factors and presenting complaints.
-
-In conclusion, this study contributes to our understanding of ED operations by quantifying the causal impacts of different diagnostic testing strategies.
-By highlighting the nuanced trade-offs involved in batch ordering, we provide a foundation for more informed decision-making in emergency care.
-As healthcare systems continue to grapple with the dual challenges of providing high-quality care and managing limited resources, insights from studies like this one will be crucial in shaping evidence-based policies and practices in emergency medicine.
-
-\newpage

### 2024-09-09 v2/manuscriptv2.tex
getting going

--- 2024-09-09 manuscript/CopyOfmanuscript.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
-%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
-%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
-%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[,,nonblindrev]{informs}
-
-\OneAndAHalfSpacedXI
-%%\OneAndAHalfSpacedXII % Current default line spacing
-%%\DoubleSpacedXII
-%%\DoubleSpacedXI
-
-%% BEGIN MY ADDITIONS %%
-\usepackage{hyperref}
+\documentclass{article}
+
+\usepackage{arxiv}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{graphicx}
+
+\title{Judicious Batching: The Impact of Batch Ordering Imaging Tests in
+the Emergency Department}
+
+\author{
+    Jacob Jameson
+   \\
+    Interfaculty Initiative in Health Policy \\
+    Harvard University \\
+  Cambridge, MA \\
+  \texttt{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}} \\
+   \And
+    Soroush Saghafian
+   \\
+    Kennedy School of Government \\
+    Harvard University \\
+  Cambridge, MA \\
+  \texttt{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}} \\
+   \And
+    Robert Huckman
+   \\
+    Harvard Business School \\
+    Harvard University \\
+  Boston, MA \\
+  \texttt{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}} \\
+   \And
+    Nicole Hodgson
+   \\
+    Department of Emergency Medicine \\
+    Mayo Clinic \\
+  Phoenix, AZ \\
+  \texttt{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}} \\
+  }
 
@@
 \usepackage{caption}
-
-%% END MY ADDITIONS %%
-
-% Natbib setup for author-year style
-\usepackage{natbib}
- \bibpunct[, ]{(}{)}{,}{a}{}{,}%
- \def\bibfont{\small}%
- \def\bibsep{\smallskipamount}%
- \def\bibhang{24pt}%
- \def\newblock{\ }%
- \def\BIBand{and}%
-
-%% Setup of theorem styles. Outcomment only one.
-%% Preferred default is the first option.
-\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
-%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
-\ECRepeatTheorems
-
-%% Setup of the equation numbering system. Outcomment only one.
-%% Preferred default is the first option.
-\EquationsNumberedThrough    % Default: (1), (2), ...
-%\EquationsNumberedBySection % (1.1), (1.2), ...
-
-% For new submissions, leave this number blank.
-% For revisions, input the manuscript number assigned by the on-line
-% system along with a suffix ".Rx" where x is the revision number.
-\MANUSCRIPTNO{}
-
-%%%%%%%%%%%%%%%%
 \begin{document}
-%%%%%%%%%%%%%%%%
-
-% Outcomment only when entries are known. Otherwise leave as is and
-%   default values will be used.
-%\setcounter{page}{1}
-%\VOLUME{00}%
-%\NO{0}%
-%\MONTH{Xxxxx}% (month or a similar seasonal id)
-%\YEAR{0000}% e.g., 2005
-%\FIRSTPAGE{000}%
-%\LASTPAGE{000}%
-%\SHORTYEAR{00}% shortened year (two-digit)
-%\ISSUE{0000} %
-%\LONGFIRSTPAGE{0001} %
-%\DOI{10.1287/xxxx.0000.0000}%
-
-% Author's names for the running heads
-% Sample depending on the number of authors;
-\RUNAUTHOR{%
-Jameson, Saghafian, Huckman
- and Hodgson
+\maketitle
+
+\begin{abstract}
+We use practice variation across physicians to uncover the role of
+test-ordering on care delivery in the ED. Using records of over 45,000
+Mayo Clinic emergency department visits\ldots{}
+\end{abstract}
+
+\keywords{
+    Emergency Department
+   \and
+    Operational Efficiency
+   \and
+    Diagnostic Testing
   }
-% \RUNAUTHOR{Jones and Wilson}
-% \RUNAUTHOR{Jones, Miller, and Wilson}
-% \RUNAUTHOR{Jones et al.} % for four or more authors
-% Enter authors following the given pattern:
-%\RUNAUTHOR{}
-
-\RUNTITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}
-
-\TITLE{To Batch or Not to Batch: Test-Ordering Variability in the ED}
-
-\ARTICLEAUTHORS{%
-\AUTHOR{Jacob Jameson}
-\AFF{Interfaculty Initiative in Health Policy, Harvard
-University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}
-
-\AUTHOR{Soroush Saghafian}
-\AFF{Kennedy School of Government, Harvard
-University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}
-
-\AUTHOR{Robert Huckman}
-\AFF{Harvard Business
-School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}
-
-\AUTHOR{Nicole Hodgson}
-\AFF{Mayo
-Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}
-
-%
-}
-
-\ABSTRACT{We use practice variation across physicians to uncover the
-role of test-ordering on care delivery in the ED. Using records of over
-45,000 Mayo Clinic emergency department visits\ldots{}}
-
-\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
-Testing}
-
-\maketitle
 
@@
 crucial for alleviating overcrowding, a severe issue associated with
-higher complication rates and increased mortality \citet{bernstein2009}.
-In this complex environment, the availability and performance of
-diagnostic tests, especially imaging studies, play a pivotal role in
-shaping both patient care and operational efficiency \citet{naseim2015}.
+higher complication rates and increased mortality
+(\textbf{bernstein2009?}). In this complex environment, the availability
+and performance of diagnostic tests, especially imaging studies, play a
+pivotal role in shaping both patient care and operational efficiency
+(\textbf{naseim2015?}).
 
@@
 increased costs, patient anxiety, and potential harm from follow-up of
-false-positive results \citet{koch2018}. Conversely, a more sequential
-approach might reduce unnecessary testing but could extend a patient's
-time in the ED, potentially contributing to overcrowding. Striking the
-right balance between comprehensive evaluation and efficient resource
-use is crucial, yet research on optimal test ordering strategies in the
-ED remains limited \citet{saghafian2015}.
+false-positive results (\textbf{koch2018?}). Conversely, a more
+sequential approach might reduce unnecessary testing but could extend a
+patient's time in the ED, potentially contributing to overcrowding.
+Striking the right balance between comprehensive evaluation and
+efficient resource use is crucial, yet research on optimal test ordering
+strategies in the ED remains limited (\textbf{saghafian2015?}).
 
@@
 assigned to physicians via a rotational patient assignment algorithm
-\citet{traub2016emergency}, which removes potential selection bias
+(\textbf{traub2016emergency?}), which removes potential selection bias
 concerns for our analyses. In essence, barring arrival time and
@@
 leave-out average batch rate. This measure is derived from two steps
-following the approaches taken by \citet{doyle2015measuring},
-\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
-obtain residuals from a regression model, which includes all ED
-encounters in our sample period.
+following the approaches taken by (\textbf{doyle2015measuring?}),
+(\textbf{dobbie2018effects?}), and (\textbf{eichmeyer2022pathways?}).
+First, we obtain residuals from a regression model, which includes all
+ED encounters in our sample period.
 
@@
 that the first stage should be weakly positive for all subsamples
-(\citet{dobbie2018effects}). The second implication asserts that the
+((\textbf{dobbie2018effects?})). The second implication asserts that the
 instrument constructed by leaving out a particular subsample has
 predictive power over that same left-out subsample
-(\citet{bhuller2020incarceration}). Appendix Table
+((\textbf{bhuller2020incarceration?})). Appendix Table
 \ref{tab:monotonicity} presents both of these tests in the two columns
@@
 \end{table}
+
+To determine whether or not the effect of batch tendency on LOS is
+mediated by the number of tests ordered, we estimate a mediation
+analysis using the approach outlined by (\textbf{imai2010general?}). Our
+analysis, scrutinizes the pathways through which physician batch
+tendency influences patient length of stay (LOS) in the emergency
+department. The analysis delineates both direct effects, which capture
+the influence of batch tendency on LOS without intermediaries, and
+indirect effects, which operate through the mediator of test ordering
+volume. Our findings underscore a significant Average Direct Effect
+(ADE) of batch tendency on LOS, estimated at -0.15 (p = 0.046).
 
@@
 To that end, we restrict attention to ED visits for complaints where
-batching occurs no more than 5 percent of the time (recall that our
-baseline sample only includes complaints with a \(>5\) percent batching
-rate). We estimate a reduced-form regression of each main outcome on
-physician batch tendency for the subsample, following equation (6). The
-results of this exercise are displayed in Appendix Table 1. They show
-that in contrast to results for our main sample, the association between
-physician tendency to batch and a given outcome is statistically
-indistinguishable from zero and much smaller in magnitude for the
-samples of patients who visit the ED with health conditions that are
-rarely batched.
+batching occurs no more than 10 percent of the time (recall that our
+baseline sample only includes complaints with a \textgreater10 percent
+batching rate). We estimate a reduced-form regression of each main
+outcome on physician batch tendency for the subsample, following
+equation (3). The results of this exercise are displayed in Appendix
+Table 1. They show that in contrast to results for our main sample, the
+association between physician tendency to batch and a given outcome is
+statistically indistinguishable from zero and much smaller in magnitude
+for the samples of patients who visit the ED with health conditions that
+are rarely batched.
 
@@
 
-This study provides empirical evidence on the impacts of batch ordering
-imaging tests in emergency departments, illuminating the trade-offs
-between comprehensive initial assessments and efficient resource
-utilization. Our findings, derived from a quasi-experimental design
-leveraging physician practice variation, align closely with our
-theoretical model of ED decision-making. We demonstrate that batch
-ordering, while leading to more thorough initial evaluations, results in
-increased resource use and longer ED stays.
-
-The fundamental tension we uncover is between the information gain
-approach of sequential testing and the comprehensive but potentially
-excessive nature of batch ordering. Our results indicate that batch
-ordering leads to approximately 70 additional imaging tests per 100
-patient encounters compared to a sequential approach. This substantial
-increase in testing volume translates to longer ED stays but also
-slightly reduced 72-hour return rates with admission. These findings
-underscore the complex decision-making environment in EDs, where
-physicians must balance the need for thorough diagnostics against
-operational efficiency and patient flow.
-
-Importantly, our subgroup analysis reveals significant heterogeneity in
-the effects of batch ordering across different presenting complaints.
-For conditions with high diagnostic uncertainty or potential for serious
-underlying issues, such as dizziness or syncope, the benefits of
-comprehensive initial testing may outweigh the costs of increased
-resource utilization. Conversely, for more straightforward
-presentations, a sequential, information gain approach appears more
-appropriate. This variation in effects emphasizes the need for nuanced,
-complaint-specific approaches to test ordering in emergency settings.
-
-Our results have significant implications for ED management and health
-policy. They suggest that a one-size-fits-all approach to diagnostic
-testing in the ED is suboptimal. Instead, ED managers and policymakers
-should consider implementing targeted batch ordering strategies that
-account for the specific nature of patient presentations and the
-relative risks of missed diagnoses versus overcrowding and resource
-strain. Future research should focus on developing and validating
-decision support tools that can guide clinicians in choosing between
-batch and sequential ordering based on patient-specific factors and
-presenting complaints.
-
-In conclusion, this study contributes to our understanding of ED
-operations by quantifying the causal impacts of different diagnostic
-testing strategies. By highlighting the nuanced trade-offs involved in
-batch ordering, we provide a foundation for more informed
-decision-making in emergency care. As healthcare systems continue to
-grapple with the dual challenges of providing high-quality care and
-managing limited resources, insights from studies like this one will be
-crucial in shaping evidence-based policies and practices in emergency
-medicine.
-
-\newpage
-
-% Appendix here
-% Options are (1) APPENDIX (with or without general title) or
-%             (2) APPENDICES (if it has more than one unrelated sections)
-% Outcomment the appropriate case if necessary
-%
-% \begin{APPENDIX}{<Title of the Appendix>}
-% \end{APPENDIX}
-%
-%   or
-%
-% \begin{APPENDICES}
-% \section{<Title of Section A>}
-% \section{<Title of Section B>}
-% etc
-% \end{APPENDICES}
-
-% Acknowledgments here
-\ACKNOWLEDGMENT{}
-
-\bibliographystyle{informs2014}
+\bibliographystyle{unsrt}
 \bibliography{references.bib}

### 2024-09-09 src/IV Validation.R

@@ -8,8 +8,28 @@ library(stargazer)
 library(texreg)
 library(xtable)
 library(tidyverse)
+library(sandwich)
+library(lmtest)
 
-data <- read_csv('outputs/data/final.csv')
+data <- read_csv('outputs/data/final.csv') %>%
+  filter(ARRIVAL_AGE_DI > 18) %>%
+  group_by(complaint_esi) %>%
+  filter(n() > 1) %>%
+  ungroup() %>%
+  group_by(CHIEF_COMPLAINT) %>%
+  mutate(batchmean = mean(batched)) %>%
+  filter(batchmean > 0.05, imaging == 1)
+
+data <- data %>%
+  mutate(
+    Time_arrival_to_triage = TRIAGE_COMPLETED_REL - ARRIVAL_DTTM_REL,
+    Time_triage_to_firstcontact = FIRST_CONTACT_DTTM_REL - TRIAGE_COMPLETED_REL,
+    Time_to_Result_Lab = LAB_RESULT_DTTM_REL - LAB_ORDER_REL,
+    Time_to_Result_XRay = PLAIN_XRAY_RESULT_DTTM_REL - XR_ORDER_REL,
+    Time_to_Result_Ultrasound = US_RESULT_DTTM_REL - US_ORDER_DTTM_REL,
+    Time_to_Result_CTcon = CT_WITH_CONTR_RESULT_DTTM_REL - CON_CT_ORDER_REL,
+    Time_to_Result_CTnon = CT_WITHOUT_CONTR_RESULT_DTTM_REL - NON_CON_CT_ORDER_REL
+  )
 
 ##########################################################################
 #=========================================================================
@@ -27,29 +47,29 @@ data <- read_csv('outputs/data/final.csv')
 ## First Stage results
 #=========================================================================
 ##########################################################################
+# Save the results to a .txt file
+sink("outputs/tables/First Stage.txt")
 
 # Shift-level FE
 first_stage_final <- felm(
-  any.batch ~ batch.tendency + test.inclination | 
+  batched ~ batch.tendency | 
               dayofweekt + month_of_year |0| ED_PROVIDER, 
   data = data)
 
 # Shift-level + complaint FE
 first_stage_final_d <- felm(
-  any.batch ~ batch.tendency + test.inclination | 
-              dayofweekt + month_of_year + age_groups + 
-              complaint_esi | 0 | ED_PROVIDER,
+  batched ~ batch.tendency  | 
+              dayofweekt + month_of_year + 
[…]

### 2024-09-09 src/Results new clean.R

@@ -1,0 +1,184 @@
+#=========================================================================
+# Purpose: Validate IV + Main Results
+# Author: Jacob Jameson 
+#=========================================================================
+
+library(lfe)
+library(stargazer)
+library(texreg)
+library(xtable)
+library(tidyverse)
+library(sandwich)
+library(lmtest)
+
+data <- read_csv('outputs/data/final.csv') %>%
+  filter(ARRIVAL_AGE_DI > 18) %>%
+  group_by(complaint_esi) %>%
+  filter(n() > 1) %>%
+  ungroup() %>%
+  group_by(CHIEF_COMPLAINT) %>%
+  mutate(batchmean = mean(batched)) %>%
+  filter(batchmean > 0.05, imaging == 1)
+
+################################################################################
+#-------------------------------------------------------------------------------
+# Physician factors to explain batching
+#-------------------------------------------------------------------------------
+################################################################################
+
+# Model 1 ---- Baseline model --------------------------------------------------
+model.1 <- felm(
+  batched ~ EXPERIENCE + PROVIDER_SEX + patients_tbs + patients_in_hospital | 
+    dayofweekt + month_of_year |0| ED_PROVIDER, 
+  data = data)
+
+# Model 2 ---- Add patient characteristics -------------------------------------
+model.2 <- felm(
+  batched ~ EXPERIENCE + PROVIDER_SEX + patients_tbs + patients_in_hospital |
+    dayofweekt + month_of_year + 
+    complaint_esi + LAB_PERF  | 0 | ED_PROVIDER, 
+  data = data)
+
+stargazer(model.1, model.2, type = "text", single.row = TRUE, digits = 3,
+          style = "qje", header = T,
+          column.labels = c("Model 1", "Model 2", "Model 3"),
+          dep.var.caption = "Dependent Variable: Batched", 
+          dep.var.labels.include = FALSE)
+      
+################################################################################
+
+################################################################################
+#-------------------------------------------------------------------------------
+# First Stage 
+#-------------------------------------------------------------------------------
+################################################################################
+
+# Model 1 ---- Baseline model --------------------------------------------------
+fs.1 <- felm(
[…]

### 2024-09-09 src/clean.R

@@ -19,6 +19,7 @@ df <- read.csv(paste0(path, 'deidentified_FINAL.csv'))
 test_columns = c("US_PERF", "NON_CON_CT_PERF", "CON_CT_PERF", 
                  "LAB_PERF", "XR_PERF")
 
+
 colnames(df)[colnames(df) == "PLAIN_XRAY"] = "XR_PERF"
 colnames(df)[colnames(df) == "US_ORDER_DTTM_REL"] ="US_ORDER_DTTM_REL"
 colnames(df)[colnames(df) == "CT_WITHOUT_CONTR_ORDER_DTTM_REL"] = "NON_CON_CT_ORDER_REL"
@@ -28,13 +29,13 @@ colnames(df)[colnames(df) == "PLAIN_XRAY_ORDER_DTTM_REL"] ="XR_ORDER_REL"
 
 df$CT_PERF = ifelse(df$NON_CON_CT_PERF=='Y' | df$CON_CT_PERF=='Y', 1, 0)
 
-df$ESI = as.character(df$ESI)
-
 for (i in test_columns){
   df[[i]] = ifelse(df[[i]] =='Y', 1, 0) 
 }
 
 df$nEDTests = rowSums(df[test_columns])
+
+df$imgTests = df$nEDTests - df$LAB_PERF
 
 # Identify columns with *_REL suffix
 rel_cols <- grep("_REL$", names(df), value = TRUE)
@@ -50,21 +51,33 @@ for (col in rel_cols) {
 df <- df %>%
   select(-matches("_hours$|_minutes$"))
 
+
 #=========================================================================
 # Determine batching
 #   - criteria: ordered within 5 minutes of each other
 #               first tests ordered were in a batch
 #=========================================================================
 
-# Determine which was the first test ordered -----------------------------
-# Columns of interest
 
+# Columns of interest
 cols_of_interest <- c('US_ORDER_DTTM_REL', 'NON_CON_CT_ORDER_REL', 
-                      'CON_CT_ORDER_REL',
-                      'LAB_ORDER_REL', 'XR_ORDER_REL')
+                      'CON_CT_ORDER_REL', 'XR_ORDER_REL')
+
+cols_res <- c("CT_WITHOUT_CONTR_RESULT_DTTM_REL",
+              "CT_WITH_CONTR_RESULT_DTTM_REL", "PLAIN_XRAY_RESULT_DTTM_REL" ,     
+              "US_RESULT_DTTM_REL" )
+
+# determine earliest test order to latest test result
+
[…]

### 2024-09-09 src/summary tables.R

@@ -8,8 +8,11 @@ library(stargazer)
 library(texreg)
 library(xtable)
 library(tidyverse)
+library(caret)
+library(sandwich)
+library(lmtest)
 
-data <- read_csv('outputs/data/final.csv')
+data <- read_csv('outputs/data/all_clean.csv')
 
 ##########################################################################
 ##########################################################################
@@ -18,70 +21,103 @@ data <- read_csv('outputs/data/final.csv')
 #=========================================================================
 ##########################################################################
 
+# Adjust the dataset and calculate the time differences
+data <- data %>%
+  mutate(
+    Time_arrival_to_triage = TRIAGE_COMPLETED_REL - ARRIVAL_DTTM_REL,
+    Time_triage_to_firstcontact = FIRST_CONTACT_DTTM_REL - TRIAGE_COMPLETED_REL,
+    Time_to_Result_Lab = LAB_RESULT_DTTM_REL - LAB_ORDER_REL,
+    Time_to_Result_XRay = PLAIN_XRAY_RESULT_DTTM_REL - XR_ORDER_REL,
+    Time_to_Result_Ultrasound = US_RESULT_DTTM_REL - US_ORDER_DTTM_REL,
+    Time_to_Result_CTcon = CT_WITH_CONTR_RESULT_DTTM_REL - CON_CT_ORDER_REL,
+    Time_to_Result_CTnon = CT_WITHOUT_CONTR_RESULT_DTTM_REL - NON_CON_CT_ORDER_REL
+  )
+
+
 # Identify the five most frequent chief complaints
 top_chief_complaints <- data %>%
   count(CHIEF_COMPLAINT, sort = TRUE) %>%
-  slice_max(n = 5, order_by = n) %>%
+  top_n(5, wt = n) %>%
   pull(CHIEF_COMPLAINT)
 
-# Calculate the mean for each of the top chief complaints
+# Calculate the mean proportion for each of the top chief complaints
 means_chief_complaints <- sapply(
   top_chief_complaints, 
   function(cc) mean(data$CHIEF_COMPLAINT == cc, na.rm = TRUE))
 
-# Panel 1: Emergency department characteristics ===================================
+# Panel 1: Emergency department characteristics ====================================
[…]
